---
title: "Age distribution of fatal opioid-detected overdoses in New York [excluding NYC], 2018-2023"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 5
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r sourceData, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(stringr)
library(dplyr)
library(lubridate)
library(patchwork)
library(tidyverse)
library(sf)
library(multimode) #to calculate modes for the age plot
library(diptest)
library(ggrepel)

# CHANGE PATH AS NEEDED -- get the most updated data files
source("data_preprocessing/ny_processing.R")
# CHANGE PATH AS NEEDED -- set working directory, I have to do this after sourcing the data because it loses the path otherwise
setwd("C:\\Users\\ar2529\\Desktop\\overdose_age")

```

```{r importData, echo=FALSE, message=FALSE, warning=FALSE}

# import ny data 
ny_data <- read.csv("data_preprocessing/processed_data/ny_processed.csv")

# split the dataset by year [so we can get the age distribution by year], from https://stackoverflow.com/questions/57454085/assign-a-variable-in-r-using-another-variable

for(i in unique(ny_data$Year)){
  assign(paste0("ny_", as.character(i)), subset(ny_data, Year == as.character(i)))
}

```

### Age distribution for all fatal overdoses {.tabset}

#### 2018

```{r age2018, echo=FALSE, message=FALSE, warning=FALSE}
# calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
# looks like there are some NAs in the 2018 data so I have to get rid of them
bandwidth <- bw.nrd0(na.omit(ny_2018$age)) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(ny_2018$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(ny_2018$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# I will need these values for a separate graph that I can generate without data access
print("Modes:")
modes
print("Antimodes:")
antimodes

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
ny_2018_young <- subset(ny_2018, age <= antimodes[1])
ny_2018_old <- subset(ny_2018, age > antimodes[1])

# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

ny_2018$aftermode <- ny_2018$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid

# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=ny_2018, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(ny_2018$age), max(ny_2018$age), by = 5), labels = seq(min(ny_2018$age), max(ny_2018$age), by = 5), minor_breaks = seq(min(ny_2018$age), max(ny_2018$age), by = 1)) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2018 NY opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

# plot with mode lines
age_2018 <- ggplot(data=ny_2018, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(ny_2018$age), max(ny_2018$age), by = 5), labels = seq(min(ny_2018$age), max(ny_2018$age), by = 5), minor_breaks = seq(min(ny_2018$age), max(ny_2018$age), by = 1)) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2018 NY opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2018

# CHANGE PATH OR COMMENT THIS PART OUT
ggsave(plot=age_2018, filename = "images/ny/age_2018.png", width = 10, height = 8)


sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(ny_2018$age)$p.value) 

```

#### 2019

```{r age2019, echo=FALSE, message=FALSE, warning=FALSE}

# calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
bandwidth <- bw.nrd0(ny_2019$age) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(ny_2019$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(ny_2019$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# I will need these values for a separate graph that I can generate without data access
print("Modes:")
modes
print("Antimodes:")
antimodes

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
ny_2019_young <- subset(ny_2019, age <= antimodes[1])
ny_2019_old <- subset(ny_2019, age > antimodes[1])

# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

ny_2019$aftermode <- ny_2019$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=ny_2019, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") +scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(ny_2019$age), max(ny_2019$age), by = 5), labels = seq(min(ny_2019$age), max(ny_2019$age), by = 5), minor_breaks = seq(min(ny_2019$age), max(ny_2019$age), by = 1)) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2019 NY opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

# plot with mode lines
age_2019 <- ggplot(data=ny_2019, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(ny_2019$age), max(ny_2019$age), by = 5), labels = seq(min(ny_2019$age), max(ny_2019$age), by = 5), minor_breaks = seq(min(ny_2019$age), max(ny_2019$age), by = 1)) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2019 NY opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2019

# CHANGE PATH OR COMMENT THIS PART OUT
ggsave(plot=age_2019, filename = "images/ny/age_2019.png", width = 10, height = 8)

sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(ny_2019$age)$p.value) 


```

#### 2020

```{r age2020, echo=FALSE, message=FALSE, warning=FALSE}

# calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
bandwidth <- bw.nrd0(ny_2020$age) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(ny_2020$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(ny_2020$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# I will need these values for a separate graph that I can generate without data access
print("Modes:")
modes
print("Antimodes:")
antimodes

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
ny_2020_young <- subset(ny_2020, age <= antimodes[1])
ny_2020_old <- subset(ny_2020, age > antimodes[1])

# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

ny_2020$aftermode <- ny_2020$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=ny_2020, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = aftermode), position = "dodge") +  scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(ny_2020$age), max(ny_2020$age), by = 5), labels = seq(min(ny_2020$age), max(ny_2020$age), by = 5), minor_breaks = seq(min(ny_2020$age), max(ny_2020$age), by = 1)) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2020 NY opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values =  c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

# plot with mode lines
age_2020 <- ggplot(data=ny_2020, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = aftermode), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(ny_2020$age), max(ny_2020$age), by = 5), labels = seq(min(ny_2020$age), max(ny_2020$age), by = 5), minor_breaks = seq(min(ny_2020$age), max(ny_2020$age), by = 1)) + theme_light() + geom_vline(xintercept = modes, size=1, color="black") + ggtitle("Age distribution for 2020 NY opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values =  c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2020

# CHANGE PATH OR COMMENT THIS PART OUT
ggsave(plot=age_2020, filename = "images/ny/age_2020.png", width = 10, height = 8)

sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(ny_2020$age)$p.value) 


```

#### 2021

```{r age2021, echo=FALSE, message=FALSE, warning=FALSE}

# calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
bandwidth <- bw.nrd0(ny_2021$age) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(ny_2021$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(ny_2021$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# I will need these values for a separate graph that I can generate without data access
print("Modes:")
modes
print("Antimodes:")
antimodes

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
ny_2021_young <- subset(ny_2021, age <= antimodes[1])
ny_2021_old <- subset(ny_2021, age > antimodes[1])


# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

ny_2021$aftermode <- ny_2021$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=ny_2021, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") +  scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(ny_2021$age), max(ny_2021$age), by = 5), labels = seq(min(ny_2021$age), max(ny_2021$age), by = 5), minor_breaks = seq(min(ny_2021$age), max(ny_2021$age), by = 1))  + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2021 NY opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))

# plot with mode lines
age_2021 <- ggplot(data=ny_2021, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(ny_2021$age), max(ny_2021$age), by = 5), labels = seq(min(ny_2021$age), max(ny_2021$age), by = 5), minor_breaks = seq(min(ny_2021$age), max(ny_2021$age), by = 1))  + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2021 NY opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))

age_2021

# CHANGE PATH OR COMMENT THIS PART OUT
ggsave(plot=age_2021, filename = "images/ny/age_2021.png", width = 10, height = 8)

sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(ny_2021$age)$p.value) 

 
```

#### 2022

```{r age2022, echo=FALSE, message=FALSE, warning=FALSE}

# calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
bandwidth <- bw.nrd0(ny_2022$age) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(ny_2022$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(ny_2022$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# I will need these values for a separate graph that I can generate without data access
print("Modes:")
modes
print("Antimodes:")
antimodes

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
# in this case we have to use the second antimode because it bifurcates the distribution better
ny_2022_young <- subset(ny_2022, age <= antimodes[2])
ny_2022_old <- subset(ny_2022, age > antimodes[2])

# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

ny_2022$aftermode <- ny_2022$age > antimodes[2]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=ny_2022, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(ny_2022$age), max(ny_2022$age), by = 5), labels = seq(min(ny_2022$age), max(ny_2022$age), by = 5), minor_breaks = seq(min(ny_2022$age), max(ny_2022$age), by = 1)) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2022 NY opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

# plot with mode lines
age_2022 <- ggplot(data=ny_2022, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(ny_2022$age), max(ny_2022$age), by = 5), labels = seq(min(ny_2022$age), max(ny_2022$age), by = 5), minor_breaks = seq(min(ny_2022$age), max(ny_2022$age), by = 1)) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2022 NY opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")
age_2022

# CHANGE PATH OR COMMENT THIS PART OUT
ggsave(plot=age_2022, filename = "images/ny/age_2022.png", width = 10, height = 8)

sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(ny_2022$age)$p.value) 

```

#### 2023

```{r age2023, echo=FALSE, message=FALSE, warning=FALSE}

# calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
bandwidth <- bw.nrd0(ny_2023$age) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(ny_2023$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(ny_2023$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# I will need these values for a separate graph that I can generate without data access
print("Modes:")
modes
print("Antimodes:")
antimodes

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
ny_2023_young <- subset(ny_2023, age <= antimodes[1])
ny_2023_old <- subset(ny_2023, age > antimodes[1])

# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

ny_2023$aftermode <- ny_2023$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=ny_2023, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(ny_2023$age), max(ny_2023$age), by = 5), labels = seq(min(ny_2023$age), max(ny_2023$age), by = 5), minor_breaks = seq(min(ny_2023$age), max(ny_2023$age), by = 1)) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2023 NY opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

# plot with mode lines
age_2023 <- ggplot(data=ny_2023, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") +scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(ny_2023$age), max(ny_2023$age), by = 5), labels = seq(min(ny_2023$age), max(ny_2023$age), by = 5), minor_breaks = seq(min(ny_2023$age), max(ny_2023$age), by = 1)) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2023 NY opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2023

# CHANGE PATH OR COMMENT THIS PART OUT
ggsave(plot=age_2023, filename = "images/ny/age_2023.png", width = 10, height = 8)

sprintf("Using Hartigan's dip test for multimodality, the resulting p value is %f.", dip.test(ny_2023$age)$p.value) 

 
```

### Median age

```{r median, echo=FALSE, message=FALSE, warning=FALSE}
# COMMENT OUT AS NEEDED -- Will produce NA for years without bimodality, you can comment out those

sprintf("The median age of decedents in NY in 2018 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(ny_2018$age), median(ny_2018_young$age), median(ny_2018_old$age))

sprintf("The median age of decedents in NY in 2019 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(ny_2019$age), median(ny_2019_young$age), median(ny_2019_old$age))

sprintf("The median age of decedents in NY in 2020 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(ny_2020$age), median(ny_2020_young$age), median(ny_2020_old$age))

sprintf("The median age of decedents in NY in 2021 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(ny_2021$age), median(ny_2021_young$age), median(ny_2021_old$age))

sprintf("The median age of decedents in NY in 2022 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(ny_2022$age), median(ny_2022_young$age), median(ny_2022_old$age))

sprintf("The median age of decedents in NY in 2023 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(ny_2023$age), median(ny_2023_young$age), median(ny_2023_old$age))
```

### Proportion of older people by year

```{r propOld, echo=FALSE, message=FALSE, warning=FALSE}

# COMMENT OUT AS NEEDED -- Will say the proportion is 0 for no detectable bimodality, can comment out those lines

sprintf("The proportion of older decedents in 2018 was %f, with %d out of %d deaths.", nrow(ny_2018_old)/nrow(ny_2018), nrow(ny_2018_old), nrow(ny_2018))
sprintf("The proportion of older decedents in 2019 was %f, with %d out of %d deaths.", nrow(ny_2019_old)/nrow(ny_2019), nrow(ny_2019_old), nrow(ny_2019))
sprintf("The proportion of older decedents in 2020 was %f, with %d out of %d deaths.", nrow(ny_2020_old)/nrow(ny_2020), nrow(ny_2020_old), nrow(ny_2020))
sprintf("The proportion of older decedents in 2021 was %f, with %d out of %d deaths.", nrow(ny_2021_old)/nrow(ny_2021), nrow(ny_2021_old), nrow(ny_2021))
sprintf("The proportion of older decedents in 2022 was %f, with %d out of %d deaths.", nrow(ny_2022_old)/nrow(ny_2022), nrow(ny_2022_old), nrow(ny_2022))
sprintf("The proportion of older decedents in 2023 was %f, with %d out of %d deaths.", nrow(ny_2023_old)/nrow(ny_2023), nrow(ny_2023_old), nrow(ny_2023))

```

<!-- ny lists -- for analysis purposes only --> 

```{r nyLists, echo=FALSE, message=FALSE, warning=FALSE}

# CHANGE AS NEEDED -- this is the list of years with a detectable bimodal distribution. Should only include years with identifiable bimodality
yearList <- c(2018, 2019, 2020, 2021, 2022, 2023)

# CHANGE AS NEEDED -- Should only include years with identifiable bimodality

nylistYoung <- list(ny_2018_young, ny_2019_young, ny_2020_young, ny_2021_young, ny_2022_young, ny_2023_young)

nylistOld <- list(ny_2018_old, ny_2019_old, ny_2020_old, ny_2021_old, ny_2022_old, ny_2023_old)

```

### Race of decedents {.tabset}

#### Young

```{r raceNumericYoung, echo=FALSE, message=FALSE, warning=FALSE}
# Bar chart of race/ethnicity per year for younger subpopulation

for(i in 1:length(nylistYoung)){
  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  ny <- nylistYoung[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- ny$Year[1]
  
# Automating the title, from https://stackoverflow.com/questions/59095237/automate-ggplots-while-using-variable-labels-as-title-and-axis-titles
racetitle <- paste("Racial distribution for", year, "NY opioid-detected overdose deaths in younger people")

# Plot with number of decedents per category
race <- ggplot(data=ny, aes(x=Race, ..count..)) + 
      geom_bar(aes(fill = Race), position = "dodge")  + scale_fill_manual(name = "Race", values = c("#E69F00", "#56B4E9", "#009E73", "#CC79A7"), labels = c("Black", "Hispanic", "Other", "White")) + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle(racetitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7)) 

print(race)

# Automating file names, from https://www.edureka.co/community/41963/r-programming-naming-the-output-file-using-a-variable
ggsave(plot=race, filename = paste0("images/ny/race_", year, "_young.png"), width = 10, height = 8)

# Plot with proportion of decedents per category
raceProp <- ggplot(data=ny, aes(x=Race, ..count..)) + 
      geom_bar(aes(fill = Race), position = "dodge")  + scale_fill_manual(name = "Race", values = c("#E69F00", "#56B4E9", "#009E73", "#CC79A7"), labels = c("Black", "Hispanic", "Other", "White")) + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(ny), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle(racetitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7)) 

print(raceProp)

# Save bar plot with proportions
ggsave(plot=raceProp, filename = paste0("images/ny/race_proportions_", year, "_young.png"), width = 10, height = 8)
  
}


```

#### Old 

```{r raceNumericOld, echo=FALSE, message=FALSE, warning=FALSE}
# Bar chart of race/ethnicity per year for older subpopulation

for(i in 1:length(nylistOld)){
  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  ny <- nylistOld[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- ny$Year[1]
  
  
# see if I can automate the title, from https://stackoverflow.com/questions/59095237/automate-ggplots-while-using-variable-labels-as-title-and-axis-titles
racetitle <- paste("Racial distribution for", year, "NY opioid-detected overdose deaths in older people")

# Plot with number of decedents per category
race <- ggplot(data=ny, aes(x=Race, ..count..)) + 
      geom_bar(aes(fill = Race), position = "dodge")  + scale_fill_manual(name = "Race", values = c("#E69F00", "#56B4E9", "#009E73", "#CC79A7"), labels = c("Black", "Hispanic", "Other", "White")) + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle(racetitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7)) 

# seem to need to use print() when in a for loop, from https://stackoverflow.com/questions/29053223/plots-not-working-in-for-loop
print(race)

# try to automate the filenames? from https://www.edureka.co/community/41963/r-programming-naming-the-output-file-using-a-variable
ggsave(plot=race, filename = paste0("images/ny/race_", year, "_old.png"), width = 10, height = 8)

# Plot with proportion of decedents per category
raceProp <- ggplot(data=ny, aes(x=Race, ..count..)) + 
      geom_bar(aes(fill = Race), position = "dodge")  + scale_fill_manual(name = "Race", values = c("#E69F00", "#56B4E9", "#009E73", "#CC79A7"), labels = c("Black", "Hispanic", "Other", "White")) + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(ny), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle(racetitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7)) 

print(raceProp)

# Save plot with proportions
ggsave(plot=raceProp, filename = paste0("images/ny/race_proportions_", year, "_old.png"), width = 10, height = 8)
  
}

```


### Significance testing for race {.tabset}


```{r chiSquaredRace, echo=FALSE, message=FALSE, warning=FALSE}
# Since this is categorical data, I use Fisher's exact test if >20% of cell counts are <5, if not use chi squared

# initialize empty matrix, this will hold all the p values
chisq_race <- data.frame(matrix(ncol = 2, nrow = 0))
cols <- c("Year", "p value")

for(i in 1:length(nylistYoung)){
  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  young <- nylistYoung[[i]] 
  old <- nylistOld[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- young$Year[1]
  
  # omit NAs if needed
  young <- young[!is.na(young$Race),]
  old <- old[!is.na(old$Race),]
  
  # figure out the number of young/old decedents per race/ethnicity

  racefreqtable <- young %>% group_by(Race) %>% tally()
  names(racefreqtable)[names(racefreqtable) == 'n'] <- 'Young'

  racetableold <- old %>% group_by(Race) %>% tally()
  names(racetableold)[names(racetableold) == 'n'] <- 'Old'

  racetable <- merge(racefreqtable, racetableold, by="Race")
  
  # convert the "race" column into row names so I get a table with only numeric values

  racemat <- racetable[,-1]
  rownames(racemat) <- racetable[,1]

  print(as.matrix(racemat))
  
   # append year and p value to a vector, rounding to make it more legible
  # I'm using Fisher's exact test if >20% of cell counts are <5, if not use chi squared
  # I get >20% <5 from https://www.ccjm.org/content/84/9_suppl_2/e20
  pvalue <- ifelse(sum(racemat < 5)/(ncol(racemat) * nrow(racemat)) > 0.2, round(fisher.test(as.matrix(racemat))$p.value, 4), round(chisq.test(as.matrix(racemat))$p.value, 4))
  
  results <- c(year, pvalue)
  
  # add this to the chi squared df 
  chisq_race <- rbind(chisq_race, results)

}

# for some reason it doesn't work if I do this before adding rows?
colnames(chisq_race) <- cols

chisq_race

```


### Proportion of white deaths by year

```{r raceWhite, echo=FALSE, message=FALSE, warning=FALSE}
# CHANGE AS NEEDED: whiteyoung and whiteold should only include years with detectable bimodality

# get proportion of white decedents for each year for younger people
whiteyoung <- c(sum(ny_2018_young$Race == "White")/nrow(ny_2018_young), sum(ny_2019_young$Race == "White")/nrow(ny_2019_young), sum(ny_2020_young$Race == "White")/nrow(ny_2020_young),  sum(ny_2021_young$Race == "White")/nrow(ny_2021_young), sum(ny_2022_young$Race == "White")/nrow(ny_2022_young), sum(ny_2023_young$Race == "White")/nrow(ny_2023_young))

# get proportion for older people
whiteold <- c(sum(ny_2018_old$Race == "White")/nrow(ny_2018_old), sum(ny_2019_old$Race == "White")/nrow(ny_2019_old), sum(ny_2020_old$Race == "White")/nrow(ny_2020_old),  sum(ny_2021_old$Race == "White")/nrow(ny_2021_old), sum(ny_2022_old$Race == "White")/nrow(ny_2022_old), sum(ny_2023_old$Race == "White")/nrow(ny_2023_old))

# combine vectors into df
propRace <- data.frame(yearList, whiteyoung, whiteold)

# change the col names to be more informative
names(propRace) <- c("Year", "Younger", "Older")

# Reshape data to long format
pracelong  <- propRace |>
  pivot_longer(Younger:Older)

# create labels with the value of each proportion
pracelong$labels <- round(pracelong$value, 2)

# Create a line graph with repelling labels
pRace <-  ggplot(data = pracelong, aes(x = Year, y = value, fill = name, group=name, color=factor(name))) +
  geom_line(stat = "identity", size = 1) + geom_point(size = 2) + theme_light() + ggtitle("Proportion of White opioid-detected overdose deaths \n by age group in NY, 2018-2023") + ylab("Proportion") + theme(plot.title = element_text(hjust = 0.5)) + scale_fill_discrete(guide="none") + scale_x_continuous(breaks=propRace$Year) + xlab("Year")+ coord_cartesian(ylim = c(0, 1)) + guides(color = guide_legend(title = "Age group"))+ geom_label_repel(aes(label = labels, x = Year, y = value), nudge_x = 0.1, nudge_y = 0.1, na.rm = FALSE, color = "black", size = 2.5, label.size = NA, fill = NA)

pRace

ggsave(plot=pRace, filename = "images/ny/prop_white_allyears.png", width = 10, height = 8)

```


### Significance testing for white deaths

```{r chiSquaredWhite, echo=FALSE, message=FALSE, warning=FALSE}

# initialize empty matrix, this will hold all the p values

chisq_white <- data.frame(matrix(ncol = 2, nrow = 0))
cols <- c("Year", "Younger", "Older")

for(i in 1:length(nylistYoung)){
  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  young <- nylistYoung[[i]] 
  old <- nylistOld[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- young$Year[1]
  
  # figure out the number of white young/old decedents

  racefreqtable <- subset(young, Race == "White") %>% group_by(Race) %>% tally()
  names(racefreqtable)[names(racefreqtable) == 'n'] <- 'Young'

  racetableold <- subset(old, Race == "White") %>% group_by(Race) %>% tally()
  names(racetableold)[names(racetableold) == 'n'] <- 'Old'

  racetable <- merge(racefreqtable, racetableold, by="Race")
  
  # convert the first column into row names so I get a table with only numeric values

  racemat <- racetable[,-1]
  
  # add to matrix with all the white results 
  
  results <- cbind(year, racemat)
  
  # add this to the chi squared df 
  chisq_white <- rbind(chisq_white, results)

}

# for some reason it doesn't work if I do this before adding rows?
colnames(chisq_white) <- cols
chisq_white

# get rid of the first column [row labels for readability]

racemat <- as.matrix(chisq_white[,-1])
  
  # run Fisher's exact test if >20% of cell counts are <5, if not use chi squared
  pvalue <- ifelse(sum(racemat < 5)/(ncol(racemat) * nrow(racemat)) > 0.2, round(fisher.test(as.matrix(racemat))$p.value, 4), round(chisq.test(as.matrix(racemat))$p.value, 4))

pvalue

```



### Proportion of Latine deaths

```{r raceLatine, echo=FALSE, message=FALSE, warning=FALSE}
# CHANGE AS NEEDED: latineyoung and latineold should only include years with detectable bimodality

# get proportion of Latine decedents for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
latineyoung <- c(sum(ny_2018_young$Race == "Hispanic")/nrow(ny_2018_young), sum(ny_2019_young$Race == "Hispanic")/nrow(ny_2019_young), sum(ny_2020_young$Race == "Hispanic")/nrow(ny_2020_young),  sum(ny_2021_young$Race == "Hispanic")/nrow(ny_2021_young), sum(ny_2022_young$Race == "Hispanic")/nrow(ny_2022_young), sum(ny_2023_young$Race == "Hispanic")/nrow(ny_2023_young))

# get proportion for older people
latineold <- c(sum(ny_2018_old$Race == "Hispanic")/nrow(ny_2018_old), sum(ny_2019_old$Race == "Hispanic")/nrow(ny_2019_old), sum(ny_2020_old$Race == "Hispanic")/nrow(ny_2020_old),  sum(ny_2021_old$Race == "Hispanic")/nrow(ny_2021_old), sum(ny_2022_old$Race == "Hispanic")/nrow(ny_2022_old), sum(ny_2023_old$Race == "Hispanic")/nrow(ny_2023_old))

# combine vectors into df
propRace <- data.frame(yearList, latineyoung, latineold)

# change the col names to be more informative
names(propRace) <- c("Year", "Younger", "Older")

# Reshape data to long format

pracelong  <- propRace |>
  pivot_longer(Younger:Older)

# create labels with the value of each proportion
pracelong$labels <- round(pracelong$value, 2)

# Create a line graph with repelling labels
pRace <-  ggplot(data = pracelong, aes(x = Year, y = value, fill = name, group=name, color=factor(name))) +
  geom_line(stat = "identity", size = 1) + geom_point(size = 2) + theme_light() + ggtitle("Proportion of Latine opioid-detected overdose deaths \n by age group in NY, 2018-2023") + ylab("Proportion") + theme(plot.title = element_text(hjust = 0.5)) + scale_fill_discrete(guide="none") + scale_x_continuous(breaks=propRace$Year) + xlab("Year")+ coord_cartesian(ylim = c(0, 1)) + guides(color = guide_legend(title = "Age group"))+ geom_label_repel(aes(label = labels, x = Year, y = value), nudge_x = 0.1, nudge_y = 0.1, na.rm = FALSE, color = "black", size = 2.5, label.size = NA, fill = NA)

pRace

ggsave(plot=pRace, filename = "images/ny/prop_latine_allyears.png", width = 10, height = 8)

```


### Significance testing for Latine deaths

```{r chiSquaredLatine, echo=FALSE, message=FALSE, warning=FALSE}

# initialize empty matrix, this will hold all the p values
chisq_latine <- data.frame(matrix(ncol = 2, nrow = 0))
cols <- c("Year", "Younger", "Older")

for(i in 1:length(nylistYoung)){
  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  young <- nylistYoung[[i]] 
  old <- nylistOld[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- young$Year[1]
  
  # figure out the number of Latine young/old decedents

  racefreqtable <- subset(young, Race == "Hispanic") %>% group_by(Race) %>% tally()
  names(racefreqtable)[names(racefreqtable) == 'n'] <- 'Young'

  racetableold <- subset(old, Race == "Hispanic") %>% group_by(Race) %>% tally()
  names(racetableold)[names(racetableold) == 'n'] <- 'Old'

  racetable <- merge(racefreqtable, racetableold, by="Race")
  
 # convert the first column into row names so I get a table with only numeric values

  racemat <- racetable[,-1]
  
  # add to matrix with all the Latine results 
  
  results <- cbind(year, racemat)
  
  # add this to the chi squared df 
  chisq_latine <- rbind(chisq_latine, results)

}

# for some reason it doesn't work if I do this before adding rows?
colnames(chisq_latine) <- cols

chisq_latine

# get rid of the first column [row labels for readability]

racemat <- as.matrix(chisq_latine[,-1])

# run Fisher's exact test if >20% of cell counts are <5, if not use chi squared

  pvalue <- ifelse(sum(racemat < 5)/(ncol(racemat) * nrow(racemat)) > 0.2, round(fisher.test(as.matrix(racemat))$p.value, 4), round(chisq.test(as.matrix(racemat))$p.value, 4))

pvalue
```

### Proportion of Black deaths

```{r raceBlack, echo=FALSE, message=FALSE, warning=FALSE}

# CHANGE AS NEEDED: blackyoung and blackold should only include years with detectable bimodality

# get proportion of Race for each year for younger people 
blackyoung <- c(sum(ny_2018_young$Race == "Black")/nrow(ny_2018_young), sum(ny_2019_young$Race == "Black")/nrow(ny_2019_young), sum(ny_2020_young$Race == "Black")/nrow(ny_2020_young),  sum(ny_2021_young$Race == "Black")/nrow(ny_2021_young), sum(ny_2022_young$Race == "Black")/nrow(ny_2022_young), sum(ny_2023_young$Race == "Black")/nrow(ny_2023_young))

# get proportion for older people
blackold <- c(sum(ny_2018_old$Race == "Black")/nrow(ny_2018_old), sum(ny_2019_old$Race == "Black")/nrow(ny_2019_old), sum(ny_2020_old$Race == "Black")/nrow(ny_2020_old),  sum(ny_2021_old$Race == "Black")/nrow(ny_2021_old), sum(ny_2022_old$Race == "Black")/nrow(ny_2022_old), sum(ny_2023_old$Race == "Black")/nrow(ny_2023_old))

# combine vectors into df
propRace <- data.frame(yearList, blackyoung, blackold)

# change the col names to be more informative
names(propRace) <- c("Year", "Younger", "Older")

# Reshape data to long format
pracelong  <- propRace |>
  pivot_longer(Younger:Older)

# create labels with the value of each proportion
pracelong$labels <- round(pracelong$value, 2)

# Create a line graph with repelling labels
pRace <-  ggplot(data = pracelong, aes(x = Year, y = value, fill = name, group=name, color=factor(name))) +
  geom_line(stat = "identity", size = 1) + geom_point(size = 2) + theme_light() + ggtitle("Proportion of Black opioid-detected overdose deaths \n by age group in NY, 2018-2023") + ylab("Proportion") + theme(plot.title = element_text(hjust = 0.5)) + scale_fill_discrete(guide="none") + scale_x_continuous(breaks=propRace$Year) + xlab("Year")+ coord_cartesian(ylim = c(0, 1)) + guides(color = guide_legend(title = "Age group"))+ geom_label_repel(aes(label = labels, x = Year, y = value), nudge_x = 0.1, nudge_y = 0.1, na.rm = FALSE, color = "black", size = 2.5, label.size = NA, fill = NA)

pRace

ggsave(plot=pRace, filename = "images/ny/prop_black_allyears.png", width = 10, height = 8)

```


### Significance testing for Black deaths

```{r chiSquaredBlack, echo=FALSE, message=FALSE, warning=FALSE}

# initialize empty matrix, this will hold all the p values
chisq_black <- data.frame(matrix(ncol = 2, nrow = 0))
cols <- c("Year", "Younger", "Older")
 
 for(i in 1:length(nylistYoung)){
   # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
   young <- nylistYoung[[i]] 
   old <- nylistOld[[i]] 
   # grab the year -- you only need the first element in the year column since all of them have the same value
   year <- young$Year[1]
   
  # figure out the number of young/old Black decedents

   racefreqtable <- subset(young, Race == "Black") %>% group_by(Race) %>% tally()
   names(racefreqtable)[names(racefreqtable) == 'n'] <- 'Young'
 
   racetableold <- subset(old, Race == "Black") %>% group_by(Race) %>% tally()
   names(racetableold)[names(racetableold) == 'n'] <- 'Old'
 
   racetable <- merge(racefreqtable, racetableold, by="Race")
   
   # convert the first column into row names so I get a table with only numeric values
 
   racemat <- racetable[,-1]

   # add to matrix with all the Black results 
  
   results <- cbind(year, racemat)
  
   # add this to the chi squared df 
   chisq_black <- rbind(chisq_black, results)
 
 }
 
 # for some reason it doesn't work if I do this before adding rows?
 colnames(chisq_black) <- cols
 
 chisq_black

# get rid of the first column [row labels for readability]

racemat <- as.matrix(chisq_black[,-1])

# run Fisher's exact test if >20% of cell counts are <5, if not use chi squared

  pvalue <- ifelse(sum(racemat < 5)/(ncol(racemat) * nrow(racemat)) > 0.2, round(fisher.test(as.matrix(racemat))$p.value, 4), round(chisq.test(as.matrix(racemat))$p.value, 4))

pvalue

```

### Sex of decedents/proportion male decedents 

```{r sexAllYears, echo=FALSE, message=FALSE, warning=FALSE}

# CHANGE AS NEEDED -- only include years with bimodality

# get proportion of Sex_revised for each year for younger people 
sexyoung <- c(sum(ny_2018_young$Sex_revised == "Male")/nrow(ny_2018_young), sum(ny_2019_young$Sex_revised == "Male")/nrow(ny_2019_young), sum(ny_2020_young$Sex_revised == "Male")/nrow(ny_2020_young), sum(ny_2021_young$Sex_revised == "Male")/nrow(ny_2021_young), sum(ny_2022_young$Sex_revised == "Male")/nrow(ny_2022_young), sum(ny_2023_young$Sex_revised == "Male")/nrow(ny_2023_young))

# get proportion for older people
sexold <- c(sum(ny_2018_old$Sex_revised == "Male")/nrow(ny_2018_old), sum(ny_2019_old$Sex_revised == "Male")/nrow(ny_2019_old), sum(ny_2020_old$Sex_revised == "Male")/nrow(ny_2020_old), sum(ny_2021_old$Sex_revised == "Male")/nrow(ny_2021_old), sum(ny_2022_old$Sex_revised == "Male")/nrow(ny_2022_old), sum(ny_2023_old$Sex_revised == "Male")/nrow(ny_2023_old))


# combine vectors into df
propSex_revised <- data.frame(yearList, sexyoung, sexold)

# change the col names to be more informative
names(propSex_revised) <- c("Year", "Younger", "Older")

# Reshape data to long format
psexlong  <- propSex_revised |>
  pivot_longer(Younger:Older)

# create labels with the value of each proportion
psexlong$labels <- round(psexlong$value, 2)

# Create line plot with repelling labels
pSex_revised <-  ggplot(data = psexlong, aes(x = Year, y = value, fill = name, group=name, color=factor(name))) +
  geom_line(stat = "identity", size = 1) + geom_point(size = 2) + theme_light() + ggtitle("Proportion of male deaths in NY by age group, 2018-2023") + ylab("Proportion") + theme(plot.title = element_text(hjust = 0.5)) + scale_fill_discrete(guide="none") + scale_x_continuous(breaks=propSex_revised$Year) + xlab("Year")+ coord_cartesian(ylim = c(0, 1)) + guides(color = guide_legend(title = "Age group"))+ geom_label_repel(aes(label = labels, x = Year, y = value), nudge_x = 0.1, na.rm = FALSE, color = "black", size = 2.5, label.size = NA, fill = NA)

pSex_revised

ggsave(plot=pSex_revised, filename = "images/ny/prop_sex_allyears.png", width = 10, height = 8)

```

### Significance testing for proportion of male decedents

```{r chiSquaredSexny, echo=FALSE, message=FALSE, warning=FALSE}
# initialize empty matrix, from https://stackoverflow.com/questions/32712301/create-empty-data-frame-with-column-names-by-assigning-a-string-vector
# this will hold all the p values

chisq_sex <- data.frame(matrix(ncol = 2, nrow = 0))
cols <- c("Year", "p value")

for(i in 1:length(nylistYoung)){
  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  young <- nylistYoung[[i]] 
  old <- nylistOld[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- young$Year[1]
  
  # figure out the number of young/old male decedents

  sexfreqtable <- young %>% group_by(Sex_revised) %>% tally()
  names(sexfreqtable)[names(sexfreqtable) == 'n'] <- 'Young'

  sextableold <- old %>% group_by(Sex_revised) %>% tally()
  names(sextableold)[names(sextableold) == 'n'] <- 'Old'

  sextable <- merge(sexfreqtable, sextableold, by="Sex_revised")
  
  # convert the first column into row names so I get a table with only numeric values
  sexmat <- sextable[,-1]
  rownames(sexmat) <- sextable[,1]

  print(as.matrix(sexmat))
  
  # append year and p value to a vector, rounding to make it more legible
  # I'm using Fisher's exact test if >20% of cell counts are <5, if not use chi squared
  # I get >20% <5 from https://www.ccjm.org/content/84/9_suppl_2/e20
  pvalue <- ifelse(sum(sexmat < 5)/(ncol(sexmat) * nrow(sexmat)) > 0.2, round(fisher.test(as.matrix(sexmat))$p.value, 4), round(chisq.test(as.matrix(sexmat))$p.value, 4))
  
  results <- c(year, pvalue)
  
  # add this to the chi squared df 
  chisq_sex <- rbind(chisq_sex, results)

}

# for some reason it doesn't work if I do this before adding rows?
colnames(chisq_sex) <- cols

chisq_sex

```

### Race and sex of decedents {.tabset}

#### Young

```{r sexRaceTableYoung, echo=FALSE, message=FALSE, warning=FALSE}

# create dummy df with race sex info for all years
racesexmat <- data.frame(matrix(ncol = 4, nrow = 0))
cols <- c("Year", "Race_sex", "Proportion", "Color")

for(i in 1:length(nylistYoung)){

  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  ny <- nylistYoung[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- ny$Year[1]
  # remove unknowns to comply with suppression guidelines, since there's only 1 decedent with unknown sex
  ny <- subset(ny, Sex_revised != "Unknown")

raceSextitle <- paste("Race/ethnicity and sex distribution for", year, "NY opioid-detected \n overdose deaths in younger people")

# bar chart with the number of decedents per race-sex  
raceSex <- ggplot(data=ny, aes(x=Race_sex, ..count..)) + 
      geom_bar(aes(fill = Race_sex), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = -0.25, colour = "black", size = 7/.pt)+ ggtitle(raceSextitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7)) 

print(raceSex)

ggsave(plot=raceSex, filename = paste0("images/ny/raceSex", year, "_young.png"), width = 10, height = 8)

# bar chart with the proportion of decedents per race-sex
raceSexProp <- ggplot(data=ny, aes(x=Race_sex, ..count..)) + 
      geom_bar(aes(fill = Race_sex), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(ny), 3)), stat = "count", vjust = -0.25, colour = "black", size = 7/.pt)+ ggtitle(raceSextitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7)) 

print(raceSexProp)
ggsave(plot=raceSexProp, filename = paste0("images/ny/raceSexProp", year, "_young.png"), width = 10, height = 8)

# pie chart of race-sex per year

# get the % of each race sex category
ny <- ny %>% 
  group_by(Race_sex) %>% 
  count() %>% 
  ungroup() %>% 
  mutate(per=`n`/sum(`n`)) %>% 
  arrange(desc(Race_sex))
# create a label out of this
ny$racesexlabel <- scales::percent(ny$per)

# create a year label [for plotting]
ny$Year <- year

# automatically assign the color based on race sex value -- not doing this in preprocessing because I don't want this to be a global change
ny$color <- ifelse(ny$Race_sex == "Black Female", "#e6d0a1", ifelse(ny$Race_sex == "Black Male", "#E69F00", ifelse(ny$Race_sex == "Hispanic Female", "#b3d4e8", ifelse(ny$Race_sex == "Hispanic Male", "#56B4E9", ifelse(ny$Race_sex == "Other Female", "#cca1b9", ifelse(ny$Race_sex == "Other Male", "#CC79A7", ifelse(ny$Race_sex == "White Female", "#a1a1a1", "#000000")))))))


# grab the values i need from ny
results <- select(ny, Year, Race_sex, per, color)
  
# add this to the chi squared df 
racesexmat <- rbind(racesexmat, results)

# plot into a pie chart -- for some reason I have to reverse the colors to get them to work right
piechart <- ggplot(data=ny)+
  geom_bar(aes(x="", y=per, fill=Race_sex), stat="identity", width = 1)+
  coord_polar("y", start=0)+
  theme_void()+ geom_label_repel(aes(label = racesexlabel, x = "", y = cumsum(per) - per/2), nudge_x = 0.02,
                   size = 3, show.legend = F)+  scale_fill_manual(name = "Race/ethnicity and sex", values=rev(ny$color), labels = rev(ny$Race_sex))+ ggtitle(raceSextitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="right")  

print(piechart)

ggsave(plot=piechart, filename = paste0("images/ny/pie", year, "_young.png"), width = 10, height = 8)

}

# rename the columns to be more informative
colnames(racesexmat) <- cols

# stacked area plot depicting the change in race-sex year by year
stackedarea <- ggplot(data = racesexmat, aes(x=Year, y=Proportion, fill=Race_sex)) + 
    geom_area(aes(position = 'stack', group = Race_sex))+  scale_fill_manual(name = "Race/ethnicity and sex", values=rev(racesexmat$Color), labels = rev(racesexmat$Race_sex)) + ggtitle("Race/ethnicity and sex distribution for NY opioid-detected \n overdose deaths in younger people")+theme_bw()+ scale_x_continuous(n.breaks = length(yearList))+theme(plot.title = element_text(hjust = 0.5), legend.position="right")

print(stackedarea)

ggsave(plot=stackedarea, filename = paste0("images/ny/rs_area_", year, "_young.png"), width = 10, height = 8)
```


#### Old

```{r sexRaceTableOld, echo=FALSE, message=FALSE, warning=FALSE}

# create dummy df with race sex info for all years
racesexmat <- data.frame(matrix(ncol = 4, nrow = 0))
cols <- c("Year", "Race_sex", "Proportion", "Color")

for(i in 1:length(nylistOld)){

  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  ny <- nylistOld[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- ny$Year[1]
  # remove unknowns to comply with suppression guidelines, since there's only 1 decedent with unknown sex
  ny <- subset(ny, Sex_revised != "Unknown")

raceSextitle <- paste("Race/ethnicity and sex distribution for", year, "NY opioid-detected \n overdose deaths in older people")

# bar chart with the number of decedents per race-sex
raceSex <- ggplot(data=ny, aes(x=Race_sex, ..count..)) + 
      geom_bar(aes(fill = Race_sex), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = -0.25, colour = "black", size = 7/.pt)+ ggtitle(raceSextitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7)) 

print(raceSex)

ggsave(plot=raceSex, filename = paste0("images/ny/raceSex", year, "_old.png"), width = 10, height = 8)

# bar chart with the proportion of decedents per race-sex
raceSexProp <- ggplot(data=ny, aes(x=Race_sex, ..count..)) + 
      geom_bar(aes(fill = Race_sex), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(ny), 3)), stat = "count", vjust = -0.25, colour = "black", size = 7/.pt)+ ggtitle(raceSextitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7)) 

print(raceSexProp)

ggsave(plot=raceSexProp, filename = paste0("images/ny/raceSexProp", year, "_old.png"), width = 10, height = 8)


# pie chart of race-sex per year

# get the % of each race sex category
ny <- ny %>% 
  group_by(Race_sex) %>% 
  count() %>% 
  ungroup() %>% 
  mutate(per=`n`/sum(`n`)) %>% 
  arrange(desc(Race_sex))
# create a label out of this
ny$racesexlabel <- scales::percent(ny$per)

# create a year label [for plotting]
ny$Year <- year

# automatically assign the color based on race sex value -- not doing this in preprocessing because I don't want this to be a global change
ny$color <- ifelse(ny$Race_sex == "Black Female", "#e6d0a1", ifelse(ny$Race_sex == "Black Male", "#E69F00", ifelse(ny$Race_sex == "Hispanic Female", "#b3d4e8", ifelse(ny$Race_sex == "Hispanic Male", "#56B4E9", ifelse(ny$Race_sex == "Other Female", "#cca1b9", ifelse(ny$Race_sex == "Other Male", "#CC79A7", ifelse(ny$Race_sex == "White Female", "#a1a1a1", "#000000")))))))


# grab the values i need from nc
results <- select(ny, Year, Race_sex, per, color)
  
# add this to the chi squared df 
racesexmat <- rbind(racesexmat, results)

# plot into a pie chart -- for some reason I have to reverse the colors to get them to work right
piechart <- ggplot(data=ny)+
  geom_bar(aes(x="", y=per, fill=Race_sex), stat="identity", width = 1)+
  coord_polar("y", start=0)+
  theme_void()+ geom_label_repel(aes(label = racesexlabel, x = "", y = cumsum(per) - per/2), nudge_x = 0.02,
                   size = 3, show.legend = F)+  scale_fill_manual(name = "Race/ethnicity and sex", values=rev(ny$color), labels = rev(ny$Race_sex))+ ggtitle(raceSextitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="right")  

print(piechart)

ggsave(plot=piechart, filename = paste0("images/ny/pie", year, "_young.png"), width = 10, height = 8)

}

# rename the columns to be more informative
colnames(racesexmat) <- cols

# stacked area plot depicting the change in race-sex year by year
stackedarea <- ggplot(data = racesexmat, aes(x=Year, y=Proportion, fill=Race_sex)) + 
    geom_area(aes(position = 'stack', group = Race_sex))+  scale_fill_manual(name = "Race/ethnicity and sex", values=rev(racesexmat$Color), labels = rev(racesexmat$Race_sex)) + ggtitle("Race/ethnicity and sex distribution for NY opioid-detected \n overdose deaths in older people")+theme_bw()+ scale_x_continuous(n.breaks = length(yearList))+theme(plot.title = element_text(hjust = 0.5), legend.position="right")

print(stackedarea)

ggsave(plot=stackedarea, filename = paste0("images/ny/rs_area_", year, "_old.png"), width = 10, height = 8)
```

### Significance testing for race and sex {.tabset}


```{r chiSquaredRaceSex, echo=FALSE, message=FALSE, warning=FALSE}
# initialize empty matrix, from https://stackoverflow.com/questions/32712301/create-empty-data-frame-with-column-names-by-assigning-a-string-vector
# this will hold all the p values

chisq_racesex <- data.frame(matrix(ncol = 2, nrow = 0))
cols <- c("Year", "p value")

for(i in 1:length(nylistYoung)){
  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  young <- nylistYoung[[i]] 
  old <- nylistOld[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- young$Year[1]
  
  # figure out the number of young/old decedents per race-sex
  racesexfreqtable <- young %>% group_by(Race_sex) %>% tally()
  names(racesexfreqtable)[names(racesexfreqtable) == 'n'] <- 'Young'

  racesextableold <- old %>% group_by(Race_sex) %>% tally()
  names(racesextableold)[names(racesextableold) == 'n'] <- 'Old'

  racesextable <- merge(racesexfreqtable, racesextableold, by="Race_sex")
  
  # convert the first column into row names so I get a table with only numeric values

  racesexmat <- racesextable[,-1]
  rownames(racesexmat) <- racesextable[,1]

  print(as.matrix(racesexmat))
  
  # append year and p value to a vector, rounding to make it more legible
  # I'm using Fisher's exact test if >20% of cell counts are <5, if not use chi squared
  # I get >20% <5 from https://www.ccjm.org/content/84/9_suppl_2/e20
  pvalue <- ifelse(sum(racesexmat < 5)/(ncol(racesexmat) * nrow(racesexmat)) > 0.2, round(fisher.test(as.matrix(racesexmat))$p.value, 4), round(chisq.test(as.matrix(racesexmat))$p.value, 4))
  
  results <- c(year, pvalue)
  
  # add this to the chi squared df 
  chisq_racesex <- rbind(chisq_racesex, results)

}

# for some reason it doesn't work if I do this before adding rows?
colnames(chisq_racesex) <- cols

chisq_racesex

```

### Day of week {.tabset}

#### Young

```{r weekYoung, echo=FALSE, message=FALSE, warning=FALSE}
for(i in 1:length(nylistYoung)){

  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  ny <- nylistYoung[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- ny$Year[1]
  # remove unknowns, there are a lot but they don't contribute anything
  ny <- subset(ny, Weekday %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))

weekdaytitle <- paste("ODs per weekday for", year, "NY opioid-detected overdose deaths in younger people")
  
# bar plot with number of decedents per day 
weekday <- ggplot(data=ny, aes(x=factor(Weekday, level=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")), ..count..)) + geom_bar(aes(fill = as.character(Weekday)), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 0), breaks = ny$Weekday, labels = ny$Weekday) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle(weekdaytitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))+xlab("Weekday")

print(weekday)

ggsave(plot=weekday, filename = paste0("images/ny/weekday_", year, "_young.png"), width = 10, height = 8)

# bar plot with proportion of decedents per day
weekday_prop <- ggplot(data=ny, aes(x=factor(Weekday, level=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")), round(..count../nrow(ny), 3))) + geom_bar(aes(fill = as.character(Weekday)), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 0), breaks = ny$Weekday, labels = ny$Weekday) + theme_light() + geom_text(aes(label = round(..count../nrow(ny), 3)), stat = "count", vjust = 0, colour = "black") + ggtitle(weekdaytitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))+xlab("Weekday")+ylab("Proportion")

print(weekday_prop)

ggsave(plot=weekday_prop, filename = paste0("images/ny/weekday_", year, "_young_prop.png"), width = 10, height = 8)
}

```


#### Old

```{r weekOld, echo=FALSE, message=FALSE, warning=FALSE}
for(i in 1:length(nylistOld)){

  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  ny <- nylistOld[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- ny$Year[1]
  # remove unknowns, there are a lot but they don't contribute anything
  ny <- subset(ny, Weekday %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))

weekdaytitle <- paste("ODs per weekday for", year, "NY opioid-detected overdose deaths in older people")

# bar plot of number of decedents per day
weekday <- ggplot(data=ny, aes(x=factor(Weekday, level=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")), ..count..)) + geom_bar(aes(fill = as.character(Weekday)), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 0), breaks = ny$Weekday, labels = ny$Weekday) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle(weekdaytitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))+xlab("Weekday")

print(weekday)

ggsave(plot=weekday, filename = paste0("images/ny/weekday_", year, "_old.png"), width = 10, height = 8)

# bar plot of proportion of decedents per day
weekday_prop <- ggplot(data=ny, aes(x=factor(Weekday, level=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")), round(..count../nrow(ny), 3))) + geom_bar(aes(fill = as.character(Weekday)), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 0), breaks = ny$Weekday, labels = ny$Weekday) + theme_light() + geom_text(aes(label = round(..count../nrow(ny), 3)), stat = "count", vjust = 0, colour = "black") + ggtitle(weekdaytitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))+xlab("Weekday")+ylab("Proportion")

print(weekday_prop)

ggsave(plot=weekday_prop, filename = paste0("images/ny/weekday_", year, "_old_prop.png"), width = 10, height = 8)
}

```

### Significance testing for day of week {.tabset}

```{r chiSquaredWeekday, echo=FALSE, message=FALSE, warning=FALSE}
# initialize empty matrix, from https://stackoverflow.com/questions/32712301/create-empty-data-frame-with-column-names-by-assigning-a-string-vector
# this will hold all the p values

chisq_weekday <- data.frame(matrix(ncol = 2, nrow = 0))
cols <- c("Year", "p value")

for(i in 1:length(nylistYoung)){
  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  young <- nylistYoung[[i]] 
  old <- nylistOld[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- young$Year[1]
  
  # omit NAs if needed, and get rid of other values that aren't days of the week
  young <- young[!is.na(young$Weekday),]
  young <- subset(young, Weekday %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
  old <- old[!is.na(old$Weekday),]
  old <- subset(old, Weekday %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))

  # figure out the number of young/old decedents per day of week of death
  
  weekdayfreqtable <- young %>% group_by(Weekday) %>% tally()
  names(weekdayfreqtable)[names(weekdayfreqtable) == 'n'] <- 'Young'

  weekdaytableold <- old %>% group_by(Weekday) %>% tally()
  names(weekdaytableold)[names(weekdaytableold) == 'n'] <- 'Old'

  weekdaytable <- merge(weekdayfreqtable, weekdaytableold, by="Weekday")
  
  # convert the first column into row names so I get a table with only numeric values

  weekdaymat <- weekdaytable[,-1]
  rownames(weekdaymat) <- weekdaytable[,1]

  print(as.matrix(weekdaymat))
  
  # append year and p value to a vector, rounding to make it more legible
  # I'm using Fisher's exact test if >20% of cell counts are <5, if not use chi squared
  # I get >20% <5 from https://www.ccjm.org/content/84/9_suppl_2/e20
  pvalue <- ifelse(sum(weekdaymat < 5)/(ncol(weekdaymat) * nrow(weekdaymat)) > 0.2, round(fisher.test(as.matrix(weekdaymat))$p.value, 4), round(chisq.test(as.matrix(weekdaymat))$p.value, 4))
  
  results <- c(year, pvalue)
  
  # add this to the chi squared df 
  chisq_weekday <- rbind(chisq_weekday, results)

}

# for some reason it doesn't work if I do this before adding rows?
colnames(chisq_weekday) <- cols

chisq_weekday

```

### Month {.tabset}

#### Young

```{r monthYoung, echo=FALSE, message=FALSE, warning=FALSE}
for(i in 1:length(nylistYoung)){

  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  ny <- nylistYoung[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- ny$Year[1]
  
  monthtitle <- paste("ODs per month for", year, "NY opioid-detected overdose deaths in younger people")
  
  # get rid of the 99s -- not doing it in preprocessing because records with 99 months may have other useful info
  ny <- subset(ny, Month != "99")

# bar plot of number of decedents per month
month <- ggplot(data=ny, aes(x=Month, ..count..)) + geom_bar(aes(fill = as.character(Month)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0), breaks = ny$Month, labels = ny$Month) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle(monthtitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))

print(month)

ggsave(plot=month, filename = paste0("images/ny/month_", year, "_young.png"), width = 10, height = 8)

# bar plot of proportion of decedents per month
month_prop <- ggplot(data=ny, aes(x=Month, round(..count../nrow(ny), 3))) + geom_bar(aes(fill = as.character(Month)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0), breaks = ny$Month, labels = ny$Month) + theme_light() + geom_text(aes(label =  round(..count../nrow(ny), 3)), stat = "count", vjust = 0, colour = "black") + ggtitle(monthtitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))+ylab("Proportion")

print(month_prop)

ggsave(plot=month_prop, filename = paste0("images/ny/month_", year, "_young_prop.png"), width = 10, height = 8)


}

```


#### Old

```{r monthOld, echo=FALSE, message=FALSE, warning=FALSE}
for(i in 1:length(nylistOld)){

  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  ny <- nylistOld[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- ny$Year[1]
  
  monthtitle <- paste("ODs per month for", year, "NY opioid-detected overdose deaths in older people")

  # get rid of the 99s -- not doing it in preprocessing because records with 99 months may have other useful info
  ny <- subset(ny, Month != "99")
  
# bar plot of number of decedents per month
month <- ggplot(data=ny, aes(x=Month, ..count..)) + geom_bar(aes(fill = as.character(Month)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0), breaks = ny$Month, labels = ny$Month) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle(monthtitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))

print(month)

ggsave(plot=month, filename = paste0("images/ny/month_", year, "_old.png"), width = 10, height = 8)

# bar plot of proportion of decedents per month
month_prop <- ggplot(data=ny, aes(x=Month, round(..count../nrow(ny), 3))) + geom_bar(aes(fill = as.character(Month)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0), breaks = ny$Month, labels = ny$Month) + theme_light() + geom_text(aes(label =  round(..count../nrow(ny), 3)), stat = "count", vjust = 0, colour = "black") + ggtitle(monthtitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))+ylab("Proportion")

print(month_prop)

ggsave(plot=month_prop, filename = paste0("images/ny/month_", year, "_old_prop.png"), width = 10, height = 8)


}

```

### Significance testing for month {.tabset}

```{r chiSquaredMonth, echo=FALSE, message=FALSE, warning=FALSE}
# initialize empty matrix, from https://stackoverflow.com/questions/32712301/create-empty-data-frame-with-column-names-by-assigning-a-string-vector
# this will hold all the p values

chisq_month <- data.frame(matrix(ncol = 2, nrow = 0))
cols <- c("Year", "p value")

for(i in 1:length(nylistYoung)){
  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  young <- nylistYoung[[i]] 
  old <- nylistOld[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- young$Year[1]
  
  # omit NAs and 99s if needed
  young <- young[!is.na(young$Month),]
  young <- subset(young, Month != "99")
  old <- old[!is.na(old$Month),]
  old <- subset(old, Month != "99")
  
  # figure out the number of young/old decedents per month of death
  monthfreqtable <- young %>% group_by(Month) %>% tally()
  names(monthfreqtable)[names(monthfreqtable) == 'n'] <- 'Young'

  monthtableold <- old %>% group_by(Month) %>% tally()
  names(monthtableold)[names(monthtableold) == 'n'] <- 'Old'

  monthtable <- merge(monthfreqtable, monthtableold, by="Month")
  
  # convert the first column into row names so I get a table with only numeric values

  monthmat <- monthtable[,-1]
  rownames(monthmat) <- monthtable[,1]

  print(as.matrix(monthmat))

  # append year and p value to a vector, rounding to make it more legible
  # I'm using Fisher's exact test if >20% of cell counts are <5, if not use chi squared
  # I get >20% <5 from https://www.ccjm.org/content/84/9_suppl_2/e20
  pvalue <- ifelse(sum(monthmat < 5)/(ncol(monthmat) * nrow(monthmat)) > 0.2, round(fisher.test(as.matrix(monthmat))$p.value, 4), round(chisq.test(as.matrix(monthmat))$p.value, 4))
  
  results <- c(year, pvalue)
  
  # add this to the chi squared df 
  chisq_month <- rbind(chisq_month, results)

}

# for some reason it doesn't work if I do this before adding rows?
colnames(chisq_month) <- cols

chisq_month

```

