---
title: "Age distribution of fatal opioid-detected overdoses in Vermont, 2009-2023"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 5
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r sourceData, echo=FALSE, message=FALSE, warning=FALSE}
library(tigris)
library(ggplot2)
library(stringr)
library(dplyr)
library(lubridate)
library(patchwork)
library(tidyverse) #to replace NA with something else
library(sf)
library(multimode) #to calculate modes for the age plot
library(findSVI) #for SVI calculation
library(diptest)
library(ggrepel)

# get the most updated data files
source("data_preprocessing/vt_processing.R")
# set working directory -- I have to do this after sourcing the data because it loses the path otherwise
setwd("C:\\Users\\ar2529\\Desktop\\overdose_age")

```

```{r importData, echo=FALSE, message=FALSE, warning=FALSE}

# import vt data 
vt_data <- read.csv("data_preprocessing/processed_data/vt_processed.csv")


# split the dataset by year [so we can use it to split into old/young age], from https://stackoverflow.com/questions/57454085/assign-a-variable-in-r-using-another-variable

for(i in unique(vt_data$Year)){
  assign(paste0("vt_", as.character(i)), subset(vt_data, Year == as.character(i)))
}


# get town data from tigris so we can map deaths by town later

options(tigris_use_cache=TRUE)
vt_towns = county_subdivisions(state="Vermont")
vt_towns$NAME = str_to_title(vt_towns$NAME)
vt_towns = filter(vt_towns, NAME != "County Subdivisions Not Defined")
# tigris also has shapefiles for the ocean so let's get rid of that
vt_towns <- subset(vt_towns, ALAND!=0)
```

### Age distribution for all fatal overdoses {.tabset}

#### 2009

```{r age2009, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
# looks like there are some NAs in the 2009 data so I have to get rid of them
bandwidth <- bw.nrd0(na.omit(vt_2009$age)) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(vt_2009$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(vt_2009$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
vt_2009_young <- subset(vt_2009, age <= antimodes[1])
vt_2009_old <- subset(vt_2009, age > antimodes[1])


# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

vt_2009$aftermode <- vt_2009$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=vt_2009, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0), breaks = seq(min(vt_2009$age), max(vt_2009$age), by = 5), labels = seq(min(vt_2009$age), max(vt_2009$age), by = 5), minor_breaks = seq(min(vt_2009$age), max(vt_2009$age), by = 1)) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2009 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2009 <- ggplot(data=vt_2009, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0), breaks = seq(min(vt_2009$age), max(vt_2009$age), by = 5), labels = seq(min(vt_2009$age), max(vt_2009$age), by = 5), minor_breaks = seq(min(vt_2009$age), max(vt_2009$age), by = 1)) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2009 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2009

ggsave(plot=age_2009, filename = "images/vt/age_2009.png", width = 10, height = 8)


sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(vt_2009$age)$p.value) 

```

#### 2010

```{r age2010, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
# looks like there are some NAs in the 2010 data so I have to get rid of them
bandwidth <- bw.nrd0(na.omit(vt_2010$age)) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(vt_2010$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(vt_2010$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
vt_2010_young <- subset(vt_2010, age <= antimodes[1])
vt_2010_old <- subset(vt_2010, age > antimodes[1])

# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

vt_2010$aftermode <- vt_2010$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=vt_2010, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2010$age), max(vt_2010$age), by = 5), labels = seq(min(vt_2010$age), max(vt_2010$age), by = 5), minor_breaks = seq(min(vt_2010$age), max(vt_2010$age), by = 1)) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2010 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2010 <- ggplot(data=vt_2010, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2010$age), max(vt_2010$age), by = 5), labels = seq(min(vt_2010$age), max(vt_2010$age), by = 5), minor_breaks = seq(min(vt_2010$age), max(vt_2010$age), by = 1)) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2010 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2010

ggsave(plot=age_2010, filename = "images/vt/age_2010.png", width = 10, height = 8)


sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(vt_2010$age)$p.value) 

 
```


#### 2011

```{r age2011, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
# looks like there are some NAs in the 2011 data so I have to get rid of them
bandwidth <- bw.nrd0(na.omit(vt_2011$age)) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(vt_2011$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(vt_2011$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
vt_2011_young <- subset(vt_2011, age <= antimodes[1])
vt_2011_old <- subset(vt_2011, age > antimodes[1])


# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

vt_2011$aftermode <- vt_2011$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=vt_2011, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") +scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2011$age), max(vt_2011$age), by = 5), labels = seq(min(vt_2011$age), max(vt_2011$age), by = 5), minor_breaks = seq(min(vt_2011$age), max(vt_2011$age), by = 1)) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2011 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2011 <- ggplot(data=vt_2011, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2011$age), max(vt_2011$age), by = 5), labels = seq(min(vt_2011$age), max(vt_2011$age), by = 5), minor_breaks = seq(min(vt_2011$age), max(vt_2011$age), by = 1)) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2011 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2011

ggsave(plot=age_2011, filename = "images/vt/age_2011.png", width = 10, height = 8)

sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(vt_2011$age)$p.value) 

 
```


#### 2012

```{r age2012, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
# looks like there are some NAs in the 2012 data so I have to get rid of them
bandwidth <- bw.nrd0(na.omit(vt_2012$age)) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(vt_2012$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(vt_2012$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
vt_2012_young <- subset(vt_2012, age <= antimodes[1])
vt_2012_old <- subset(vt_2012, age > antimodes[1])

# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

vt_2012$aftermode <- vt_2012$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=vt_2012, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") +  scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2012$age), max(vt_2012$age), by = 5), labels = seq(min(vt_2012$age), max(vt_2012$age), by = 5), minor_breaks = seq(min(vt_2012$age), max(vt_2012$age), by = 1)) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2012 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2012 <- ggplot(data=vt_2012, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") +  scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2012$age), max(vt_2012$age), by = 5), labels = seq(min(vt_2012$age), max(vt_2012$age), by = 5), minor_breaks = seq(min(vt_2012$age), max(vt_2012$age), by = 1)) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2012 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2012

ggsave(plot=age_2012, filename = "images/vt/age_2012.png", width = 10, height = 8)


sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(vt_2012$age)$p.value) 

 
```


#### 2013

```{r age2013, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
# looks like there are some NAs in the 2013 data so I have to get rid of them
bandwidth <- bw.nrd0(na.omit(vt_2013$age)) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(vt_2013$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(vt_2013$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
vt_2013_young <- subset(vt_2013, age <= antimodes[1])
vt_2013_old <- subset(vt_2013, age > antimodes[1])


# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

vt_2013$aftermode <- vt_2013$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=vt_2013, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2013$age), max(vt_2013$age), by = 5), labels = seq(min(vt_2013$age), max(vt_2013$age), by = 5), minor_breaks = seq(min(vt_2013$age), max(vt_2013$age), by = 1)) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2013 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2013 <- ggplot(data=vt_2013, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2013$age), max(vt_2013$age), by = 5), labels = seq(min(vt_2013$age), max(vt_2013$age), by = 5), minor_breaks = seq(min(vt_2013$age), max(vt_2013$age), by = 1)) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2013 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2013

ggsave(plot=age_2013, filename = "images/vt/age_2013.png", width = 10, height = 8)


sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(vt_2013$age)$p.value) 

 
```


#### 2014

```{r age2014, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
# looks like there are some NAs in the 2014 data so I have to get rid of them
bandwidth <- bw.nrd0(na.omit(vt_2014$age)) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(vt_2014$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(vt_2014$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
# I am using the second antimode here because the first one is not informative!!
vt_2014_young <- subset(vt_2014, age <= antimodes[1])
vt_2014_old <- subset(vt_2014, age > antimodes[1])


# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

vt_2014$aftermode <- vt_2014$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=vt_2014, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2014$age), max(vt_2014$age), by = 5), labels = seq(min(vt_2014$age), max(vt_2014$age), by = 5), minor_breaks = seq(min(vt_2014$age), max(vt_2014$age), by = 1)) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2014 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2014 <- ggplot(data=vt_2014, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2014$age), max(vt_2014$age), by = 5), labels = seq(min(vt_2014$age), max(vt_2014$age), by = 5), minor_breaks = seq(min(vt_2014$age), max(vt_2014$age), by = 1)) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2014 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2014

ggsave(plot=age_2014, filename = "images/vt/age_2014.png", width = 10, height = 8)

sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(vt_2014$age)$p.value) 

```

#### 2015


```{r age2015, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
# looks like there are some NAs in the 2015 data so I have to get rid of them
bandwidth <- bw.nrd0(na.omit(vt_2015$age)) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(vt_2015$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(vt_2015$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode

# THIS IS A UNIMODAL DISTRIBUTION! NOT INCLUDING? 
vt_2015_young <- subset(vt_2015, age <= antimodes[1])
vt_2015_old <- subset(vt_2015, age > antimodes[1])



# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

vt_2015$aftermode <- vt_2015$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=vt_2015, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2015$age), max(vt_2015$age), by = 5), labels = seq(min(vt_2015$age), max(vt_2015$age), by = 5), minor_breaks = seq(min(vt_2015$age), max(vt_2015$age), by = 1)) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2015 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2015 <- ggplot(data=vt_2015, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") +scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2015$age), max(vt_2015$age), by = 5), labels = seq(min(vt_2015$age), max(vt_2015$age), by = 5), minor_breaks = seq(min(vt_2015$age), max(vt_2015$age), by = 1)) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2015 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2015

ggsave(plot=age_2015, filename = "images/vt/age_2015.png", width = 10, height = 8)


sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(vt_2015$age)$p.value) 

 
```


#### 2016

```{r age2016, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
# looks like there are some NAs in the 2016 data so I have to get rid of them
bandwidth <- bw.nrd0(na.omit(vt_2016$age)) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(vt_2016$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(vt_2016$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
vt_2016_young <- subset(vt_2016, age <= antimodes[1])
vt_2016_old <- subset(vt_2016, age > antimodes[1])


# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

vt_2016$aftermode <- vt_2016$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=vt_2016, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2016$age), max(vt_2016$age), by = 5), labels = seq(min(vt_2016$age), max(vt_2016$age), by = 5), minor_breaks = seq(min(vt_2016$age), max(vt_2016$age), by = 1)) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2016 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2016 <- ggplot(data=vt_2016, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2016$age), max(vt_2016$age), by = 5), labels = seq(min(vt_2016$age), max(vt_2016$age), by = 5), minor_breaks = seq(min(vt_2016$age), max(vt_2016$age), by = 1)) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2016 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2016

ggsave(plot=age_2016, filename = "images/vt/age_2016.png", width = 10, height = 8)


sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(vt_2016$age)$p.value) 

 
```


#### 2017

```{r age2017, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
# looks like there are some NAs in the 2017 data so I have to get rid of them
bandwidth <- bw.nrd0(na.omit(vt_2017$age)) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(vt_2017$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(vt_2017$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
vt_2017_young <- subset(vt_2017, age <= antimodes[1])
vt_2017_old <- subset(vt_2017, age > antimodes[1])


# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

vt_2017$aftermode <- vt_2017$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=vt_2017, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2017$age), max(vt_2017$age), by = 5), labels = seq(min(vt_2017$age), max(vt_2017$age), by = 5), minor_breaks = seq(min(vt_2017$age), max(vt_2017$age), by = 1)) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2017 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2017 <- ggplot(data=vt_2017, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2017$age), max(vt_2017$age), by = 5), labels = seq(min(vt_2017$age), max(vt_2017$age), by = 5), minor_breaks = seq(min(vt_2017$age), max(vt_2017$age), by = 1)) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2017 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2017

ggsave(plot=age_2017, filename = "images/vt/age_2017.png", width = 10, height = 8)


sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(vt_2017$age)$p.value) 

```

#### 2018

```{r age2018, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
# looks like there are some NAs in the 2018 data so I have to get rid of them
bandwidth <- bw.nrd0(na.omit(vt_2018$age)) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(vt_2018$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(vt_2018$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
vt_2018_young <- subset(vt_2018, age <= antimodes[1])
vt_2018_old <- subset(vt_2018, age > antimodes[1])


# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

vt_2018$aftermode <- vt_2018$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=vt_2018, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2018$age), max(vt_2018$age), by = 5), labels = seq(min(vt_2018$age), max(vt_2018$age), by = 5), minor_breaks = seq(min(vt_2018$age), max(vt_2018$age), by = 1)) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2018 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2018 <- ggplot(data=vt_2018, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2018$age), max(vt_2018$age), by = 5), labels = seq(min(vt_2018$age), max(vt_2018$age), by = 5), minor_breaks = seq(min(vt_2018$age), max(vt_2018$age), by = 1)) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2018 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2018

ggsave(plot=age_2018, filename = "images/vt/age_2018.png", width = 10, height = 8)


sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(vt_2018$age)$p.value) 

```

#### 2019

```{r age2019, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
bandwidth <- bw.nrd0(vt_2019$age) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(vt_2019$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(vt_2019$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
vt_2019_young <- subset(vt_2019, age <= antimodes[1])
vt_2019_old <- subset(vt_2019, age > antimodes[1])


# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

vt_2019$aftermode <- vt_2019$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=vt_2019, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2019$age), max(vt_2019$age), by = 5), labels = seq(min(vt_2019$age), max(vt_2019$age), by = 5), minor_breaks = seq(min(vt_2019$age), max(vt_2019$age), by = 1)) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2019 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2019 <- ggplot(data=vt_2019, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2019$age), max(vt_2019$age), by = 5), labels = seq(min(vt_2019$age), max(vt_2019$age), by = 5), minor_breaks = seq(min(vt_2019$age), max(vt_2019$age), by = 1)) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2019 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2019

ggsave(plot=age_2019, filename = "images/vt/age_2019.png", width = 10, height = 8)

sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(vt_2019$age)$p.value) 


```

#### 2020

```{r age2020, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
bandwidth <- bw.nrd0(vt_2020$age) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(vt_2020$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(vt_2020$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
vt_2020_young <- subset(vt_2020, age <= antimodes[1])
vt_2020_old <- subset(vt_2020, age > antimodes[1])

# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

vt_2020$aftermode <- vt_2020$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=vt_2020, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = aftermode), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2020$age), max(vt_2020$age), by = 5), labels = seq(min(vt_2020$age), max(vt_2020$age), by = 5), minor_breaks = seq(min(vt_2020$age), max(vt_2020$age), by = 1)) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2020 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values =  c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2020 <- ggplot(data=vt_2020, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = aftermode), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2020$age), max(vt_2020$age), by = 5), labels = seq(min(vt_2020$age), max(vt_2020$age), by = 5), minor_breaks = seq(min(vt_2020$age), max(vt_2020$age), by = 1)) + theme_light() + geom_vline(xintercept = modes, size=1, color="black") + ggtitle("Age distribution for 2020 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values =  c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2020

ggsave(plot=age_2020, filename = "images/vt/age_2020.png", width = 10, height = 8)

sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(vt_2020$age)$p.value) 


```

#### 2021

```{r age2021, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
bandwidth <- bw.nrd0(vt_2021$age) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(vt_2021$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(vt_2021$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
vt_2021_young <- subset(vt_2021, age <= antimodes[1])
vt_2021_old <- subset(vt_2021, age > antimodes[1])


# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

vt_2021$aftermode <- vt_2021$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=vt_2021, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2021$age), max(vt_2021$age), by = 5), labels = seq(min(vt_2021$age), max(vt_2021$age), by = 5), minor_breaks = seq(min(vt_2021$age), max(vt_2021$age), by = 1)) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2021 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2021 <- ggplot(data=vt_2021, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2021$age), max(vt_2021$age), by = 5), labels = seq(min(vt_2021$age), max(vt_2021$age), by = 5), minor_breaks = seq(min(vt_2021$age), max(vt_2021$age), by = 1)) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2021 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2021

ggsave(plot=age_2021, filename = "images/vt/age_2021.png", width = 10, height = 8)

sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(vt_2021$age)$p.value) 

 
```

#### 2022

```{r age2022, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
bandwidth <- bw.nrd0(vt_2022$age) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(vt_2022$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(vt_2022$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
# in this case we have to use the second antimode because the first one seems to represent some infant deaths
vt_2022_young <- subset(vt_2022, age <= antimodes[1])
vt_2022_old <- subset(vt_2022, age > antimodes[1])

# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

vt_2022$aftermode <- vt_2022$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=vt_2022, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2022$age), max(vt_2022$age), by = 5), labels = seq(min(vt_2022$age), max(vt_2022$age), by = 5), minor_breaks = seq(min(vt_2022$age), max(vt_2022$age), by = 1)) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2022 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

# using modes[-1] to get rid of that first mode which is for a baby death
age_2022 <- ggplot(data=vt_2022, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2022$age), max(vt_2022$age), by = 5), labels = seq(min(vt_2022$age), max(vt_2022$age), by = 5), minor_breaks = seq(min(vt_2022$age), max(vt_2022$age), by = 1)) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2022 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2022


ggsave(plot=age_2022, filename = "images/vt/age_2022.png", width = 10, height = 8)

sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(vt_2022$age)$p.value) 

```

#### 2023

```{r age2023, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
bandwidth <- bw.nrd0(vt_2023$age) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(vt_2023$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(vt_2023$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
vt_2023_young <- subset(vt_2023, age <= antimodes[1])
vt_2023_old <- subset(vt_2023, age > antimodes[1])

# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

vt_2023$aftermode <- vt_2023$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=vt_2023, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2023$age), max(vt_2023$age), by = 5), labels = seq(min(vt_2023$age), max(vt_2023$age), by = 5), minor_breaks = seq(min(vt_2023$age), max(vt_2023$age), by = 1))  + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2023 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2023 <- ggplot(data=vt_2023, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0),  breaks = seq(min(vt_2023$age), max(vt_2023$age), by = 5), labels = seq(min(vt_2023$age), max(vt_2023$age), by = 5), minor_breaks = seq(min(vt_2023$age), max(vt_2023$age), by = 1))  + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2023 VT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 9)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2023


ggsave(plot=age_2023, filename = "images/vt/age_2023.png", width = 10, height = 8)

sprintf("Using Hartigan's dip test for multimodality, the resulting p value is %f.", dip.test(vt_2023$age)$p.value) 

 
```

### Median age

I have only included information for years where I can identify a "younger" and "older" subpopulation -- take these determinations with a grain of salt, though, because the sample size is small and the kde could be erroneously identifying [or erroneously missing] subpopulations. 

```{r median, echo=FALSE, message=FALSE, warning=FALSE}
sprintf("The median age of decedents in VT in 2009 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(vt_2009$age), median(vt_2009_young$age), median(vt_2009_old$age))

sprintf("The median age of decedents in VT in 2010 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(vt_2010$age), median(vt_2010_young$age), median(vt_2010_old$age))

sprintf("The median age of decedents in VT in 2011 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(vt_2011$age), median(vt_2011_young$age), median(vt_2011_old$age))

sprintf("The median age of decedents in VT in 2012 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(vt_2012$age), median(vt_2012_young$age), median(vt_2012_old$age))

sprintf("The median age of decedents in VT in 2013 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(vt_2013$age), median(vt_2013_young$age), median(vt_2013_old$age))

sprintf("The median age of decedents in VT in 2014 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(vt_2014$age), median(vt_2014_young$age), median(vt_2014_old$age))

sprintf("The median age of decedents in VT in 2015 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(vt_2015$age), median(vt_2015_young$age), median(vt_2015_old$age))

sprintf("The median age of decedents in VT in 2016 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(vt_2016$age), median(vt_2016_young$age), median(vt_2016_old$age))

sprintf("The median age of decedents in VT in 2017 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(vt_2017$age), median(vt_2017_young$age), median(vt_2017_old$age))

sprintf("The median age of decedents in VT in 2018 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(vt_2018$age), median(vt_2018_young$age), median(vt_2018_old$age))

sprintf("The median age of decedents in VT in 2019 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(vt_2019$age), median(vt_2019_young$age), median(vt_2019_old$age))

sprintf("The median age of decedents in VT in 2020 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(vt_2020$age), median(vt_2020_young$age), median(vt_2020_old$age))

sprintf("The median age of decedents in VT in 2021 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(vt_2021$age), median(vt_2021_young$age), median(vt_2021_old$age))

sprintf("The median age of decedents in VT in 2022 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(vt_2022$age), median(vt_2022_young$age), median(vt_2022_old$age))

sprintf("The median age of decedents in VT in 2023 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(vt_2023$age), median(vt_2023_young$age), median(vt_2023_old$age))
```

### Proportion of older people by year

Once more, only using years for which I can bifurcate the distribution. 

```{r propOld, echo=FALSE, message=FALSE, warning=FALSE}

sprintf("The proportion of older decedents in 2009 was %f, with %d out of %d deaths.", nrow(vt_2009_old)/nrow(vt_2009), nrow(vt_2009_old), nrow(vt_2009))
#sprintf("The proportion of older decedents in 2010 was %f, with %d out of %d deaths.", nrow(vt_2010_old)/nrow(vt_2010), nrow(vt_2010_old), nrow(vt_2010))
#sprintf("The proportion of older decedents in 2011 was %f, with %d out of %d deaths.", nrow(vt_2011_old)/nrow(vt_2011), nrow(vt_2011_old), nrow(vt_2011))
sprintf("The proportion of older decedents in 2012 was %f, with %d out of %d deaths.", nrow(vt_2012_old)/nrow(vt_2012), nrow(vt_2012_old), nrow(vt_2012))
sprintf("The proportion of older decedents in 2013 was %f, with %d out of %d deaths.", nrow(vt_2013_old)/nrow(vt_2013), nrow(vt_2013_old), nrow(vt_2013))
sprintf("The proportion of older decedents in 2014 was %f, with %d out of %d deaths.", nrow(vt_2014_old)/nrow(vt_2014), nrow(vt_2014_old), nrow(vt_2014))
sprintf("The proportion of older decedents in 2015 was %f, with %d out of %d deaths.", nrow(vt_2015_old)/nrow(vt_2015), nrow(vt_2015_old), nrow(vt_2015))
#sprintf("The proportion of older decedents in 2016 was %f, with %d out of %d deaths.", nrow(vt_2016_old)/nrow(vt_2016), nrow(vt_2016_old), nrow(vt_2016))
sprintf("The proportion of older decedents in 2017 was %f, with %d out of %d deaths.", nrow(vt_2017_old)/nrow(vt_2017), nrow(vt_2017_old), nrow(vt_2017))
#sprintf("The proportion of older decedents in 2018 was %f, with %d out of %d deaths.", nrow(vt_2018_old)/nrow(vt_2018), nrow(vt_2018_old), nrow(vt_2018))
sprintf("The proportion of older decedents in 2019 was %f, with %d out of %d deaths.", nrow(vt_2019_old)/nrow(vt_2019), nrow(vt_2019_old), nrow(vt_2019))
sprintf("The proportion of older decedents in 2020 was %f, with %d out of %d deaths.", nrow(vt_2020_old)/nrow(vt_2020), nrow(vt_2020_old), nrow(vt_2020))
#sprintf("The proportion of older decedents in 2021 was %f, with %d out of %d deaths.", nrow(vt_2021_old)/nrow(vt_2021), nrow(vt_2021_old), nrow(vt_2021))
#sprintf("The proportion of older decedents in 2022 was %f, with %d out of %d deaths.", nrow(vt_2022_old)/nrow(vt_2022), nrow(vt_2022_old), nrow(vt_2022))
#sprintf("The proportion of older decedents in 2023 was %f, with %d out of %d deaths.", nrow(vt_2023_old)/nrow(vt_2023), nrow(vt_2023_old), nrow(vt_2023))

```

<!-- VT lists -- for analysis purposes only --> 

```{r vtLists, echo=FALSE, message=FALSE, warning=FALSE}

# this will come in handy for the tox graphs
yearList <- c(2009, 2010, 2012, 2013, 2014, 2015, 2017, 2019, 2020)

# make lists of the different vt datasets so I can use them later
# create a list of dfs so I can iterate over them -- from https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
vtlist <- list(vt_2009, vt_2010, vt_2012, vt_2013, vt_2014, vt_2015, vt_2017, vt_2019, vt_2020)

# excluding years with unimodal distributions

vtlistYoung <- list(vt_2009_young, vt_2010_young, vt_2012_young, vt_2013_young, vt_2014_young, vt_2015_young, vt_2017_young, vt_2019_young, vt_2020_young)

vtlistOld <- list(vt_2009_old, vt_2010_old, vt_2012_old, vt_2013_old, vt_2014_old, vt_2015_old, vt_2017_old, vt_2019_old, vt_2020_old)

```

### Race of decedents {.tabset}

#### Young

```{r raceNumericYoung, echo=FALSE, message=FALSE, warning=FALSE}
for(i in 1:length(vtlistYoung)){
  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  vt <- vtlistYoung[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- vt$Year[1]
  
# see if I can automate the title, from https://stackoverflow.com/questions/59095237/automate-ggplots-while-using-variable-labels-as-title-and-axis-titles
racetitle <- paste("Racial distribution for", year, "VT opioid-detected overdose deaths in younger people")

race <- ggplot(data=vt, aes(x=Race, ..count..)) + 
      geom_bar(aes(fill = Race), position = "dodge")  + scale_fill_manual(name = "Race", values = c("#E69F00", "#56B4E9", "#009E73", "#CC79A7"), labels = c("Black", "Hispanic", "Other", "White")) + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle(racetitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7)) 

# seem to need to use print() when in a for loop, from https://stackoverflow.com/questions/29053223/plots-not-working-in-for-loop
print(race)

# try to automate the filenames? from https://www.edureka.co/community/41963/r-programming-naming-the-output-file-using-a-variable
ggsave(plot=race, filename = paste0("images/vt/race_", year, "_young.png"), width = 10, height = 8)

raceProp <- ggplot(data=vt, aes(x=Race, ..count..)) + 
      geom_bar(aes(fill = Race), position = "dodge")  + scale_fill_manual(name = "Race", values = c("#E69F00", "#56B4E9", "#009E73", "#CC79A7"), labels = c("Black", "Hispanic", "Other", "White")) + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(vt), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle(racetitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7)) 

print(raceProp)

#save this one for the presentation
ggsave(plot=raceProp, filename = paste0("images/vt/race_proportions_", year, "_young.png"), width = 10, height = 8)
  
}


```

#### Old 

```{r raceNumericOld, echo=FALSE, message=FALSE, warning=FALSE}
for(i in 1:length(vtlistOld)){
  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  vt <- vtlistOld[[i]] 
  # grab the year -- you only need the first element in the year column sivte all of them have the same value
  year <- vt$Year[1]
  
  
# see if I can automate the title, from https://stackoverflow.com/questions/59095237/automate-ggplots-while-using-variable-labels-as-title-and-axis-titles
racetitle <- paste("Racial distribution for", year, "VT opioid-detected overdose deaths in older people")

race <- ggplot(data=vt, aes(x=Race, ..count..)) + 
      geom_bar(aes(fill = Race), position = "dodge")  + scale_fill_manual(name = "Race", values = c("#E69F00", "#56B4E9", "#009E73", "#CC79A7"), labels = c("Black", "Hispanic", "Other", "White")) + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle(racetitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7)) 

# seem to need to use print() when in a for loop, from https://stackoverflow.com/questions/29053223/plots-not-working-in-for-loop
print(race)

# try to automate the filenames? from https://www.edureka.co/community/41963/r-programming-naming-the-output-file-using-a-variable
ggsave(plot=race, filename = paste0("images/vt/race_", year, "_old.png"), width = 10, height = 8)

raceProp <- ggplot(data=vt, aes(x=Race, ..count..)) + 
      geom_bar(aes(fill = Race), position = "dodge")  + scale_fill_manual(name = "Race", values = c("#E69F00", "#56B4E9", "#009E73", "#CC79A7"), labels = c("Black", "Hispanic", "Other", "White")) + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(vt), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle(racetitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7)) 

print(raceProp)

#save this one for the presentation
ggsave(plot=raceProp, filename = paste0("images/vt/race_proportions_", year, "_old.png"), width = 10, height = 8)
  
}

```


### Significance testing for race {.tabset}


```{r chiSquaredRace, echo=FALSE, message=FALSE, warning=FALSE}
# initialize empty matrix, from https://stackoverflow.com/questions/32712301/create-empty-data-frame-with-column-names-by-assigning-a-string-vector
# this will hold all the p values

chisq_race <- data.frame(matrix(ncol = 2, nrow = 0))
cols <- c("Year", "p value")

for(i in 1:length(vtlistYoung)){
  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  young <- vtlistYoung[[i]] 
  old <- vtlistOld[[i]] 
  # grab the year -- you only need the first element in the year column sivte all of them have the same value
  year <- young$Year[1]
  
  racefreqtable <- young %>% group_by(Race) %>% tally()
  names(racefreqtable)[names(racefreqtable) == 'n'] <- 'Young'

  racetableold <- old %>% group_by(Race) %>% tally()
  names(racetableold)[names(racetableold) == 'n'] <- 'Old'

  racetable <- merge(racefreqtable, racetableold, by="Race")
  
  # convert the "race" column into row names so I get a table with only numeric values, from   https://stackoverflow.com/questions/5555408/convert-the-values-in-a-column-into-row-names-in-an-existing-data-frame

  racemat <- racetable[,-1]
  rownames(racemat) <- racetable[,1]

  print(as.matrix(racemat))
  
   # append year and p value to a vector, rounding to make it more legible
  # I'm using Fisher's exact test if >20% of cell counts are <5, if not use chi squared
  # I get >20% <5 from https://www.ccjm.org/content/84/9_suppl_2/e20
  pvalue <- ifelse(sum(racemat < 5)/(ncol(racemat) * nrow(racemat)) > 0.2, round(fisher.test(as.matrix(racemat))$p.value, 4), round(chisq.test(as.matrix(racemat))$p.value, 4))
  
  results <- c(year, pvalue)
  
  # add this to the chi squared df 
  chisq_race <- rbind(chisq_race, results)

}

# for some reason it doesn't work if I do this before adding rows?
colnames(chisq_race) <- cols

chisq_race

```


### Proportion of white deaths

```{r raceWhite, echo=FALSE, message=FALSE, warning=FALSE}

# unfortunately there are different variable names for each vt file so a for loop won't work :/

# get proportion of Race for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
whiteyoung <- c(sum(vt_2009_young$Race == "White")/nrow(vt_2009_young), sum(vt_2010_young$Race == "White")/nrow(vt_2010_young), sum(vt_2012_young$Race == "White")/nrow(vt_2012_young),  sum(vt_2013_young$Race == "White")/nrow(vt_2013_young), sum(vt_2014_young$Race == "White")/nrow(vt_2014_young), sum(vt_2015_young$Race == "White")/nrow(vt_2015_young), sum(vt_2017_young$Race == "White")/nrow(vt_2017_young), sum(vt_2019_young$Race == "White")/nrow(vt_2019_young), sum(vt_2020_young$Race == "White")/nrow(vt_2020_young))

# get proportion for older people
whiteold <- c(sum(vt_2009_old$Race == "White")/nrow(vt_2009_old), sum(vt_2010_old$Race == "White")/nrow(vt_2010_old), sum(vt_2012_old$Race == "White")/nrow(vt_2012_old),  sum(vt_2013_old$Race == "White")/nrow(vt_2013_old), sum(vt_2014_old$Race == "White")/nrow(vt_2014_old), sum(vt_2015_old$Race == "White")/nrow(vt_2015_old), sum(vt_2017_old$Race == "White")/nrow(vt_2017_old), sum(vt_2019_old$Race == "White")/nrow(vt_2019_old), sum(vt_2020_old$Race == "White")/nrow(vt_2020_old))

# combine vectors into df
propRace <- data.frame(yearList, whiteyoung, whiteold)

# change the col names to be more informative
names(propRace) <- c("Year", "Younger", "Older")

# create grouped bar plot, from https://stackoverflow.com/questions/72449952/making-a-grouped-bar-chart-in-r
# getting the labels in the right position from https://stackoverflow.com/questions/37587288/ggplot-label-bars-in-grouped-bar-plot
# info on changing legend title from here https://stackoverflow.com/questions/41649929/how-to-modify-a-legend-title-in-a-barplot-with-ggplot
# the scale x continuous is there because for some reason adding 2023 data messed up the x axis, see https://stackoverflow.com/questions/72051789/how-to-display-all-values-on-x-and-y-axes-in-r-ggplot2

# ivtreasing line size from https://stackoverflow.com/questions/14794599/how-to-change-line-width-in-ggplot
# line graph info from https://www.sthda.com/english/wiki/ggplot2-line-plot-quick-start-guide-r-software-and-data-visualization

pracelong  <- propRace |>
  # Reshape data to long format
  pivot_longer(Younger:Older)

# create labels with the value of each proportion
pracelong$labels <- round(pracelong$value, 2)

pRace <-  ggplot(data = pracelong, aes(x = Year, y = value, fill = name, group=name, color=factor(name))) +
  # Implement a grouped bar chart
  # Implement a line graph
  geom_line(stat = "identity", size = 1) + geom_point(size = 2) + theme_light() + ggtitle("Proportion of White opioid-detected overdose deaths \n by age group in VT, 2009-2023") + ylab("Proportion") + theme(plot.title = element_text(hjust = 0.5)) + scale_fill_discrete(guide="none") + scale_x_continuous(breaks=propRace$Year) + xlab("Year")+ coord_cartesian(ylim = c(0, 1)) + guides(color = guide_legend(title = "Age group"))+ geom_label_repel(aes(label = labels, x = Year, y = value), nudge_x = 0.1, na.rm = FALSE, color = "black", size = 2.5, label.size = NA, fill = NA)

pRace

ggsave(plot=pRace, filename = "images/vt/prop_white_allyears.png", width = 10, height = 8)

```


### Significance testing for white deaths

```{r chiSquaredWhite, echo=FALSE, message=FALSE, warning=FALSE}
# initialize empty matrix, from https://stackoverflow.com/questions/32712301/create-empty-data-frame-with-column-names-by-assigning-a-string-vector
# this will hold all the p values

chisq_white <- data.frame(matrix(ncol = 2, nrow = 0))
cols <- c("Year", "Younger", "Older")

for(i in 1:length(vtlistYoung)){
  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  young <- vtlistYoung[[i]] 
  old <- vtlistOld[[i]] 
  # grab the year -- you only need the first element in the year column sivte all of them have the same value
  year <- young$Year[1]
  
  racefreqtable <- subset(young, Race == "White") %>% group_by(Race) %>% tally()
  names(racefreqtable)[names(racefreqtable) == 'n'] <- 'Young'

  racetableold <- subset(old, Race == "White") %>% group_by(Race) %>% tally()
  names(racetableold)[names(racetableold) == 'n'] <- 'Old'

  racetable <- merge(racefreqtable, racetableold, by="Race")
  
  # convert the "race" column into row names so I get a table with only numeric values, from   https://stackoverflow.com/questions/5555408/convert-the-values-in-a-column-into-row-names-in-an-existing-data-frame

  racemat <- racetable[,-1]
  
  # add to matrix with all the white results 
  
  results <- cbind(year, racemat)
  
  # add this to the chi squared df 
  chisq_white <- rbind(chisq_white, results)

}

# for some reason it doesn't work if I do this before adding rows?
colnames(chisq_white) <- cols
chisq_white

# compute chi squared

racemat <- as.matrix(chisq_white[,-1])

  pvalue <- ifelse(sum(racemat < 5)/(ncol(racemat) * nrow(racemat)) > 0.2, round(fisher.test(as.matrix(racemat))$p.value, 4), round(chisq.test(as.matrix(racemat))$p.value, 4))

pvalue

```



### Proportion of Latine deaths

```{r raceLatine, echo=FALSE, message=FALSE, warning=FALSE}

# unfortunately there are different variable names for each vt file so a for loop won't work :/

# get proportion of Race for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
latineyoung <- c(sum(vt_2009_young$Race == "Hispanic")/nrow(vt_2009_young), sum(vt_2010_young$Race == "Hispanic")/nrow(vt_2010_young), sum(vt_2012_young$Race == "Hispanic")/nrow(vt_2012_young),  sum(vt_2013_young$Race == "Hispanic")/nrow(vt_2013_young), sum(vt_2014_young$Race == "Hispanic")/nrow(vt_2014_young), sum(vt_2015_young$Race == "Hispanic")/nrow(vt_2015_young), sum(vt_2017_young$Race == "Hispanic")/nrow(vt_2017_young), sum(vt_2019_young$Race == "Hispanic")/nrow(vt_2019_young), sum(vt_2020_young$Race == "Hispanic")/nrow(vt_2020_young))

# get proportion for older people
latineold <- c(sum(vt_2009_old$Race == "Hispanic")/nrow(vt_2009_old), sum(vt_2010_old$Race == "Hispanic")/nrow(vt_2010_old), sum(vt_2012_old$Race == "Hispanic")/nrow(vt_2012_old),  sum(vt_2013_old$Race == "Hispanic")/nrow(vt_2013_old), sum(vt_2014_old$Race == "Hispanic")/nrow(vt_2014_old), sum(vt_2015_old$Race == "Hispanic")/nrow(vt_2015_old), sum(vt_2017_old$Race == "Hispanic")/nrow(vt_2017_old), sum(vt_2019_old$Race == "Hispanic")/nrow(vt_2019_old), sum(vt_2020_old$Race == "Hispanic")/nrow(vt_2020_old))

# combine vectors into df
propRace <- data.frame(yearList, latineyoung, latineold)

# change the col names to be more informative
names(propRace) <- c("Year", "Younger", "Older")

# create grouped bar plot, from https://stackoverflow.com/questions/72449952/making-a-grouped-bar-chart-in-r
# getting the labels in the right position from https://stackoverflow.com/questions/37587288/ggplot-label-bars-in-grouped-bar-plot
# info on changing legend title from here https://stackoverflow.com/questions/41649929/how-to-modify-a-legend-title-in-a-barplot-with-ggplot
# the scale x continuous is there because for some reason adding 2023 data messed up the x axis, see https://stackoverflow.com/questions/72051789/how-to-display-all-values-on-x-and-y-axes-in-r-ggplot2

# ivtreasing line size from https://stackoverflow.com/questions/14794599/how-to-change-line-width-in-ggplot
# line graph info from https://www.sthda.com/english/wiki/ggplot2-line-plot-quick-start-guide-r-software-and-data-visualization

pracelong  <- propRace |>
  # Reshape data to long format
  pivot_longer(Younger:Older)

# create labels with the value of each proportion
pracelong$labels <- round(pracelong$value, 2)

pRace <-  ggplot(data = pracelong, aes(x = Year, y = value, fill = name, group=name, color=factor(name))) +
  # Implement a grouped bar chart
  # Implement a line graph
  geom_line(stat = "identity", size = 1) + geom_point(size = 2) + theme_light() + ggtitle("Proportion of Latine opioid-detected overdose deaths \n by age group in VT, 2009-2023") + ylab("Proportion") + theme(plot.title = element_text(hjust = 0.5)) + scale_fill_discrete(guide="none") + scale_x_continuous(breaks=propRace$Year) + xlab("Year")+ coord_cartesian(ylim = c(0, 1)) + guides(color = guide_legend(title = "Age group"))+ geom_label_repel(aes(label = labels, x = Year, y = value), nudge_x = 0.1, na.rm = FALSE, color = "black", size = 2.5, label.size = NA, fill = NA)

pRace

ggsave(plot=pRace, filename = "images/vt/prop_latine_allyears.png", width = 10, height = 8)

```


### No chi squared testing for Latine deaths--basically everything is 0

### Proportion of Black deaths

```{r raceBlack, echo=FALSE, message=FALSE, warning=FALSE}

# unfortunately there are different variable names for each vt file so a for loop won't work :/

# get proportion of Race for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
blackyoung <- c(sum(vt_2009_young$Race == "Black")/nrow(vt_2009_young), sum(vt_2010_young$Race == "Black")/nrow(vt_2010_young), sum(vt_2012_young$Race == "Black")/nrow(vt_2012_young),  sum(vt_2013_young$Race == "Black")/nrow(vt_2013_young), sum(vt_2014_young$Race == "Black")/nrow(vt_2014_young), sum(vt_2015_young$Race == "Black")/nrow(vt_2015_young), sum(vt_2017_young$Race == "Black")/nrow(vt_2017_young), sum(vt_2019_young$Race == "Black")/nrow(vt_2019_young), sum(vt_2020_young$Race == "Black")/nrow(vt_2020_young))

# get proportion for older people
blackold <- c(sum(vt_2009_old$Race == "Black")/nrow(vt_2009_old), sum(vt_2010_old$Race == "Black")/nrow(vt_2010_old), sum(vt_2012_old$Race == "Black")/nrow(vt_2012_old),  sum(vt_2013_old$Race == "Black")/nrow(vt_2013_old), sum(vt_2014_old$Race == "Black")/nrow(vt_2014_old), sum(vt_2015_old$Race == "Black")/nrow(vt_2015_old), sum(vt_2017_old$Race == "Black")/nrow(vt_2017_old), sum(vt_2019_old$Race == "Black")/nrow(vt_2019_old), sum(vt_2020_old$Race == "Black")/nrow(vt_2020_old))

# combine vectors into df
propRace <- data.frame(yearList, blackyoung, blackold)

# change the col names to be more informative
names(propRace) <- c("Year", "Younger", "Older")

# create grouped bar plot, from https://stackoverflow.com/questions/72449952/making-a-grouped-bar-chart-in-r
# getting the labels in the right position from https://stackoverflow.com/questions/37587288/ggplot-label-bars-in-grouped-bar-plot
# info on changing legend title from here https://stackoverflow.com/questions/41649929/how-to-modify-a-legend-title-in-a-barplot-with-ggplot
# the scale x continuous is there because for some reason adding 2023 data messed up the x axis, see https://stackoverflow.com/questions/72051789/how-to-display-all-values-on-x-and-y-axes-in-r-ggplot2

# ivtreasing line size from https://stackoverflow.com/questions/14794599/how-to-change-line-width-in-ggplot
# line graph info from https://www.sthda.com/english/wiki/ggplot2-line-plot-quick-start-guide-r-software-and-data-visualization

pracelong  <- propRace |>
  # Reshape data to long format
  pivot_longer(Younger:Older)

# create labels with the value of each proportion
pracelong$labels <- round(pracelong$value, 2)

pRace <-  ggplot(data = pracelong, aes(x = Year, y = value, fill = name, group=name, color=factor(name))) +
  # Implement a grouped bar chart
  # Implement a line graph
  geom_line(stat = "identity", size = 1) + geom_point(size = 2) + theme_light() + ggtitle("Proportion of Black opioid-detected overdose deaths \n by age group in VT, 2009-2023") + ylab("Proportion") + theme(plot.title = element_text(hjust = 0.5)) + scale_fill_discrete(guide="none") + scale_x_continuous(breaks=propRace$Year) + xlab("Year")+ coord_cartesian(ylim = c(0, 1)) + guides(color = guide_legend(title = "Age group"))+ geom_label_repel(aes(label = labels, x = Year, y = value), nudge_x = 0.1, na.rm = FALSE, color = "black", size = 2.5, label.size = NA, fill = NA)

pRace

ggsave(plot=pRace, filename = "images/vt/prop_black_allyears.png", width = 10, height = 8)

```


### Significance testing for Black deaths -- I don't think this is worth it either, there are too few [but maybe change to fisher's exact test?]

```{r chiSquaredBlack, echo=FALSE, message=FALSE, warning=FALSE}
# initialize empty matrix, from https://stackoverflow.com/questions/32712301/create-empty-data-frame-with-column-names-by-assigning-a-string-vector
# this will hold all the p values

# chisq_black <- data.frame(matrix(ncol = 2, nrow = 0))
# cols <- c("Year", "p value")
# 
# for(i in 1:length(vtlistYoung)){
#   # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
#   young <- vtlistYoung[[i]] 
#   old <- vtlistOld[[i]] 
#   # grab the year -- you only need the first element in the year column sivte all of them have the same value
#   year <- young$Year[1]
#   
#   racefreqtable <- subset(young, Race == "Black") %>% group_by(Race) %>% tally()
#   names(racefreqtable)[names(racefreqtable) == 'n'] <- 'Young'
# 
#   racetableold <- subset(old, Race == "Black") %>% group_by(Race) %>% tally()
#   names(racetableold)[names(racetableold) == 'n'] <- 'Old'
# 
#   racetable <- merge(racefreqtable, racetableold, by="Race")
#   
#   # convert the "race" column into row names so I get a table with only numeric values, from   https://stackoverflow.com/questions/5555408/convert-the-values-in-a-column-into-row-names-in-an-existing-data-frame
# 
#   racemat <- racetable[,-1]
#   rownames(racemat) <- racetable[,1]
# 
#   print(as.matrix(racemat))
#   
#   # append year and p value to a vector, rounding to make it more legible
#   results <- c(year, round(chisq.test(as.matrix(racemat))$p.value, 4))
#   
#   # add this to the chi squared df 
#   chisq_black <- rbind(chisq_black, results)
# 
# }
# 
# # for some reason it doesn't work if I do this before adding rows?
# colnames(chisq_black) <- cols
# 
# chisq_black

```

### Sex of decedents {.tabset}

#### Young

```{r sexTableYoung, echo=FALSE, message=FALSE, warning=FALSE}
for(i in 1:length(vtlistYoung)){

  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  vt <- vtlistYoung[[i]] 
  # grab the year -- you only need the first element in the year column sivte all of them have the same value
  year <- vt$Year[1]
  
  sextitle <- paste("Sex distribution for", year, "VT opioid-detected overdose deaths in younger people")

  
sex <- ggplot(data=vt, aes(x=Sex_revised, ..count..)) + 
      geom_bar(aes(fill = Sex_revised), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle(sextitle)   + theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

print(sex)

ggsave(plot=sex, filename = paste0("images/vt/sex", year, "_young.png"), width = 10, height = 8)


sexProp <- ggplot(data=vt, aes(x=Sex_revised, ..count..)) + 
      geom_bar(aes(fill = Sex_revised), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(vt), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle(sextitle)   + theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

print(sexProp)

ggsave(plot=sexProp, filename = paste0("images/vt/sexProp", year, "_young.png"), width = 10, height = 8)

}

```

#### Old

```{r sexTableOld, echo=FALSE, message=FALSE, warning=FALSE}
for(i in 1:length(vtlistOld)){

  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  vt <- vtlistOld[[i]] 
  # grab the year -- you only need the first element in the year column sivte all of them have the same value
  year <- vt$year[1]
  
  sextitle <- paste("Sex distribution for", year, "VT opioid-detected overdose deaths in older people")

  
sex <- ggplot(data=vt, aes(x=Sex_revised, ..count..)) + 
      geom_bar(aes(fill = Sex_revised), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle(sextitle)   + theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

print(sex)

ggsave(plot=sex, filename = paste0("images/vt/sex", year, "_old.png"), width = 10, height = 8)


sexProp <- ggplot(data=vt, aes(x=Sex_revised, ..count..)) + 
      geom_bar(aes(fill = Sex_revised), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(vt), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle(sextitle)  + theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

print(sexProp)

ggsave(plot=sexProp, filename = paste0("images/vt/sexProp", year, "_old.png"), width = 10, height = 8)

}

```

### Significance testing for sex {.tabset}

```{r chiSquaredSex, echo=FALSE, message=FALSE, warning=FALSE}
# initialize empty matrix, from https://stackoverflow.com/questions/32712301/create-empty-data-frame-with-column-names-by-assigning-a-string-vector
# this will hold all the p values

chisq_sex <- data.frame(matrix(ncol = 2, nrow = 0))
cols <- c("Year", "p value")

for(i in 1:length(vtlistYoung)){
  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  young <- vtlistYoung[[i]] 
  old <- vtlistOld[[i]] 
  # grab the year -- you only need the first element in the year column sivte all of them have the same value
  year <- young$Year[1]
  
  sexfreqtable <- young %>% group_by(Sex_revised) %>% tally()
  names(sexfreqtable)[names(sexfreqtable) == 'n'] <- 'Young'

  sextableold <- old %>% group_by(Sex_revised) %>% tally()
  names(sextableold)[names(sextableold) == 'n'] <- 'Old'

  sextable <- merge(sexfreqtable, sextableold, by="Sex_revised")
  
  # convert the "race" column into row names so I get a table with only numeric values, from   https://stackoverflow.com/questions/5555408/convert-the-values-in-a-column-into-row-names-in-an-existing-data-frame

  sexmat <- sextable[,-1]
  rownames(sexmat) <- sextable[,1]

  print(as.matrix(sexmat))
  
   # append year and p value to a vector, rounding to make it more legible
  # I'm using Fisher's exact test if >20% of cell counts are <5, if not use chi squared
  # I get >20% <5 from https://www.ccjm.org/content/84/9_suppl_2/e20
  pvalue <- ifelse(sum(sexmat < 5)/(ncol(sexmat) * nrow(sexmat)) > 0.2, round(fisher.test(as.matrix(sexmat))$p.value, 4), round(chisq.test(as.matrix(sexmat))$p.value, 4))
  
  results <- c(year, pvalue)
  
  # add this to the chi squared df 
  chisq_sex <- rbind(chisq_sex, results)

}

# for some reason it doesn't work if I do this before adding rows?
colnames(chisq_sex) <- cols

chisq_sex

```

### Sex over all years

```{r sexAllYears, echo=FALSE, message=FALSE, warning=FALSE}

# unfortunately there are different variable names for each mass file so a for loop won't work :/

# get proportion of Sex_revised for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
sexyoung <- c(sum(vt_2009_young$Sex_revised == "Male")/nrow(vt_2009_young), sum(vt_2010_young$Sex_revised == "Male")/nrow(vt_2010_young), sum(vt_2012_young$Sex_revised == "Male")/nrow(vt_2012_young), sum(vt_2013_young$Sex_revised == "Male")/nrow(vt_2013_young), sum(vt_2014_young$Sex_revised == "Male")/nrow(vt_2014_young), sum(vt_2015_young$Sex_revised == "Male")/nrow(vt_2015_young), sum(vt_2017_young$Sex_revised == "Male")/nrow(vt_2017_young), sum(vt_2019_young$Sex_revised == "Male")/nrow(vt_2019_young), sum(vt_2020_young$Sex_revised == "Male")/nrow(vt_2020_young))

# get proportion for older people
sexold <- c(sum(vt_2009_old$Sex_revised == "Male")/nrow(vt_2009_old), sum(vt_2010_old$Sex_revised == "Male")/nrow(vt_2010_old), sum(vt_2012_old$Sex_revised == "Male")/nrow(vt_2012_old), sum(vt_2013_old$Sex_revised == "Male")/nrow(vt_2013_old), sum(vt_2014_old$Sex_revised == "Male")/nrow(vt_2014_old), sum(vt_2015_old$Sex_revised == "Male")/nrow(vt_2015_old), sum(vt_2017_old$Sex_revised == "Male")/nrow(vt_2017_old), sum(vt_2019_old$Sex_revised == "Male")/nrow(vt_2019_old), sum(vt_2020_old$Sex_revised == "Male")/nrow(vt_2020_old))


# combine vectors into df
propSex_revised <- data.frame(yearList, sexyoung, sexold)

# change the col names to be more informative
names(propSex_revised) <- c("Year", "Younger", "Older")

# create grouped bar plot, from https://stackoverflow.com/questions/72449952/making-a-grouped-bar-chart-in-r
# getting the labels in the right position from https://stackoverflow.com/questions/37587288/ggplot-label-bars-in-grouped-bar-plot
# info on changing legend title from here https://stackoverflow.com/questions/41649929/how-to-modify-a-legend-title-in-a-barplot-with-ggplot
# the scale x continuous is there because for some reason adding 2023 data messed up the x axis, see https://stackoverflow.com/questions/72051789/how-to-display-all-values-on-x-and-y-axes-in-r-ggplot2

psexlong  <- propSex_revised |>
  # Reshape data to long format
  pivot_longer(Younger:Older)

# create labels with the value of each proportion
psexlong$labels <- round(psexlong$value, 2)

pSex_revised <-  ggplot(data = psexlong, aes(x = Year, y = value, fill = name, group=name, color=factor(name))) +
  # Implement a grouped bar chart
  # Implement a line graph
  geom_line(stat = "identity", size = 1) + geom_point(size = 2) + theme_light() + ggtitle("Proportion of male deaths in VT by age group, 2009-2023") + ylab("Proportion") + theme(plot.title = element_text(hjust = 0.5)) + scale_fill_discrete(guide="none") + scale_x_continuous(breaks=propSex_revised$Year) + xlab("Year")+ coord_cartesian(ylim = c(0, 1)) + guides(color = guide_legend(title = "Age group"))+ geom_label_repel(aes(label = labels, x = Year, y = value), nudge_x = 0.1, na.rm = FALSE, color = "black", size = 2.5, label.size = NA, fill = NA)

pSex_revised

ggsave(plot=pSex_revised, filename = "images/vt/prop_sex_allyears.png", width = 10, height = 8)

```

### Significance testing for sex over all years

```{r chiSquaredSexVT, echo=FALSE, message=FALSE, warning=FALSE}
# initialize empty matrix, from https://stackoverflow.com/questions/32712301/create-empty-data-frame-with-column-names-by-assigning-a-string-vector
# this will hold all the p values

chisq_sex <- data.frame(matrix(ncol = 2, nrow = 0))
cols <- c("Year", "p value")

for(i in 1:length(vtlistYoung)){
  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  young <- vtlistYoung[[i]] 
  old <- vtlistOld[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- young$Year[1]
  
  sexfreqtable <- young %>% group_by(Sex_revised) %>% tally()
  names(sexfreqtable)[names(sexfreqtable) == 'n'] <- 'Young'

  sextableold <- old %>% group_by(Sex_revised) %>% tally()
  names(sextableold)[names(sextableold) == 'n'] <- 'Old'

  sextable <- merge(sexfreqtable, sextableold, by="Sex_revised")
  
  # convert the "race" column into row names so I get a table with only numeric values, from   https://stackoverflow.com/questions/5555408/convert-the-values-in-a-column-into-row-names-in-an-existing-data-frame

  sexmat <- sextable[,-1]
  rownames(sexmat) <- sextable[,1]

  print(as.matrix(sexmat))
  
  # append year and p value to a vector, rounding to make it more legible
  # I'm using Fisher's exact test if >20% of cell counts are <5, if not use chi squared
  # I get >20% <5 from https://www.ccjm.org/content/84/9_suppl_2/e20
  pvalue <- ifelse(sum(sexmat < 5)/(ncol(sexmat) * nrow(sexmat)) > 0.2, round(fisher.test(as.matrix(sexmat))$p.value, 4), round(chisq.test(as.matrix(sexmat))$p.value, 4))
  
  results <- c(year, pvalue)
  
  # add this to the chi squared df 
  chisq_sex <- rbind(chisq_sex, results)

}

# for some reason it doesn't work if I do this before adding rows?
colnames(chisq_sex) <- cols

chisq_sex

```

### Race and sex of decedents {.tabset}

#### Young

```{r sexRaceTableYoung, echo=FALSE, message=FALSE, warning=FALSE}
# create dummy df with race sex info for all years

racesexmat <- data.frame(matrix(ncol = 4, nrow = 0))
cols <- c("Year", "Race_sex", "Proportion", "Color")

for(i in 1:length(vtlistYoung)){

  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  vt <- vtlistYoung[[i]] 
  # grab the year -- you only need the first element in the year column sivte all of them have the same value
  year <- vt$Year[1]

raceSextitle <- paste("Race/ethnicity and sex distribution for", year, "VT opioid-detected \n overdose deaths in younger people")

  
raceSex <- ggplot(data=vt, aes(x=Race_sex, ..count..)) + 
      geom_bar(aes(fill = Race_sex), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = -0.25, colour = "black", size = 7/.pt)+ ggtitle(raceSextitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7)) 

print(raceSex)

ggsave(plot=raceSex, filename = paste0("images/vt/raceSex", year, "_young.png"), width = 10, height = 8)

raceSexProp <- ggplot(data=vt, aes(x=Race_sex, ..count..)) + 
      geom_bar(aes(fill = Race_sex), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(vt), 3)), stat = "count", vjust = -0.25, colour = "black", size = 7/.pt)+ ggtitle(raceSextitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7)) 

print(raceSexProp)
ggsave(plot=raceSexProp, filename = paste0("images/vt/raceSexProp", year, "_young.png"), width = 10, height = 8)


# get the % of each race sex category
vt <- vt %>% 
  group_by(Race_sex) %>% 
  count() %>% 
  ungroup() %>% 
  mutate(per=`n`/sum(`n`)) %>% 
  arrange(desc(Race_sex))
# create a label out of this
vt$racesexlabel <- scales::percent(vt$per)

# create a year label [for plotting]
vt$Year <- year

# automatically assign the color based on race sex value -- not doing this in preprocessing because I don't want this to be a global change
# from https://stackoverflow.com/questions/39142596/r-ggplot-set-colour-for-specific-value

vt$color <- ifelse(vt$Race_sex == "Black Female", "#e6d0a1", ifelse(vt$Race_sex == "Black Male", "#E69F00", ifelse(vt$Race_sex == "Hispanic Female", "#b3d4e8", ifelse(vt$Race_sex == "Hispanic Male", "#56B4E9", ifelse(vt$Race_sex == "Other Female", "#cca1b9", ifelse(vt$Race_sex == "Other Male", "#CC79A7", ifelse(vt$Race_sex == "White Female", "#a1a1a1", "#000000")))))))

# grab the values i need from nc
results <- select(vt, Year, Race_sex, per, color)
  
# add this to the chi squared df 
racesexmat <- rbind(racesexmat, results)


# plot into a pie chart -- for some reason I have to reverse the colors to get them to work right
piechart <- ggplot(data=vt)+
  geom_bar(aes(x="", y=per, fill=Race_sex), stat="identity", width = 1)+
  coord_polar("y", start=0)+
  theme_void()+ geom_label_repel(aes(label = racesexlabel, x = "", y = cumsum(per) - per/2), nudge_x = 0.02,
                   size = 3, show.legend = F)+  scale_fill_manual(name = "Race/ethnicity and sex", values=rev(vt$color), labels = rev(vt$Race_sex))+ ggtitle(raceSextitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="right")  
# geom_text(aes(x=1, y = cumsum(per) - per/2, label=label)) for regular labels [not repelling]

print(piechart)

ggsave(plot=piechart, filename = paste0("images/vt/pie", year, "_young.png"), width = 10, height = 8)

}

# for some reason it doesn't work if I do this before adding rows?
colnames(racesexmat) <- cols

#racesexmat


# fill in missing years with NAs so they won't show up interpolated -- from https://stackoverflow.com/questions/65097273/r-how-to-fill-up-missing-year-values-in-a-data-frame
#racesexmat <- racesexmat %>% complete(Year = 2009:2023)

stackedarea <- ggplot(data = racesexmat, aes(x=Year, y=Proportion, fill=Race_sex)) + 
    geom_area(aes(position = 'stack', group = Race_sex))+  scale_fill_manual(name = "Race/ethnicity and sex", values=rev(racesexmat$Color), labels = rev(racesexmat$Race_sex)) + ggtitle("Race/ethnicity and sex distribution for VT opioid-detected \n overdose deaths in younger people")+theme_bw()+ scale_x_continuous(n.breaks = length(yearList))+theme(plot.title = element_text(hjust = 0.5), legend.position="right")

print(stackedarea)

ggsave(plot=stackedarea, filename = paste0("images/vt/rs_area_", year, "_young.png"), width = 10, height = 8)



```


#### Old

```{r sexRaceTableOld, echo=FALSE, message=FALSE, warning=FALSE}
# create dummy df with race sex info for all years

racesexmat <- data.frame(matrix(ncol = 4, nrow = 0))
cols <- c("Year", "Race_sex", "Proportion", "Color")

for(i in 1:length(vtlistOld)){

  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  vt <- vtlistOld[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- vt$Year[1]

raceSextitle <- paste("Race/ethnicity and sex distribution for", year, "VT opioid-detected \n overdose deaths in older people")

  
raceSex <- ggplot(data=vt, aes(x=Race_sex, ..count..)) + 
      geom_bar(aes(fill = Race_sex), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = -0.25, colour = "black", size = 7/.pt)+ ggtitle(raceSextitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7)) 

print(raceSex)

ggsave(plot=raceSex, filename = paste0("images/vt/raceSex", year, "_old.png"), width = 10, height = 8)

raceSexProp <- ggplot(data=vt, aes(x=Race_sex, ..count..)) + 
      geom_bar(aes(fill = Race_sex), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(vt), 3)), stat = "count", vjust = -0.25, colour = "black", size = 7/.pt)+ ggtitle(raceSextitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7)) 

print(raceSexProp)

ggsave(plot=raceSexProp, filename = paste0("images/vt/raceSexProp", year, "_old.png"), width = 10, height = 8)


# get the % of each race sex category
vt <- vt %>% 
  group_by(Race_sex) %>% 
  count() %>% 
  ungroup() %>% 
  mutate(per=`n`/sum(`n`)) %>% 
  arrange(desc(Race_sex))
# create a label out of this
vt$racesexlabel <- scales::percent(vt$per)

# create a year label [for plotting]
vt$Year <- year

# automatically assign the color based on race sex value -- not doing this in preprocessing because I don't want this to be a global change
# from https://stackoverflow.com/questions/39142596/r-ggplot-set-colour-for-specific-value

vt$color <- ifelse(vt$Race_sex == "Black Female", "#e6d0a1", ifelse(vt$Race_sex == "Black Male", "#E69F00", ifelse(vt$Race_sex == "Hispanic Female", "#b3d4e8", ifelse(vt$Race_sex == "Hispanic Male", "#56B4E9", ifelse(vt$Race_sex == "Other Female", "#cca1b9", ifelse(vt$Race_sex == "Other Male", "#CC79A7", ifelse(vt$Race_sex == "White Female", "#a1a1a1", "#000000")))))))

# grab the values i need from nc
results <- select(vt, Year, Race_sex, per, color)
  
# add this to the chi squared df 
racesexmat <- rbind(racesexmat, results)


# plot into a pie chart -- for some reason I have to reverse the colors to get them to work right
piechart <- ggplot(data=vt)+
  geom_bar(aes(x="", y=per, fill=Race_sex), stat="identity", width = 1)+
  coord_polar("y", start=0)+
  theme_void()+ geom_label_repel(aes(label = racesexlabel, x = "", y = cumsum(per) - per/2), nudge_x = 0.02,
                   size = 3, show.legend = F)+  scale_fill_manual(name = "Race/ethnicity and sex", values=rev(vt$color), labels = rev(vt$Race_sex))+ ggtitle(raceSextitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="right")  
# geom_text(aes(x=1, y = cumsum(per) - per/2, label=label)) for regular labels [not repelling]

print(piechart)

ggsave(plot=piechart, filename = paste0("images/vt/pie", year, "_old.png"), width = 10, height = 8)


}

# for some reason it doesn't work if I do this before adding rows?
colnames(racesexmat) <- cols

# racesexmat


# fill in missing years with NAs so they won't show up interpolated -- from https://stackoverflow.com/questions/65097273/r-how-to-fill-up-missing-year-values-in-a-data-frame
#racesexmat <- racesexmat %>% complete(Year = 2009:2023)

stackedarea <- ggplot(data = racesexmat, aes(x=Year, y=Proportion, fill=Race_sex)) + 
    geom_area(aes(position = 'stack', group = Race_sex))+  scale_fill_manual(name = "Race/ethnicity and sex", values=rev(racesexmat$Color), labels = rev(racesexmat$Race_sex)) + ggtitle("Race/ethnicity and sex distribution for VT opioid-detected \n overdose deaths in older people")+theme_bw() + theme(plot.title = element_text(hjust = 0.5), legend.position="right")+ scale_x_continuous(n.breaks = length(yearList))

print(stackedarea)

ggsave(plot=stackedarea, filename = paste0("images/vt/rs_area_", year, "_old.png"), width = 10, height = 8)



```

### Significance testing for race and sex {.tabset}


```{r chiSquaredRaceSex, echo=FALSE, message=FALSE, warning=FALSE}
# initialize empty matrix, from https://stackoverflow.com/questions/32712301/create-empty-data-frame-with-column-names-by-assigning-a-string-vector
# this will hold all the p values

chisq_racesex <- data.frame(matrix(ncol = 2, nrow = 0))
cols <- c("Year", "p value")

for(i in 1:length(vtlistYoung)){
  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  young <- vtlistYoung[[i]] 
  old <- vtlistOld[[i]] 
  # grab the year -- you only need the first element in the year column sivte all of them have the same value
  year <- young$Year[1]
  
  racesexfreqtable <- young %>% group_by(Race_sex) %>% tally()
  names(racesexfreqtable)[names(racesexfreqtable) == 'n'] <- 'Young'

  racesextableold <- old %>% group_by(Race_sex) %>% tally()
  names(racesextableold)[names(racesextableold) == 'n'] <- 'Old'

  racesextable <- merge(racesexfreqtable, racesextableold, by="Race_sex")
  
  # convert the "race" column into row names so I get a table with only numeric values, from   https://stackoverflow.com/questions/5555408/convert-the-values-in-a-column-into-row-names-in-an-existing-data-frame

  racesexmat <- racesextable[,-1]
  rownames(racesexmat) <- racesextable[,1]

  print(as.matrix(racesexmat))
  
  # append year and p value to a vector, rounding to make it more legible
  # I'm using Fisher's exact test if >20% of cell counts are <5, if not use chi squared
  # I get >20% <5 from https://www.ccjm.org/content/84/9_suppl_2/e20
  pvalue <- ifelse(sum(racesexmat < 5)/(ncol(racesexmat) * nrow(racesexmat)) > 0.2, round(fisher.test(as.matrix(racesexmat))$p.value, 4), round(chisq.test(as.matrix(racesexmat))$p.value, 4))
  
  results <- c(year, pvalue)
  
  # add this to the chi squared df 
  chisq_racesex <- rbind(chisq_racesex, results)

}

# for some reason it doesn't work if I do this before adding rows?
colnames(chisq_racesex) <- cols

chisq_racesex

racesexmat

```

### Day of week {.tabset}

#### Young

```{r weekYoung, echo=FALSE, message=FALSE, warning=FALSE}
for(i in 1:length(vtlistYoung)){

  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  vt <- vtlistYoung[[i]] 
  # grab the year -- you only need the first element in the year column sivte all of them have the same value
  year <- vt$Year[1]

weekdaytitle <- paste("ODs per weekday for", year, "VT opioid-detected overdose deaths in younger people")

  
# ordering of variables on x axis from https://www.statology.org/ggplot-order-x-axis/
# make sure to use scale_x_discrete!

weekday <- ggplot(data=vt, aes(x=factor(Weekday, level=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")), ..count..)) + geom_bar(aes(fill = as.character(Weekday)), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 0), breaks = vt$Weekday, labels = vt$Weekday) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle(weekdaytitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))+xlab("Weekday")

print(weekday)

ggsave(plot=weekday, filename = paste0("images/vt/weekday_", year, "_young.png"), width = 10, height = 8)


weekday_prop <- ggplot(data=vt, aes(x=factor(Weekday, level=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")), round(..count../nrow(vt), 3))) + geom_bar(aes(fill = as.character(Weekday)), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 0), breaks = vt$Weekday, labels = vt$Weekday) + theme_light() + geom_text(aes(label = round(..count../nrow(vt), 3)), stat = "count", vjust = 0, colour = "black") + ggtitle(weekdaytitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))+xlab("Weekday")+ylab("Proportion")

print(weekday_prop)

ggsave(plot=weekday_prop, filename = paste0("images/vt/weekday_", year, "_young_prop.png"), width = 10, height = 8)
}
```


#### Old

```{r weekOld, echo=FALSE, message=FALSE, warning=FALSE}
for(i in 1:length(vtlistOld)){

  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  vt <- vtlistOld[[i]] 
  # grab the year -- you only need the first element in the year column sivte all of them have the same value
  year <- vt$year[1]

weekdaytitle <- paste("ODs per weekday for", year, "VT opioid-detected overdose deaths in older people")

  
# ordering of variables on x axis from https://www.statology.org/ggplot-order-x-axis/
# make sure to use scale_x_discrete!

weekday <- ggplot(data=vt, aes(x=factor(Weekday, level=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")), ..count..)) + geom_bar(aes(fill = as.character(Weekday)), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 0), breaks = vt$Weekday, labels = vt$Weekday) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle(weekdaytitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))+xlab("Weekday")

print(weekday)

ggsave(plot=weekday, filename = paste0("images/vt/weekday_", year, "_old.png"), width = 10, height = 8)


weekday_prop <- ggplot(data=vt, aes(x=factor(Weekday, level=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")), round(..count../nrow(vt), 3))) + geom_bar(aes(fill = as.character(Weekday)), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 0), breaks = vt$Weekday, labels = vt$Weekday) + theme_light() + geom_text(aes(label = round(..count../nrow(vt), 3)), stat = "count", vjust = 0, colour = "black") + ggtitle(weekdaytitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))+xlab("Weekday")+ylab("Proportion")

print(weekday_prop)

ggsave(plot=weekday_prop, filename = paste0("images/vt/weekday_", year, "_old_prop.png"), width = 10, height = 8)
}
```

### Significance testing for day of week {.tabset}

```{r chiSquaredWeekday, echo=FALSE, message=FALSE, warning=FALSE}
# initialize empty matrix, from https://stackoverflow.com/questions/32712301/create-empty-data-frame-with-column-names-by-assigning-a-string-vector
# this will hold all the p values

chisq_weekday <- data.frame(matrix(ncol = 2, nrow = 0))
cols <- c("Year", "p value")

for(i in 1:length(vtlistYoung)){
  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  young <- vtlistYoung[[i]] 
  old <- vtlistOld[[i]] 
  # grab the year -- you only need the first element in the year column sivte all of them have the same value
  year <- young$Year[1]
  
  # omit NAs because it seems to be having issues with that -- from https://stackoverflow.com/questions/11254524/omit-rows-containing-specific-column-of-na
  young <- young[!is.na(young$Weekday),]
  old <- old[!is.na(old$Weekday),]

  
  weekdayfreqtable <- young %>% group_by(Weekday) %>% tally()
  names(weekdayfreqtable)[names(weekdayfreqtable) == 'n'] <- 'Young'

  weekdaytableold <- old %>% group_by(Weekday) %>% tally()
  names(weekdaytableold)[names(weekdaytableold) == 'n'] <- 'Old'

  weekdaytable <- merge(weekdayfreqtable, weekdaytableold, by="Weekday")
  
  # convert the "race" column into row names so I get a table with only numeric values, from   https://stackoverflow.com/questions/5555408/convert-the-values-in-a-column-into-row-names-in-an-existing-data-frame

  weekdaymat <- weekdaytable[,-1]
  rownames(weekdaymat) <- weekdaytable[,1]

  print(as.matrix(weekdaymat))
  
  # append year and p value to a vector, rounding to make it more legible
  # I'm using Fisher's exact test if >20% of cell counts are <5, if not use chi squared
  # I get >20% <5 from https://www.ccjm.org/content/84/9_suppl_2/e20
  pvalue <- ifelse(sum(weekdaymat < 5)/(ncol(weekdaymat) * nrow(weekdaymat)) > 0.2, round(fisher.test(as.matrix(weekdaymat))$p.value, 4), round(chisq.test(as.matrix(weekdaymat))$p.value, 4))
  
  results <- c(year, pvalue)
  
  # add this to the chi squared df 
  chisq_weekday <- rbind(chisq_weekday, results)

}

# for some reason it doesn't work if I do this before adding rows?
colnames(chisq_weekday) <- cols

chisq_weekday

```

### Month {.tabset}

#### Young

```{r monthYoung, echo=FALSE, message=FALSE, warning=FALSE}
for(i in 1:length(vtlistYoung)){

  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  vt <- vtlistYoung[[i]] 
  # grab the year -- you only need the first element in the year column sivte all of them have the same value
  year <- vt$Year[1]
  
  monthtitle <- paste("ODs per month for", year, "VT opioid-detected overdose deaths in younger people")

  
month <- ggplot(data=vt, aes(x=Month, ..count..)) + geom_bar(aes(fill = as.character(Month)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0), breaks = vt$Month, labels = vt$Month) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle(monthtitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))

print(month)

ggsave(plot=month, filename = paste0("images/vt/month_", year, "_young.png"), width = 10, height = 8)

month_prop <- ggplot(data=vt, aes(x=Month, round(..count../nrow(vt), 3))) + geom_bar(aes(fill = as.character(Month)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0), breaks = vt$Month, labels = vt$Month) + theme_light() + geom_text(aes(label =  round(..count../nrow(vt), 3)), stat = "count", vjust = 0, colour = "black") + ggtitle(monthtitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))+ylab("Proportion")

print(month_prop)

ggsave(plot=month_prop, filename = paste0("images/vt/month_", year, "_young_prop.png"), width = 10, height = 8)


}

```


#### Old

```{r monthOld, echo=FALSE, message=FALSE, warning=FALSE}
for(i in 1:length(vtlistOld)){

  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  vt <- vtlistOld[[i]] 
  # grab the year -- you only need the first element in the year column sivte all of them have the same value
  year <- vt$year[1]
  
  monthtitle <- paste("ODs per month for", year, "VT opioid-detected overdose deaths in older people")

  
month <- ggplot(data=vt, aes(x=Month, ..count..)) + geom_bar(aes(fill = as.character(Month)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0), breaks = vt$Month, labels = vt$Month) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle(monthtitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))

print(month)

ggsave(plot=month, filename = paste0("images/vt/month_", year, "_old.png"), width = 10, height = 8)

month_prop <- ggplot(data=vt, aes(x=Month, round(..count../nrow(vt), 3))) + geom_bar(aes(fill = as.character(Month)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0), breaks = vt$Month, labels = vt$Month) + theme_light() + geom_text(aes(label =  round(..count../nrow(vt), 3)), stat = "count", vjust = 0, colour = "black") + ggtitle(monthtitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))+ylab("Proportion")

print(month_prop)

ggsave(plot=month_prop, filename = paste0("images/vt/month_", year, "_old_prop.png"), width = 10, height = 8)


}

```

### Significance testing for month {.tabset}

```{r chiSquaredMonth, echo=FALSE, message=FALSE, warning=FALSE}
# initialize empty matrix, from https://stackoverflow.com/questions/32712301/create-empty-data-frame-with-column-names-by-assigning-a-string-vector
# this will hold all the p values

chisq_month <- data.frame(matrix(ncol = 2, nrow = 0))
cols <- c("Year", "p value")

for(i in 1:length(vtlistYoung)){
  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  young <- vtlistYoung[[i]] 
  old <- vtlistOld[[i]] 
  # grab the year -- you only need the first element in the year column sivte all of them have the same value
  year <- young$Year[1]
  
  monthfreqtable <- young %>% group_by(Month) %>% tally()
  names(monthfreqtable)[names(monthfreqtable) == 'n'] <- 'Young'

  monthtableold <- old %>% group_by(Month) %>% tally()
  names(monthtableold)[names(monthtableold) == 'n'] <- 'Old'

  monthtable <- merge(monthfreqtable, monthtableold, by="Month")
  
  # convert the "race" column into row names so I get a table with only numeric values, from   https://stackoverflow.com/questions/5555408/convert-the-values-in-a-column-into-row-names-in-an-existing-data-frame

  monthmat <- monthtable[,-1]
  rownames(monthmat) <- monthtable[,1]

  print(as.matrix(monthmat))
  
  # append year and p value to a vector, rounding to make it more legible
  # I'm using Fisher's exact test if >20% of cell counts are <5, if not use chi squared
  # I get >20% <5 from https://www.ccjm.org/content/84/9_suppl_2/e20
  pvalue <- ifelse(sum(monthmat < 5)/(ncol(monthmat) * nrow(monthmat)) > 0.2, round(fisher.test(as.matrix(monthmat))$p.value, 4), round(chisq.test(as.matrix(monthmat))$p.value, 4))
  
  results <- c(year, pvalue)
  
  # add this to the chi squared df 
  chisq_month <- rbind(chisq_month, results)

}

# for some reason it doesn't work if I do this before adding rows?
colnames(chisq_month) <- cols

chisq_month

```


### Proportion of deaths involving fentanyl

```{r propFent, echo=FALSE, message=FALSE, warning=FALSE}
# unfortunately there are different variable names for each vt file so a for loop won't work :/

# get proportion of fent for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
fentyoung <- c(sum(vt_2009_young$fentonly == 1)/nrow(vt_2009_young), sum(vt_2010_young$fentonly == 1)/nrow(vt_2010_young), sum(vt_2012_young$fentonly == 1)/nrow(vt_2012_young), sum(vt_2013_young$fentonly == 1)/nrow(vt_2013_young), sum(vt_2014_young$fentonly == 1)/nrow(vt_2014_young), sum(vt_2015_young$fentonly == 1)/nrow(vt_2015_young), sum(vt_2017_young$fentonly == 1)/nrow(vt_2017_young), sum(vt_2019_young$fentonly == 1)/nrow(vt_2019_young), sum(vt_2020_young$fentonly == 1)/nrow(vt_2020_young))

# get proportion for older people
fentold <- c(sum(vt_2009_old$fentonly == 1)/nrow(vt_2009_old), sum(vt_2010_old$fentonly == 1)/nrow(vt_2010_old), sum(vt_2012_old$fentonly == 1)/nrow(vt_2012_old), sum(vt_2013_old$fentonly == 1)/nrow(vt_2013_old), sum(vt_2014_old$fentonly == 1)/nrow(vt_2014_old), sum(vt_2015_old$fentonly == 1)/nrow(vt_2015_old), sum(vt_2017_old$fentonly == 1)/nrow(vt_2017_old), sum(vt_2019_old$fentonly == 1)/nrow(vt_2019_old), sum(vt_2020_old$fentonly == 1)/nrow(vt_2020_old))

# combine vectors into df
propfent <- data.frame(yearList, fentyoung, fentold)

# change the col names to be more informative
names(propfent) <- c("Year", "Younger", "Older")

# create grouped bar plot, from https://stackoverflow.com/questions/72449952/making-a-grouped-bar-chart-in-r
# getting the labels in the right position from https://stackoverflow.com/questions/37587288/ggplot-label-bars-in-grouped-bar-plot
# info on changing legend title from here https://stackoverflow.com/questions/41649929/how-to-modify-a-legend-title-in-a-barplot-with-ggplot
# the scale x continuous is there because for some reason adding 2023 data messed up the x axis, see https://stackoverflow.com/questions/72051789/how-to-display-all-values-on-x-and-y-axes-in-r-ggplot2

# ivtreasing line size from https://stackoverflow.com/questions/14794599/how-to-change-line-width-in-ggplot
# line graph info from https://www.sthda.com/english/wiki/ggplot2-line-plot-quick-start-guide-r-software-and-data-visualization


pfentlong  <- propfent |>
  # Reshape data to long format
  pivot_longer(Younger:Older)

# create labels with the value of each proportion
pfentlong$labels <- round(pfentlong$value, 2)

pfent <-  ggplot(data = pfentlong, aes(x = Year, y = value, fill = name, group=name, color=factor(name))) +
  # Implement a grouped bar chart
  # Implement a line graph
  geom_line(stat = "identity", size = 1) + geom_point(size = 2) + theme_light() + ggtitle("Proportion of fentanyl-detected deaths in VT \n by age group, 2009-2023") + ylab("Proportion") + theme(plot.title = element_text(hjust = 0.5)) + scale_fill_discrete(guide="none") + scale_x_continuous(breaks=propfent$Year) + xlab("Year")+ coord_cartesian(ylim = c(0, 1)) + guides(color = guide_legend(title = "Age group"))+ geom_label_repel(aes(label = labels, x = Year, y = value), nudge_x = 0.05, na.rm = FALSE, color = "black", size = 2.5, label.size = NA, fill = NA)

pfent

ggsave(plot=pfent, filename = "images/vt/prop_fent.png", width = 10, height = 8)
```


### Significance testing for the number of fentanyl-detected deaths

```{r chiSquaredFent, echo=FALSE, message=FALSE, warning=FALSE}
# get frequency of each race for old and young, from https://stackoverflow.com/questions/48562288/frequevty-of-each-unique-combination-in-data-frame

# get proportion of fent for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
fentyoungcount <- c(sum(vt_2009_young$fentonly == 1), sum(vt_2010_young$fentonly == 1), sum(vt_2012_young$fentonly == 1), sum(vt_2013_young$fentonly == 1), sum(vt_2014_young$fentonly == 1), sum(vt_2015_young$fentonly == 1), sum(vt_2017_young$fentonly == 1), sum(vt_2019_young$fentonly == 1), sum(vt_2020_young$fentonly == 1))

# get proportion for older people
fentoldcount <- c(sum(vt_2009_old$fentonly == 1), sum(vt_2010_old$fentonly == 1), sum(vt_2012_old$fentonly == 1), sum(vt_2013_old$fentonly == 1), sum(vt_2014_old$fentonly == 1), sum(vt_2015_old$fentonly == 1), sum(vt_2017_old$fentonly == 1), sum(vt_2019_old$fentonly == 1), sum(vt_2020_old$fentonly == 1))

# combine vectors into df
fentcount <- data.frame(yearList, fentyoungcount, fentoldcount)

# change the col names to be more informative
names(fentcount) <- c("Year", "Younger", "Older")

# convert the "race_sex" column into row names so I get a table with only numeric values, from https://stackoverflow.com/questions/5555408/convert-the-values-in-a-column-into-row-names-in-an-existing-data-frame

fentchi <- fentcount[,-1]
rownames(fentchi) <- fentcount[,1]

as.matrix(fentchi)

# looks like I can just use any matrix for chisq.test, from https://researchguides.library.vanderbilt.edu/c.php?g=156859&p=3058795

  
  # append year and p value to a vector, rounding to make it more legible
  # I'm using Fisher's exact test if >20% of cell counts are <5, if not use chi squared
  # I get >20% <5 from https://www.ccjm.org/content/84/9_suppl_2/e20
  pvalue <- ifelse(sum(fentchi < 5)/(ncol(fentchi) * nrow(fentchi)) > 0.2, fisher.test(as.matrix(fentchi), workspace=2e7)$p.value, chisq.test(as.matrix(fentchi))$p.value)
  
  pvalue
  

```

### Proportion of deaths involving xylazine

```{r propXyl, echo=FALSE, message=FALSE, warning=FALSE}

# unfortunately there are different variable names for each vt file so a for loop won't work :/

# get proportion of fent for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
xylyoung <- c(sum(vt_2009_young$xylazine == 1)/nrow(vt_2009_young), sum(vt_2010_young$xylazine == 1)/nrow(vt_2010_young),  sum(vt_2012_young$xylazine == 1)/nrow(vt_2012_young), sum(vt_2013_young$xylazine == 1)/nrow(vt_2013_young), sum(vt_2014_young$xylazine == 1)/nrow(vt_2014_young), sum(vt_2015_young$xylazine == 1)/nrow(vt_2015_young), sum(vt_2017_young$xylazine == 1)/nrow(vt_2017_young), sum(vt_2019_young$xylazine == 1)/nrow(vt_2019_young), sum(vt_2020_young$xylazine == 1)/nrow(vt_2020_young))

# get proportion for older people
xylold <- c(sum(vt_2009_old$xylazine == 1)/nrow(vt_2009_old), sum(vt_2010_old$xylazine == 1)/nrow(vt_2010_old),  sum(vt_2012_old$xylazine == 1)/nrow(vt_2012_old), sum(vt_2013_old$xylazine == 1)/nrow(vt_2013_old), sum(vt_2014_old$xylazine == 1)/nrow(vt_2014_old), sum(vt_2015_old$xylazine == 1)/nrow(vt_2015_old), sum(vt_2017_old$xylazine == 1)/nrow(vt_2017_old), sum(vt_2019_old$xylazine == 1)/nrow(vt_2019_old), sum(vt_2020_old$xylazine == 1)/nrow(vt_2020_old))


# combine vectors into df
propxyl <- data.frame(yearList, xylyoung, xylold)

# change the col names to be more informative
names(propxyl) <- c("Year", "Younger", "Older")

# create grouped bar plot, from https://stackoverflow.com/questions/72449952/making-a-grouped-bar-chart-in-r
# getting the labels in the right position from https://stackoverflow.com/questions/37587288/ggplot-label-bars-in-grouped-bar-plot
# info on changing legend title from here https://stackoverflow.com/questions/41649929/how-to-modify-a-legend-title-in-a-barplot-with-ggplot

# ivtreasing line size from https://stackoverflow.com/questions/14794599/how-to-change-line-width-in-ggplot
# line graph info from https://www.sthda.com/english/wiki/ggplot2-line-plot-quick-start-guide-r-software-and-data-visualization


pxyllong  <- propxyl |>
  # Reshape data to long format
  pivot_longer(Younger:Older)

# create labels with the value of each proportion
pxyllong$labels <- round(pxyllong$value, 2)

pxyl <-  ggplot(data = pxyllong, aes(x = Year, y = value, fill = name, group=name, color=factor(name))) +
  # Implement a grouped bar chart
  # Implement a line graph
  geom_line(stat = "identity", size = 1) + geom_point(size = 2) + theme_light() + ggtitle("Proportion of xylazine in opioid-detected overdose deaths \n in VT by age group, 2009-2023") + ylab("Proportion") + theme(plot.title = element_text(hjust = 0.5)) + scale_fill_discrete(guide="none") + scale_x_continuous(breaks=propxyl$Year) + xlab("Year")+ coord_cartesian(ylim = c(0, 1)) + guides(color = guide_legend(title = "Age group"))+ geom_label_repel(aes(label = labels, x = Year, y = value), nudge_x = 0.05, na.rm = FALSE, color = "black", size = 2.5, label.size = NA, fill = NA)

pxyl

ggsave(plot=pxyl, filename = "images/vt/prop_xyl_old.png", width = 10, height = 8)

```

### Significance testing for the number of xylazine-detected deaths

Didn't expect this to be informative and it wasn't, but I've done it for the sake of completeness. 

```{r chiSquaredXyl, echo=FALSE, message=FALSE, warning=FALSE}
# get frequevty of each race for old and young, from https://stackoverflow.com/questions/48562288/frequevty-of-each-unique-combination-in-data-frame

# get proportion of fent for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
xylyoungcount <- c(sum(vt_2009_young$xylazine == 1), sum(vt_2010_young$xylazine == 1), sum(vt_2012_young$xylazine == 1), sum(vt_2013_young$xylazine == 1), sum(vt_2014_young$xylazine == 1), sum(vt_2015_young$xylazine == 1), sum(vt_2017_young$xylazine == 1), sum(vt_2019_young$xylazine == 1), sum(vt_2020_young$xylazine == 1))

# get proportion for older people
xyloldcount <- c(sum(vt_2009_old$xylazine == 1), sum(vt_2010_old$xylazine == 1), sum(vt_2012_old$xylazine == 1), sum(vt_2013_old$xylazine == 1), sum(vt_2014_old$xylazine == 1), sum(vt_2015_old$xylazine == 1), sum(vt_2017_old$xylazine == 1), sum(vt_2019_old$xylazine == 1), sum(vt_2020_old$xylazine == 1))

# combine vectors into df
xylcount <- data.frame(yearList, xylyoungcount, xyloldcount)

# change the col names to be more informative
names(xylcount) <- c("Year", "Younger", "Older")

# convert the "race_sex" column into row names so I get a table with only numeric values, from https://stackoverflow.com/questions/5555408/convert-the-values-in-a-column-into-row-names-in-an-existing-data-frame

xylchi <- xylcount[,-1]
rownames(xylchi) <- xylcount[,1]

as.matrix(xylchi)

# looks like I can just use any matrix for chisq.test, from https://researchguides.library.vanderbilt.edu/c.php?g=156859&p=3058795

  # I'm using Fisher's exact test if >20% of cell counts are <5, if not use chi squared
  # I get >20% <5 from https://www.ccjm.org/content/84/9_suppl_2/e20
  pvalue <- ifelse(sum(xylchi < 5)/(ncol(xylchi) * nrow(xylchi)) > 0.2, fisher.test(as.matrix(xylchi), workspace=2e6)$p.value, chisq.test(as.matrix(xylchi))$p.value)
  
  pvalue
```

### Proportion of deaths involving gabapentin

```{r propGab, echo=FALSE, message=FALSE, warning=FALSE}

# unfortunately there are different variable names for each vt file so a for loop won't work :/

# get proportion of gabapentin for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
gabyoung <- c(sum(vt_2009_young$gabapentin == 1)/nrow(vt_2009_young), sum(vt_2010_young$gabapentin == 1)/nrow(vt_2010_young), sum(vt_2012_young$gabapentin == 1)/nrow(vt_2012_young), sum(vt_2013_young$gabapentin == 1)/nrow(vt_2013_young), sum(vt_2014_young$gabapentin == 1)/nrow(vt_2014_young), sum(vt_2015_young$gabapentin == 1)/nrow(vt_2015_young), sum(vt_2017_young$gabapentin == 1)/nrow(vt_2017_young), sum(vt_2019_young$gabapentin == 1)/nrow(vt_2019_young), sum(vt_2020_young$gabapentin == 1)/nrow(vt_2020_young))

# get proportion for older people
gabold <- c(sum(vt_2009_old$gabapentin == 1)/nrow(vt_2009_old), sum(vt_2010_old$gabapentin == 1)/nrow(vt_2010_old), sum(vt_2012_old$gabapentin == 1)/nrow(vt_2012_old), sum(vt_2013_old$gabapentin == 1)/nrow(vt_2013_old), sum(vt_2014_old$gabapentin == 1)/nrow(vt_2014_old), sum(vt_2015_old$gabapentin == 1)/nrow(vt_2015_old), sum(vt_2017_old$gabapentin == 1)/nrow(vt_2017_old), sum(vt_2019_old$gabapentin == 1)/nrow(vt_2019_old), sum(vt_2020_old$gabapentin == 1)/nrow(vt_2020_old))

# combine vectors into df
propgab <- data.frame(yearList, gabyoung, gabold)

# change the col names to be more informative
names(propgab) <- c("Year", "Younger", "Older")

# create grouped bar plot, from https://stackoverflow.com/questions/72449952/making-a-grouped-bar-chart-in-r
# getting the labels in the right position from https://stackoverflow.com/questions/37587288/ggplot-label-bars-in-grouped-bar-plot
# info on changing legend title from here https://stackoverflow.com/questions/41649929/how-to-modify-a-legend-title-in-a-barplot-with-ggplot

# ivtreasing line size from https://stackoverflow.com/questions/14794599/how-to-change-line-width-in-ggplot
# line graph info from https://www.sthda.com/english/wiki/ggplot2-line-plot-quick-start-guide-r-software-and-data-visualization


pgablong  <- propgab |>
  # Reshape data to long format
  pivot_longer(Younger:Older)

# create labels with the value of each proportion
pgablong$labels <- round(pgablong$value, 2)

pgab <-  ggplot(data = pgablong, aes(x = Year, y = value, fill = name, group=name, color=factor(name))) +
  # Implement a grouped bar chart
  # Implement a line graph
  geom_line(stat = "identity", size = 1) + geom_point(size = 2) + theme_light() + ggtitle("Proportion of gabapentin in opioid-detected overdose deaths \n in VT by age group, 2009-2023") + ylab("Proportion") + theme(plot.title = element_text(hjust = 0.5)) + scale_fill_discrete(guide="none") + scale_x_continuous(breaks=propgab$Year) + xlab("Year")+ coord_cartesian(ylim = c(0, 1)) + guides(color = guide_legend(title = "Age group"))+ geom_label_repel(aes(label = labels, x = Year, y = value), nudge_x = 0.05, na.rm = FALSE, color = "black", size = 2.5, label.size = NA, fill = NA)

pgab

ggsave(plot=pgab, filename = "images/vt/prop_gab_old.png", width = 10, height = 8)
```

### Significance testing for the number of gabepentin-detected deaths

```{r chiSquaredGab, echo=FALSE, message=FALSE, warning=FALSE}
# get frequevty of each race for old and young, from https://stackoverflow.com/questions/48562288/frequevty-of-each-unique-combination-in-data-frame

# get proportion of gab for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
gabyoungcount <- c(sum(vt_2009_young$gabapentin == 1), sum(vt_2010_young$gabapentin == 1), sum(vt_2012_young$gabapentin == 1), sum(vt_2013_young$gabapentin == 1), sum(vt_2014_young$gabapentin == 1), sum(vt_2015_young$gabapentin == 1), sum(vt_2017_young$gabapentin == 1), sum(vt_2019_young$gabapentin == 1), sum(vt_2020_young$gabapentin == 1))

# get proportion for older people
gaboldcount <- c(sum(vt_2009_old$gabapentin == 1), sum(vt_2010_old$gabapentin == 1), sum(vt_2012_old$gabapentin == 1), sum(vt_2013_old$gabapentin == 1), sum(vt_2014_old$gabapentin == 1), sum(vt_2015_old$gabapentin == 1), sum(vt_2017_old$gabapentin == 1), sum(vt_2019_old$gabapentin == 1), sum(vt_2020_old$gabapentin == 1))

# combine vectors into df
gabcount <- data.frame(yearList, gabyoungcount, gaboldcount)

# change the col names to be more informative
names(gabcount) <- c("Year", "Younger", "Older")

# convert the "race_sex" column into row names so I get a table with only numeric values, from https://stackoverflow.com/questions/5555408/convert-the-values-in-a-column-into-row-names-in-an-existing-data-frame

gabchi <- gabcount[,-1]
rownames(gabchi) <- gabcount[,1]

as.matrix(gabchi)

# looks like I can just use any matrix for chisq.test, from https://researchguides.library.vanderbilt.edu/c.php?g=156859&p=3058795
# getting rid of the nonzero entries for testing

  # I'm using Fisher's exact test if >20% of cell counts are <5, if not use chi squared
  # I get >20% <5 from https://www.ccjm.org/content/84/9_suppl_2/e20
  pvalue <- ifelse(sum(gabchi < 5)/(ncol(gabchi) * nrow(gabchi)) > 0.2, fisher.test(as.matrix(gabchi), workspace=2e6)$p.value, chisq.test(as.matrix(gabchi))$p.value)
  
  pvalue
```

### Proportion of deaths involving pharmaceutical opioids

```{r propPharma, echo=FALSE, message=FALSE, warning=FALSE}

# unfortunately there are different variable names for each vt file so a for loop won't work :/

# get proportion of pharma for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
pharmayoung <- c(sum(vt_2009_young$pharma == 1)/nrow(vt_2009_young), sum(vt_2010_young$pharma == 1)/nrow(vt_2010_young), sum(vt_2012_young$pharma == 1)/nrow(vt_2012_young), sum(vt_2013_young$pharma == 1)/nrow(vt_2013_young), sum(vt_2014_young$pharma == 1)/nrow(vt_2014_young), sum(vt_2015_young$pharma == 1)/nrow(vt_2015_young), sum(vt_2017_young$pharma == 1)/nrow(vt_2017_young), sum(vt_2019_young$pharma == 1)/nrow(vt_2019_young), sum(vt_2020_young$pharma == 1)/nrow(vt_2020_young))

# get proportion for older people
pharmaold <- c(sum(vt_2009_old$pharma == 1)/nrow(vt_2009_old), sum(vt_2010_old$pharma == 1)/nrow(vt_2010_old), sum(vt_2012_old$pharma == 1)/nrow(vt_2012_old), sum(vt_2013_old$pharma == 1)/nrow(vt_2013_old), sum(vt_2014_old$pharma == 1)/nrow(vt_2014_old), sum(vt_2015_old$pharma == 1)/nrow(vt_2015_old), sum(vt_2017_old$pharma == 1)/nrow(vt_2017_old), sum(vt_2019_old$pharma == 1)/nrow(vt_2019_old), sum(vt_2020_old$pharma == 1)/nrow(vt_2020_old))

# combine vectors into df
proppharma <- data.frame(yearList, pharmayoung, pharmaold)

# change the col names to be more informative
names(proppharma) <- c("Year", "Younger", "Older")

# create grouped bar plot, from https://stackoverflow.com/questions/72449952/making-a-grouped-bar-chart-in-r
# getting the labels in the right position from https://stackoverflow.com/questions/37587288/ggplot-label-bars-in-grouped-bar-plot
# info on changing legend title from here https://stackoverflow.com/questions/41649929/how-to-modify-a-legend-title-in-a-barplot-with-ggplot

# ivtreasing line size from https://stackoverflow.com/questions/14794599/how-to-change-line-width-in-ggplot
# line graph info from https://www.sthda.com/english/wiki/ggplot2-line-plot-quick-start-guide-r-software-and-data-visualization


ppharmalong  <- proppharma |>
  # Reshape data to long format
  pivot_longer(Younger:Older)

# create labels with the value of each proportion
ppharmalong$labels <- round(ppharmalong$value, 2)

ppharma <-  ggplot(data = ppharmalong, aes(x = Year, y = value, fill = name, group=name, color=factor(name))) +
  # Implement a grouped bar chart
  # Implement a line graph
  geom_line(stat = "identity", size = 1) + geom_point(size = 2) + theme_light() + ggtitle("Proportion of pharmaceutical opioid-detected overdose deaths in VT \n by age group, 2009-2023") + ylab("Proportion") + theme(plot.title = element_text(hjust = 0.5)) + scale_fill_discrete(guide="none") + scale_x_continuous(breaks=proppharma$Year) + xlab("Year")+ coord_cartesian(ylim = c(0, 1)) + guides(color = guide_legend(title = "Age group"))+ geom_label_repel(aes(label = labels, x = Year, y = value), nudge_x = 0.05, na.rm = FALSE, color = "black", size = 2.5, label.size = NA, fill = NA)

ppharma

ggsave(plot=ppharma, filename = "images/vt/prop_pharma_old.png", width = 10, height = 8)
```

### Significance testing for the number of pharma-detected deaths

```{r chiSquaredPharma, echo=FALSE, message=FALSE, warning=FALSE}
# get frequevty of each race for old and young, from https://stackoverflow.com/questions/48562288/frequevty-of-each-unique-combination-in-data-frame

# get proportion of gab for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
pharmayoungcount <- c(sum(vt_2009_young$pharma == 1), sum(vt_2010_young$pharma == 1), sum(vt_2012_young$pharma == 1), sum(vt_2013_young$pharma == 1), sum(vt_2014_young$pharma == 1), sum(vt_2015_young$pharma == 1), sum(vt_2017_young$pharma == 1), sum(vt_2019_young$pharma == 1), sum(vt_2020_young$pharma == 1))

# get proportion for older people
pharmaoldcount <- c(sum(vt_2009_old$pharma == 1), sum(vt_2010_old$pharma == 1), sum(vt_2012_old$pharma == 1), sum(vt_2013_old$pharma == 1), sum(vt_2014_old$pharma == 1), sum(vt_2015_old$pharma == 1), sum(vt_2017_old$pharma == 1), sum(vt_2019_old$pharma == 1), sum(vt_2020_old$pharma == 1))

# combine vectors into df
pharmacount <- data.frame(yearList, pharmayoungcount, pharmaoldcount)

# change the col names to be more informative
names(pharmacount) <- c("Year", "Younger", "Older")

# convert the "race_sex" column into row names so I get a table with only numeric values, from https://stackoverflow.com/questions/5555408/convert-the-values-in-a-column-into-row-names-in-an-existing-data-frame

pharmachi <- pharmacount[,-1]
rownames(pharmachi) <- pharmacount[,1]

as.matrix(pharmachi)

# looks like I can just use any matrix for chisq.test, from https://researchguides.library.vanderbilt.edu/c.php?g=156859&p=3058795
# getting rid of the nonzero entries for testing

  # I'm using Fisher's exact test if >20% of cell counts are <5, if not use chi squared
  # I get >20% <5 from https://www.ccjm.org/content/84/9_suppl_2/e20
  pvalue <- ifelse(sum(pharmachi < 5)/(ncol(pharmachi) * nrow(pharmachi)) > 0.2, fisher.test(as.matrix(pharmachi), workspace=2e6)$p.value, chisq.test(as.matrix(pharmachi))$p.value)
  
  pvalue
```


### Proportion of deaths involving stimulants

```{r propCoke, echo=FALSE, message=FALSE, warning=FALSE}

# unfortunately there are different variable names for each vt file so a for loop won't work :/

# get proportion of cocaine for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
cokeyoung <- c(sum(vt_2009_young$cocaine == 1)/nrow(vt_2009_young), sum(vt_2010_young$cocaine == 1)/nrow(vt_2010_young), sum(vt_2012_young$cocaine == 1)/nrow(vt_2012_young), sum(vt_2013_young$cocaine == 1)/nrow(vt_2013_young), sum(vt_2014_young$cocaine == 1)/nrow(vt_2014_young), sum(vt_2015_young$cocaine == 1)/nrow(vt_2015_young), sum(vt_2017_young$cocaine == 1)/nrow(vt_2017_young), sum(vt_2019_young$cocaine == 1)/nrow(vt_2019_young), sum(vt_2020_young$cocaine == 1)/nrow(vt_2020_young))

# get proportion for older people
cokeold <- c(sum(vt_2009_old$cocaine == 1)/nrow(vt_2009_old), sum(vt_2010_old$cocaine == 1)/nrow(vt_2010_old), sum(vt_2012_old$cocaine == 1)/nrow(vt_2012_old), sum(vt_2013_old$cocaine == 1)/nrow(vt_2013_old), sum(vt_2014_old$cocaine == 1)/nrow(vt_2014_old), sum(vt_2015_old$cocaine == 1)/nrow(vt_2015_old), sum(vt_2017_old$cocaine == 1)/nrow(vt_2017_old), sum(vt_2019_old$cocaine == 1)/nrow(vt_2019_old), sum(vt_2020_old$cocaine == 1)/nrow(vt_2020_old))

# combine vectors into df
propcoke <- data.frame(yearList, cokeyoung, cokeold)

# change the col names to be more informative
names(propcoke) <- c("Year", "Younger", "Older")

# create grouped bar plot, from https://stackoverflow.com/questions/72449952/making-a-grouped-bar-chart-in-r
# getting the labels in the right position from https://stackoverflow.com/questions/37587288/ggplot-label-bars-in-grouped-bar-plot
# info on changing legend title from here https://stackoverflow.com/questions/41649929/how-to-modify-a-legend-title-in-a-barplot-with-ggplot

# ivtreasing line size from https://stackoverflow.com/questions/14794599/how-to-change-line-width-in-ggplot
# line graph info from https://www.sthda.com/english/wiki/ggplot2-line-plot-quick-start-guide-r-software-and-data-visualization

pcokelong  <- propcoke |>
  # Reshape data to long format
  pivot_longer(Younger:Older)

# create labels with the value of each proportion
pcokelong$labels <- round(pcokelong$value, 2)

pcoke <-  ggplot(data = pcokelong, aes(x = Year, y = value, fill = name, group=name, color=factor(name))) +
  # Implement a grouped bar chart
  # Implement a line graph
  geom_line(stat = "identity", size = 1) + geom_point(size = 2) + theme_light() + ggtitle("Proportion of stimulants in opioid-detected overdose deaths \n in VT by age group, 2009-2023") + ylab("Proportion") + theme(plot.title = element_text(hjust = 0.5)) + scale_fill_discrete(guide="none") + scale_x_continuous(breaks=propcoke$Year) + xlab("Year")+ coord_cartesian(ylim = c(0, 1)) + guides(color = guide_legend(title = "Age group"))+ geom_label_repel(aes(label = labels, x = Year, y = value), nudge_x = 0.05, na.rm = FALSE, color = "black", size = 2.5, label.size = NA, fill = NA)

pcoke

ggsave(plot=pcoke, filename = "images/vt/prop_coke_old.png", width = 10, height = 8)
```

### Significance testing for the number of stimulant-detected deaths

```{r chiSquaredCoke, echo=FALSE, message=FALSE, warning=FALSE}
# get frequevty of each race for old and young, from https://stackoverflow.com/questions/48562288/frequevty-of-each-unique-combination-in-data-frame

# get proportion of coke for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
cokeyoungcount <- c(sum(vt_2009_young$cocaine == 1), sum(vt_2010_young$cocaine == 1), sum(vt_2012_young$cocaine == 1), sum(vt_2013_young$cocaine == 1), sum(vt_2014_young$cocaine == 1), sum(vt_2015_young$cocaine == 1), sum(vt_2017_young$cocaine == 1), sum(vt_2019_young$cocaine == 1), sum(vt_2020_young$cocaine == 1))

# get proportion for older people
cokeoldcount <- c(sum(vt_2009_old$cocaine == 1), sum(vt_2010_old$cocaine == 1), sum(vt_2012_old$cocaine == 1), sum(vt_2013_old$cocaine == 1), sum(vt_2014_old$cocaine == 1), sum(vt_2015_old$cocaine == 1), sum(vt_2017_old$cocaine == 1), sum(vt_2019_old$cocaine == 1), sum(vt_2020_old$cocaine == 1))

# combine vectors into df
cokecount <- data.frame(yearList, cokeyoungcount, cokeoldcount)

# change the col names to be more informative
names(cokecount) <- c("Year", "Younger", "Older")

# convert the "race_sex" column into row names so I get a table with only numeric values, from https://stackoverflow.com/questions/5555408/convert-the-values-in-a-column-into-row-names-in-an-existing-data-frame

cokechi <- cokecount[,-1]
rownames(cokechi) <- cokecount[,1]

as.matrix(cokechi)

# looks like I can just use any matrix for chisq.test, from https://researchguides.library.vanderbilt.edu/c.php?g=156859&p=3058795
  # I'm using Fisher's exact test if >20% of cell counts are <5, if not use chi squared
  # I get >20% <5 from https://www.ccjm.org/content/84/9_suppl_2/e20
  pvalue <- ifelse(sum(cokechi < 5)/(ncol(cokechi) * nrow(cokechi)) > 0.2, fisher.test(as.matrix(cokechi), workspace=2e6)$p.value, chisq.test(as.matrix(cokechi))$p.value)
  
  pvalue
```

### Proportion of deaths involving alcohol

```{r propEtOH, echo=FALSE, message=FALSE, warning=FALSE}

# unfortunately there are different variable names for each vt file so a for loop won't work :/

# get proportion of alcohol for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
etohyoung <- c(sum(vt_2009_young$etoh == 1)/nrow(vt_2009_young), sum(vt_2010_young$etoh == 1)/nrow(vt_2010_young), sum(vt_2012_young$etoh == 1)/nrow(vt_2012_young), sum(vt_2013_young$etoh == 1)/nrow(vt_2013_young), sum(vt_2014_young$etoh == 1)/nrow(vt_2014_young), sum(vt_2015_young$etoh == 1)/nrow(vt_2015_young), sum(vt_2017_young$etoh == 1)/nrow(vt_2017_young), sum(vt_2019_young$etoh == 1)/nrow(vt_2019_young), sum(vt_2020_young$etoh == 1)/nrow(vt_2020_young))

# get proportion for older people
etohold <- c(sum(vt_2009_old$etoh == 1)/nrow(vt_2009_old), sum(vt_2010_old$etoh == 1)/nrow(vt_2010_old), sum(vt_2012_old$etoh == 1)/nrow(vt_2012_old), sum(vt_2013_old$etoh == 1)/nrow(vt_2013_old), sum(vt_2014_old$etoh == 1)/nrow(vt_2014_old), sum(vt_2015_old$etoh == 1)/nrow(vt_2015_old), sum(vt_2017_old$etoh == 1)/nrow(vt_2017_old), sum(vt_2019_old$etoh == 1)/nrow(vt_2019_old), sum(vt_2020_old$etoh == 1)/nrow(vt_2020_old))

# combine vectors into df
propetoh <- data.frame(yearList, etohyoung, etohold)

# change the col names to be more informative
names(propetoh) <- c("Year", "Younger", "Older")

# create grouped bar plot, from https://stackoverflow.com/questions/72449952/making-a-grouped-bar-chart-in-r
# getting the labels in the right position from https://stackoverflow.com/questions/37587288/ggplot-label-bars-in-grouped-bar-plot
# info on changing legend title from here https://stackoverflow.com/questions/41649929/how-to-modify-a-legend-title-in-a-barplot-with-ggplot

# ivtreasing line size from https://stackoverflow.com/questions/14794599/how-to-change-line-width-in-ggplot
# line graph info from https://www.sthda.com/english/wiki/ggplot2-line-plot-quick-start-guide-r-software-and-data-visualization

petohlong  <- propetoh |>
  # Reshape data to long format
  pivot_longer(Younger:Older)

# create labels with the value of each proportion
petohlong$labels <- round(petohlong$value, 2)

petoh <-  ggplot(data = petohlong, aes(x = Year, y = value, fill = name, group=name, color=factor(name))) +
  # Implement a grouped bar chart
  # Implement a line graph
  geom_line(stat = "identity", size = 1) + geom_point(size = 2) + theme_light() + ggtitle("Proportion of alcohol in opioid-detected overdose deaths \n in VT by age group, 2009-2023") + ylab("Proportion") + theme(plot.title = element_text(hjust = 0.5)) + scale_fill_discrete(guide="none") + scale_x_continuous(breaks=propetoh$Year) + xlab("Year")+ coord_cartesian(ylim = c(0, 1)) + guides(color = guide_legend(title = "Age group"))+ geom_label_repel(aes(label = labels, x = Year, y = value), nudge_x = 0.05, na.rm = FALSE, color = "black", size = 2.5, label.size = NA, fill = NA)

petoh


ggsave(plot=petoh, filename = "images/vt/prop_etoh.png", width = 10, height = 8)
```

### Significance testing for the number of alcohol-detected deaths

```{r chiSquaredEtoh, echo=FALSE, message=FALSE, warning=FALSE}
# get frequevty of each race for old and young, from https://stackoverflow.com/questions/48562288/frequevty-of-each-unique-combination-in-data-frame

# get proportion of etoh for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
etohyoungcount <- c(sum(vt_2009_young$etoh == 1), sum(vt_2010_young$etoh == 1), sum(vt_2012_young$etoh == 1), sum(vt_2013_young$etoh == 1), sum(vt_2014_young$etoh == 1), sum(vt_2015_young$etoh == 1), sum(vt_2017_young$etoh == 1), sum(vt_2019_young$etoh == 1), sum(vt_2020_young$etoh == 1))

# get proportion for older people
etoholdcount <- c(sum(vt_2009_old$etoh == 1), sum(vt_2010_old$etoh == 1), sum(vt_2012_old$etoh == 1), sum(vt_2013_old$etoh == 1), sum(vt_2014_old$etoh == 1), sum(vt_2015_old$etoh == 1), sum(vt_2017_old$etoh == 1), sum(vt_2019_old$etoh == 1), sum(vt_2020_old$etoh == 1))

# combine vectors into df
etohcount <- data.frame(yearList, etohyoungcount, etoholdcount)

# change the col names to be more informative
names(etohcount) <- c("Year", "Younger", "Older")

# convert the "race_sex" column into row names so I get a table with only numeric values, from https://stackoverflow.com/questions/5555408/convert-the-values-in-a-column-into-row-names-in-an-existing-data-frame

etohchi <- etohcount[,-1]
rownames(etohchi) <- etohcount[,1]

as.matrix(etohchi)

# looks like I can just use any matrix for chisq.test, from https://researchguides.library.vanderbilt.edu/c.php?g=156859&p=3058795

# looks like I can just use any matrix for chisq.test, from https://researchguides.library.vanderbilt.edu/c.php?g=156859&p=3058795
  # I'm using Fisher's exact test if >20% of cell counts are <5, if not use chi squared
  # I get >20% <5 from https://www.ccjm.org/content/84/9_suppl_2/e20
  pvalue <- ifelse(sum(etohchi < 5)/(ncol(etohchi) * nrow(etohchi)) > 0.2, fisher.test(as.matrix(etohchi), workspace=2e6)$p.value, chisq.test(as.matrix(etohchi))$p.value)
  
  pvalue
```

### Proportion of deaths involving buprenorphine/methadone


```{r propBupeMethadone, echo=FALSE, message=FALSE, warning=FALSE}

# unfortunately there are different variable names for each vt file so a for loop won't work :/

# get proportion of cocaine for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
methbupeyoung <- c(sum(sum(vt_2009_young$bupe == 1), sum(vt_2009_young$methadone == 1))/nrow(vt_2009_young), sum(sum(vt_2010_young$bupe == 1), sum(vt_2010_young$methadone == 1))/nrow(vt_2010_young), sum(sum(vt_2012_young$bupe == 1), sum(vt_2012_young$methadone == 1))/nrow(vt_2012_young), sum(sum(vt_2013_young$bupe == 1), sum(vt_2013_young$methadone == 1))/nrow(vt_2013_young), sum(sum(vt_2014_young$bupe == 1), sum(vt_2014_young$methadone == 1))/nrow(vt_2014_young), sum(sum(vt_2015_young$bupe == 1), sum(vt_2015_young$methadone == 1))/nrow(vt_2015_young), sum(sum(vt_2017_young$bupe == 1), sum(vt_2017_young$methadone == 1))/nrow(vt_2017_young), sum(sum(vt_2019_young$bupe == 1), sum(vt_2019_young$methadone == 1))/nrow(vt_2019_young), sum(sum(vt_2020_young$bupe == 1), sum(vt_2020_young$methadone == 1))/nrow(vt_2020_young))

# get proportion for older people
methbupeold <- c(sum(sum(vt_2009_old$bupe == 1), sum(vt_2009_old$methadone == 1))/nrow(vt_2009_old), sum(sum(vt_2010_old$bupe == 1), sum(vt_2010_old$methadone == 1))/nrow(vt_2010_old), sum(sum(vt_2012_old$bupe == 1), sum(vt_2012_old$methadone == 1))/nrow(vt_2012_old), sum(sum(vt_2013_old$bupe == 1), sum(vt_2013_old$methadone == 1))/nrow(vt_2013_old), sum(sum(vt_2014_old$bupe == 1), sum(vt_2014_old$methadone == 1))/nrow(vt_2014_old), sum(sum(vt_2015_old$bupe == 1), sum(vt_2015_old$methadone == 1))/nrow(vt_2015_old), sum(sum(vt_2017_old$bupe == 1), sum(vt_2017_old$methadone == 1))/nrow(vt_2017_old), sum(sum(vt_2019_old$bupe == 1), sum(vt_2019_old$methadone == 1))/nrow(vt_2019_old), sum(sum(vt_2020_old$bupe == 1), sum(vt_2020_old$methadone == 1))/nrow(vt_2020_old))

# combine vectors into df
propmethbupe <- data.frame(yearList, methbupeyoung, methbupeold)

# change the col names to be more informative
names(propmethbupe) <- c("Year", "Younger", "Older")

# create grouped bar plot, from https://stackoverflow.com/questions/72449952/making-a-grouped-bar-chart-in-r
# getting the labels in the right position from https://stackoverflow.com/questions/37587288/ggplot-label-bars-in-grouped-bar-plot
# info on changing legend title from here https://stackoverflow.com/questions/41649929/how-to-modify-a-legend-title-in-a-barplot-with-ggplot

# ivtreasing line size from https://stackoverflow.com/questions/14794599/how-to-change-line-width-in-ggplot
# line graph info from https://www.sthda.com/english/wiki/ggplot2-line-plot-quick-start-guide-r-software-and-data-visualization


pmethbupelong  <- propmethbupe |>
  # Reshape data to long format
  pivot_longer(Younger:Older)

# create labels with the value of each proportion
pmethbupelong$labels <- round(pmethbupelong$value, 2)

pmethbupe <-  ggplot(data = pmethbupelong, aes(x = Year, y = value, fill = name, group=name, color=factor(name))) +
  # Implement a grouped bar chart
  # Implement a line graph
  geom_line(stat = "identity", size = 1) + geom_point(size = 2) + theme_light() + ggtitle("Proportion of methadone/buprenorphine-detected \n deaths in VT by age group, 2009-2023") + ylab("Proportion") + theme(plot.title = element_text(hjust = 0.5)) + scale_fill_discrete(guide="none") + scale_x_continuous(breaks=propmethbupe$Year) + xlab("Year")+ coord_cartesian(ylim = c(0, 1)) + guides(color = guide_legend(title = "Age group"))+ geom_label_repel(aes(label = labels, x = Year, y = value), nudge_x = 0.1, na.rm = FALSE, color = "black", size = 2.5, label.size = NA, fill = NA)

pmethbupe

ggsave(plot=pmethbupe, filename = "images/vt/prop_meth_bupe.png", width = 10, height = 8)

```

### Significance testing for the number of buprenorphine/methadone-detected deaths


```{r chiSquaredMethBupe, echo=FALSE, message=FALSE, warning=FALSE}
# get frequevty of each race for old and young, from https://stackoverflow.com/questions/48562288/frequevty-of-each-unique-combination-in-data-frame

# get proportion of cocaine for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
methbupeyoungcount <- c(sum(sum(vt_2009_young$buprenorphine == 1), sum(vt_2009_young$methadone == 1)), sum(sum(vt_2010_young$buprenorphine == 1), sum(vt_2010_young$methadone == 1)), sum(sum(vt_2012_young$buprenorphine == 1), sum(vt_2012_young$methadone == 1)), sum(sum(vt_2013_young$buprenorphine == 1), sum(vt_2013_young$methadone == 1)), sum(sum(vt_2014_young$buprenorphine == 1), sum(vt_2014_young$methadone == 1)), sum(sum(vt_2015_young$buprenorphine == 1), sum(vt_2015_young$methadone == 1)), sum(sum(vt_2017_young$buprenorphine == 1), sum(vt_2017_young$methadone == 1)), sum(sum(vt_2019_young$buprenorphine == 1), sum(vt_2019_young$methadone == 1)), sum(sum(vt_2020_young$buprenorphine == 1), sum(vt_2020_young$methadone == 1)))

# get proportion for older people
methbupeoldcount <-  c(sum(sum(vt_2009_old$buprenorphine == 1), sum(vt_2009_old$methadone == 1)), sum(sum(vt_2010_old$buprenorphine == 1), sum(vt_2010_old$methadone == 1)), sum(sum(vt_2012_old$buprenorphine == 1), sum(vt_2012_old$methadone == 1)), sum(sum(vt_2013_old$buprenorphine == 1), sum(vt_2013_old$methadone == 1)), sum(sum(vt_2014_old$buprenorphine == 1), sum(vt_2014_old$methadone == 1)), sum(sum(vt_2015_old$buprenorphine == 1), sum(vt_2015_old$methadone == 1)), sum(sum(vt_2017_old$buprenorphine == 1), sum(vt_2017_old$methadone == 1)), sum(sum(vt_2019_old$buprenorphine == 1), sum(vt_2019_old$methadone == 1)), sum(sum(vt_2020_old$buprenorphine == 1), sum(vt_2020_old$methadone == 1)))


# combine vectors into df
methbupecount <- data.frame(yearList, methbupeyoungcount, methbupeoldcount)

# change the col names to be more informative
names(methbupecount) <- c("Year", "Younger", "Older")

# convert the "race_sex" column into row names so I get a table with only numeric values, from https://stackoverflow.com/questions/5555408/convert-the-values-in-a-column-into-row-names-in-an-existing-data-frame

methbupechi <- methbupecount[,-1]
rownames(methbupechi) <- methbupecount[,1]

as.matrix(methbupechi)

# looks like I can just use any matrix for chisq.test, from https://researchguides.library.vanderbilt.edu/c.php?g=156859&p=3058795

# looks like I can just use any matrix for chisq.test, from https://researchguides.library.vanderbilt.edu/c.php?g=156859&p=3058795
  # I'm using Fisher's exact test if >20% of cell counts are <5, if not use chi squared
  # I get >20% <5 from https://www.ccjm.org/content/84/9_suppl_2/e20

# I KEPT RUNNING INTO AN ISSUE WITH WORKSPACE SIZE [NEVER BIG ENOUGH] SO I SET SIMULATE P VALUE TO BE TRUE
  pvalue <- ifelse(sum(methbupechi < 5)/(ncol(methbupechi) * nrow(methbupechi)) > 0.2, fisher.test(as.matrix(methbupechi), workspace=2e6)$p.value, chisq.test(as.matrix(methbupechi))$p.value)
  
  pvalue
```
