---
title: "Fatal opioid-detected overdoses in Connecticut, 2018-2022"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 5
    theme: lumen
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<a href="https://ctoverdosedeath.github.io">
  <button>Home</button>
</a>
<a href="https://ctoverdosedeath.github.io/allstate.html">
  <button>Statewide</button>
</a>
<a href="https://ctoverdosedeath.github.io/region1.html">
  <button>Region 1</button>
</a>
<a href="https://ctoverdosedeath.github.io/region2.html">
  <button>Region 2</button>
</a>
<a href="https://ctoverdosedeath.github.io/region3.html">
  <button>Region 3</button>
</a>
<a href="https://ctoverdosedeath.github.io/region4.html">
  <button>Region 4</button>
</a>
<a href="https://ctoverdosedeath.github.io/region5.html">
  <button>Region 5</button>
</a>
<a href="https://ctoverdosedeath.github.io/region6.html">
  <button>Region 6</button>
</a>

  
```{r importData, echo=FALSE, message=FALSE, warning=FALSE}
library(tigris)
library(ggplot2)
library(stringr)
library(dplyr)
library(lubridate)
library(patchwork)
library(tidyverse) #to replace NA with something else
library(sf)
library(multimode) #to calculate modes for the age plot
library(findSVI) #for SVI calculation

# import OCME data and convert the date into a "date" data type so we can work with it

ocme_2018 <- read.csv("ocme_2018.csv")
ocme_2018$DOD = as.Date(ocme_2018$DOD, "%m/%d/%Y")
#add in day of week -- from https://stackoverflow.com/questions/9216138/find-the-day-of-a-week
ocme_2018$Weekday <- weekdays(ocme_2018$DOD)

ocme_2019 <- read.csv("ocme_2019.csv")
ocme_2019$DOD = as.Date(ocme_2019$DOD, "%m/%d/%Y")
#add in day of week
ocme_2019$Weekday <- weekdays(ocme_2019$DOD)

ocme_2020 <- read.csv("OCME_2020.csv")
ocme_2020$DOD = as.Date(ocme_2020$DOD, "%m/%d/%Y")
#add in day of week
ocme_2020$Weekday <- weekdays(ocme_2020$DOD)

ocme_2021 <- read.csv("OCME_2021.csv")
ocme_2021$DateofDeath = as.Date(ocme_2021$DateofDeath, "%m/%d/%Y")
#add in day of week
ocme_2021$Weekday <- weekdays(ocme_2021$DateofDeath)

ocme_2022 <- read.csv("OCME_2022.csv")
ocme_2022$Date.of.Death = as.Date(ocme_2022$Date.of.Death, "%m/%d/%Y")
#add in day of week
ocme_2022$Weekday <- weekdays(ocme_2022$Date.of.Death)

# adding in county/city population data from https://portal.ct.gov/DPH/Health-Information-Systems--Reporting/Population/Annual-Town-and-County-Population-for-Connecticut -- I converted to csv and isolated county/city in separate files

county_pop_2018 <- read.csv("county_pop_2018.csv")
# I think the population numbers will be read as strings, so convert them to numbers so we can manipulate
# the commas were causing NAs when converting to numeric, solution to remove from https://stackoverflow.com/questions/57656023/as-numeric-returns-na-for-no-apparent-reason-for-some-of-the-values-in-a-column
county_pop_2018$Pop <- as.numeric(gsub(",","",county_pop_2018$Pop))

county_pop_2019 <- read.csv("county_pop_2019.csv")
# I think the population numbers will be read as strings, so convert them to numbers so we can manipulate
county_pop_2019$Pop <- as.numeric(gsub(",","",county_pop_2019$Pop)) 

county_pop_2020 <- read.csv("county_pop_2020.csv")
# I think the population numbers will be read as strings, so convert them to numbers so we can manipulate
county_pop_2020$Pop <- as.numeric(gsub(",","",county_pop_2020$Pop))

county_pop_2021 <- read.csv("county_pop_2021.csv")
# I think the population numbers will be read as strings, so convert them to numbers so we can manipulate
county_pop_2021$Pop <- as.numeric(gsub(",","",county_pop_2021$Pop))

# for 2022 I had to manually calculate the county-level population estimates using [this](https://www1.ctdol.state.ct.us/lmi/misc/counties.asp) since CT had switched to reporting "planning region" estimates instead of county. I will continue using county for the sake of consistency
county_pop_2022 <- read.csv("county_pop_2022.csv")
# I think the population numbers will be read as strings, so convert them to numbers so we can manipulate
county_pop_2022$Pop <- as.numeric(gsub(",","",county_pop_2022$Pop))


city_pop_2018 <- read.csv("city_pop_2018.csv")
# I think the population numbers will be read as strings, so convert them to numbers so we can manipulate
city_pop_2018$Pop <- as.numeric(gsub(",","",city_pop_2018$Pop))

city_pop_2019 <- read.csv("city_pop_2019.csv")
# I think the population numbers will be read as strings, so convert them to numbers so we can manipulate
city_pop_2019$Pop <- as.numeric(gsub(",","",city_pop_2019$Pop))

city_pop_2020 <- read.csv("city_pop_2020.csv")
# I think the population numbers will be read as strings, so convert them to numbers so we can manipulate
city_pop_2020$Pop <- as.numeric(gsub(",","",city_pop_2020$Pop))

city_pop_2021 <- read.csv("city_pop_2021.csv")
# I think the population numbers will be read as strings, so convert them to numbers so we can manipulate
city_pop_2021$Pop <- as.numeric(gsub(",","",city_pop_2021$Pop))

city_pop_2022 <- read.csv("city_pop_2022.csv")
# I think the population numbers will be read as strings, so convert them to numbers so we can manipulate
city_pop_2022$Pop <- as.numeric(gsub(",","",city_pop_2022$Pop))

# get town data from tigris so we can map deaths by town later

options(tigris_use_cache=TRUE)
ct_towns = county_subdivisions(state="Connecticut")
ct_towns$NAME = str_to_title(ct_towns$NAME)
ct_towns = filter(ct_towns, NAME != "County Subdivisions Not Defined")
# tigris also has shapefiles for the long island sound so let's get rid of that
ct_towns <- subset(ct_towns, ALAND!=0)
```

  
### Race {.tabset}

#### 2018

```{r raceNumeric2018, echo=FALSE, message=FALSE, warning=FALSE}
# we don't want to see the numerical labels but rather the actual race values, so let's create another variable that turns race_numeric into words

ocme_2018$race_numeric_words <- ifelse(ocme_2018$race_.numeric == "1", "White", ifelse(ocme_2018$race_.numeric == "2", "Black", ifelse(ocme_2018$race_.numeric == "3", "Hispanic", ifelse(ocme_2018$race_.numeric == "4", "Other", ifelse(ocme_2018$race_.numeric == "7", "Unknown", ocme_2018$race_.numeric)))))

ggplot(data=ocme_2018, aes(x=race_numeric_words, ..count..)) + 
      geom_bar(aes(fill = race_numeric_words), position = "dodge") + scale_fill_brewer(palette="Spectral") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Racial distribution for 2018 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

raceProp2018 <- ggplot(data=ocme_2018, aes(x=race_numeric_words, ..count..)) + 
      geom_bar(aes(fill = race_numeric_words), position = "dodge") + scale_fill_brewer(palette="Spectral") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme_2018), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Racial distribution for 2018 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

raceProp2018

#save this one for the presentation
ggsave(plot=raceProp2018, filename = "images/statewide/race_proportions_2018.png", width = 10, height = 8)
```

#### 2019

```{r raceNumeric2019, echo=FALSE, message=FALSE, warning=FALSE}
# we don't want to see the numerical labels but rather the actual race values, so let's create another variable that turns race_numeric into words

ocme_2019$race_numeric_words <- ifelse(ocme_2019$race_numeric...1.White..........2.Black....3.Hispanic...4.Other....7.UNK == "1", "White", ifelse(ocme_2019$race_numeric...1.White..........2.Black....3.Hispanic...4.Other....7.UNK == "2", "Black", ifelse(ocme_2019$race_numeric...1.White..........2.Black....3.Hispanic...4.Other....7.UNK == "3", "Hispanic", ifelse(ocme_2019$race_numeric...1.White..........2.Black....3.Hispanic...4.Other....7.UNK == "4", "Other", ifelse(ocme_2019$race_numeric...1.White..........2.Black....3.Hispanic...4.Other....7.UNK == "7", "Unknown", ocme_2019$race_numeric...1.White..........2.Black....3.Hispanic...4.Other....7.UNK)))))

ggplot(data=ocme_2019, aes(x=race_numeric_words, ..count..)) + 
      geom_bar(aes(fill = race_numeric_words), position = "dodge") + scale_fill_brewer(palette="Spectral") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Racial distribution for 2019 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

raceProp2019 <- ggplot(data=ocme_2019, aes(x=race_numeric_words, ..count..)) + 
      geom_bar(aes(fill = race_numeric_words), position = "dodge") + scale_fill_brewer(palette="Spectral") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme_2019), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Racial distribution for 2019 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

raceProp2019

#save this one for the presentation
ggsave(plot=raceProp2019, filename = "images/statewide/race_proportions_2019.png", width = 10, height = 8)

```

#### 2020


```{r raceNumeric2020, echo=FALSE, message=FALSE, warning=FALSE}
# we don't want to see the numerical labels but rather the actual race values, so let's create another variable that turns race_numeric into words

ocme_2020$race_numeric_words <- ifelse(ocme_2020$race_numeric...1.White..........2.Black....3.Hispanic...4.Other....7.UNK == "1", "White", ifelse(ocme_2020$race_numeric...1.White..........2.Black....3.Hispanic...4.Other....7.UNK == "2", "Black", ifelse(ocme_2020$race_numeric...1.White..........2.Black....3.Hispanic...4.Other....7.UNK == "3", "Hispanic", ifelse(ocme_2020$race_numeric...1.White..........2.Black....3.Hispanic...4.Other....7.UNK == "4", "Other", ifelse(ocme_2020$race_numeric...1.White..........2.Black....3.Hispanic...4.Other....7.UNK == "7", "Unknown", ocme_2020$race_numeric...1.White..........2.Black....3.Hispanic...4.Other....7.UNK)))))

ggplot(data=ocme_2020, aes(x=race_numeric_words, ..count..)) + 
      geom_bar(aes(fill = race_numeric_words), position = "dodge") + scale_fill_brewer(palette="Spectral") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Racial distribution for 2020 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

raceProp2020 <- ggplot(data=ocme_2020, aes(x=race_numeric_words, ..count..)) + 
      geom_bar(aes(fill = race_numeric_words), position = "dodge") + scale_fill_brewer(palette="Spectral") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme_2020), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Racial distribution for 2020 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

raceProp2020

#save this one for the presentation
ggsave(plot=raceProp2020, filename = "images/statewide/race_proportions_2020.png", width = 10, height = 8)

```


#### 2021

```{r raceNumeric2021, echo=FALSE, message=FALSE, warning=FALSE}
# we don't want to see the numerical labels but rather the actual race values, so let's create another variable that turns race_numeric into words

ocme_2021$race_numeric_words <- ifelse(ocme_2021$race_numeric...1.White..........2.Black....3.Hispanic...4.Other....7.UNK == "1", "White", ifelse(ocme_2021$race_numeric...1.White..........2.Black....3.Hispanic...4.Other....7.UNK == "2", "Black", ifelse(ocme_2021$race_numeric...1.White..........2.Black....3.Hispanic...4.Other....7.UNK == "3", "Hispanic", ifelse(ocme_2021$race_numeric...1.White..........2.Black....3.Hispanic...4.Other....7.UNK == "4", "Other", ifelse(ocme_2021$race_numeric...1.White..........2.Black....3.Hispanic...4.Other....7.UNK == "7", "Unknown", ocme_2021$race_numeric...1.White..........2.Black....3.Hispanic...4.Other....7.UNK)))))

ggplot(data=ocme_2021, aes(x=race_numeric_words, ..count..)) + 
      geom_bar(aes(fill = race_numeric_words), position = "dodge") + scale_fill_brewer(palette="Spectral") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Racial distribution for 2021 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

raceProp2021 <- ggplot(data=ocme_2021, aes(x=race_numeric_words, ..count..)) + 
      geom_bar(aes(fill = race_numeric_words), position = "dodge") + scale_fill_brewer(palette="Spectral") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme_2021), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Racial distribution for 2021 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

raceProp2021

#save this one for the presentation
ggsave(plot=raceProp2021, filename = "images/statewide/race_proportions_2021.png", width = 10, height = 8)

```


#### 2022

```{r raceNumeric2022, echo=FALSE, message=FALSE, warning=FALSE}
# we don't want to see the numerical labels but rather the actual race values, so let's create another variable that turns race_numeric into words
ocme_2022$race_numeric_words <- ifelse(ocme_2022$race_.numeric...1.White..........2.Black....3.Hispanic...4.Other....7.UNK == "1", "White", ifelse(ocme_2022$race_.numeric...1.White..........2.Black....3.Hispanic...4.Other....7.UNK == "2", "Black", ifelse(ocme_2022$race_.numeric...1.White..........2.Black....3.Hispanic...4.Other....7.UNK == "3", "Hispanic", ifelse(ocme_2022$race_.numeric...1.White..........2.Black....3.Hispanic...4.Other....7.UNK == "4", "Other", ifelse(ocme_2022$race_.numeric...1.White..........2.Black....3.Hispanic...4.Other....7.UNK == "7", "Unknown", ocme_2022$race_.numeric...1.White..........2.Black....3.Hispanic...4.Other....7.UNK)))))

# there's a NA in here and for some reason replacing it using the line above didn't work, so we'll try this
# from https://stackoverflow.com/questions/55910824/replace-na-with-unknown
ocme_2022 <- ocme_2022 %>% replace_na(list(race_numeric_words = "Unknown"))

ggplot(data=ocme_2022, aes(x=race_numeric_words, ..count..)) + 
      geom_bar(aes(fill = race_numeric_words), position = "dodge") + scale_fill_brewer(palette="Spectral") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Racial distribution for 2022 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

raceProp2022 <- ggplot(data=ocme_2022, aes(x=race_numeric_words, ..count..)) + 
      geom_bar(aes(fill = race_numeric_words), position = "dodge") + scale_fill_brewer(palette="Spectral") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme_2022), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Racial distribution for 2022 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

raceProp2022

#save this one for the presentation
ggsave(plot=raceProp2022, filename = "images/statewide/race_proportions_2022.png", width = 10, height = 8)

```

### Sex {.tabset}

#### 2018

```{r sexTable2018, echo=FALSE, message=FALSE, warning=FALSE}
ocme_2018$Sex_revised <- ifelse(ocme_2018$sex == "", "Unknown", ocme_2018$sex)

sex2018 <- ggplot(data=ocme_2018, aes(x=Sex_revised, ..count..)) + 
      geom_bar(aes(fill = Sex_revised), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Sex distribution for 2018 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

sex2018


sexProp2018 <- ggplot(data=ocme_2018, aes(x=Sex_revised, ..count..)) + 
      geom_bar(aes(fill = Sex_revised), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme_2018), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Sex distribution for 2018 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

sexProp2018

ggsave(plot=sexProp2018, filename = "images/statewide/sexProp2018.png", width = 10, height = 8)
```

#### 2019


```{r sexTable2019, echo=FALSE, message=FALSE, warning=FALSE}

sex2019 <- ggplot(data=ocme_2019, aes(x=sex, ..count..)) + 
      geom_bar(aes(fill = sex), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Sex distribution for 2019 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

sex2019

sexProp2019 <- ggplot(data=ocme_2019, aes(x=sex, ..count..)) + 
      geom_bar(aes(fill = sex), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme_2019), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Sex distribution for 2019 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

sexProp2019

ggsave(plot=sexProp2019, filename = "images/statewide/sexProp2019.png", width = 10, height = 8)

```

#### 2020

```{r sexTable2020, echo=FALSE, message=FALSE, warning=FALSE}
# There was one person whose race was just listed as blank, so I making it "Unknown"
ocme_2020$Sex_revised <- ifelse(ocme_2020$Sex == "", "Unknown", ifelse(ocme_2020$Sex == " Male ", "Male", ocme_2020$Sex))

sex2020 <- ggplot(data=ocme_2020, aes(x=Sex_revised, ..count..)) + 
      geom_bar(aes(fill = Sex_revised), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Sex distribution for 2020 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

sex2020

#show the proportion of each sex, for easy comparison
sexProp2020 <- ggplot(data=ocme_2020, aes(x=Sex_revised, ..count..)) + 
      geom_bar(aes(fill = Sex_revised), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme_2020), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Sex distribution for 2020 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

sexProp2020

ggsave(plot=sexProp2020, filename = "images/statewide/sexProp2020.png", width = 10, height = 8)

```


#### 2021

```{r sexTable2021, echo=FALSE, message=FALSE, warning=FALSE}
# There was one person whose race was just listed as blank, so I making it "Unknown"
ocme_2021$Sex_revised <- ifelse(ocme_2021$Sex == "", "Unknown", ocme_2021$Sex)

sex2021 <- ggplot(data=ocme_2021, aes(x=Sex_revised, ..count..)) + 
      geom_bar(aes(fill = Sex_revised), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Sex distribution for 2021 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

sex2021

sexProp2021 <- ggplot(data=ocme_2021, aes(x=Sex_revised, ..count..)) + 
      geom_bar(aes(fill = Sex_revised), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme_2021), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Sex distribution for 2021 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

sexProp2021

ggsave(plot=sexProp2021, filename = "images/statewide/sexProp2021.png", width = 10, height = 8)
```

#### 2022

```{r sexTable2022, echo=FALSE, message=FALSE, warning=FALSE}
sex2022 <- ggplot(data=ocme_2022, aes(x=Sex, ..count..)) + 
      geom_bar(aes(fill = Sex), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Sex distribution for 2022 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

sex2022

sexProp2022 <- ggplot(data=ocme_2022, aes(x=Sex, ..count..)) + 
      geom_bar(aes(fill = Sex), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme_2022), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Sex distribution for 2022 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

sexProp2022

ggsave(plot=sexProp2022, filename = "images/statewide/sexProp2022.png", width = 10, height = 8)
```

### Race and sex {.tabset}

#### 2018

```{r sexRaceTable2018, echo=FALSE, message=FALSE, warning=FALSE}
#concatenate the race and sex variables, so "Black" and "female" becomes "Black female"
ocme_2018$race_sex <- paste(ocme_2018$race_numeric_words, ocme_2018$sex, sep = " ")

raceSex2018 <- ggplot(data=ocme_2018, aes(x=race_sex, ..count..)) + 
      geom_bar(aes(fill = race_sex), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Race/sex distribution for 2018 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

raceSex2018

raceSexProp2018 <- ggplot(data=ocme_2018, aes(x=race_sex, ..count..)) + 
      geom_bar(aes(fill = race_sex), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme_2018), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Race/sex distribution for 2018 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

raceSexProp2018

ggsave(plot=raceSexProp2018, filename = "images/statewide/raceSexProp2018.png", width = 10, height = 8)

```

#### 2019

```{r sexRaceTable2019, echo=FALSE, message=FALSE, warning=FALSE}
#concatenate the race and sex variables, so "Black" and "female" becomes "Black female"
ocme_2019$race_sex <- paste(ocme_2019$race_numeric_words, ocme_2019$sex, sep = " ")

raceSex2019 <- ggplot(data=ocme_2019, aes(x=race_sex, ..count..)) + 
      geom_bar(aes(fill = race_sex), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Race/sex distribution for 2019 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

raceSex2019

raceSexProp2019 <- ggplot(data=ocme_2019, aes(x=race_sex, ..count..)) + 
      geom_bar(aes(fill = race_sex), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme_2019), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Race/sex distribution for 2019 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

raceSexProp2019

ggsave(plot=raceSexProp2019, filename = "images/statewide/raceSexProp2019.png", width = 10, height = 8)

```


#### 2020

```{r sexRaceTable2020, echo=FALSE, message=FALSE, warning=FALSE}
# There was one person whose race was just listed as blank, so I making it "Unknown"
ocme_2020$race_sex <- paste(ocme_2020$race_numeric_words, ocme_2020$Sex_revised, sep = " ")

raceSex2020 <- ggplot(data=ocme_2020, aes(x=race_sex, ..count..)) + 
      geom_bar(aes(fill = race_sex), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Race/sex distribution for 2020 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

raceSex2020

raceSexProp2020 <- ggplot(data=ocme_2020, aes(x=race_sex, ..count..)) + 
      geom_bar(aes(fill = race_sex), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme_2020), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Race/sex distribution for 2020 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

raceSexProp2020

ggsave(plot=raceSexProp2020, filename = "images/statewide/raceSexProp2020.png", width = 10, height = 8)

```


#### 2021

```{r sexRaceTable2021, echo=FALSE, message=FALSE, warning=FALSE}
ocme_2021$race_sex <- paste(ocme_2021$race_numeric_words, ocme_2021$Sex_revised, sep = " ")

raceSex2021 <- ggplot(data=ocme_2021, aes(x=race_sex, ..count..)) + 
      geom_bar(aes(fill = race_sex), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Race/sex distribution for 2021 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

raceSex2021

raceSexProp2021 <- ggplot(data=ocme_2021, aes(x=race_sex, ..count..)) + 
      geom_bar(aes(fill = race_sex), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme_2021), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Race/sex distribution for 2021 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

raceSexProp2021

ggsave(plot=raceSexProp2021, filename = "images/statewide/raceSexProp2021.png", width = 10, height = 8)

```


#### 2022


```{r sexRaceTable2022, echo=FALSE, message=FALSE, warning=FALSE}
ocme_2022$race_sex <- paste(na.omit(ocme_2022$race_numeric_words), ocme_2022$Sex, sep = " ")

raceSex2022 <- ggplot(data=ocme_2022, aes(x=race_sex, ..count..)) + 
      geom_bar(aes(fill = race_sex), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Race/sex distribution for 2022 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

raceSex2022

raceSexProp2022 <- ggplot(data=ocme_2022, aes(x=race_sex, ..count..)) + 
      geom_bar(aes(fill = race_sex), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme_2022), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Race/sex distribution for 2022 opioid-detected overdose deaths")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

raceSexProp2022

ggsave(plot=raceSexProp2022, filename = "images/statewide/raceSexProp2022.png", width = 10, height = 8)

```

### Age {.tabset}

The black lines on the plots represent modes. The approach I used to calculate modes creates a continuous approximation of the age distribution [which admittedly is not the best approach here since age is a discrete variable]. I'll be looking for a better approach. 

#### 2018

```{r age2018, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
# looks like there are some NAs in the 2018 data so I have to get rid of them
bandwidth <- bw.nrd0(na.omit(ocme_2018$age)) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(ocme_2018$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(ocme_2018$age, mod0 = n)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]


#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=ocme_2018, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(age)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2018$age, labels = ocme_2018$age) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2018 opioid-detected deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))

age_2018 <- ggplot(data=ocme_2018, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(age)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2018$age, labels = ocme_2018$age) + theme_light() + geom_vline(xintercept = modes[-length(modes)], size=1) + ggtitle("Age distribution for 2018 opioid-detected deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))

age_2018

ggsave(plot=age_2018, filename = "images/statewide/age_2018.png", width = 10, height = 8)
 
```

#### 2019

```{r age2019, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
bandwidth <- bw.nrd0(ocme_2019$age) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(ocme_2019$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(ocme_2019$age, mod0 = n)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]


#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=ocme_2019, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(age)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2019$age, labels = ocme_2019$age) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2019 opioid-detected deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))

age_2019 <- ggplot(data=ocme_2019, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(age)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2019$age, labels = ocme_2019$age) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2019 opioid-detected deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))

age_2019

ggsave(plot=age_2019, filename = "images/statewide/age_2019.png", width = 10, height = 8)

```

#### 2020

```{r age2020, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
bandwidth <- bw.nrd0(ocme_2020$Age) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(ocme_2020$Age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(ocme_2020$Age, mod0 = n)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]


#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=ocme_2020, aes(x=Age, ..count..)) + 
      geom_bar(aes(fill = as.character(Age)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2020$Age, labels = ocme_2020$Age) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2020 opioid-detected deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))

age_2020 <- ggplot(data=ocme_2020, aes(x=Age, ..count..)) + 
      geom_bar(aes(fill = as.character(Age)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2020$Age, labels = ocme_2020$Age) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2020 opioid-detected deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))

age_2020

ggsave(plot=age_2020, filename = "images/statewide/age_2020.png", width = 10, height = 8)

```


#### 2021

```{r age2021, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
bandwidth <- bw.nrd0(ocme_2021$Age) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(ocme_2021$Age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(ocme_2021$Age, mod0 = n)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]


#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=ocme_2021, aes(x=Age, ..count..)) + 
      geom_bar(aes(fill = as.character(Age)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2021$Age, labels = ocme_2021$Age) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2021 opioid-detected deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))

age_2021 <- ggplot(data=ocme_2021, aes(x=Age, ..count..)) + 
      geom_bar(aes(fill = as.character(Age)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2021$Age, labels = ocme_2021$Age) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2021 opioid-detected deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))
age_2021

ggsave(plot=age_2021, filename = "images/statewide/age_2021.png", width = 10, height = 8)
 
```


#### 2022

```{r age2022, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
bandwidth <- bw.nrd0(ocme_2022$Age) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(ocme_2022$Age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(ocme_2022$Age, mod0 = n)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]


#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=ocme_2022, aes(x=Age, ..count..)) + 
      geom_bar(aes(fill = as.character(Age)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2022$Age, labels = ocme_2022$Age) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2022 opioid-detected deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))

age_2022 <- ggplot(data=ocme_2022, aes(x=Age, ..count..)) + 
      geom_bar(aes(fill = as.character(Age)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2022$Age, labels = ocme_2022$Age) + theme_light() + geom_vline(xintercept = modes[-length(modes)], size=1) + ggtitle("Age distribution for 2022 opioid-detected deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))
age_2022


ggsave(plot=age_2022, filename = "images/statewide/age_2022.png", width = 10, height = 8)
 
```

### Day of week {.tabset}

#### 2018

```{r week2018, echo=FALSE, message=FALSE, warning=FALSE}
#table(ocme_2018$month)

# ordering of variables on x axis from https://www.statology.org/ggplot-order-x-axis/
# make sure to use scale_x_discrete!

weekday_2018 <- ggplot(data=ocme_2018, aes(x=factor(Weekday, level=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")), ..count..)) + geom_bar(aes(fill = as.character(Weekday)), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 0), breaks = ocme_2018$Weekday, labels = ocme_2018$Weekday) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle("ODs per weekday for 2018 opioid-detected deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))+xlab("Weekday")

weekday_2018

ggsave(plot=weekday_2018, filename = "images/statewide/weekday_2018.png", width = 10, height = 8)

```

#### 2019

```{r week2019, echo=FALSE, message=FALSE, warning=FALSE}
#table(ocme_2019$Month)

weekday_2019 <- ggplot(data=ocme_2019, aes(x=factor(Weekday, level=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")), ..count..)) + geom_bar(aes(fill = as.character(Weekday)), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 0), breaks = ocme_2019$Weekday, labels = ocme_2019$Weekday) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle("ODs per weekday for 2019 opioid-detected deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))+xlab("Weekday")

weekday_2019

ggsave(plot=weekday_2019, filename = "images/statewide/weekday_2019.png", width = 10, height = 8)

```

#### 2020

```{r week2020, echo=FALSE, message=FALSE, warning=FALSE}
weekday_2020 <- ggplot(data=ocme_2020, aes(x=factor(Weekday, level=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")), ..count..)) + geom_bar(aes(fill = as.character(Weekday)), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 0), breaks = ocme_2020$Weekday, labels = ocme_2020$Weekday) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle("ODs per weekday for 2020 opioid-detected deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))+xlab("Weekday")

weekday_2020

ggsave(plot=weekday_2020, filename = "images/statewide/weekday_2020.png", width = 10, height = 8)

```


#### 2021

```{r week2021, echo=FALSE, message=FALSE, warning=FALSE}
#table(ocme_2021$Month)

weekday_2021 <- ggplot(data=ocme_2021, aes(x=factor(Weekday, level=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")), ..count..)) + geom_bar(aes(fill = as.character(Weekday)), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 0), breaks = ocme_2021$Weekday, labels = ocme_2021$Weekday) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle("ODs per weekday for 2021 opioid-detected deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))+xlab("Weekday")

weekday_2021

ggsave(plot=weekday_2021, filename = "images/statewide/weekday_2021.png", width = 10, height = 8)

```


#### 2022

```{r week2022, echo=FALSE, message=FALSE, warning=FALSE}

weekday_2022 <- ggplot(data=ocme_2022, aes(x=factor(Weekday, level=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")), ..count..)) + geom_bar(aes(fill = as.character(Weekday)), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 0), breaks = ocme_2022$Weekday, labels = ocme_2022$Weekday) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle("ODs per weekday for 2022 opioid-detected deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))+xlab("Weekday")

weekday_2022

ggsave(plot=weekday_2022, filename = "images/statewide/weekday_2022.png", width = 10, height = 8)

```


### Month {.tabset}

#### 2018

```{r month2018, echo=FALSE, message=FALSE, warning=FALSE}
#table(ocme_2018$month)

month_2018 <- ggplot(data=ocme_2018, aes(x=month, ..count..)) + geom_bar(aes(fill = as.character(month)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0), breaks = ocme_2018$month, labels = ocme_2018$month) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle("ODs per month for 2018 opioid-detected deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))

month_2018

ggsave(plot=month_2018, filename = "images/statewide/month_2018.png", width = 10, height = 8)

```

#### 2019

```{r month2019, echo=FALSE, message=FALSE, warning=FALSE}
#table(ocme_2019$Month)

month_2019 <- ggplot(data=ocme_2019, aes(x=Month, ..count..)) + geom_bar(aes(fill = as.character(Month)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0), breaks = ocme_2019$Month, labels = ocme_2019$Month) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle("ODs per month for 2019 opioid-detected deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))

month_2019

ggsave(plot=month_2019, filename = "images/statewide/month_2019.png", width = 10, height = 8)

```

#### 2020

```{r month2020, echo=FALSE, message=FALSE, warning=FALSE}
month_2020 <- ggplot(data=ocme_2020, aes(x=Month, ..count..)) + geom_bar(aes(fill = as.character(Month)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0), breaks = ocme_2020$Month, labels = ocme_2020$Month) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle("ODs per month for 2020 opioid-detected deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))

month_2020

ggsave(plot=month_2020, filename = "images/statewide/month_2020.png", width = 10, height = 8)

```


#### 2021

```{r month2021, echo=FALSE, message=FALSE, warning=FALSE}
#table(ocme_2021$Month)

month_2021 <- ggplot(data=ocme_2021, aes(x=Month, ..count..)) + geom_bar(aes(fill = as.character(Month)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0), breaks = ocme_2021$Month, labels = ocme_2021$Month) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle("ODs per month for 2021 opioid-detected deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))

month_2021

ggsave(plot=month_2021, filename = "images/statewide/month_2021.png", width = 10, height = 8)

```


#### 2022

```{r month2022, echo=FALSE, message=FALSE, warning=FALSE}

month_2022 <- ggplot(data=ocme_2022, aes(x=Month, ..count..)) + geom_bar(aes(fill = as.character(Month)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0), breaks = ocme_2022$Month, labels = ocme_2022$Month) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle("ODs per month for 2022 opioid-detected deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))

month_2022

ggsave(plot=month_2022, filename = "images/statewide/month_2022.png", width = 10, height = 8)

```

### Town of OD {.tabset}

#### 2018


```{r town2018, echo=FALSE, message=FALSE, warning=FALSE}
#table(ocme_2018$injury_city_updated)

ocme_2018$fatal <- 1

# turn "" into "Unknown" so we can plot it, and convert Willimantic to Windham because it messes up the out-of-town calculations otherwise

ocme_2018$injury_city_revised <- ifelse(ocme_2018$injury_city_updated == "", "Unknown", ifelse(ocme_2018$injury_city_updated == "Willimantic", "Windham", ocme_2018$injury_city_updated))

#fixed axis text size in theme() using https://stackoverflow.com/questions/65956833/ggplot2-how-to-dynamically-wrap-resize-rescale-x-axis-labels-so-they-wont-over

town_2018 <- ggplot(data=ocme_2018, aes(x=injury_city_revised, ..count..)) + 
      geom_bar(aes(fill = ..count..), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "red", size = 3/.pt) + ggtitle("City/town of OD for 2018 opioid-detected deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 3))

town_2018

ggsave(plot=town_2018, filename = "images/statewide/overdoses_by_town_2018.png", width = 10, height = 8)

#find the number of ODs in each town by aggregating the indicator var for each town, from https://stackoverflow.com/questions/60961960/how-do-i-sum-a-column-with-categorical-vairables-in-r

ocme_2018_counts <- ocme_2018 %>% group_by(injury_city_revised) %>% summarise_at(vars(fatal), sum)

# rename the town_pop_2020 column names to be more descriptive--"NAME" is so we can map it easily

colnames(city_pop_2018) <- c('Town', 'Pop')

# rename the injury county variable so we can merge it with the county_pop_2020 df

colnames(ocme_2018_counts) <- c('Town', 'Fatal')

#merge with ct_towns so that way we can use the fatal OD counts for the fill

ct_town_odcounts_2018 <- merge(city_pop_2018, ocme_2018_counts,by="Town")

# get the per-capita death rate, multiply by 10000 so the numbers aren't super small
ct_town_odcounts_2018$Prop <- round((ct_town_odcounts_2018$Fatal/ct_town_odcounts_2018$Pop)*10000, digits=2)

# Plot the per-capita death rate
# had an issue with the fill, edited with http://www.sthda.com/english/wiki/ggplot2-barplots-quick-start-guide-r-software-and-data-visualization
# needed to change the font size for the text overlay, used https://stackoverflow.com/questions/25061822/ggplot-geom-text-font-size-control

town_2018_capita <- ggplot(data=ct_town_odcounts_2018, aes(x=Town, y=Prop, fill=Prop, position = "dodge")) + 
      geom_bar(stat="Identity") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = Prop), vjust = 0, colour = "red", size = 3/.pt) + ggtitle("Town of OD for 2018 opioid-detected deaths \n per 10000 residents")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 3))

town_2018_capita

ggsave(plot=town_2018_capita, filename = "images/statewide/overdoses_by_town_capita_2018.png", width = 10, height = 8)
 
```

#### 2019


```{r town2019, echo=FALSE, message=FALSE, warning=FALSE}
#table(ocme_2019$Injury.City)

ocme_2019$fatal <- 1

ocme_2019$injury_city_revised <- ifelse(ocme_2019$Injury.City == "", "Unknown", ifelse(ocme_2019$Injury.City == "Willimantic", "Windham", ocme_2019$Injury.City))

#fixed axis text size in theme() using https://stackoverflow.com/questions/65956833/ggplot2-how-to-dynamically-wrap-resize-rescale-x-axis-labels-so-they-wont-over

town_2019 <- ggplot(data=ocme_2019, aes(x=injury_city_revised, ..count..)) + 
      geom_bar(aes(fill = ..count..), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "red", size = 3/.pt) + ggtitle("City/town of OD for 2019 opioid-detected deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 3))

town_2019

ggsave(plot=town_2019, filename = "images/statewide/overdoses_by_town_2019.png", width = 10, height = 8)

#find the number of ODs in each town by aggregating the indicator var for each town, from https://stackoverflow.com/questions/60961960/how-do-i-sum-a-column-with-categorical-vairables-in-r

ocme_2019_counts <- ocme_2019 %>% group_by(Injury.City) %>% summarise_at(vars(fatal), sum)

# rename the town_pop_2020 column names to be more descriptive--"NAME" is so we can map it easily

colnames(city_pop_2019) <- c('Town', 'Pop')

# rename the injury county variable so we can merge it with the county_pop_2020 df

colnames(ocme_2019_counts) <- c('Town', 'Fatal')

#merge with ct_towns so that way we can use the fatal OD counts for the fill

ct_town_odcounts_2019 <- merge(city_pop_2019, ocme_2019_counts,by="Town")

# get the per-capita death rate, multiply by 10000 so the numbers aren't super small
ct_town_odcounts_2019$Prop <- round((ct_town_odcounts_2019$Fatal/ct_town_odcounts_2019$Pop)*10000, digits=2)

# Plot the per-capita death rate
# had an issue with the fill, edited with http://www.sthda.com/english/wiki/ggplot2-barplots-quick-start-guide-r-software-and-data-visualization
# needed to change the font size for the text overlay, used https://stackoverflow.com/questions/25061822/ggplot-geom-text-font-size-control

town_2019_capita <- ggplot(data=ct_town_odcounts_2019, aes(x=Town, y=Prop, fill=Prop, position = "dodge")) + 
      geom_bar(stat="Identity") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = Prop), vjust = 0, colour = "red", size = 3/.pt) + ggtitle("Town of OD for 2019 opioid-detected deaths \n per 10000 residents")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 3))

town_2019_capita

ggsave(plot=town_2019_capita, filename = "images/statewide/overdoses_by_town_capita_2019.png", width = 10, height = 8)
```

#### 2020



```{r town2020, echo=FALSE, message=FALSE, warning=FALSE}
#table(ocme_2020$Injury.City)

ocme_2020$fatal <- 1

ocme_2020$injury_city_revised <- ifelse(ocme_2020$Injury.City == "", "Unknown", ifelse(ocme_2020$Injury.City == "Willimantic", "Windham", ocme_2020$Injury.City))

#fixed axis text size in theme() using https://stackoverflow.com/questions/65956833/ggplot2-how-to-dynamically-wrap-resize-rescale-x-axis-labels-so-they-wont-over

town_2020 <- ggplot(data=ocme_2020, aes(x=injury_city_revised, ..count..)) + 
      geom_bar(aes(fill = ..count..), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "red", size = 3/.pt)  + ggtitle("City/town of OD for 2020 opioid-detected deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 3))

town_2020

ggsave(plot=town_2020, filename = "images/statewide/overdoses_by_town_2020.png", width = 10, height = 8)


#find the number of ODs in each town by aggregating the indicator var for each town, from https://stackoverflow.com/questions/60961960/how-do-i-sum-a-column-with-categorical-vairables-in-r

ocme_2020_counts <- ocme_2020 %>% group_by(Injury.City) %>% summarise_at(vars(fatal), sum)

# rename the town_pop_2020 column names to be more descriptive--"NAME" is so we can map it easily

colnames(city_pop_2020) <- c('Town', 'Pop')

# rename the injury county variable so we can merge it with the county_pop_2020 df

colnames(ocme_2020_counts) <- c('Town', 'Fatal')

#merge with ct_towns so that way we can use the fatal OD counts for the fill

ct_town_odcounts_2020 <- merge(city_pop_2020, ocme_2020_counts,by="Town")

# get the per-capita death rate, multiply by 10000 so the numbers aren't super small
ct_town_odcounts_2020$Prop <- round((ct_town_odcounts_2020$Fatal/ct_town_odcounts_2020$Pop)*10000, digits=2)

# not all towns experienced an OD in 2020. Check which: https://stackoverflow.com/questions/72274729/find-elements-in-a-column-that-is-not-in-another-column-of-another-dataframe-in
#ocme_2020_odcounts <- data.frame(setdiff(ct_towns$NAME, ct_town_odcounts_2020$NAME), 0)
# give this the same column names as ocme_2020_counts
#colnames(ocme_2020_odcounts) <- c('NAME', 'Fatal', 'Prop')
#merge the two datasets
#ocme_2020_counts <- rbind(ocme_2020_counts, ct_town_odcounts_2020)

# Plot the per-capita death rate
# had an issue with the fill, edited with http://www.sthda.com/english/wiki/ggplot2-barplots-quick-start-guide-r-software-and-data-visualization
# needed to change the font size for the text overlay, used https://stackoverflow.com/questions/25061822/ggplot-geom-text-font-size-control

town_2020_capita <- ggplot(data=ct_town_odcounts_2020, aes(x=Town, y=Prop, fill=Prop, position = "dodge")) + 
      geom_bar(stat="Identity") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = Prop), vjust = 0, colour = "red", size = 3/.pt) + ggtitle("Town of OD for 2020 opioid-detected deaths \n per 10000 residents")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 3))

town_2020_capita

ggsave(plot=town_2020_capita, filename = "images/statewide/overdoses_by_town_capita_2020.png", width = 10, height = 8)

```

#### 2021

```{r town2021, echo=FALSE, message=FALSE, warning=FALSE}
#table(ocme_2021$Injury.City)

ocme_2021$fatal <- 1

ocme_2021$injury_city_revised <- ifelse(ocme_2021$Injury.City == "", "Unknown", ifelse(ocme_2021$Injury.City == "Willimantic", "Windham", ocme_2021$Injury.City))

#fixed axis text size in theme() using https://stackoverflow.com/questions/65956833/ggplot2-how-to-dynamically-wrap-resize-rescale-x-axis-labels-so-they-wont-over

town_2021 <- ggplot(data=ocme_2021, aes(x=injury_city_revised, ..count..)) + 
      geom_bar(aes(fill = ..count..), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "red", size = 3/.pt) + ggtitle("City/town of OD for 2021 opioid-detected deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 3))

town_2021


ggsave(plot=town_2021, filename = "images/statewide/overdoses_by_town_2021.png", width = 10, height = 8)

#find the number of ODs in each town by aggregating the indicator var for each town, from https://stackoverflow.com/questions/60961960/how-do-i-sum-a-column-with-categorical-vairables-in-r

ocme_2021_counts <- ocme_2021 %>% group_by(Injury.City) %>% summarise_at(vars(fatal), sum)

# rename the town_pop_2020 column names to be more descriptive--"NAME" is so we can map it easily

colnames(city_pop_2021) <- c('Town', 'Pop')

# rename the injury county variable so we can merge it with the county_pop_2020 df

colnames(ocme_2021_counts) <- c('Town', 'Fatal')

#merge with ct_towns so that way we can use the fatal OD counts for the fill

ct_town_odcounts_2021 <- merge(city_pop_2021, ocme_2021_counts,by="Town")

# get the per-capita death rate, multiply by 10000 so the numbers aren't super small
ct_town_odcounts_2021$Prop <- round((ct_town_odcounts_2021$Fatal/ct_town_odcounts_2021$Pop)*10000, digits=2)

# Plot the per-capita death rate
# had an issue with the fill, edited with http://www.sthda.com/english/wiki/ggplot2-barplots-quick-start-guide-r-software-and-data-visualization
# needed to change the font size for the text overlay, used https://stackoverflow.com/questions/25061822/ggplot-geom-text-font-size-control

town_2021_capita <- ggplot(data=ct_town_odcounts_2021, aes(x=Town, y=Prop, fill=Prop, position = "dodge")) + 
      geom_bar(stat="Identity") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = Prop), vjust = 0, colour = "red", size = 3/.pt) + ggtitle("Town of OD for 2021 opioid-detected deaths \n per 10000 residents")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 3))

town_2021_capita

ggsave(plot=town_2021_capita, filename = "images/statewide/overdoses_by_town_capita_2021.png", width = 10, height = 8)
```

#### 2022
```{r town2022, echo=FALSE, message=FALSE, warning=FALSE}
#table(ocme_2021$Injury.City)

ocme_2022$fatal <- 1

ocme_2022$injury_city_revised <- ifelse(ocme_2022$Injury.City == "", "Unknown", ifelse(ocme_2022$Injury.City == "Willimantic", "Windham", ocme_2022$Injury.City))

#fixed axis text size in theme() using https://stackoverflow.com/questions/65956833/ggplot2-how-to-dynamically-wrap-resize-rescale-x-axis-labels-so-they-wont-over

town_2022 <- ggplot(data=ocme_2022, aes(x=injury_city_revised, ..count..)) + 
      geom_bar(aes(fill = ..count..), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "red", size = 3/.pt) + ggtitle("City/town of OD for 2022 opioid-detected deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 3))

town_2022

ggsave(plot=town_2022, filename = "images/statewide/overdoses_by_town_2022.png", width = 10, height = 8)

#find the number of ODs in each town by aggregating the indicator var for each town, from https://stackoverflow.com/questions/60961960/how-do-i-sum-a-column-with-categorical-vairables-in-r

ocme_2022_counts <- ocme_2022 %>% group_by(Injury.City) %>% summarise_at(vars(fatal), sum)

# rename the town_pop_2020 column names to be more descriptive--"NAME" is so we can map it easily

colnames(city_pop_2022) <- c('Town', 'Pop')

# rename the injury county variable so we can merge it with the county_pop_2020 df

colnames(ocme_2022_counts) <- c('Town', 'Fatal')

#merge with ct_towns so that way we can use the fatal OD counts for the fill

ct_town_odcounts_2022 <- merge(city_pop_2022, ocme_2022_counts,by="Town")

# get the per-capita death rate, multiply by 10000 so the numbers aren't super small
ct_town_odcounts_2022$Prop <- round((ct_town_odcounts_2022$Fatal/ct_town_odcounts_2022$Pop)*10000, digits=2)

# Plot the per-capita death rate
# had an issue with the fill, edited with http://www.sthda.com/english/wiki/ggplot2-barplots-quick-start-guide-r-software-and-data-visualization
# needed to change the font size for the text overlay, used https://stackoverflow.com/questions/25061822/ggplot-geom-text-font-size-control

town_2022_capita <- ggplot(data=ct_town_odcounts_2022, aes(x=Town, y=Prop, fill=Prop, position = "dodge")) + 
      geom_bar(stat="Identity") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = Prop), vjust = 0, colour = "red", size = 3/.pt) + ggtitle("Town of OD for 2022 opioid-detected deaths \n per 10000 residents")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 3))

town_2022_capita

ggsave(plot=town_2022_capita, filename = "images/statewide/overdoses_by_town_capita_2022.png", width = 10, height = 8)
```

### Maps of OD towns {.tabset} 

#### 2018

```{r town2018Map, echo=FALSE, message=FALSE, warning=FALSE}
#create an indicator variable so we can aggregate the number of ODs per city

ocme_2018$fatal <- 1

#find the number of ODs in each town by aggregating the indicator var for each town, from https://stackoverflow.com/questions/60961960/how-do-i-sum-a-column-with-categorical-vairables-in-r

ocme_2018_counts<- ocme_2018 %>% group_by(injury_city_revised) %>% summarise_at(vars(fatal), sum)

#rename the injury city variable so we can merge it with the ct_towns df

colnames(ocme_2018_counts) <- c('NAME', 'Fatal')

# not all towns experienced an OD in 2020. Check which: https://stackoverflow.com/questions/72274729/find-elements-in-a-column-that-is-not-in-another-column-of-another-dataframe-in

ocme_2018_odcounts <- data.frame(setdiff(ct_towns$NAME, ocme_2018_counts$NAME), 0)

# give this the same column names as ocme_2020_counts

colnames(ocme_2018_odcounts) <- c('NAME', 'Fatal')

# merge the two datasets

ocme_2018_counts <- rbind(ocme_2018_counts, ocme_2018_odcounts)

#merge with ct_towns so that way we can use the fatal OD counts for the fill

ct_town_odcounts_2018 <- merge(ct_towns, ocme_2018_counts,by="NAME")

# Plot! And use the fatal data to color the plot in. Info from https://geanders.github.io/navy_public_health/3-2-basic-mapping.html

map_2018 <- ggplot(ocme_2018) + theme_void() + geom_sf(data=ct_town_odcounts_2018, size=0.1, aes(fill=Fatal))+ ggtitle("Fatal opioid-detected overdoses by town in 2018")+scale_fill_viridis_c(option="plasma")+theme(plot.title = element_text(hjust = 0.5), legend.position="right")

map_2018

#save this one for the presentation
ggsave(plot=map_2018, filename = "images/statewide/map_2018.png", width = 10, height = 8)

#save this one for the presentation
ggsave(plot=map_2018, filename = "images/statewide/map_2018.png", width = 10, height = 8)

```

#### 2019

```{r town2019Map, echo=FALSE, message=FALSE, warning=FALSE}
#create an indicator variable so we can aggregate the number of ODs per city

ocme_2019$fatal <- 1

#find the number of ODs in each town by aggregating the indicator var for each town, from https://stackoverflow.com/questions/60961960/how-do-i-sum-a-column-with-categorical-vairables-in-r

ocme_2019_counts<- ocme_2019 %>% group_by(injury_city_revised) %>% summarise_at(vars(fatal), sum)

#rename the injury city variable so we can merge it with the ct_towns df

colnames(ocme_2019_counts) <- c('NAME', 'Fatal')

# not all towns experienced an OD in 2020. Check which: https://stackoverflow.com/questions/72274729/find-elements-in-a-column-that-is-not-in-another-column-of-another-dataframe-in

ocme_2019_odcounts <- data.frame(setdiff(ct_towns$NAME, ocme_2019_counts$NAME), 0)

# give this the same column names as ocme_2020_counts

colnames(ocme_2019_odcounts) <- c('NAME', 'Fatal')

# merge the two datasets

ocme_2019_counts <- rbind(ocme_2019_counts, ocme_2019_odcounts)

#merge with ct_towns so that way we can use the fatal OD counts for the fill

ct_town_odcounts_2019 <- merge(ct_towns, ocme_2019_counts,by="NAME")

# Plot! And use the fatal data to color the plot in. Info from https://geanders.github.io/navy_public_health/3-2-basic-mapping.html

map_2019 <- ggplot(ocme_2019) + theme_void() + geom_sf(data=ct_town_odcounts_2019, size=0.1, aes(fill=Fatal))+ ggtitle("Fatal opioid-detected overdoses by town in 2019")+scale_fill_viridis_c(option="plasma")+theme(plot.title = element_text(hjust = 0.5), legend.position="right")

map_2019

#save this one for the presentation
ggsave(plot=map_2019, filename = "images/statewide/map_2019.png", width = 10, height = 8)

#save this one for the presentation
ggsave(plot=map_2019, filename = "images/statewide/map_2019.png", width = 10, height = 8)

```

#### 2020

```{r town2020Map, echo=FALSE, message=FALSE, warning=FALSE}
#create an indicator variable so we can aggregate the number of ODs per city

ocme_2020$fatal <- 1

#find the number of ODs in each town by aggregating the indicator var for each town, from https://stackoverflow.com/questions/60961960/how-do-i-sum-a-column-with-categorical-vairables-in-r

ocme_2020_counts<- ocme_2020 %>% group_by(injury_city_revised) %>% summarise_at(vars(fatal), sum)

#rename the injury city variable so we can merge it with the ct_towns df

colnames(ocme_2020_counts) <- c('NAME', 'Fatal')

# not all towns experienced an OD in 2020. Check which: https://stackoverflow.com/questions/72274729/find-elements-in-a-column-that-is-not-in-another-column-of-another-dataframe-in

ocme_2020_odcounts <- data.frame(setdiff(ct_towns$NAME, ocme_2020_counts$NAME), 0)

# give this the same column names as ocme_2020_counts

colnames(ocme_2020_odcounts) <- c('NAME', 'Fatal')

# merge the two datasets

ocme_2020_counts <- rbind(ocme_2020_counts, ocme_2020_odcounts)

#merge with ct_towns so that way we can use the fatal OD counts for the fill

ct_town_odcounts_2020 <- merge(ct_towns, ocme_2020_counts,by="NAME")

# Plot! And use the fatal data to color the plot in. Info from https://geanders.github.io/navy_public_health/3-2-basic-mapping.html

map_2020 <- ggplot(ocme_2020) + theme_void() + geom_sf(data=ct_town_odcounts_2020, size=0.1, aes(fill=Fatal))+ ggtitle("Fatal opioid-detected overdoses by town in 2020")+scale_fill_viridis_c(option="plasma")+theme(plot.title = element_text(hjust = 0.5), legend.position="right")

map_2020

#save this one for the presentation
ggsave(plot=map_2020, filename = "images/statewide/map_2020.png", width = 10, height = 8)

#save this one for the presentation
ggsave(plot=map_2020, filename = "images/statewide/map_2020.png", width = 10, height = 8)

```

#### 2021

```{r town2021Map, echo=FALSE, message=FALSE, warning=FALSE}
#create an indicator variable so we can aggregate the number of ODs per city

ocme_2021$fatal <- 1

#find the number of ODs in each town by aggregating the indicator var for each town, from https://stackoverflow.com/questions/60961960/how-do-i-sum-a-column-with-categorical-vairables-in-r

ocme_2021_counts<- ocme_2021 %>% group_by(injury_city_revised) %>% summarise_at(vars(fatal), sum)

#rename the injury city variable so we can merge it with the ct_towns df

colnames(ocme_2021_counts) <- c('NAME', 'Fatal')

# not all towns experienced an OD in 2020. Check which: https://stackoverflow.com/questions/72274729/find-elements-in-a-column-that-is-not-in-another-column-of-another-dataframe-in

ocme_2021_odcounts <- data.frame(setdiff(ct_towns$NAME, ocme_2021_counts$NAME), 0)

# give this the same column names as ocme_2020_counts

colnames(ocme_2021_odcounts) <- c('NAME', 'Fatal')

# merge the two datasets

ocme_2021_counts <- rbind(ocme_2021_counts, ocme_2021_odcounts)

#merge with ct_towns so that way we can use the fatal OD counts for the fill

ct_town_odcounts_2021 <- merge(ct_towns, ocme_2021_counts,by="NAME")

# Plot! And use the fatal data to color the plot in. Info from https://geanders.github.io/navy_public_health/3-2-basic-mapping.html

map_2021<- ggplot(ocme_2021) + theme_void() + geom_sf(data=ct_town_odcounts_2021, size=0.1, aes(fill=Fatal))+ ggtitle("Fatal opioid-detected overdoses by town in 2021")+scale_fill_viridis_c(option="plasma")+theme(plot.title = element_text(hjust = 0.5), legend.position="right")

map_2021

#save this one for the presentation
ggsave(plot=map_2021, filename = "images/statewide/map_2021.png", width = 10, height = 8)

#save this one for the presentation
ggsave(plot=map_2021, filename = "images/statewide/map_2021.png", width = 10, height = 8)
```

#### 2022 {.tabset}

```{r town2022Map, echo=FALSE, message=FALSE, warning=FALSE}
#create an indicator variable so we can aggregate the number of ODs per city

ocme_2022$fatal <- 1

#find the number of ODs in each town by aggregating the indicator var for each town, from https://stackoverflow.com/questions/60961960/how-do-i-sum-a-column-with-categorical-vairables-in-r

ocme_2022_counts<- ocme_2022 %>% group_by(injury_city_revised) %>% summarise_at(vars(fatal), sum)

#rename the injury city variable so we can merge it with the ct_towns df

colnames(ocme_2022_counts) <- c('NAME', 'Fatal')

# not all towns experienced an OD in 2020. Check which: https://stackoverflow.com/questions/72274729/find-elements-in-a-column-that-is-not-in-another-column-of-another-dataframe-in

ocme_2022_odcounts <- data.frame(setdiff(ct_towns$NAME, ocme_2022_counts$NAME), 0)

# give this the same column names as ocme_2020_counts

colnames(ocme_2022_odcounts) <- c('NAME', 'Fatal')

# merge the two datasets

ocme_2022_counts <- rbind(ocme_2022_counts, ocme_2022_odcounts)

#merge with ct_towns so that way we can use the fatal OD counts for the fill

ct_town_odcounts_2022 <- merge(ct_towns, ocme_2022_counts,by="NAME")

# Plot! And use the fatal data to color the plot in. Info from https://geanders.github.io/navy_public_health/3-2-basic-mapping.html

map_2022 <- ggplot(ocme_2022) + theme_void() + geom_sf(data=ct_town_odcounts_2022, size=0.1, aes(fill=Fatal))+ ggtitle("Fatal opioid-detected overdoses by town in 2022")+scale_fill_viridis_c(option="plasma")+theme(plot.title = element_text(hjust = 0.5), legend.position="right")

map_2022

#save this one for the presentation
ggsave(plot=map_2022, filename = "images/statewide/map_2022.png", width = 10, height = 8)

#save this one for the presentation
ggsave(plot=map_2022, filename = "images/statewide/map_2022.png", width = 10, height = 8)

```

### Injury town v. residence town {.tabset}

I chose towns based on either what regional harm reduction workers wanted, based on towns with overdose prevention infrastructure or based on towns with the most ODs in the past few years. I can change this easily to show all towns. I'm looking at instances where people who OD'd in a town didn't live there -- I am referring to this as "out-of-town ODs."

#### 2018

```{r injuryResidence2018, echo=FALSE, message=FALSE, warning=FALSE}
# get rid of blank/unknown injury and residence locations, from https://www.r-bloggers.com/2022/05/subsetting-with-multiple-conditions-in-r/

ocme_2018_noblank <- ocme_2018 %>% filter(residence_city_updated != "" & injury_city_revised!="Unknown")

# convert all Willimantic residences to Windham, because otherwise the out of town calculations get messed up
ocme_2018_noblank$residence_city_updated <- ifelse(ocme_2018_noblank$residence_city_updated == "Willimantic", "Windham", ocme_2018_noblank$residence_city_updated)

# print a statement with some function output -- from https://stackoverflow.com/questions/13036649/how-do-i-concatenate-string-and-an-output-evaluated-from-a-function-in-r [the %d stands for an integer]

sprintf("There were %d instances where a person's residence town didn't match up with their injury town, out of %d ODs in 2018 where the injury and residence town were both known. This is %f of all such ODs in this year.", nrow(subset(ocme_2018_noblank, residence_city_updated != injury_city_revised)), nrow(ocme_2018_noblank), nrow(subset(ocme_2018_noblank, residence_city_updated != injury_city_revised))/nrow(ocme_2018_noblank))

#print("Here are the towns where someone OD'd but didn't live there:")

#table(subset(ocme_2018_noblank, residence_city_updated != injury_city_revised)$injury_city_revised)

print("Here's a more in-depth look at out-of-town ODs in specific towns:")

# initialize empty matrix, from https://stackoverflow.com/questions/32712301/create-empty-data-frame-with-column-names-by-assigning-a-string-vector

out_of_town_2018 <- data.frame(matrix(ncol = 4, nrow = 0))
cols <- c("Town name","Out-of-town ODs","Total ODs","Percent of out-of-town ODs")

# iteratively add info from these towns to the df, from https://www.projectpro.io/recipes/append-output-from-for-loop-dataframe-r

for (i in c("Danbury", "Waterbury", "Torrington", "New Haven", "West Haven", "Hartford", "Norwich", "New London", "Bridgeport", "New Britain", "Litchfield", "Meriden", "Norwalk", "Stamford", "Windham")) {
  
  # subset for the number of ODs in that town 
  
  ocme_injury_2018 <- subset(ocme_2018_noblank, injury_city_revised == i)
  
  # create a results vector to append to out_of_town 2020 df
  # contains town name, number of ODs from out of town residents, total ODs and fraction of out of town ODs in that   order
  
  results <- c(i, nrow(subset(ocme_injury_2018, residence_city_updated != injury_city_revised)), nrow(ocme_injury_2018), round(nrow(subset(ocme_injury_2018, residence_city_updated != injury_city_revised))/nrow(ocme_injury_2018), 3)*100)

  # add this to the out of town df 
  out_of_town_2018 = rbind(out_of_town_2018, results)
}

# for some reason it doesn't work if I do this before adding rows?
colnames(out_of_town_2018) <- cols

out_of_town_2018

```


#### 2019

```{r injuryResidence2019, echo=FALSE, message=FALSE, warning=FALSE}

# get rid of blank/unknown injury and residence locations, from https://www.r-bloggers.com/2022/05/subsetting-with-multiple-conditions-in-r/

ocme_2019_noblank <- ocme_2019 %>% filter(Residence.City != "" & injury_city_revised!="Unknown")

# convert all Willimantic residences to Windham, because otherwise the out of town calculations get messed up
ocme_2019_noblank$Residence.City <- ifelse(ocme_2019_noblank$Residence.City == "Willimantic", "Windham", ocme_2019_noblank$Residence.City)

# print a statement with some function output -- from https://stackoverflow.com/questions/13036649/how-do-i-concatenate-string-and-an-output-evaluated-from-a-function-in-r [the %d stands for an integer]
sprintf("There were %d instances where a person's residence town didn't match up with their injury town, out of %d ODs in 2019 where the injury and residence town were both known. This is %f of all such ODs in this year.", nrow(subset(ocme_2019_noblank, Residence.City != injury_city_revised)), nrow(ocme_2019_noblank), nrow(subset(ocme_2019_noblank, Residence.City != injury_city_revised))/nrow(ocme_2019_noblank))

#print("Here are the towns where someone OD'd but didn't live there:")

#table(subset(ocme_2019_noblank, Residence.City != injury_city_revised)$injury_city_revised)


print("Here's a more in-depth look at out-of-town ODs in specific towns:")

# initialize empty matrix, from https://stackoverflow.com/questions/32712301/create-empty-data-frame-with-column-names-by-assigning-a-string-vector

out_of_town_2019 <- data.frame(matrix(ncol = 4, nrow = 0))
cols <- c("Town name","Out-of-town ODs","Total ODs","Percent of out-of-town ODs")

# iteratively add info from these towns to the df, from https://www.projectpro.io/recipes/append-output-from-for-loop-dataframe-r

for (i in c("Danbury", "Waterbury", "Torrington", "New Haven", "West Haven", "Hartford", "Norwich", "New London", "Bridgeport", "New Britain", "Litchfield", "Meriden", "Norwalk", "Stamford", "Windham")) {
  
  # subset for the number of ODs in that town 
  
  ocme_injury_2019 <- subset(ocme_2019_noblank, injury_city_revised == i)
  
  # create a results vector to append to out_of_town 2020 df
  # contains town name, number of ODs from out of town residents, total ODs and fraction of out of town ODs in that   order
  
  results <- c(i, nrow(subset(ocme_injury_2019, Residence.City != injury_city_revised)), nrow(ocme_injury_2019), round(nrow(subset(ocme_injury_2019, Residence.City != injury_city_revised))/nrow(ocme_injury_2019), 3)*100)

  # add this to the out of town df 
  out_of_town_2019 = rbind(out_of_town_2019, results)
}

# for some reason it doesn't work if I do this before adding rows?
colnames(out_of_town_2019) <- cols

out_of_town_2019

```


#### 2020


```{r injuryResidence2020, echo=FALSE, message=FALSE, warning=FALSE}
# get rid of blank/unknown injury and residence locations, from https://www.r-bloggers.com/2022/05/subsetting-with-multiple-conditions-in-r/

ocme_2020_noblank <- ocme_2020 %>% filter(Residence.City != "" & injury_city_revised!="Unknown")

# convert all Willimantic residences to Windham, because otherwise the out of town calculations get messed up
ocme_2020_noblank$Residence.City <- ifelse(ocme_2020_noblank$Residence.City == "Willimantic", "Windham", ocme_2020_noblank$Residence.City)

# print a statement with some function output -- from https://stackoverflow.com/questions/13036649/how-do-i-concatenate-string-and-an-output-evaluated-from-a-function-in-r [the %d stands for an integer]
sprintf("There were %d instances where a person's residence town didn't match up with their injury town, out of %d ODs in 2020 where the injury and residence town were both known. This is %f of all such ODs in this year.", nrow(subset(ocme_2020_noblank, Residence.City != injury_city_revised)), nrow(ocme_2020_noblank), nrow(subset(ocme_2020_noblank, Residence.City != injury_city_revised))/nrow(ocme_2020_noblank))

#print("Here are the towns where someone OD'd but didn't live there:")

#table(subset(ocme_2020_noblank, Residence.City != injury_city_revised)$injury_city_revised)

print("Here's a more in-depth look at out-of-town ODs in specific towns:")

# initialize empty matrix, from https://stackoverflow.com/questions/32712301/create-empty-data-frame-with-column-names-by-assigning-a-string-vector

out_of_town_2020 <- data.frame(matrix(ncol = 4, nrow = 0))
cols <- c("Town name","Out-of-town ODs","Total ODs","Percent of out-of-town ODs")

# iteratively add info from these towns to the df, from https://www.projectpro.io/recipes/append-output-from-for-loop-dataframe-r

for (i in c("Danbury", "Waterbury", "Torrington", "New Haven", "West Haven", "Hartford", "Norwich", "New London", "Bridgeport", "New Britain", "Litchfield", "Meriden", "Norwalk", "Stamford", "Windham")) {
  
  # subset for the number of ODs in that town 
  
  ocme_injury_2020 <- subset(ocme_2020_noblank, injury_city_revised == i)
  
  # create a results vector to append to out_of_town 2020 df
  # contains town name, number of ODs from out of town residents, total ODs and fraction of out of town ODs in that   order
  
  results <- c(i, nrow(subset(ocme_injury_2020, Residence.City != injury_city_revised)), nrow(ocme_injury_2020), round(nrow(subset(ocme_injury_2020, Residence.City != injury_city_revised))/nrow(ocme_injury_2020), 3)*100)

  # add this to the out of town df 
  out_of_town_2020 = rbind(out_of_town_2020, results)
}

# for some reason it doesn't work if I do this before adding rows?
colnames(out_of_town_2020) <- cols

out_of_town_2020

```

#### 2021

```{r injuryResidence2021, echo=FALSE, message=FALSE, warning=FALSE}
# get rid of blank/unknown injury and residence locations, from https://www.r-bloggers.com/2022/05/subsetting-with-multiple-conditions-in-r/ 
# to get rid of people who aren't CT residents add in  & !(ct_residence.2.HOMELESS.777.UNK %in% c("0", "777"))

ocme_2021_noblank <- ocme_2021 %>% filter(Res.City != "" & injury_city_revised!="Unknown")

# convert all Willimantic residences to Windham, because otherwise the out of town calculations get messed up
ocme_2021_noblank$Res.City <- ifelse(ocme_2021_noblank$Res.City == "Willimantic", "Windham", ocme_2021_noblank$Res.City)

# print a statement with some function output -- from https://stackoverflow.com/questions/13036649/how-do-i-concatenate-string-and-an-output-evaluated-from-a-function-in-r [the %d stands for an integer]
sprintf("There were %d instances where a person's residence town didn't match up with their injury town, out of %d ODs in 2021 where the injury and residence town were both known. This is %f of all such ODs in this year.", nrow(subset(ocme_2021_noblank, Res.City != injury_city_revised)), nrow(ocme_2021_noblank), nrow(subset(ocme_2021_noblank, Res.City != injury_city_revised))/nrow(ocme_2021_noblank))

#print("Here are the towns where someone OD'd but didn't live there:")

#table(subset(ocme_2021_noblank, Res.City != injury_city_revised)$injury_city_revised)

print("Here's a more in-depth look at out-of-town ODs in specific towns:")

# initialize empty matrix, from https://stackoverflow.com/questions/32712301/create-empty-data-frame-with-column-names-by-assigning-a-string-vector

out_of_town_2021 <- data.frame(matrix(ncol = 4, nrow = 0))
cols <- c("Town name","Out-of-town ODs","Total ODs","Percent of out-of-town ODs")

# iteratively add info from these towns to the df, from https://www.projectpro.io/recipes/append-output-from-for-loop-dataframe-r

for (i in c("Danbury", "Waterbury", "Torrington", "New Haven", "West Haven", "Hartford", "Norwich", "New London", "Bridgeport", "New Britain", "Litchfield", "Meriden", "Norwalk", "Stamford", "Windham")) {
  
  # subset for the number of ODs in that town 
  
  ocme_injury_2021 <- subset(ocme_2021_noblank, injury_city_revised == i)
  
  # create a results vector to append to out_of_town 2020 df
  # contains town name, number of ODs from out of town residents, total ODs and fraction of out of town ODs in that   order
  
  results <- c(i, nrow(subset(ocme_injury_2021, Res.City != injury_city_revised)), nrow(ocme_injury_2021), round(nrow(subset(ocme_injury_2021, Res.City != injury_city_revised))/nrow(ocme_injury_2021), 3)*100)

  # add this to the out of town df 
  out_of_town_2021 = rbind(out_of_town_2021, results)
}

# for some reason it doesn't work if I do this before adding rows?
colnames(out_of_town_2021) <- cols

out_of_town_2021
```


#### 2022

```{r injuryResidence2022, echo=FALSE, message=FALSE, warning=FALSE}
# get rid of blank/unknown injury and residence locations, from https://www.r-bloggers.com/2022/05/subsetting-with-multiple-conditions-in-r/ -- also adding to get rid of people who aren't CT residents

ocme_2022_noblank <- ocme_2022 %>% filter(Res.City != "" & injury_city_revised!="Unknown")

# convert all Willimantic residences to Windham, because otherwise the out of town calculations get messed up
ocme_2022_noblank$Res.City <- ifelse(ocme_2022_noblank$Res.City == "Willimantic", "Windham", ocme_2022_noblank$Res.City)

# print a statement with some function output -- from https://stackoverflow.com/questions/13036649/how-do-i-concatenate-string-and-an-output-evaluated-from-a-function-in-r [the %d stands for an integer]
sprintf("There were %d instances where a person's residence town didn't match up with their injury town, out of %d ODs in 2022 where the injury and residence town were both known. This is %f of all such ODs in this year.", nrow(subset(ocme_2022_noblank, Res.City != injury_city_revised)), nrow(ocme_2022_noblank), nrow(subset(ocme_2022_noblank, Res.City != injury_city_revised))/nrow(ocme_2022_noblank))

#print("Here are the towns where someone OD'd but didn't live there:")

#table(subset(ocme_2022_noblank, Res.City != injury_city_revised)$injury_city_revised)

#export data for Hartford, New Haven, Bridgeport and Waterbury

#oot_2022 <- subset(ocme_2022_noblank, Res.City != injury_city_revised)
            
#big4_2022 <- subset(oot_2022, injury_city_revised %in% c("New Haven", "Hartford", "Bridgeport", "Waterbury"))

#big4_2022 <- big4_2022 %>% select(Res.City, injury.locn_revised, injury.location_details, injury_city_revised)

# for some reason for this you need double backslashes instead of single
#write.csv(big4_2022,"C:\\Users\\ar2529\\Desktop\\opioid_causal_effect\\2022_out_of_town.csv", row.names = FALSE) 

print("Here's a more in-depth look at out-of-town ODs in specific towns:")

# initialize empty matrix, from https://stackoverflow.com/questions/32712301/create-empty-data-frame-with-column-names-by-assigning-a-string-vector

out_of_town_2022 <- data.frame(matrix(ncol = 4, nrow = 0))
cols <- c("Town name","Out-of-town ODs","Total ODs","Percent of out-of-town ODs")

# iteratively add info from these towns to the df, from https://www.projectpro.io/recipes/append-output-from-for-loop-dataframe-r

# change to unique(ocme_2022$injury_city_revised) to get all towns -- from https://stackoverflow.com/questions/45235422/does-r-have-a-set-data-structure

for (i in c("Danbury", "Waterbury", "Torrington", "New Haven", "West Haven", "Hartford", "Norwich", "New London", "Bridgeport", "New Britain", "Litchfield", "Meriden", "Norwalk", "Stamford", "Windham")) {
  
  # subset for the number of ODs in that town 
  
  ocme_injury_2022 <- subset(ocme_2022_noblank, injury_city_revised == i)
  
  # create a results vector to append to out_of_town 2020 df
  # contains town name, number of ODs from out of town residents, total ODs and fraction of out of town ODs in that   order
  
  results <- c(i, nrow(subset(ocme_injury_2022, Res.City != injury_city_revised)), nrow(ocme_injury_2022), round(nrow(subset(ocme_injury_2022, Res.City != injury_city_revised))/nrow(ocme_injury_2022), 3)*100)

  # add this to the out of town df 
  out_of_town_2022 = rbind(out_of_town_2022, results)
}

# for some reason it doesn't work if I do this before adding rows?
colnames(out_of_town_2022) <- cols

out_of_town_2022
```

### Injury location {.tabset}

#### 2018

```{r injury2018, echo=FALSE, message=FALSE, warning=FALSE}
ocme_2018$injury_location <- ifelse(ocme_2018$injury.location_revised %in% c("", "Unknown", "unclear"), "unknown",ifelse(ocme_2018$injury.location_revised == "other,indoors", "other, indoors", ifelse(ocme_2018$injury.location_revised %in% c("Other, Residential Institution", "other residence"), "residence", ifelse(ocme_2018$injury.location_revised == "other, outdoors", "outdoors", ocme_2018$injury.location_revised))))

residence_2018 <- ggplot(data=ocme_2018, aes(x=injury_location, ..count..)) + geom_bar(aes(fill = injury_location), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45), breaks = ocme_2018$injury_location, labels = ocme_2018$injury_location) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle("Opioid-detected overdose injury locations for 2018")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 7))

residence_2018

ggsave(plot=residence_2018, filename = "images/statewide/injury_collapsed_2018.png", width = 10, height = 8)

residence_2018_prop <- ggplot(data=ocme_2018, aes(x=injury_location, ..count..)) + geom_bar(aes(fill = injury_location), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45), breaks = ocme_2018$injury_location, labels = ocme_2018$injury_location) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme_2018), 3)), stat = "count", vjust = 0, colour = "black", size=3) + ggtitle("Opioid-detected overdose injury locations for 2018")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 10))

residence_2018_prop

ggsave(plot=residence_2018_prop, filename = "images/statewide/injury_collapsed_2018_prop.png", width = 10, height = 8)

```

#### 2019

```{r injuryRevised2019, echo=FALSE, message=FALSE, warning=FALSE}
ocme_2019$injury_collapsed <- ifelse(ocme_2019$injury.locn_revised == "", "unknown", ifelse(ocme_2019$injury.locn_revised == "other residence", "residence", ocme_2019$injury.locn_revised))

ocme_2019$injury_collapsed <- ifelse(ocme_2019$injury.locn_revised %in% c("other residence", "residence?"), "residence", ifelse(ocme_2019$injury.locn_revised == "other, outdoors", "outdoors", ifelse(ocme_2019$injury.locn_revised=="store", "store/shopping area", ifelse(ocme_2019$injury.locn_revised == "work site", "workplace", ifelse(ocme_2019$injury.locn_revised %in% c("", "unclear", "unspecified", "missing", "unidentified"), "unknown", ifelse(ocme_2019$injury.locn_revised == "other, indoors", "other indoors", ocme_2019$injury.locn_revised))))))


injury_collapsed_2019 <- ggplot(data=ocme_2019, aes(x=injury_collapsed, ..count..)) + geom_bar(aes(fill = injury_collapsed), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45), breaks = ocme_2019$injury_collapsed, labels = ocme_2019$injury_collapsed) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle("Opioid-detected overdose injury locations for 2019")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 10))

injury_collapsed_2019

ggsave(plot=injury_collapsed_2019, filename = "images/statewide/injury_collapsed_2019.png", width = 10, height = 8)

injury_collapsed_2019_prop <- ggplot(data=ocme_2019, aes(x=injury_collapsed, ..count..)) + geom_bar(aes(fill = injury_collapsed), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45), breaks = ocme_2019$injury_collapsed, labels = ocme_2019$injury_collapsed) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme_2019), 3)), stat = "count", vjust = 0, colour = "black") + ggtitle("Opioid-detected overdose injury locations for 2019")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 10))

injury_collapsed_2019_prop

ggsave(plot=injury_collapsed_2019_prop, filename = "images/statewide/injury_collapsed_2019_prop.png", width = 10, height = 8)

```


#### 2020


```{r injury2020, echo=FALSE, message=FALSE, warning=FALSE}
ocme_2020$injury_location <- ifelse(ocme_2020$injury.locn_revised == " residence ", "residence", ocme_2020$injury.locn_revised)

ocme_2020$injury_collapsed <- ifelse(ocme_2020$injury_location %in% c("apartment complex", "former residence", "other residence", "residence or drug spot"), "residence", ifelse(ocme_2020$injury_location %in% c("hospital ER", "hospital "), "hospital", ifelse(ocme_2020$injury_location == "other, indoors", "other indoors", ifelse(ocme_2020$injury_location == "unspecified", "unknown", ocme_2020$injury_location))))

injury_collapsed_2020 <- ggplot(data=ocme_2020, aes(x=injury_collapsed, ..count..)) + geom_bar(aes(fill = injury_collapsed), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45), breaks = ocme_2020$injury_collapsed, labels = ocme_2020$injury_collapsed) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle("Opioid-detected overdose injury locations for 2020")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 10))

injury_collapsed_2020

ggsave(plot=injury_collapsed_2020, filename = "images/statewide/injury_collapsed_2020.png", width = 10, height = 8)

injury_collapsed_2020_prop <- ggplot(data=ocme_2020, aes(x=injury_collapsed, ..count..)) + geom_bar(aes(fill = injury_collapsed), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45), breaks = ocme_2020$injury_collapsed, labels = ocme_2020$injury_collapsed) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme_2020), 3)), stat = "count", vjust = 0, colour = "black") + ggtitle("Opioid-detected overdose injury locations for 2020")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 10))

injury_collapsed_2020_prop

ggsave(plot=injury_collapsed_2020_prop, filename = "images/statewide/injury_collapsed_2020_prop.png", width = 10, height = 8)

```


#### 2021


```{r injuryRevised2021, echo=FALSE, message=FALSE, warning=FALSE}
ocme_2021$injury_location <- ifelse(ocme_2021$injury.locn_revised %in% c("residence ", "resdience"), "residence", ifelse(ocme_2021$injury.locn_revised == "other,indoors", "other, indoors", ifelse(ocme_2021$injury.locn_revised == "outdoors ", "outdoors", ifelse(ocme_2021$injury.locn_revised == "store/shoppping area", "store/shopping area", ifelse(ocme_2021$injury.locn_revised == "vehicle ", "vehicle", ocme_2021$injury.locn_revised)))))

ocme_2021$injury_collapsed <- ifelse(ocme_2021$injury_location %in% c("residence for p4m)", "house", "unclear residence", "other residence"), "residence", ifelse(ocme_2021$injury_location == "DOC transitional site", "halfway house", ifelse(ocme_2021$injury_location %in% c("outdoors?", "street"), "outdoors", ifelse(ocme_2021$injury_location == "vehicle", "vehicle/parking lot", ifelse(ocme_2021$injury_location %in% c("unclear", "unspecified"), "unknown", ifelse(ocme_2021$injury_location %in% c("indoors", "elementary school", "truck cabin"), "other, indoors", ocme_2021$injury_location))))))

injury_collapsed_2021 <- ggplot(data=ocme_2021, aes(x=injury_collapsed, ..count..)) + geom_bar(aes(fill = injury_collapsed), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45), breaks = ocme_2021$injury_collapsed, labels = ocme_2021$injury_collapsed) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle("Opioid-detected overdose injury locations for 2021")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 10))

injury_collapsed_2021

ggsave(plot=injury_collapsed_2021, filename = "images/statewide/injury_collapsed_2021.png", width = 10, height = 8)

injury_collapsed_2021_prop <- ggplot(data=ocme_2021, aes(x=injury_collapsed, ..count..)) + geom_bar(aes(fill = injury_collapsed), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45), breaks = ocme_2021$injury_collapsed, labels = ocme_2021$injury_collapsed) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme_2021), 3)), stat = "count", vjust = 0, colour = "black") + ggtitle("Opioid-detected overdose injury locations for 2021")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 10))

injury_collapsed_2021_prop

ggsave(plot=injury_collapsed_2021_prop, filename = "images/statewide/injury_collapsed_2021_prop.png", width = 10, height = 8)
```

#### 2022


```{r injuryRevised2022, echo=FALSE, message=FALSE, warning=FALSE}

ocme_2022$injury_collapsed <- ifelse(ocme_2022$injury.locn_revised %in% c("other residence", "residence?", "residence ", "residential building", "unknown residence", "home"), "residence", ifelse(ocme_2022$injury.locn_revised %in% c("other, indoors", "other, indoors ", "other building"), "other indoors", ifelse(ocme_2022$injury.locn_revised %in% c("boat", "camper IFO address"), "vehicle", ifelse(ocme_2022$injury.locn_revised %in% c("", "unknown", "unspecified", "unspecified place"), "unknown", ifelse(ocme_2022$injury.locn_revised == "workpkace", "workplace", ifelse(ocme_2022$injury.locn_revised =="other ", "other", ifelse(ocme_2022$injury.locn_revised %in% c("outdoors ", "street"), "outdoors", ifelse(ocme_2022$injury.locn_revised == "parking garage", "parking lot", ocme_2022$injury.locn_revised))))))))

injury_collapsed_2022 <- ggplot(data=ocme_2022, aes(x=injury_collapsed, ..count..)) + geom_bar(aes(fill = injury_collapsed), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45), breaks = ocme_2022$injury_collapsed, labels = ocme_2022$injury_collapsed) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle("Opioid-detected overdose injury locations for 2022")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 10))

injury_collapsed_2022

ggsave(plot=injury_collapsed_2022, filename = "images/statewide/injury_collapsed_2022.png", width = 10, height = 8)

injury_collapsed_2022_prop <- ggplot(data=ocme_2022, aes(x=injury_collapsed, ..count..)) + geom_bar(aes(fill = injury_collapsed), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45), breaks = ocme_2022$injury_collapsed, labels = ocme_2022$injury_collapsed) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme_2022), 3)), stat = "count", vjust = 0, colour = "black") + ggtitle("Opioid-detected overdose injury locations for 2022")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 10))

injury_collapsed_2022_prop

ggsave(plot=injury_collapsed_2022_prop, filename = "images/statewide/injury_collapsed_2022_prop.png", width = 10, height = 8)
```

### Are people dying in their own homes?  {.tabset}

#### 2018

```{r ownHome2018, echo=FALSE, message=FALSE, warning=FALSE}

sprintf("There were %d people who OD'd in their own residence.", sum(ocme_2018$injury_address_updated == ocme_2018$residence_address_updated))

sprintf("The proportion of decedents ODing in their own residence was %f.", sum(ocme_2018$injury_address_updated == ocme_2018$residence_address_updated)/nrow(ocme_2018))

sprintf("Out of everyone who OD'd in a residence, %f of people OD'd in their own residence.", sum(ocme_2018$injury_address_updated == ocme_2018$residence_address_updated)/sum(ocme_2018$injury_location=="residence"))

```

#### 2019

```{r ownHome2019, echo=FALSE, message=FALSE, warning=FALSE}

sprintf("There were %d people who OD'd in their own residence.", sum(ocme_2019$Injury.Address1 == ocme_2019$Residence.Address1))

sprintf("The proportion of decedents ODing in their own residence was %f.", sum(ocme_2019$Injury.Address1 == ocme_2019$Residence.Address1)/nrow(ocme_2019))

sprintf("Out of everyone who OD'd in a residence, %f of people OD'd in their own residence.", sum(ocme_2019$Injury.Address1 == ocme_2019$Residence.Address1)/sum(ocme_2019$injury_collapsed=="residence"))

```

#### 2020

```{r ownHome2020, echo=FALSE, message=FALSE, warning=FALSE}

sprintf("There were %d people who OD'd in their own residence.", sum(ocme_2020$Injury.Address1 == ocme_2020$Residence.Address1))

sprintf("The proportion of decedents ODing in their own residence was %f.", sum(ocme_2020$Injury.Address1 == ocme_2020$Residence.Address1)/nrow(ocme_2020))

sprintf("Out of everyone who OD'd in a residence, %f of people OD'd in their own residence.", sum(ocme_2020$Injury.Address1 == ocme_2020$Residence.Address1)/sum(ocme_2020$injury_collapsed=="residence"))

```

#### 2021

```{r ownHome2021, echo=FALSE, message=FALSE, warning=FALSE}

sprintf("There were %d people who OD'd in their own residence.", sum(ocme_2021$ResStreet.Name %in% c(ocme_2021$InjuryPost.Directional, ocme_2021$InjuryStreet.Name)))

sprintf("The proportion of decedents ODing in their own residence was %f.", sum(ocme_2021$ResStreet.Name %in% c(ocme_2021$InjuryPost.Directional, ocme_2021$InjuryStreet.Name))/nrow(ocme_2021))

sprintf("Out of everyone who OD'd in a residence, %f of people OD'd in their own residence.", sum(ocme_2021$ResStreet.Name %in% c(ocme_2021$InjuryPost.Directional, ocme_2021$InjuryStreet.Name))/sum(ocme_2021$injury_collapsed=="residence"))

```

#### 2022

```{r ownHome2022, echo=FALSE, message=FALSE, warning=FALSE}
sprintf("There were %d people who OD'd in their own residence.", sum(ocme_2022$Residential.Street.Address_REVISED == ocme_2022$Injury.Street.Address_REVISED))

sprintf("The proportion of decedents ODing in their own residence was %f.",  sum(ocme_2022$Residential.Street.Address_REVISED == ocme_2022$Injury.Street.Address_REVISED)/nrow(ocme_2022))

sprintf("Out of everyone who OD'd in a residence, %f of people OD'd in their own residence.",  sum(ocme_2022$Residential.Street.Address_REVISED == ocme_2022$Injury.Street.Address_REVISED)/sum(ocme_2022$injury_collapsed=="residence"))
```


### Most common substances {.tabset}

#### 2018
```{r drugsUsed2018, echo=FALSE, message=FALSE, warning=FALSE}
# got original idea from https://stackoverflow.com/questions/48562288/frequency-of-each-unique-combination-in-data-frame
# the n column wasn't in order, which i fixed with https://dplyr.tidyverse.org/reference/arrange.html
# arrange() provides ascending order so I used desc(n) [instead of just n] to fix that, from https://www.rdocumentation.org/packages/dplyr/versions/0.7.6/topics/desc

mostCommon2018 <- arrange(ocme_2018 %>% group_by(cocaine, combo..her.pharm.or.fent..OR.pharm.fent, heronly, pharmonly, fentonly, heroin, X6.mam, morphine, hermor_nocod, codeine, cod_w_no_hermor, di.H.codeine, hydromorphone, oxymorphone, hydrocodone, oxycodone, methadone, buprenorphine, fentanyl..4ANPP.too., Frankens, fent.or.frankens, tramadol, opioid.analogs..e.g...U47700., otherop, pharma_w_meth_no_fent.or.op.analogs.or.cod.or..other.op., pharma_nobupnometh, other, benzos, amphetamine, EtOH, THC) %>% tally, desc(n))

# show the top 5 combinations, from https://stackoverflow.com/questions/22471256/overriding-variables-not-shown-in-dplyr-to-display-all-columns-from-df

head(mostCommon2018, 5) %>% glimpse()

```

The top 5 substance combinations for 2018 are: 

1) fent only
2) cocaine, fent only 
3) fent only, weed
4) fent only, other
5) cocaine, fent only, other 


#### 2019

```{r drugsUsed2019, echo=FALSE, message=FALSE, warning=FALSE}
# got original idea from https://stackoverflow.com/questions/48562288/frequency-of-each-unique-combination-in-data-frame
# the n column wasn't in order, which i fixed with https://dplyr.tidyverse.org/reference/arrange.html
# arrange() provides ascending order so I used desc(n) [instead of just n] to fix that, from https://www.rdocumentation.org/packages/dplyr/versions/0.7.6/topics/desc

# turn the xylazine column into a binary variable--I treat any non-null character as a positive indication of xyl presence. One issue is the "*" characters -- not sure what to do with them.  

ocme_2019$xyla <- replace(ocme_2019$xyla, ocme_2019$xyla == "", 0)
ocme_2019$xyla <- replace(ocme_2019$xyla, !(ocme_2019$xyla == 0), 1)

# convert into integers since for some reason the 0 and 1 are characters
ocme_2019$xyla <- sapply(ocme_2019$xyla, as.integer)

mostCommon2019 <- arrange(ocme_2019 %>% group_by(xyla, combo, heronly, pharmonly, fentonly, Heroin, X6.mam , Morphine, hermor_nocod, Codeine, cod.w.no.hermor, di.H.codeine, Hydromorphone, oxymorphone, Hydrocodone, Oxycodone, Methadone, bup, fentanyl..4.ANPP.too., frankens, fent...frankens, tramadol, opioid.analogs..e.g...U47700., Other.Op, pharma..w..meth.bup_no.fent.or.op.analogs.or.cod.or..other.op.., pharma_nobupnometh, other, benzos, cocaine, amphetamine, EtOH, THC) %>% tally, desc(n))

# rename some vars for easier viewing, from https://stackoverflow.com/questions/7531868/how-to-rename-a-single-column-in-a-data-frame

names(mostCommon2019)[names(mostCommon2019) == 'pharma..w..meth.bup_no.fent.or.op.analogs.or.cod.or..other.op..'] <- 'pharma.w.meth.bup.no.fent.or.other.op'

# show the top 5 combinations
head(mostCommon2019, 5) %>% glimpse()
```

The top 5 substance combinations for 2019 are: 

1) fent only, cocaine
2) fent only 
3) fent only, other, cocaine
4) fent only, alcohol
5) fent only, other, cocaine, alcohol 

#### 2020

```{r drugsUsed2020, echo=FALSE, message=FALSE, warning=FALSE}
# got original idea from https://stackoverflow.com/questions/48562288/frequency-of-each-unique-combination-in-data-frame
# the n column wasn't in order, which i fixed with https://dplyr.tidyverse.org/reference/arrange.html
# arrange() provides ascending order so I used desc(n) [instead of just n] to fix that, from https://www.rdocumentation.org/packages/dplyr/versions/0.7.6/topics/desc

mostCommon2020 <- arrange(ocme_2020 %>% group_by(combo, heronly, pharmonly, fentonly, Heroin, X6.mam , Morphine, hermor_nocod, Codeine, cod.w.no.hermor, di.H.codeine, Hydromorphone, oxymorphone, Hydrocodone, Oxycodone, Methadone, bup, fentanyl..4.ANPP.too., frankens, fent...frankens, tramadol, opioid.analogs..e.g...U47700., Other.Op, pharma..w..meth.bup_no.fent.or.op.analogs.or.cod.or..other.op.., pharma_nobupnometh, other, benzos, cocaine, amphetamine, EtOH, THC) %>% tally, desc(n))

# rename some vars for easier viewing, from https://stackoverflow.com/questions/7531868/how-to-rename-a-single-column-in-a-data-frame

names(mostCommon2020)[names(mostCommon2020) == 'pharma..w..meth.bup_no.fent.or.op.analogs.or.cod.or..other.op..'] <- 'pharma.w.meth.bup.no.fent.or.other.op'

# show the top 5 combinations
head(mostCommon2020, 5) %>% glimpse()

```

The top 5 substance combinations for 2020 are: 

1) fent only, cocaine
2) fent only 
3) fent only, other, cocaine
4) fent only, other
5) fent only, alcohol

#### 2021

```{r drugsUsed2021, echo=FALSE, message=FALSE, warning=FALSE}
# got original idea from https://stackoverflow.com/questions/48562288/frequency-of-each-unique-combination-in-data-frame
# the n column wasn't in order, which i fixed with https://dplyr.tidyverse.org/reference/arrange.html
# arrange() provides ascending order so I used desc(n) [instead of just n] to fix that, from https://www.rdocumentation.org/packages/dplyr/versions/0.7.6/topics/desc

mostCommon2021 <- arrange(ocme_2021 %>% group_by(combo, her.only, pharm.only, fent.only, Heroin, X6.mam , Morphine, hermor_nocod, Codeine, cod.w.no.hermor, di.H.codeine, Hydromorphone, oxymorphone, Hydrocodone, Oxycodone, Methadone, bup, fentanyl..4.ANPP.despropionyl.fent.too., frankens, fent...frankens, tramadol, opioid.analogs..e.g...U47700..mitragynine.KRATOM..metonitazene...syr.and.cooker.etc.on.nightstand.in.photos...Me.prednisone, Other.Op, pharma..w..meth.bup_no.fent.or.op.analogs.or.cod.or..other.op.., pharma_nobupnometh, other, benzos, cocaine, amphetamine..including.eutylone., EtOH, THC) %>% tally, desc(n))

# rename some vars for easier viewing, from https://stackoverflow.com/questions/7531868/how-to-rename-a-single-column-in-a-data-frame
# for some reason I can't combine these statements?

names(mostCommon2021)[names(mostCommon2021) == 'opioid.analogs..e.g...U47700..mitragynine.KRATOM..metonitazene...syr.and.cooker.etc.on.nightstand.in.photos...Me.prednisone'] <- 'opioid.analogs.e.g.mitragynine'

names(mostCommon2021)[names(mostCommon2021) == 'pharma..w..meth.bup_no.fent.or.op.analogs.or.cod.or..other.op..'] <- 'pharma.w.meth.bup.no.fent.or.other.op'

# show the top 5 combinations
head(mostCommon2021, 5) %>% glimpse()

```

The top 5 substance combinations for 2021 are: 

1) fent only, other, cocaine
2) fent only, cocaine
3) fent only, other
4) fent only, cocaine, alcohol
5) fent only, other, cocaine, alcohol


#### 2022

```{r drugsUsed2022, echo=FALSE, message=FALSE, warning=FALSE}
# got original idea from https://stackoverflow.com/questions/48562288/frequency-of-each-unique-combination-in-data-frame
# the n column wasn't in order, which i fixed with https://dplyr.tidyverse.org/reference/arrange.html
# arrange() provides ascending order so I used desc(n) [instead of just n] to fix that, from https://www.rdocumentation.org/packages/dplyr/versions/0.7.6/topics/desc

mostCommon2022 <- arrange(ocme_2022 %>% group_by(combo, her.only, pharm.only, fent.only, Heroin, X6.mam , Morphine, hermor_nocod, Codeine, cod.w.no.hermor, di.H.codeine, Hydro.morphone, oxy.morphone, Hydro.codone, Oxy.codone, Methadone, bup, fentanyl..4.ANPP.despropionyl.fent.too., frankens, fent...frankens, tramadol, opioid.analogs..e.g...U47700..mitragynine.KRATOM..metonitazene...syr.and.cooker.etc.on.nightstand.in.photos...Me.prednisone, Other.Op, pharma..w..meth.bup_no.fent.or.op.analogs.or.cod.or..other.op.., pharma_nobupnometh, other, benzos, cocaine, amphetamine.other.stimulants..including.eutylone..MDMA..N.N.diMe.pentylone., EtOH, THC) %>% tally, desc(n))

# rename some vars for easier viewing, from https://stackoverflow.com/questions/7531868/how-to-rename-a-single-column-in-a-data-frame

names(mostCommon2022)[names(mostCommon2022) == 'opioid.analogs..e.g...U47700..mitragynine.KRATOM..metonitazene...syr.and.cooker.etc.on.nightstand.in.photos...Me.prednisone'] <- 'opioid.analogs.e.g.mitragynine'

names(mostCommon2022)[names(mostCommon2022) == 'pharma..w..meth.bup_no.fent.or.op.analogs.or.cod.or..other.op..'] <- 'pharma.w.meth.bup.no.fent.or.other.op'

names(mostCommon2022)[names(mostCommon2022) == 'amphetamine.other.stimulants..including.eutylone..MDMA..N.N.diMe.pentylone.'] <- 'amphetamine.including.eutylone.MDMA'


# show the top 5 combinations
head(mostCommon2022, 5) %>% glimpse()

```

The top 5 substance combinations for 2022 are: 

1) fent only, other, cocaine
2) fent only, cocaine
3) fent only, other
4) fent only, cocaine, alcohol
5) fent only, other, cocaine, alcohol

### How many people were using chronically? {.tabset}

I identify chronic users crudely -- I look for entries where the words "chronic" [can sometimes indicate chronic pain, meaning opioid scripts], "Chronic", "user", "drug use", "drug user", "drug abuse", "snort", "snorts", "drug abuse", "addict", "addicted", "rehab", "sober house", "clean" or "abuse" pops up in either the notes field or immediate cause of death, because that generally means that the person had chronic drug use [based on my look at the data]

#### 2018

```{r chronicUse2018, echo=FALSE, message=FALSE, warning=FALSE}
# find entries in either "MY.NOTES" or "AmendedImmediateCauseofDeath" with the words "chronic", "drug user", "drug use", "user", "snort", "shorts" or "drug abuse"
# I tried doing this as grep(c("chronic", "drug user", ...)) but that literally didn't work so I'm stuck with this ugliness

chronic2018 <- subset(ocme_2018, grepl("chronic", ocme_2018$MY.NOTES, fixed=TRUE) | grepl( "Chronic", ocme_2018$MY.NOTES, fixed=TRUE) | grepl( "drug user", ocme_2018$MY.NOTES, fixed=TRUE) | grepl( "user", ocme_2018$MY.NOTES, fixed=TRUE) | grepl( "snort", ocme_2018$MY.NOTES, fixed=TRUE) | grepl( "snorts", ocme_2018$MY.NOTES, fixed=TRUE) | grepl( "drug abuse", ocme_2018$MY.NOTES, fixed=TRUE) | grepl( "addict", ocme_2018$MY.NOTES, fixed=TRUE) | grepl( "addicted", ocme_2018$MY.NOTES, fixed=TRUE) | grepl( "rehab", ocme_2018$MY.NOTES, fixed=TRUE) | grepl( "sober house ", ocme_2018$MY.NOTES, fixed=TRUE) | grepl( "clean", ocme_2018$MY.NOTES, fixed=TRUE) | grepl( "abuse", ocme_2018$MY.NOTES, fixed=TRUE) | grepl("chronic", ocme_2018$AmendedImmediateCauseofDeath, fixed=TRUE) | grepl( "Chronic", ocme_2018$AmendedImmediateCauseofDeath, fixed=TRUE) | grepl( "drug user", ocme_2018$AmendedImmediateCauseofDeath, fixed=TRUE) | grepl( "user", ocme_2018$AmendedImmediateCauseofDeath, fixed=TRUE) | grepl( "snort", ocme_2018$AmendedImmediateCauseofDeath, fixed=TRUE) | grepl( "snorts", ocme_2018$AmendedImmediateCauseofDeath, fixed=TRUE) | grepl( "drug abuse", ocme_2018$AmendedImmediateCauseofDeath, fixed=TRUE) | grepl( "addict", ocme_2018$AmendedImmediateCauseofDeath, fixed=TRUE) | grepl( "addicted", ocme_2018$AmendedImmediateCauseofDeath, fixed=TRUE) | grepl( "rehab", ocme_2018$AmendedImmediateCauseofDeath, fixed=TRUE) | grepl( "sober house ", ocme_2018$AmendedImmediateCauseofDeath, fixed=TRUE) | grepl( "clean", ocme_2018$AmendedImmediateCauseofDeath, fixed=TRUE) | grepl( "abuse", ocme_2018$AmendedImmediateCauseofDeath, fixed=TRUE))

sprintf("There were at least %d people with known chronic use out of %d decedents in 2018, which is %f percent of all decedents.", nrow(chronic2018), nrow(ocme_2018), (nrow(chronic2018)/nrow(ocme_2018))*100)

```


#### 2019

```{r chronicUse2019, echo=FALSE, message=FALSE, warning=FALSE}
# find entries in either "MY.NOTES" or "AmendedImmediateCauseofDeath" with the words "chronic", "drug user", "drug use", "user", "snort", "shorts" or "drug abuse"

chronic2019 <- subset(ocme_2019, grepl("chronic", ocme_2019$MY.NOTES, fixed=TRUE) | grepl( "Chronic", ocme_2019$MY.NOTES, fixed=TRUE) | grepl( "drug user", ocme_2019$MY.NOTES, fixed=TRUE) | grepl( "user", ocme_2019$MY.NOTES, fixed=TRUE) | grepl( "snort", ocme_2019$MY.NOTES, fixed=TRUE) | grepl( "snorts", ocme_2019$MY.NOTES, fixed=TRUE) | grepl( "drug abuse", ocme_2019$MY.NOTES, fixed=TRUE) | grepl( "addict", ocme_2019$MY.NOTES, fixed=TRUE) | grepl( "addicted", ocme_2019$MY.NOTES, fixed=TRUE) | grepl( "rehab", ocme_2019$MY.NOTES, fixed=TRUE) | grepl( "sober house ", ocme_2019$MY.NOTES, fixed=TRUE) | grepl( "clean", ocme_2019$MY.NOTES, fixed=TRUE) | grepl( "abuse", ocme_2019$MY.NOTES, fixed=TRUE) | grepl("chronic", ocme_2019$Amended.Immediate.Cause.of.Death, fixed=TRUE) | grepl( "Chronic", ocme_2019$Amended.Immediate.Cause.of.Death, fixed=TRUE) | grepl( "drug user", ocme_2019$Amended.Immediate.Cause.of.Death, fixed=TRUE) | grepl( "user", ocme_2019$Amended.Immediate.Cause.of.Death, fixed=TRUE) | grepl( "snort", ocme_2019$Amended.Immediate.Cause.of.Death, fixed=TRUE) | grepl( "snorts", ocme_2019$Amended.Immediate.Cause.of.Death, fixed=TRUE) | grepl( "drug abuse", ocme_2019$Amended.Immediate.Cause.of.Death, fixed=TRUE) | grepl( "addict", ocme_2019$Amended.Immediate.Cause.of.Death, fixed=TRUE) | grepl( "addicted", ocme_2019$Amended.Immediate.Cause.of.Death, fixed=TRUE) | grepl( "rehab", ocme_2019$Amended.Immediate.Cause.of.Death, fixed=TRUE) | grepl( "sober house ", ocme_2019$Amended.Immediate.Cause.of.Death, fixed=TRUE) | grepl( "clean", ocme_2019$Amended.Immediate.Cause.of.Death, fixed=TRUE) | grepl( "abuse", ocme_2019$Amended.Immediate.Cause.of.Death, fixed=TRUE))

sprintf("There were at least %d people with known chronic use out of %d decedents in 2019, which is %f percent of all decedents.", nrow(chronic2019), nrow(ocme_2019), (nrow(chronic2019)/nrow(ocme_2019))*100)


```
#### 2020

```{r chronicUse2020, echo=FALSE, message=FALSE, warning=FALSE}
# find entries in either "MY.NOTES" or "AmendedImmediateCauseofDeath" with the words "chronic", "drug user", "drug use", "user", "snort", "shorts" or "drug abuse"

chronic2020 <- subset(ocme_2020, grepl("chronic", ocme_2020$MY.NOTES, fixed=TRUE) | grepl( "Chronic", ocme_2020$MY.NOTES, fixed=TRUE) | grepl( "drug user", ocme_2020$MY.NOTES, fixed=TRUE) | grepl( "user", ocme_2020$MY.NOTES, fixed=TRUE) | grepl( "snort", ocme_2020$MY.NOTES, fixed=TRUE) | grepl( "snorts", ocme_2020$MY.NOTES, fixed=TRUE) | grepl( "drug abuse", ocme_2020$MY.NOTES, fixed=TRUE) | grepl( "addict", ocme_2020$MY.NOTES, fixed=TRUE) | grepl( "addicted", ocme_2020$MY.NOTES, fixed=TRUE) | grepl( "rehab", ocme_2020$MY.NOTES, fixed=TRUE) | grepl( "sober house ", ocme_2020$MY.NOTES, fixed=TRUE) | grepl( "clean", ocme_2020$MY.NOTES, fixed=TRUE) | grepl( "abuse", ocme_2020$MY.NOTES, fixed=TRUE) | grepl("chronic", ocme_2020$Ammended.Immediate.Cause.of.Death, fixed=TRUE) | grepl( "Chronic", ocme_2020$Ammended.Immediate.Cause.of.Death, fixed=TRUE) | grepl( "drug user", ocme_2020$Ammended.Immediate.Cause.of.Death, fixed=TRUE) | grepl( "user", ocme_2020$Ammended.Immediate.Cause.of.Death, fixed=TRUE) | grepl( "snort", ocme_2020$Ammended.Immediate.Cause.of.Death, fixed=TRUE) | grepl( "snorts", ocme_2020$Ammended.Immediate.Cause.of.Death, fixed=TRUE) | grepl( "drug abuse", ocme_2020$Ammended.Immediate.Cause.of.Death, fixed=TRUE) | grepl( "addict", ocme_2020$Ammended.Immediate.Cause.of.Death, fixed=TRUE) | grepl( "addicted", ocme_2020$Ammended.Immediate.Cause.of.Death, fixed=TRUE) | grepl( "rehab", ocme_2020$Ammended.Immediate.Cause.of.Death, fixed=TRUE) | grepl( "sober house ", ocme_2020$Ammended.Immediate.Cause.of.Death, fixed=TRUE) | grepl( "clean", ocme_2020$Ammended.Immediate.Cause.of.Death, fixed=TRUE) | grepl( "abuse", ocme_2020$Ammended.Immediate.Cause.of.Death, fixed=TRUE))

sprintf("There were at least %d people with known chronic use out of %d decedents in 2020, which is %f percent of all decedents.", nrow(chronic2020), nrow(ocme_2020), (nrow(chronic2020)/nrow(ocme_2020))*100)


```

#### 2021

```{r chronicUse2021, echo=FALSE, message=FALSE, warning=FALSE}
# find entries in either "MY.NOTES" or "AmendedImmediateCauseofDeath" with the words "chronic", "drug user", "drug use", "user", "snort", "shorts" or "drug abuse"

chronic2021 <- subset(ocme_2021, grepl("chronic", ocme_2021$MY.NOTES.1, fixed=TRUE) | grepl( "Chronic", ocme_2021$MY.NOTES.1, fixed=TRUE) | grepl( "drug user", ocme_2021$MY.NOTES.1, fixed=TRUE) | grepl( "user", ocme_2021$MY.NOTES.1, fixed=TRUE) | grepl( "snort", ocme_2021$MY.NOTES.1, fixed=TRUE) | grepl( "snorts", ocme_2021$MY.NOTES.1, fixed=TRUE) | grepl( "drug abuse", ocme_2021$MY.NOTES.1, fixed=TRUE) | grepl( "addict", ocme_2021$MY.NOTES.1, fixed=TRUE) | grepl( "addicted", ocme_2021$MY.NOTES.1, fixed=TRUE) | grepl( "rehab", ocme_2021$MY.NOTES.1, fixed=TRUE) | grepl( "sober house ", ocme_2021$MY.NOTES.1, fixed=TRUE) | grepl( "clean", ocme_2021$MY.NOTES.1, fixed=TRUE) | grepl( "abuse", ocme_2021$MY.NOTES.1, fixed=TRUE) | grepl("chronic", ocme_2021$CauseA, fixed=TRUE) | grepl( "Chronic", ocme_2021$CauseA, fixed=TRUE) | grepl( "drug user", ocme_2021$CauseA, fixed=TRUE) | grepl( "user", ocme_2021$CauseA, fixed=TRUE) | grepl( "snort", ocme_2021$CauseA, fixed=TRUE) | grepl( "snorts", ocme_2021$CauseA, fixed=TRUE) | grepl( "drug abuse", ocme_2021$CauseA, fixed=TRUE) | grepl( "addict", ocme_2021$CauseA, fixed=TRUE) | grepl( "addicted", ocme_2021$CauseA, fixed=TRUE) | grepl( "rehab", ocme_2021$CauseA, fixed=TRUE) | grepl( "sober house ", ocme_2021$CauseA, fixed=TRUE) | grepl( "clean", ocme_2021$CauseA, fixed=TRUE) | grepl( "abuse", ocme_2021$CauseA, fixed=TRUE))

sprintf("There were at least %d people with known chronic use out of %d decedents in 2021, which is %f percent of all decedents.", nrow(chronic2021), nrow(ocme_2021), (nrow(chronic2021)/nrow(ocme_2021))*100)


```

#### 2022

```{r chronicUse2022, echo=FALSE, message=FALSE, warning=FALSE}
# find entries in either "MY.NOTES" or "AmendedImmediateCauseofDeath" with the words "chronic", "drug user", "drug use", "user", "snort", "shorts" or "drug abuse"

chronic2022 <- subset(ocme_2022, grepl("chronic", ocme_2022$MY.NOTES, fixed=TRUE) | grepl( "Chronic", ocme_2022$MY.NOTES, fixed=TRUE) | grepl( "drug user", ocme_2022$MY.NOTES, fixed=TRUE) | grepl( "user", ocme_2022$MY.NOTES, fixed=TRUE) | grepl( "snort", ocme_2022$MY.NOTES, fixed=TRUE) | grepl( "snorts", ocme_2022$MY.NOTES, fixed=TRUE) | grepl( "drug abuse", ocme_2022$MY.NOTES, fixed=TRUE) | grepl( "addict", ocme_2022$MY.NOTES, fixed=TRUE) | grepl( "addicted", ocme_2022$MY.NOTES, fixed=TRUE) | grepl( "rehab", ocme_2022$MY.NOTES, fixed=TRUE) | grepl( "sober house ", ocme_2022$MY.NOTES, fixed=TRUE) | grepl( "clean", ocme_2022$MY.NOTES, fixed=TRUE) | grepl( "abuse", ocme_2022$MY.NOTES, fixed=TRUE) | grepl("chronic", ocme_2022$Cause.A, fixed=TRUE) | grepl( "Chronic", ocme_2022$Cause.A, fixed=TRUE) | grepl( "drug user", ocme_2022$Cause.A, fixed=TRUE) | grepl( "user", ocme_2022$Cause.A, fixed=TRUE) | grepl( "snort", ocme_2022$Cause.A, fixed=TRUE) | grepl( "snorts", ocme_2022$Cause.A, fixed=TRUE) | grepl( "drug abuse", ocme_2022$Cause.A, fixed=TRUE) | grepl( "addict", ocme_2022$Cause.A, fixed=TRUE) | grepl( "addicted", ocme_2022$Cause.A, fixed=TRUE) | grepl( "rehab", ocme_2022$Cause.A, fixed=TRUE) | grepl( "sober house ", ocme_2022$Cause.A, fixed=TRUE) | grepl( "clean", ocme_2022$Cause.A, fixed=TRUE) | grepl( "abuse", ocme_2022$Cause.A, fixed=TRUE))

sprintf("There were at least %d people with known chronic use out of %d decedents in 2022, which is %f percent of all decedents.", nrow(chronic2022), nrow(ocme_2022), (nrow(chronic2022)/nrow(ocme_2022))*100)


```

### Who did decedents live with? {.tabset} 

Please note: this field and all fields explored below are only available up to 2020. 

#### 2018

```{r liveAlone2018, echo=FALSE, message=FALSE, warning=FALSE}
#subset all cases where people died in their own house
residence2018 <- subset(ocme_2018, residence_address_updated == injury_address_updated)

#table(residence2018$OTHERS...INJURY.LOCATION..A..alone.F..fam.ptner.N.nonfamM.missing.N.A.)

residence2018$roommates <- ifelse(residence2018$OTHERS...INJURY.LOCATION..A..alone.F..fam.ptner.N.nonfamM.missing.N.A. %in% c("", "777", "999", "M"), "Unknown", ifelse(residence2018$OTHERS...INJURY.LOCATION..A..alone.F..fam.ptner.N.nonfamM.missing.N.A. %in% c("A", "A?"), "Alone", ifelse(residence2018$OTHERS...INJURY.LOCATION..A..alone.F..fam.ptner.N.nonfamM.missing.N.A. == "F", "Family/partner", ifelse(residence2018$OTHERS...INJURY.LOCATION..A..alone.F..fam.ptner.N.nonfamM.missing.N.A. == "N", "Non-family", ifelse(residence2018$OTHERS...INJURY.LOCATION..A..alone.F..fam.ptner.N.nonfamM.missing.N.A. %in% c("F/N", "N/F"), "Family and non-family", residence2018$OTHERS...INJURY.LOCATION..A..alone.F..fam.ptner.N.nonfamM.missing.N.A.)))))

liveAlone_2018 <-ggplot(data=residence2018, aes(x=roommates, ..count..)) + 
      geom_bar(aes(fill = roommates), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Living arrangements for 2018 opioid-detected deaths in own residence")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none")

liveAlone_2018

ggsave(plot=liveAlone_2018, filename = "images/statewide/live_alone_2018.png", width = 10, height = 8)

liveAlone_2018_prop <-ggplot(data=residence2018, aes(x=roommates, ..count..)) + 
      geom_bar(aes(fill = roommates), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(residence2018),3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Living arrangements for 2018 opioid-detected deaths in own residence")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none")

liveAlone_2018_prop

ggsave(plot=liveAlone_2018_prop, filename = "images/statewide/live_alone_2018_prop.png", width = 10, height = 8)

```

#### 2019

```{r liveAlone2019, echo=FALSE, message=FALSE, warning=FALSE}
#subset all cases where people died in their own house
residence2019 <- subset(ocme_2019, Injury.Address1 == Residence.Address1)

#table(residence2019$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfamM.missing.N.A.)

residence2019$roommates <- ifelse(residence2019$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfamM.missing.N.A. %in% c("lived w str & her family", "F?", "F", "TOC 59; pronounced 119"), "Family/partner", ifelse(residence2019$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfamM.missing.N.A. %in% c("unclear", "unspecified", "", "2", "999"), "Unknown", ifelse(residence2019$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfamM.missing.N.A. == "A", "Alone", ifelse(residence2019$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfamM.missing.N.A. == "M", "Unknown", ifelse(residence2019$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfamM.missing.N.A. %in% c("N"), "Non-family", ifelse(residence2019$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfamM.missing.N.A. == "F,N", "Family and non-family", residence2019$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfamM.missing.N.A.))))))

liveAlone_2019 <-ggplot(data=residence2019, aes(x=roommates, ..count..)) + 
      geom_bar(aes(fill = roommates), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Living arrangements for 2019 opioid-detected deaths in own residence")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none")

liveAlone_2019

ggsave(plot=liveAlone_2019, filename = "images/statewide/live_alone_2019.png", width = 10, height = 8)

liveAlone_2019_prop <-ggplot(data=residence2019, aes(x=roommates, ..count..)) + 
      geom_bar(aes(fill = roommates), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(residence2019),3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Living arrangements for 2019 opioid-detected deaths in own residence")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none")

liveAlone_2019_prop

ggsave(plot=liveAlone_2019_prop, filename = "images/statewide/live_alone_2019_prop.png", width = 10, height = 8)

```

#### 2020

```{r liveAlone2020, echo=FALSE, message=FALSE, warning=FALSE}
#subset all cases where people died in their own house
residence2020 <- subset(ocme_2020, Injury.Address1 == Residence.Address1)

#table(residence2020$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfam.M.missing.N.A.)

residence2020$roommates <- ifelse(residence2020$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfam.M.missing.N.A. %in% c(" F ", "F?", "F"), "Family/partner", ifelse(residence2020$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfam.M.missing.N.A. %in% c("unclear", "unspecified", "", "2"), "Unknown", ifelse(residence2020$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfam.M.missing.N.A. == "A", "Alone", ifelse(residence2020$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfam.M.missing.N.A. == "M", "Unknown", ifelse(residence2020$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfam.M.missing.N.A. %in% c("N", "N "), "Non-family", ifelse(residence2020$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfam.M.missing.N.A. == "F/N", "Family and non-family", residence2020$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfam.M.missing.N.A.))))))

liveAlone_2020 <-ggplot(data=residence2020, aes(x=roommates, ..count..)) + 
      geom_bar(aes(fill = roommates), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Living arrangements for 2020 opioid-detected deaths in own residence")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none")

liveAlone_2020

ggsave(plot=liveAlone_2020, filename = "images/statewide/live_alone_2020.png", width = 10, height = 8)

liveAlone_2020_prop <-ggplot(data=residence2020, aes(x=roommates, ..count..)) + 
      geom_bar(aes(fill = roommates), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(residence2020),3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Living arrangements for 2020 opioid-detected deaths in own residence")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none")

liveAlone_2020_prop

ggsave(plot=liveAlone_2020_prop, filename = "images/statewide/live_alone_2020_prop.png", width = 10, height = 8)

```


### % cases found in <24h and >=24h {.tabset}

#### 2018


```{r within24h2018, echo=FALSE, message=FALSE, warning=FALSE}
#remove missing/unknown values -- the "1814" entry also seems to be missing values
ocme_times_2018 <- subset(ocme_2018, !(Difference..CALCULATED..Number.for..24h.based.on.either.pronounce.time.or.info.in.narrative..whichever.is.least..ROUND.TO.NEAREST.30.MIN..TIME.ELAPSED.AS.NUMBER..min..h..d..w.mo. %in% c("777", "999", "1814", "unclear", "unknown")))

#it's easier to find items indicating >24h than <24h, so I'll just use that to subset for simplicity

within24h2018 <- subset(ocme_times_2018, !(Difference..CALCULATED..Number.for..24h.based.on.either.pronounce.time.or.info.in.narrative..whichever.is.least..ROUND.TO.NEAREST.30.MIN..TIME.ELAPSED.AS.NUMBER..min..h..d..w.mo. %in% c("27.5", "~1d", "~1.5d", "18h", "~1 day", "~2 days", "~2wks", "~7 months", "< 1 week", "<2d", "> 1 week", "> 24h", ">1d (44h)", ">24 h (days)", ">24h", ">24h (25.5h)", ">2d", "1 week", "1 wk prior", "1-2d", "1w", "2 days", "17d", "2 days, lka 1700 3/9/2018", "2 or 3d", "2 weeks", "2.5d", "2-3 days", "2-4 days", "25.5h", "2-5d?", "25h", "26.5h", "26h", "28h", "29.5h", "29h", "2d", "2d prior", "2wk", "2wks", "3 days", "3.5d", "3-21d prior", "32h", "3-4 days", "3-4d", "3d", "3d prior", "4 days", "4.5d", "4d", "4d prior", "5d", "6d", "8 days", "days", "likely >24hr from walking into bookstore to disc")))

over24h2018 <- subset(ocme_times_2018, Difference..CALCULATED..Number.for..24h.based.on.either.pronounce.time.or.info.in.narrative..whichever.is.least..ROUND.TO.NEAREST.30.MIN..TIME.ELAPSED.AS.NUMBER..min..h..d..w.mo. %in% c("27.5", "~1d", "~1.5d", "18h", "~1 day", "~2 days", "~2wks", "~7 months", "< 1 week", "<2d", "> 1 week", "> 24h", ">1d (44h)", ">24 h (days)", ">24h", ">24h (25.5h)", ">2d", "1 week", "1 wk prior", "1-2d", "1w", "2 days", "17d", "2 days, lka 1700 3/9/2018", "2 or 3d", "2 weeks", "2.5d", "2-3 days", "2-4 days", "25.5h", "2-5d?", "25h", "26.5h", "26h", "28h", "29.5h", "29h", "2d", "2d prior", "2wk", "2wks", "3 days", "3.5d", "3-21d prior", "32h", "3-4 days", "3-4d", "3d", "3d prior", "4 days", "4.5d", "4d", "4d prior", "5d", "6d", "8 days", "days", "likely >24hr from walking into bookstore to disc"))


sprintf("There were %d people found within 24 hours, which is %f of all decedents.", nrow(within24h2018), nrow(within24h2018)/nrow(ocme_2018))

sprintf("There were %d people found in over 24 hours, which is %f of all decedents.", nrow(over24h2018), nrow(over24h2018)/nrow(ocme_2018))

```

#### 2019

```{r within24h2019, echo=FALSE, message=FALSE, warning=FALSE}
#remove missing/unknown values -- the "<1h but decomp" entry seems like someone didn't realize the person was dead when they reported them as "last alive" because decomposition will work slower than that, even in the summer
ocme_times_2019 <- subset(ocme_2019, !(Difference..CALCULATED..Number.for..24h.based.on.either.pronounce.time.or.info.in.narrative..whichever.is.least..ROUND.TO.NEAREST.30.MIN..TIME.ELAPSED.AS.NUMBER..min..h..d..w.mo. %in% c("777", "999", "??", "unclear", "unknown", "<1h (but decomp)", "")))

#it's easier to find items indicating >24h than <24h, so I'll just use that to subset for simplicity

#to make my life easier this time i subsetted everything using filter on excel, and then I used https://www.commontools.org/tool/replace-new-lines-with-commas-40 to turn into a comma separated list, then I put quotes around it using https://onlinetools.com/list/quote-list-items. While I'm aware I could've done this with regexes I didn't feel like it lol

within24h2019 <- subset(ocme_times_2019, !(Difference..CALCULATED..Number.for..24h.based.on.either.pronounce.time.or.info.in.narrative..whichever.is.least..ROUND.TO.NEAREST.30.MIN..TIME.ELAPSED.AS.NUMBER..min..h..d..w.mo. %in% unique(c("38.5h", "2d", "2d", "4d", "almost 2d prior", "2d prior", "25h", "3d prior", ">1 mo", ">1d", "5d", "3d", "3d", "1wk", "4d", "10d", "4d", "2d prior", "2d prior", "4d", "6d", "4d", "3d", "days", "10d prior", "25.5h", "3d", "3d", "26h", "2 weeks", "3d prior", "7d prior", "3d", "2d prior", "6d prior", "3d", "2d prior", "2d prior", "2d", "3d", "5d", "25.5h", "2wks prior", "25.5h", "several days", "8d", "23d", "2d prior", "2d prior", "~26h", "9d prior", "few days ago", "<4d", "26h", "4d", "36h", ">36h", "4d", "25h", "8d", "30h", "long time", "5d", "3d", "8d", "3d prior", "3d", "3d", "3d", "2d prior", "2d prior", "2d prior", "3d prior", "~2d prior", "4d prior", "few days ago", "3d prior", "3d prior", "few days ago", "2d prior", "2d prior", "several days", "3d prior", "3d", ">24h", "2d prior", ">2d", "4d prior", "21 or 46h", "3d prior", "2d prior", "3.5d", "3d prior", "28.5h", "3d prior", "3d prior", "2d prior", "2d prior", "5d priot", "5d prior", "7d", "4d prior", "26.5h", "2wks prior", "6d prior", "3d prior", "3d prior", ">24h", "days?", "3d prior", "2d prior", "26h", "3d prior", "2d prior", "2d prior", "5-6d prior", "3d prior", "5d prior", "2d?", "31h", "3d prior", "2d prior", "3d prior", "2d prior", "2d prior", "2d prior", "3d prior", "2d prior", "2d prior", "~28h", "several days", ">24h", "8d", "2d prior", "2d prior", "3d prior", "3d prior", "2d prior", "4d prior", "25h", "6d prior", "2d prior", "2wks prior", "2d prior", "2d prior", "32.5h", "2d prior", "5d", "2d prior", "5d prior", "4d prior", "37h", "2d prior", "2d prior", "2d prior", "2d prior", "3d prior", "3d prior", "2d", "3d", "2d prior", "6d prior", "2d prior", "4d", "24.5h", "3d prior", "2d prior", "5d prior", "2-3 wks", "3d prior", "4d prior", "2d prior", "3d prior", "6d prior", "3d prior", "2d prior", "2d prior", "25h", "~1.5d", ">24h", "several days", "2d prior", "1 month", "32.5h", "9d prior", "3d prior", "4d", "3d prior", "2d prior", "30h", "2d prior", "2 wks", "3d prior", "2d prior", "2d prior", "2d prior", "4d", ""))))

over24h2019 <- subset(ocme_times_2019, Difference..CALCULATED..Number.for..24h.based.on.either.pronounce.time.or.info.in.narrative..whichever.is.least..ROUND.TO.NEAREST.30.MIN..TIME.ELAPSED.AS.NUMBER..min..h..d..w.mo. %in% unique(c("38.5h", "2d", "2d", "4d", "almost 2d prior", "2d prior", "25h", "3d prior", ">1 mo", ">1d", "5d", "3d", "3d", "1wk", "4d", "10d", "4d", "2d prior", "2d prior", "4d", "6d", "4d", "3d", "days", "10d prior", "25.5h", "3d", "3d", "26h", "2 weeks", "3d prior", "7d prior", "3d", "2d prior", "6d prior", "3d", "2d prior", "2d prior", "2d", "3d", "5d", "25.5h", "2wks prior", "25.5h", "several days", "8d", "23d", "2d prior", "2d prior", "~26h", "9d prior", "few days ago", "<4d", "26h", "4d", "36h", ">36h", "4d", "25h", "8d", "30h", "long time", "5d", "3d", "8d", "3d prior", "3d", "3d", "3d", "2d prior", "2d prior", "2d prior", "3d prior", "~2d prior", "4d prior", "few days ago", "3d prior", "3d prior", "few days ago", "2d prior", "2d prior", "several days", "3d prior", "3d", ">24h", "2d prior", ">2d", "4d prior", "21 or 46h", "3d prior", "2d prior", "3.5d", "3d prior", "28.5h", "3d prior", "3d prior", "2d prior", "2d prior", "5d priot", "5d prior", "7d", "4d prior", "26.5h", "2wks prior", "6d prior", "3d prior", "3d prior", ">24h", "days?", "3d prior", "2d prior", "26h", "3d prior", "2d prior", "2d prior", "5-6d prior", "3d prior", "5d prior", "2d?", "31h", "3d prior", "2d prior", "3d prior", "2d prior", "2d prior", "2d prior", "3d prior", "2d prior", "2d prior", "~28h", "several days", ">24h", "8d", "2d prior", "2d prior", "3d prior", "3d prior", "2d prior", "4d prior", "25h", "6d prior", "2d prior", "2wks prior", "2d prior", "2d prior", "32.5h", "2d prior", "5d", "2d prior", "5d prior", "4d prior", "37h", "2d prior", "2d prior", "2d prior", "2d prior", "3d prior", "3d prior", "2d", "3d", "2d prior", "6d prior", "2d prior", "4d", "24.5h", "3d prior", "2d prior", "5d prior", "2-3 wks", "3d prior", "4d prior", "2d prior", "3d prior", "6d prior", "3d prior", "2d prior", "2d prior", "25h", "~1.5d", ">24h", "several days", "2d prior", "1 month", "32.5h", "9d prior", "3d prior", "4d", "3d prior", "2d prior", "30h", "2d prior", "2 wks", "3d prior", "2d prior", "2d prior", "2d prior", "4d", "")))

sprintf("There were %d people found within 24 hours, which is %f of all decedents.", nrow(within24h2019), nrow(within24h2019)/nrow(ocme_2019))

sprintf("There were %d people found in over 24 hours, which is %f of all decedents.", nrow(over24h2019), nrow(over24h2019)/nrow(ocme_2019))

```

#### 2020

```{r within24h2020, echo=FALSE, message=FALSE, warning=FALSE}
ocme_times_2020 <- subset(ocme_2020, !(Difference..CALCULATED..Number.for..24h.based.on.either.pronounce.time.or.info.in.narrative..whichever.is.least..ROUND.TO.NEAREST.30.MIN..TIME.ELAPSED.AS.NUMBER..min..h..d..w.mo. %in% c("N/A", "unclear", "unknown", "")))


#it's easier to find items indicating >24h than <24h, so I'll just use that to subset for simplicity

within24h2020 <- subset(ocme_times_2020, !(Difference..CALCULATED..Number.for..24h.based.on.either.pronounce.time.or.info.in.narrative..whichever.is.least..ROUND.TO.NEAREST.30.MIN..TIME.ELAPSED.AS.NUMBER..min..h..d..w.mo. %in% c(" 2d ", "\"a few days\"", "~1 day", "~24h", "<2d", ">24h", ">3d", ">5d", "1-2 weeks ago", "1 week", "1 week ", "1.5d", "10d", "10d prior", "11d", "12 d", "12d", "15d prior", "19d", "1d", "1wk", "2-3 days", "2-3d prior", "2-7d", "2 weeks prior", "2.5mos", "21d", "24 h 18 min", "24.5 h", "24.5h", "25.5h", "25h", "26h", "27h", "28h", "29.5h", "2d", "2d prior", "2d?", "2wks prior", "3-4d", "30.5h", "30h", "37h", "3d prior", "40h", "41d", "4d", "4d prior", "5d", "5d prior", "6d prior", "7d", "7d prior", "7mos", "9d", "9d prior", "at least 2d", "days", "few days ago", "probably 25h")))

over24h2020 <- subset(ocme_times_2020, Difference..CALCULATED..Number.for..24h.based.on.either.pronounce.time.or.info.in.narrative..whichever.is.least..ROUND.TO.NEAREST.30.MIN..TIME.ELAPSED.AS.NUMBER..min..h..d..w.mo. %in% c(" 2d ", "\"a few days\"", "~1 day", "~24h", "<2d", ">24h", ">3d", ">5d", "1-2 weeks ago", "1 week", "1 week ", "1.5d", "10d", "10d prior", "11d", "12 d", "12d", "15d prior", "19d", "1d", "1wk", "2-3 days", "2-3d prior", "2-7d", "2 weeks prior", "2.5mos", "21d", "24 h 18 min", "24.5 h", "24.5h", "25.5h", "25h", "26h", "27h", "28h", "29.5h", "2d", "2d prior", "2d?", "2wks prior", "3-4d", "30.5h", "30h", "37h", "3d prior", "40h", "41d", "4d", "4d prior", "5d", "5d prior", "6d prior", "7d", "7d prior", "7mos", "9d", "9d prior", "at least 2d", "days", "few days ago", "probably 25h"))

sprintf("There were %d people found within 24 hours, which is %f of all decedents.", nrow(within24h2020), nrow(within24h2020)/nrow(ocme_2020))

sprintf("There were %d people found in over 24 hours, which is %f of all decedents.", nrow(over24h2020), nrow(over24h2020)/nrow(ocme_2020))
```


### Race of <24h and >=24h decedents {.tabset}

#### 2018

**Under 24h**

```{r within24hrace2018, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data=within24h2018, aes(x=race_numeric_words, ..count..)) + 
      geom_bar(aes(fill = race_numeric_words), position = "dodge") + scale_fill_brewer(palette="Spectral") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Racial distribution for 2018 opioid-detected overdose deaths found within 24h")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

withinRaceProp2018 <- ggplot(data=within24h2018, aes(x=race_numeric_words, ..count..)) + 
      geom_bar(aes(fill = race_numeric_words), position = "dodge") + scale_fill_brewer(palette="Spectral") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(within24h2018), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Racial distribution for 2018 opioid-detected overdose deaths found within 24h")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

withinRaceProp2018

#save this one for the presentation
ggsave(plot=withinRaceProp2018, filename = "images/statewide/withinRaceProp2018.png", width = 10, height = 8)

#save this one for robert mtg
#ggsave(plot=raceProp2020, filename = "images/statewide/race_proportions_2020.png", width = 10, height = 8)

```

**Over 24h**

```{r over24hrace2018, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data=over24h2018, aes(x=race_numeric_words, ..count..)) + 
      geom_bar(aes(fill = race_numeric_words), position = "dodge") + scale_fill_brewer(palette="Spectral") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Racial distribution for 2018 opioid-detected overdose deaths found in >24h")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

overRaceProp2018 <- ggplot(data=over24h2018, aes(x=race_numeric_words, ..count..)) + 
      geom_bar(aes(fill = race_numeric_words), position = "dodge") + scale_fill_brewer(palette="Spectral") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(over24h2018), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Racial distribution for 2018 opioid-detected overdose deaths found in >24h")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

overRaceProp2018

#save this one for the presentation
ggsave(plot=overRaceProp2018, filename = "images/statewide/overRaceProp2018.png", width = 10, height = 8)

#save this one for robert mtg
#ggsave(plot=raceProp2020, filename = "images/statewide/race_proportions_2020.png", width = 10, height = 8)

```

#### 2019

**Under 24h**

```{r within24hrace2019, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data=within24h2019, aes(x=race_numeric_words, ..count..)) + 
      geom_bar(aes(fill = race_numeric_words), position = "dodge") + scale_fill_brewer(palette="Spectral") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Racial distribution for 2019 opioid-detected overdose deaths found within 24h")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

withinRaceProp2019 <- ggplot(data=within24h2019, aes(x=race_numeric_words, ..count..)) + 
      geom_bar(aes(fill = race_numeric_words), position = "dodge") + scale_fill_brewer(palette="Spectral") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(within24h2019), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Racial distribution for 2019 opioid-detected overdose deaths found within 24h")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

withinRaceProp2019

#save this one for the presentation
ggsave(plot=withinRaceProp2019, filename = "images/statewide/withinRaceProp2019.png", width = 10, height = 8)

#save this one for robert mtg
#ggsave(plot=raceProp2020, filename = "images/statewide/race_proportions_2020.png", width = 10, height = 8)

```

**Over 24h**

```{r over24hrace2019, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data=over24h2019, aes(x=race_numeric_words, ..count..)) + 
      geom_bar(aes(fill = race_numeric_words), position = "dodge") + scale_fill_brewer(palette="Spectral") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Racial distribution for 2019 opioid-detected overdose deaths found in >24h")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

overRaceProp2019 <- ggplot(data=over24h2019, aes(x=race_numeric_words, ..count..)) + 
      geom_bar(aes(fill = race_numeric_words), position = "dodge") + scale_fill_brewer(palette="Spectral") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(over24h2019), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Racial distribution for 2019 opioid-detected overdose deaths found in >24h")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

overRaceProp2019

#save this one for the presentation
ggsave(plot=overRaceProp2019, filename = "images/statewide/overRaceProp2019.png", width = 10, height = 8)

#save this one for robert mtg
#ggsave(plot=raceProp2020, filename = "images/statewide/race_proportions_2020.png", width = 10, height = 8)

```

#### 2020

**Under 24h**

```{r within24hrace2020, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data=within24h2020, aes(x=race_numeric_words, ..count..)) + 
      geom_bar(aes(fill = race_numeric_words), position = "dodge") + scale_fill_brewer(palette="Spectral") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Racial distribution for 2020 opioid-detected overdose deaths found within 24h")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

withinRaceProp2020 <- ggplot(data=within24h2020, aes(x=race_numeric_words, ..count..)) + 
      geom_bar(aes(fill = race_numeric_words), position = "dodge") + scale_fill_brewer(palette="Spectral") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(within24h2020), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Racial distribution for 2020 opioid-detected overdose deaths found within 24h")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

withinRaceProp2020

#save this one for the presentation
ggsave(plot=withinRaceProp2020, filename = "images/statewide/withinRaceProp2020.png", width = 10, height = 8)

#save this one for robert mtg
#ggsave(plot=raceProp2020, filename = "images/statewide/race_proportions_2020.png", width = 10, height = 8)

```

**Over 24h**

```{r over24hrace2020, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data=over24h2020, aes(x=race_numeric_words, ..count..)) + 
      geom_bar(aes(fill = race_numeric_words), position = "dodge") + scale_fill_brewer(palette="Spectral") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Racial distribution for 2020 opioid-detected overdose deaths found in >24h")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

overRaceProp2020 <- ggplot(data=over24h2020, aes(x=race_numeric_words, ..count..)) + 
      geom_bar(aes(fill = race_numeric_words), position = "dodge") + scale_fill_brewer(palette="Spectral") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(over24h2020), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Racial distribution for 2020 opioid-detected overdose deaths found in >24h")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

overRaceProp2020

#save this one for the presentation
ggsave(plot=overRaceProp2020, filename = "images/statewide/overRaceProp2020.png", width = 10, height = 8)

#save this one for robert mtg
#ggsave(plot=raceProp2020, filename = "images/statewide/race_proportions_2020.png", width = 10, height = 8)

```

### Town of OD of <24h and >=24h decedents {.tabset}

#### 2018

**Under 24h**

```{r town24h2018, echo=FALSE, message=FALSE, warning=FALSE}
#fixed axis text size in theme() using https://stackoverflow.com/questions/65956833/ggplot2-how-to-dynamically-wrap-resize-rescale-x-axis-labels-so-they-wont-over

town_24h_2018 <- ggplot(data=within24h2018, aes(x=injury_city_revised, ..count..)) + 
      geom_bar(aes(fill = ..count..), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = -0.3, colour = "red", size = 3/.pt) + ggtitle("City/town of OD for 2018 opioid-detected deaths found within 24h")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 3))

town_24h_2018

ggsave(plot=town_24h_2018, filename = "images/statewide/overdoses_by_town_24h_2018.png", width = 10, height = 8)

#find the number of ODs in each town by aggregating the indicator var for each town, from https://stackoverflow.com/questions/60961960/how-do-i-sum-a-column-with-categorical-vairables-in-r

# create a new dataframe that we can manipulate without messing up within24h2018

within24h2018prop <- within24h2018 %>% group_by(injury_city_revised) %>% summarise_at(vars(fatal), sum)

# find towns that don't have anyone discovered within 24h and add them into the df
# i have to use unique() so I don't get repeats

missingtowns2018 <- setdiff(unique(ocme_2018$injury_city_revised), within24h2018prop$injury_city_revised)

for (i in missingtowns2018) {
  
  # create row for that town and a 0 next to number of fatalities discovered in 24h
  newtown <- c(i, 0)
  
  # add to within24h2018prop
  within24h2018prop = rbind(within24h2018prop, newtown)
  
}

# convert the within24h2018prop fatals into numerics because for some reason they're strings
within24h2018prop$fatal <- as.numeric(within24h2018prop$fatal)

# aggregate fatalities for overall deaths [not just discovered in 24h]
ocme_2018_agg24h <- ocme_2018 %>% group_by(injury_city_revised) %>% summarise_at(vars(fatal), sum)

# rename the danbury_2018_agg24h "fatal" column because it'll get confusing otherwise when merged
names(ocme_2018_agg24h)[names(ocme_2018_agg24h) == 'fatal'] <- 'total'

# merge with within24h2018prop to have a column with the actual number of deaths
ocme_2018_24h_merged <- merge(within24h2018prop, ocme_2018_agg24h, by="injury_city_revised")

# get the proportion of fatalities discovered in 24h, multiply by 100 so we have a percentage
ocme_2018_24h_merged$Prop <- round((ocme_2018_24h_merged$fatal/ocme_2018_24h_merged$total)*100, digits=2)

# convert any NaN to 0 -- from https://stackoverflow.com/questions/18142117/how-to-replace-nan-value-with-zero-in-a-huge-data-frame

ocme_2018_24h_merged$Prop[is.nan(ocme_2018_24h_merged$Prop)] <- 0

# Plot the proportion of people discovered within 24h
# had an issue with the fill, edited with http://www.sthda.com/english/wiki/ggplot2-barplots-quick-start-guide-r-software-and-data-visualization
# needed to change the font size for the text overlay, used https://stackoverflow.com/questions/25061822/ggplot-geom-text-font-size-control

town_2018_24h_capita <- ggplot(data=ocme_2018_24h_merged, aes(x=injury_city_revised, y=Prop, fill=Prop, position = "dodge")) + 
      geom_bar(stat="Identity") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = Prop), vjust = -0.3, colour = "red", size = 3/.pt) + ggtitle("Percent of fatalities discovered in under 24 hours by town in 2018")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 3))

town_2018_24h_capita

ggsave(plot=town_2018_24h_capita, filename = "images/statewide/overdoses_by_town_prop_24h_2018.png", width = 10, height = 8)

```

**Over 24h**

```{r townover24h2018, echo=FALSE, message=FALSE, warning=FALSE}
#fixed axis text size in theme() using https://stackoverflow.com/questions/65956833/ggplot2-how-to-dynamically-wrap-resize-rescale-x-axis-labels-so-they-wont-over

town_over24h_2018 <- ggplot(data=over24h2018, aes(x=injury_city_revised, ..count..)) + 
      geom_bar(aes(fill = ..count..), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = -0.3, colour = "red", size =6/.pt) + ggtitle("City/town of OD for 2018 opioid-detected deaths found in over 24h")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size =6))

town_over24h_2018

ggsave(plot=town_over24h_2018, filename = "images/statewide/overdoses_by_town_over24h_2018.png", width = 10, height = 8)

#find the number of ODs in each town by aggregating the indicator var for each town, from https://stackoverflow.com/questions/60961960/how-do-i-sum-a-column-with-categorical-vairables-in-r

# create a new dataframe that we can manipulate without messing up within24h2018

over24h2018prop <- over24h2018 %>% group_by(injury_city_revised) %>% summarise_at(vars(fatal), sum)

# find towns that don't have anyone discovered within 24h and add them into the df

missingtowns2018 <- setdiff(unique(ocme_2018$injury_city_revised), over24h2018prop$injury_city_revised)

for (i in missingtowns2018) {
  
  # create row for that town and a 0 next to number of fatalities discovered in 24h
  newtown <- c(i, 0)
  
  # add to within24h2018prop
  over24h2018prop = rbind(over24h2018prop, newtown)
  
}

# convert the within24h2018prop fatals into numerics because for some reason they're strings
over24h2018prop$fatal <- as.numeric(over24h2018prop$fatal)

# aggregate fatalities for overall deaths [not just discovered in 24h]
ocme_2018_agg24h <- ocme_2018 %>% group_by(injury_city_revised) %>% summarise_at(vars(fatal), sum)

# rename the danbury_2018_agg24h "fatal" column because it'll get confusing otherwise when merged
names(ocme_2018_agg24h)[names(ocme_2018_agg24h) == 'fatal'] <- 'total'

# merge with within24h2018prop to have a column with the actual number of deaths
ocme_2018_over24h_merged <- merge(over24h2018prop, ocme_2018_agg24h, by="injury_city_revised")

# get the proportion of fatalities discovered in 24h, multiply by 100 so we have a percentage
ocme_2018_over24h_merged$Prop <- round((ocme_2018_over24h_merged$fatal/ocme_2018_over24h_merged$total)*100, digits=2)

# convert any NaN to 0 -- from https://stackoverflow.com/questions/18142117/how-to-replace-nan-value-with-zero-in-a-huge-data-frame

ocme_2018_over24h_merged$Prop[is.nan(ocme_2018_over24h_merged$Prop)] <- 0

# Plot the proportion of people discovered within 24h
# had an issue with the fill, edited with http://www.sthda.com/english/wiki/ggplot2-barplots-quick-start-guide-r-software-and-data-visualization
# needed to change the font size for the text overlay, used https://stackoverflow.com/questions/25061822/ggplot-geom-text-font-size-control

town_2018_over24h_capita <- ggplot(data=ocme_2018_over24h_merged, aes(x=injury_city_revised, y=Prop, fill=Prop, position = "dodge")) + 
      geom_bar(stat="Identity") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = Prop), vjust = -0.3, colour = "red", size = 3/.pt) + ggtitle("Percent of fatalities discovered in over 24 hours by town in 2018")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 3))

town_2018_over24h_capita

ggsave(plot=town_2018_over24h_capita, filename = "images/statewide/overdoses_by_town_prop_over24h_2018.png", width = 10, height = 8)
 
```

#### 2019

**Under 24h**

```{r town24h2019, echo=FALSE, message=FALSE, warning=FALSE}
#fixed axis text size in theme() using https://stackoverflow.com/questions/65956833/ggplot2-how-to-dynamically-wrap-resize-rescale-x-axis-labels-so-they-wont-over

town_24h_2019 <- ggplot(data=within24h2019, aes(x=injury_city_revised, ..count..)) + 
      geom_bar(aes(fill = ..count..), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = -0.3, colour = "red", size = 3/.pt) + ggtitle("City/town of OD for 2019 opioid-detected deaths found within 24h")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 3))

town_24h_2019

ggsave(plot=town_24h_2019, filename = "images/statewide/overdoses_by_town_24h_2019.png", width = 10, height = 8)

#find the number of ODs in each town by aggregating the indicator var for each town, from https://stackoverflow.com/questions/60961960/how-do-i-sum-a-column-with-categorical-vairables-in-r

# create a new dataframe that we can manipulate without messing up within24h2018

within24h2019prop <- within24h2019 %>% group_by(injury_city_revised) %>% summarise_at(vars(fatal), sum)

# find towns that don't have anyone discovered within 24h and add them into the df

missingtowns2019 <- setdiff(unique(ocme_2019$injury_city_revised) , within24h2019prop$injury_city_revised)

for (i in missingtowns2019) {
  
  # create row for that town and a 0 next to number of fatalities discovered in 24h
  newtown <- c(i, 0)
  
  # add to within24h2018prop
  within24h2019prop = rbind(within24h2019prop, newtown)
  
}

# convert the within24h2018prop fatals into numerics because for some reason they're strings
within24h2019prop$fatal <- as.numeric(within24h2019prop$fatal)

# aggregate fatalities for overall deaths [not just discovered in 24h]
ocme_2019_agg24h <- ocme_2019 %>% group_by(injury_city_revised) %>% summarise_at(vars(fatal), sum)

# rename the danbury_2018_agg24h "fatal" column because it'll get confusing otherwise when merged
names(ocme_2019_agg24h)[names(ocme_2019_agg24h) == 'fatal'] <- 'total'

# merge with within24h2018prop to have a column with the actual number of deaths
ocme_2019_24h_merged <- merge(within24h2019prop, ocme_2019_agg24h, by="injury_city_revised")

# get the proportion of fatalities discovered in 24h, multiply by 100 so we have a percentage
ocme_2019_24h_merged$Prop <- round((ocme_2019_24h_merged$fatal/ocme_2019_24h_merged$total)*100, digits=2)

# convert any NaN to 0 -- from https://stackoverflow.com/questions/18142117/how-to-replace-nan-value-with-zero-in-a-huge-data-frame

ocme_2019_24h_merged$Prop[is.nan(ocme_2019_24h_merged$Prop)] <- 0

# Plot the proportion of people discovered within 24h
# had an issue with the fill, edited with http://www.sthda.com/english/wiki/ggplot2-barplots-quick-start-guide-r-software-and-data-visualization
# needed to change the font size for the text overlay, used https://stackoverflow.com/questions/25061822/ggplot-geom-text-font-size-control

town_2019_24h_capita <- ggplot(data=ocme_2019_24h_merged, aes(x=injury_city_revised, y=Prop, fill=Prop, position = "dodge")) + 
      geom_bar(stat="Identity") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = Prop), vjust = -0.3, colour = "red", size = 3/.pt) + ggtitle("Percent of fatalities discovered in under 24 hours by town in 2019")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 3))

town_2019_24h_capita

ggsave(plot=town_2019_24h_capita, filename = "images/statewide/overdoses_by_town_prop_24h_2019.png", width = 10, height = 8)
 
```

**Over 24h**

```{r townover24h2019, echo=FALSE, message=FALSE, warning=FALSE}
#fixed axis text size in theme() using https://stackoverflow.com/questions/65956833/ggplot2-how-to-dynamically-wrap-resize-rescale-x-axis-labels-so-they-wont-over

town_over24h_2019 <- ggplot(data=over24h2019, aes(x=injury_city_revised, ..count..)) + 
      geom_bar(aes(fill = ..count..), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = -0.3, colour = "red", size =6/.pt) + ggtitle("City/town of OD for 2019 opioid-detected deaths found in over 24h")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size =6))

town_over24h_2019

ggsave(plot=town_over24h_2019, filename = "images/statewide/overdoses_by_town_over24h_2019.png", width = 10, height = 8)

#find the number of ODs in each town by aggregating the indicator var for each town, from https://stackoverflow.com/questions/60961960/how-do-i-sum-a-column-with-categorical-vairables-in-r

# create a new dataframe that we can manipulate without messing up within24h2018

over24h2019prop <- over24h2019 %>% group_by(injury_city_revised) %>% summarise_at(vars(fatal), sum)

# find towns that don't have anyone discovered within 24h and add them into the df

missingtowns2019 <- setdiff(unique(ocme_2019$injury_city_revised) , over24h2019prop$injury_city_revised)

for (i in missingtowns2019) {
  
  # create row for that town and a 0 next to number of fatalities discovered in 24h
  newtown <- c(i, 0)
  
  # add to within24h2018prop
  over24h2019prop = rbind(over24h2019prop, newtown)
  
}

# convert the within24h2018prop fatals into numerics because for some reason they're strings
over24h2019prop$fatal <- as.numeric(over24h2019prop$fatal)

# aggregate fatalities for overall deaths [not just discovered in 24h]
ocme_2019_agg24h <- ocme_2019 %>% group_by(injury_city_revised) %>% summarise_at(vars(fatal), sum)

# rename the danbury_2018_agg24h "fatal" column because it'll get confusing otherwise when merged
names(ocme_2019_agg24h)[names(ocme_2019_agg24h) == 'fatal'] <- 'total'

# merge with within24h2018prop to have a column with the actual number of deaths
ocme_2019_over24h_merged <- merge(over24h2019prop, ocme_2019_agg24h, by="injury_city_revised")

# get the proportion of fatalities discovered in 24h, multiply by 100 so we have a percentage
ocme_2019_over24h_merged$Prop <- round((ocme_2019_over24h_merged$fatal/ocme_2019_over24h_merged$total)*100, digits=2)

# convert any NaN to 0 -- from https://stackoverflow.com/questions/18142117/how-to-replace-nan-value-with-zero-in-a-huge-data-frame

ocme_2019_over24h_merged$Prop[is.nan(ocme_2019_over24h_merged$Prop)] <- 0

# Plot the proportion of people discovered within 24h
# had an issue with the fill, edited with http://www.sthda.com/english/wiki/ggplot2-barplots-quick-start-guide-r-software-and-data-visualization
# needed to change the font size for the text overlay, used https://stackoverflow.com/questions/25061822/ggplot-geom-text-font-size-control

town_2019_over24h_capita <- ggplot(data=ocme_2019_over24h_merged, aes(x=injury_city_revised, y=Prop, fill=Prop, position = "dodge")) + 
      geom_bar(stat="Identity") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = Prop), vjust = -0.3, colour = "red", size = 3/.pt) + ggtitle("Percent of fatalities discovered in over 24 hours by town in 2019")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 3))

town_2019_over24h_capita

ggsave(plot=town_2019_over24h_capita, filename = "images/statewide/overdoses_by_town_prop_over24h_2019.png", width = 10, height = 8)
 
```

#### 2020

**Under 24h**

```{r town24h2020, echo=FALSE, message=FALSE, warning=FALSE}
#fixed axis text size in theme() using https://stackoverflow.com/questions/65956833/ggplot2-how-to-dynamically-wrap-resize-rescale-x-axis-labels-so-they-wont-over

town_24h_2020 <- ggplot(data=within24h2020, aes(x=injury_city_revised, ..count..)) + 
      geom_bar(aes(fill = ..count..), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = -0.3, colour = "red", size = 3/.pt) + ggtitle("City/town of OD for 2020 opioid-detected deaths found within 24h")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 3))

town_24h_2020

ggsave(plot=town_24h_2020, filename = "images/statewide/overdoses_by_town_24h_2020.png", width = 10, height = 8)

#find the number of ODs in each town by aggregating the indicator var for each town, from https://stackoverflow.com/questions/60961960/how-do-i-sum-a-column-with-categorical-vairables-in-r

# create a new dataframe that we can manipulate without messing up within24h2018

within24h2020prop <- within24h2020 %>% group_by(injury_city_revised) %>% summarise_at(vars(fatal), sum)

# find towns that don't have anyone discovered within 24h and add them into the df

missingtowns2020 <- setdiff(unique(ocme_2020$injury_city_revised) , within24h2020prop$injury_city_revised)

for (i in missingtowns2020) {
  
  # create row for that town and a 0 next to number of fatalities discovered in 24h
  newtown <- c(i, 0)
  
  # add to within24h2018prop
  within24h2020prop = rbind(within24h2020prop, newtown)
  
}

# convert the within24h2018prop fatals into numerics because for some reason they're strings
within24h2020prop$fatal <- as.numeric(within24h2020prop$fatal)

# aggregate fatalities for overall deaths [not just discovered in 24h]
ocme_2020_agg24h <- ocme_2020 %>% group_by(injury_city_revised) %>% summarise_at(vars(fatal), sum)

# rename the danbury_2018_agg24h "fatal" column because it'll get confusing otherwise when merged
names(ocme_2020_agg24h)[names(ocme_2020_agg24h) == 'fatal'] <- 'total'

# merge with within24h2018prop to have a column with the actual number of deaths
ocme_2020_24h_merged <- merge(within24h2020prop, ocme_2020_agg24h, by="injury_city_revised")

# get the proportion of fatalities discovered in 24h, multiply by 100 so we have a percentage
ocme_2020_24h_merged$Prop <- round((ocme_2020_24h_merged$fatal/ocme_2020_24h_merged$total)*100, digits=2)

# convert any NaN to 0 -- from https://stackoverflow.com/questions/18142117/how-to-replace-nan-value-with-zero-in-a-huge-data-frame

ocme_2020_24h_merged$Prop[is.nan(ocme_2020_24h_merged$Prop)] <- 0

# Plot the proportion of people discovered within 24h
# had an issue with the fill, edited with http://www.sthda.com/english/wiki/ggplot2-barplots-quick-start-guide-r-software-and-data-visualization
# needed to change the font size for the text overlay, used https://stackoverflow.com/questions/25061822/ggplot-geom-text-font-size-control

town_2020_24h_capita <- ggplot(data=ocme_2020_24h_merged, aes(x=injury_city_revised, y=Prop, fill=Prop, position = "dodge")) + 
      geom_bar(stat="Identity") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = Prop), vjust = -0.3, colour = "red", size = 3/.pt) + ggtitle("Percent of fatalities discovered in under 24 hours by town in 2020")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 3))

town_2020_24h_capita

ggsave(plot=town_2020_24h_capita, filename = "images/statewide/overdoses_by_town_prop_24h_2020.png", width = 10, height = 8)
 
```

**Over 24h**

```{r townover24h2020, echo=FALSE, message=FALSE, warning=FALSE}
#fixed axis text size in theme() using https://stackoverflow.com/questions/65956833/ggplot2-how-to-dynamically-wrap-resize-rescale-x-axis-labels-so-they-wont-over

town_over24h_2020 <- ggplot(data=over24h2020, aes(x=injury_city_revised, ..count..)) + 
      geom_bar(aes(fill = ..count..), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = -0.3, colour = "red", size =6/.pt) + ggtitle("City/town of OD for 2020 opioid-detected deaths found in over 24h")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size =6))

town_over24h_2020

ggsave(plot=town_over24h_2020, filename = "images/statewide/overdoses_by_town_over24h_2020.png", width = 10, height = 8)

#find the number of ODs in each town by aggregating the indicator var for each town, from https://stackoverflow.com/questions/60961960/how-do-i-sum-a-column-with-categorical-vairables-in-r

# create a new dataframe that we can manipulate without messing up within24h2018

over24h2020prop <- over24h2020 %>% group_by(injury_city_revised) %>% summarise_at(vars(fatal), sum)

# find towns that don't have anyone discovered within 24h and add them into the df

missingtowns2020 <- setdiff(unique(ocme_2020$injury_city_revised) , over24h2020prop$injury_city_revised)

for (i in missingtowns2020) {
  
  # create row for that town and a 0 next to number of fatalities discovered in 24h
  newtown <- c(i, 0)
  
  # add to within24h2018prop
  over24h2020prop = rbind(over24h2020prop, newtown)
  
}

# convert the within24h2018prop fatals into numerics because for some reason they're strings
over24h2020prop$fatal <- as.numeric(over24h2020prop$fatal)

# aggregate fatalities for overall deaths [not just discovered in 24h]
ocme_2020_agg24h <- ocme_2020 %>% group_by(injury_city_revised) %>% summarise_at(vars(fatal), sum)

# rename the danbury_2018_agg24h "fatal" column because it'll get confusing otherwise when merged
names(ocme_2020_agg24h)[names(ocme_2020_agg24h) == 'fatal'] <- 'total'

# merge with within24h2018prop to have a column with the actual number of deaths
ocme_2020_over24h_merged <- merge(over24h2020prop, ocme_2020_agg24h, by="injury_city_revised")

# get the proportion of fatalities discovered in 24h, multiply by 100 so we have a percentage
ocme_2020_over24h_merged$Prop <- round((ocme_2020_over24h_merged$fatal/ocme_2020_over24h_merged$total)*100, digits=2)

# convert any NaN to 0 -- from https://stackoverflow.com/questions/18142117/how-to-replace-nan-value-with-zero-in-a-huge-data-frame

ocme_2020_over24h_merged$Prop[is.nan(ocme_2020_over24h_merged$Prop)] <- 0

# Plot the proportion of people discovered within 24h
# had an issue with the fill, edited with http://www.sthda.com/english/wiki/ggplot2-barplots-quick-start-guide-r-software-and-data-visualization
# needed to change the font size for the text overlay, used https://stackoverflow.com/questions/25061822/ggplot-geom-text-font-size-control

town_2020_over24h_capita <- ggplot(data=ocme_2020_over24h_merged, aes(x=injury_city_revised, y=Prop, fill=Prop, position = "dodge")) + 
      geom_bar(stat="Identity") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = Prop), vjust = -0.3, colour = "red", size = 3/.pt) + ggtitle("Percent of fatalities discovered in over 24 hours by town in 2020")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 3))

town_2020_over24h_capita

ggsave(plot=town_2020_over24h_capita, filename = "images/statewide/overdoses_by_town_prop_over24h_2020.png", width = 10, height = 8)
 
```

### Injury location for <24h and >=24h {.tabset}

#### 2018

**Under 24h**

```{r residenceUnder242018, echo=FALSE, message=FALSE, warning=FALSE}
residenceUnder24_2018 <- ggplot(data=within24h2018, aes(x=injury_location, ..count..)) + geom_bar(aes(fill = injury_location), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45), breaks = within24h2018$injury_location, labels = within24h2018$injury_location) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle("Opioid-detected overdose injury locations  \n for 2018 decedents found within 24 hours")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 10))

residenceUnder24_2018

#ggsave(plot=injury_collapsed_2019, filename = "images/statewide/injury_collapsed_2019.png", width = 10, height = 8)

ggsave(plot=residenceUnder24_2018, filename = "images/statewide/residenceUnder24_2020.png", width = 10, height = 8)

residenceUnder24_2018_prop <- ggplot(data=within24h2018, aes(x=injury_location, ..count..)) + geom_bar(aes(fill = injury_location), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45), breaks = within24h2018$injury_location, labels = within24h2018$injury_location) + theme_light() + geom_text(aes(label = round(..count../nrow(within24h2018), 3)), stat = "count", vjust = 0, colour = "black") + ggtitle("Opioid-detected overdose injury locations  \n for 2018 decedents found within 24 hours")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 10))

residenceUnder24_2018_prop

#ggsave(plot=injury_collapsed_2019_prop, filename = "images/statewide/injury_collapsed_2019_prop.png", width = 10, height = 8)

ggsave(plot=residenceUnder24_2018_prop, filename = "images/statewide/residenceUnder24_2018_prop.png", width = 10, height = 8)


```

**Over 24h**

```{r residenceOver242018, echo=FALSE, message=FALSE, warning=FALSE}

residenceOver24_2018 <- ggplot(data=over24h2018, aes(x=injury_location, ..count..)) + geom_bar(aes(fill = injury_location), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45), breaks = over24h2018$injury_location, labels = over24h2018$injury_location) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle("Opioid-detected overdose injury locations  \n for 2018 decedents found after >24 hours")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 10))

residenceOver24_2018

#ggsave(plot=injury_collapsed_2019, filename = "images/statewide/injury_collapsed_2019.png", width = 10, height = 8)

ggsave(plot=residenceOver24_2018, filename = "images/statewide/residenceOver24_2018.png", width = 10, height = 8)

residenceOver24_2018_prop <- ggplot(data=over24h2018, aes(x=injury_location, ..count..)) + geom_bar(aes(fill = injury_location), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45), breaks = over24h2018$injury_location, labels = over24h2018$injury_location) + theme_light() + geom_text(aes(label = round(..count../nrow(over24h2018), 3)), stat = "count", vjust = 0, colour = "black") + ggtitle("Opioid-detected overdose injury locations  \n for 2018 decedents found after >24 hours")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 10))

residenceOver24_2018_prop

#ggsave(plot=residenceOver24_2020_prop, filename = "images/statewide/residenceOver24_2020_prop.png", width = 10, height = 8)

ggsave(plot=residenceOver24_2018_prop, filename = "images/statewide/residenceOver24_2018_prop.png", width = 10, height = 8)


```

#### 2019

**Under 24h**

```{r residenceUnder242019, echo=FALSE, message=FALSE, warning=FALSE}

residenceUnder24_2019 <- ggplot(data=within24h2019, aes(x=injury_collapsed, ..count..)) + geom_bar(aes(fill = injury_collapsed), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45), breaks = within24h2019$injury_collapsed, labels = within24h2019$injury_collapsed) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle("Opioid-detected overdose injury locations  \n for 2019 decedents found within 24 hours")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 10))

residenceUnder24_2019

#ggsave(plot=injury_collapsed_2019, filename = "images/statewide/injury_collapsed_2019.png", width = 10, height = 8)

ggsave(plot=residenceUnder24_2019, filename = "images/statewide/residenceUnder24_2019.png", width = 10, height = 8)

residenceUnder24_2019_prop <- ggplot(data=within24h2019, aes(x=injury_collapsed, ..count..)) + geom_bar(aes(fill = injury_collapsed), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45), breaks = within24h2019$injury_collapsed, labels = within24h2019$injury_collapsed) + theme_light() + geom_text(aes(label = round(..count../nrow(within24h2019), 3)), stat = "count", vjust = 0, colour = "black") + ggtitle("Opioid-detected overdose injury locations  \n for 2019 decedents found within 24 hours")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 10))

residenceUnder24_2019_prop

#ggsave(plot=injury_collapsed_2019_prop, filename = "images/statewide/injury_collapsed_2019_prop.png", width = 10, height = 8)

ggsave(plot=residenceUnder24_2019_prop, filename = "images/statewide/residenceUnder24_2019_prop.png", width = 10, height = 8)


```

**Over 24h**

```{r residenceOver242019, echo=FALSE, message=FALSE, warning=FALSE}

residenceOver24_2019 <- ggplot(data=over24h2019, aes(x=injury_collapsed, ..count..)) + geom_bar(aes(fill = injury_collapsed), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45), breaks = over24h2019$injury_collapsed, labels = over24h2019$injury_collapsed) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle("Opioid-detected overdose injury locations  \n for 2019 decedents found after >24 hours")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 10))

residenceOver24_2019

#ggsave(plot=injury_collapsed_2019, filename = "images/statewide/injury_collapsed_2019.png", width = 10, height = 8)

ggsave(plot=residenceOver24_2019, filename = "images/statewide/residenceOver24_2020.png", width = 10, height = 8)

residenceOver24_2019_prop <- ggplot(data=over24h2019, aes(x=injury_collapsed, ..count..)) + geom_bar(aes(fill = injury_collapsed), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45), breaks = over24h2019$injury_collapsed, labels = over24h2019$injury_collapsed) + theme_light() + geom_text(aes(label = round(..count../nrow(over24h2019), 3)), stat = "count", vjust = 0, colour = "black") + ggtitle("Opioid-detected overdose injury locations  \n for 2019 decedents found after >24 hours")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 10))

residenceOver24_2019_prop

#ggsave(plot=residenceOver24_2020_prop, filename = "images/statewide/residenceOver24_2020_prop.png", width = 10, height = 8)

ggsave(plot=residenceOver24_2019_prop, filename = "images/statewide/residenceOver24_2019_prop.png", width = 10, height = 8)


```

#### 2020

**Under 24h**

```{r residenceUnder242020, echo=FALSE, message=FALSE, warning=FALSE}

residenceUnder24_2020 <- ggplot(data=within24h2020, aes(x=injury_collapsed, ..count..)) + geom_bar(aes(fill = injury_collapsed), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45), breaks = within24h2020$injury_collapsed, labels = within24h2020$injury_collapsed) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle("Opioid-detected overdose injury locations  \n for 2020 decedents found within 24 hours")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 10))

residenceUnder24_2020

#ggsave(plot=injury_collapsed_2019, filename = "images/statewide/injury_collapsed_2019.png", width = 10, height = 8)

ggsave(plot=residenceUnder24_2020, filename = "images/statewide/residenceUnder24_2020.png", width = 10, height = 8)

residenceUnder24_2020_prop <- ggplot(data=within24h2020, aes(x=injury_collapsed, ..count..)) + geom_bar(aes(fill = injury_collapsed), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45), breaks = within24h2020$injury_collapsed, labels = within24h2020$injury_collapsed) + theme_light() + geom_text(aes(label = round(..count../nrow(within24h2020), 3)), stat = "count", vjust = 0, colour = "black") + ggtitle("Opioid-detected overdose injury locations  \n for 2020 decedents found within 24 hours")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 10))

residenceUnder24_2020_prop

#ggsave(plot=injury_collapsed_2019_prop, filename = "images/statewide/injury_collapsed_2019_prop.png", width = 10, height = 8)

ggsave(plot=residenceUnder24_2020_prop, filename = "images/statewide/residenceUnder24_2020_prop.png", width = 10, height = 8)


```

**Over 24h**

```{r residenceOver242020, echo=FALSE, message=FALSE, warning=FALSE}

residenceOver24_2020 <- ggplot(data=over24h2020, aes(x=injury_collapsed, ..count..)) + geom_bar(aes(fill = injury_collapsed), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45), breaks = over24h2020$injury_collapsed, labels = over24h2020$injury_collapsed) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle("Opioid-detected overdose injury locations  \n for 2020 decedents found after >24 hours")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 10))

residenceOver24_2020

#ggsave(plot=injury_collapsed_2019, filename = "images/statewide/injury_collapsed_2019.png", width = 10, height = 8)

ggsave(plot=residenceOver24_2020, filename = "images/statewide/residenceOver24_2020.png", width = 10, height = 8)

residenceOver24_2020_prop <- ggplot(data=over24h2020, aes(x=injury_collapsed, ..count..)) + geom_bar(aes(fill = injury_collapsed), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45), breaks = over24h2020$injury_collapsed, labels = over24h2020$injury_collapsed) + theme_light() + geom_text(aes(label = round(..count../nrow(over24h2020), 3)), stat = "count", vjust = 0, colour = "black") + ggtitle("Opioid-detected overdose injury locations  \n for 2020 decedents found after >24 hours")+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 10))

residenceOver24_2020_prop

#ggsave(plot=residenceOver24_2020_prop, filename = "images/statewide/residenceOver24_2020_prop.png", width = 10, height = 8)

ggsave(plot=residenceOver24_2020_prop, filename = "images/statewide/residenceOver24_2020_prop.png", width = 10, height = 8)


```

### Living arrangements for <24h and >=24h {.tabset}

#### 2018

**Under 24h**


```{r liveAloneUnder242018, echo=FALSE, message=FALSE, warning=FALSE}
#subset all cases where people died in their own house

residence2018_24h <- subset(within24h2018, injury_address_updated == residence_address_updated)

residence2018_24h$roommates <- ifelse(residence2018_24h$OTHERS...INJURY.LOCATION..A..alone.F..fam.ptner.N.nonfamM.missing.N.A. %in% c(" F ", "F?", "F", "F? lived w str & her family"), "Family/partner", ifelse(residence2018_24h$OTHERS...INJURY.LOCATION..A..alone.F..fam.ptner.N.nonfamM.missing.N.A. %in% c("unclear", "unspecified", "", "2"), "Unknown", ifelse(residence2018_24h$OTHERS...INJURY.LOCATION..A..alone.F..fam.ptner.N.nonfamM.missing.N.A. == "A", "Alone", ifelse(residence2018_24h$OTHERS...INJURY.LOCATION..A..alone.F..fam.ptner.N.nonfamM.missing.N.A. %in% c("M", "777", ""), "Unknown", ifelse(residence2018_24h$OTHERS...INJURY.LOCATION..A..alone.F..fam.ptner.N.nonfamM.missing.N.A. %in% c("N", "N "), "Non-family", ifelse(residence2018_24h$OTHERS...INJURY.LOCATION..A..alone.F..fam.ptner.N.nonfamM.missing.N.A. %in% c("F/N", "N/F"), "Family and non-family", residence2018_24h$OTHERS...INJURY.LOCATION..A..alone.F..fam.ptner.N.nonfamM.missing.N.A.))))))

liveAlone_2018_24h <-ggplot(data=residence2018_24h, aes(x=roommates, ..count..)) + 
      geom_bar(aes(fill = roommates), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Living arrangements for 2018 opioid-detected deaths in own residence \n found within 24 hours")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none")

liveAlone_2018_24h

ggsave(plot=liveAlone_2018_24h, filename = "images/statewide/live_alone_2018_24h.png", width = 10, height = 8)

liveAlone_2018_prop_24h <-ggplot(data=residence2018_24h, aes(x=roommates, ..count..)) + 
      geom_bar(aes(fill = roommates), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(residence2018_24h),3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Living arrangements for 2018 opioid-detected deaths in own residence \n found within 24h")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none")

liveAlone_2018_prop_24h

ggsave(plot=liveAlone_2018_prop_24h, filename = "images/statewide/live_alone_2018_prop_24h.png", width = 10, height = 8)

```

**Over 24h**


```{r liveAloneOver242018, echo=FALSE, message=FALSE, warning=FALSE}
#subset all cases where people died in their own house

residence2018_over24h <- subset(over24h2018, injury_address_updated == residence_address_updated)

residence2018_over24h$roommates <- ifelse(residence2018_over24h$OTHERS...INJURY.LOCATION..A..alone.F..fam.ptner.N.nonfamM.missing.N.A. %in% c(" F ", "F?", "F", "F? lived w str & her family"), "Family/partner", ifelse(residence2018_over24h$OTHERS...INJURY.LOCATION..A..alone.F..fam.ptner.N.nonfamM.missing.N.A. %in% c("unclear", "unspecified", "", "2"), "Unknown", ifelse(residence2018_over24h$OTHERS...INJURY.LOCATION..A..alone.F..fam.ptner.N.nonfamM.missing.N.A. %in% c("A", "A?"), "Alone", ifelse(residence2018_over24h$OTHERS...INJURY.LOCATION..A..alone.F..fam.ptner.N.nonfamM.missing.N.A. %in% c("M", "777", ""), "Unknown", ifelse(residence2018_over24h$OTHERS...INJURY.LOCATION..A..alone.F..fam.ptner.N.nonfamM.missing.N.A. %in% c("N", "N "), "Non-family", ifelse(residence2018_over24h$OTHERS...INJURY.LOCATION..A..alone.F..fam.ptner.N.nonfamM.missing.N.A. %in% c("F/N", "N/F"), "Family and non-family", residence2018_over24h$OTHERS...INJURY.LOCATION..A..alone.F..fam.ptner.N.nonfamM.missing.N.A.))))))

liveAlone_2018_over24h <-ggplot(data=residence2018_over24h, aes(x=roommates, ..count..)) + 
      geom_bar(aes(fill = roommates), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Living arrangements for 2018 opioid-detected deaths in own residence \n found in over 24 hours")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none")

liveAlone_2018_over24h

ggsave(plot=liveAlone_2018_over24h, filename = "images/statewide/live_alone_2018_over24h.png", width = 10, height = 8)

liveAlone_2018_prop_over24h <-ggplot(data=residence2018_over24h, aes(x=roommates, ..count..)) + 
      geom_bar(aes(fill = roommates), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(residence2018_over24h),3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Living arrangements for 2018 opioid-detected deaths in own residence \n found in over 24h")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none")

liveAlone_2018_prop_over24h

ggsave(plot=liveAlone_2018_prop_24h, filename = "images/statewide/live_alone_2018_prop_over24h.png", width = 10, height = 8)

```

#### 2019

**Under 24h**

```{r liveAloneUnder242019, echo=FALSE, message=FALSE, warning=FALSE}
#subset all cases where people died in their own house

residence2019_24h <- subset(within24h2019, Injury.Address1 == Residence.Address1)

#table(residence2020$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfam.M.missing.N.A.)

residence2019_24h$roommates <- ifelse(residence2019_24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfamM.missing.N.A. %in% c(" F ", "F?", "F", "F? lived w str & her family"), "Family/partner", ifelse(residence2019_24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfamM.missing.N.A. %in% c("unclear", "unspecified", "", "2"), "Unknown", ifelse(residence2019_24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfamM.missing.N.A. == "A", "Alone", ifelse(residence2019_24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfamM.missing.N.A. %in% c("M", "TOC 59; pronounced 119"), "Unknown", ifelse(residence2019_24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfamM.missing.N.A. %in% c("N", "N "), "Non-family", ifelse(residence2019_24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfamM.missing.N.A. == "F,N", "Family and non-family", residence2019_24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfamM.missing.N.A.))))))

liveAlone_2019_24h <-ggplot(data=residence2019_24h, aes(x=roommates, ..count..)) + 
      geom_bar(aes(fill = roommates), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Living arrangements for 2019 opioid-detected deaths in own residence \n found within 24 hours")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none")

liveAlone_2019_24h

ggsave(plot=liveAlone_2019_24h, filename = "images/statewide/live_alone_2019_24h.png", width = 10, height = 8)

liveAlone_2019_prop_24h <-ggplot(data=residence2019_24h, aes(x=roommates, ..count..)) + 
      geom_bar(aes(fill = roommates), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(residence2019_24h),3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Living arrangements for 2019 opioid-detected deaths in own residence \n found within 24h")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none")

liveAlone_2019_prop_24h

ggsave(plot=liveAlone_2019_prop_24h, filename = "images/statewide/live_alone_2019_prop_24h.png", width = 10, height = 8)

```


**Over 24h**

```{r liveAloneOver242019, echo=FALSE, message=FALSE, warning=FALSE}
#subset all cases where people died in their own house

residence2019_over24h <- subset(over24h2019, Injury.Address1 == Residence.Address1)

#table(residence2020$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfam.M.missing.N.A.)

residence2019_over24h$roommates <- ifelse(residence2019_over24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfamM.missing.N.A. %in% c(" F ", "F?", "F", "F? lived w str & her family"), "Family/partner", ifelse(residence2019_over24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfamM.missing.N.A. %in% c("unclear", "unspecified", "", "2"), "Unknown", ifelse(residence2019_over24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfamM.missing.N.A. == "A", "Alone", ifelse(residence2019_over24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfamM.missing.N.A. %in% c("M", "TOC 59; pronounced 119", "N/A"), "Unknown", ifelse(residence2019_over24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfamM.missing.N.A. %in% c("N", "N "), "Non-family", ifelse(residence2019_over24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfamM.missing.N.A. == "F,N", "Family and non-family", residence2019_over24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfamM.missing.N.A.))))))

liveAlone_2019_over24h <-ggplot(data=residence2019_over24h, aes(x=roommates, ..count..)) + 
      geom_bar(aes(fill = roommates), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Living arrangements for 2019 opioid-detected deaths in own residence \n found in over 24 hours")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none")

liveAlone_2019_over24h

ggsave(plot=liveAlone_2019_over24h, filename = "images/statewide/live_alone_2019_over24h.png", width = 10, height = 8)

liveAlone_2019_prop_over24h <-ggplot(data=residence2019_over24h, aes(x=roommates, ..count..)) + 
      geom_bar(aes(fill = roommates), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(residence2019_over24h),3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Living arrangements for 2019 opioid-detected deaths in own residence \n found in over 24h")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none")

liveAlone_2019_prop_over24h

ggsave(plot=liveAlone_2019_prop_over24h, filename = "images/statewide/live_alone_2019_prop_over24h.png", width = 10, height = 8)

```

#### 2020

**Under 24h**

```{r liveAloneUnder242020, echo=FALSE, message=FALSE, warning=FALSE}
#subset all cases where people died in their own house
residence2020_24h <- subset(within24h2020, Injury.Address1 == Residence.Address1)

#table(residence2020$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfam.M.missing.N.A.)

residence2020_24h$roommates <- ifelse(residence2020_24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfam.M.missing.N.A. %in% c(" F ", "F?", "F"), "Family/partner", ifelse(residence2020_24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfam.M.missing.N.A. %in% c("unclear", "unspecified", "", "2"), "Unknown", ifelse(residence2020_24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfam.M.missing.N.A. == "A", "Alone", ifelse(residence2020_24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfam.M.missing.N.A. == "M", "Unknown", ifelse(residence2020_24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfam.M.missing.N.A. %in% c("N", "N "), "Non-family", ifelse(residence2020_24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfam.M.missing.N.A. == "F/N", "Family and non-family", residence2020_24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfam.M.missing.N.A.))))))

liveAlone_2020_24h <-ggplot(data=residence2020_24h, aes(x=roommates, ..count..)) + 
      geom_bar(aes(fill = roommates), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Living arrangements for 2020 opioid-detected deaths in own residence \n found within 24 hours")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none")

liveAlone_2020_24h

ggsave(plot=liveAlone_2020_24h, filename = "images/statewide/live_alone_2020_24h.png", width = 10, height = 8)

liveAlone_2020_prop_24h <-ggplot(data=residence2020_24h, aes(x=roommates, ..count..)) + 
      geom_bar(aes(fill = roommates), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(residence2020_24h),3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Living arrangements for 2020 opioid-detected deaths in own residence \n found within 24h")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none")

liveAlone_2020_prop_24h

ggsave(plot=liveAlone_2020_prop_24h, filename = "images/statewide/live_alone_2020_prop_24h.png", width = 10, height = 8)

```

**Over 24h**

```{r liveAloneOver242020, echo=FALSE, message=FALSE, warning=FALSE}
#subset all cases where people died in their own house
residence2020_over24h <- subset(over24h2020, Injury.Address1 == Residence.Address1)

#table(residence2020$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfam.M.missing.N.A.)

residence2020_over24h$roommates <- ifelse(residence2020_over24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfam.M.missing.N.A. %in% c(" F ", "F?", "F"), "Family/partner", ifelse(residence2020_over24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfam.M.missing.N.A. %in% c("unclear", "unspecified", "", "2"), "Unknown", ifelse(residence2020_over24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfam.M.missing.N.A. == "A", "Alone", ifelse(residence2020_over24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfam.M.missing.N.A. == "M", "Unknown", ifelse(residence2020_over24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfam.M.missing.N.A. %in% c("N", "N "), "Non-family", ifelse(residence2020_over24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfam.M.missing.N.A. == "F/N", "Family and non-family", residence2020_over24h$OTHERS.LIVING.with.D...A..alone.F..fam.ptner.N.nonfam.M.missing.N.A.))))))

liveAlone_2020_over24h <-ggplot(data=residence2020_over24h, aes(x=roommates, ..count..)) + 
      geom_bar(aes(fill = roommates), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle("Living arrangements for 2020 opioid-detected deaths in own residence \n found in over 24 hours")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none")

liveAlone_2020_over24h

ggsave(plot=liveAlone_2020_over24h, filename = "images/statewide/live_alone_2020_over24h.png", width = 10, height = 8)

liveAlone_2020_prop_over24h <-ggplot(data=residence2020_over24h, aes(x=roommates, ..count..)) + 
      geom_bar(aes(fill = roommates), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(residence2020_over24h),3)), stat = "count", vjust = 0, colour = "black")+ ggtitle("Living arrangements for 2020 opioid-detected deaths in own residence \n found in over 24h")+ theme(plot.title = element_text(hjust = 0.5), legend.position="none")

liveAlone_2020_prop_over24h

ggsave(plot=liveAlone_2020_prop_over24h, filename = "images/statewide/live_alone_2020_prop_over24h.png", width = 10, height = 8)

```

<a href="#top">
  <button>Back to top</button>
</a>
