---
title: "Age distribution of fatal opioid-detected overdoses in Connecticut, 2009-2018"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 5
    theme: lumen
---

```{r sourceData, echo=FALSE, message=FALSE, warning=FALSE}
library(tigris)
library(ggplot2)
library(stringr)
library(dplyr)
library(lubridate)
library(patchwork)
library(tidyverse) #to replace NA with something else
library(sf)
library(multimode) #to calculate modes for the age plot
library(findSVI) #for SVI calculation
library(diptest)
library(ggrepel)


# get the most updated data files
source("data_preprocessing/ocme_processing.R")
# set working directory -- I have to do this after sourcing the data because it loses the path otherwise
setwd("C:\\Users\\ar2529\\Desktop\\overdose_age")
source("data_preprocessing/sword_processing.R")
# set working directory -- I have to do this after sourcing the data because it loses the path otherwise
setwd("C:\\Users\\ar2529\\Desktop\\overdose_age")
source("data_preprocessing/city_county_processing.R")
# set working directory -- I have to do this after sourcing the data because it loses the path otherwise
setwd("C:\\Users\\ar2529\\Desktop\\overdose_age")

```

```{r importData, echo=FALSE, message=FALSE, warning=FALSE}

# import OCME data 

ocme_old <- read.csv("data_preprocessing/processed_data/ocme_2009_2017_processed.csv")

# subset out male and female populations

male_old <- subset(ocme_old, Sex_revised == "Male")
female_old <- subset(ocme_old, Sex_revised == "Female")

# subset out populations by race

white_old <- subset(ocme_old, race_numeric_words == "White")
black_old <- subset(ocme_old, race_numeric_words == "Black")
latine_old <- subset(ocme_old, race_numeric_words == "Hispanic")


# get town data from tigris so we can map deaths by town later

options(tigris_use_cache=TRUE)
ct_towns = county_subdivisions(state="Connecticut")
ct_towns$NAME = str_to_title(ct_towns$NAME)
ct_towns = filter(ct_towns, NAME != "County Subdivisions Not Defined")
# tigris also has shapefiles for the long island sound so let's get rid of that
ct_towns <- subset(ct_towns, ALAND!=0)

for(i in unique(ocme_old$Year)){
  assign(paste0("ocme_", as.character(i)), subset(ocme_old, Year == as.character(i)))
}

```

### Age distribution for all fatal and nonfatal overdoses {.tabset}

#### 2009

```{r age2009, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
# looks like there are some NAs in the 2009 data so I have to get rid of them
bandwidth <- bw.nrd0(na.omit(ocme_2009$age)) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(ocme_2009$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(ocme_2009$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
ocme_2009_young <- subset(ocme_2009, age <= antimodes[1])
ocme_2009_old <- subset(ocme_2009, age > antimodes[1])

# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

ocme_2009$aftermode <- ocme_2009$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=ocme_2009, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2009$age, labels = ocme_2009$age) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2009 CT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 7)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2009 <- ggplot(data=ocme_2009, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2009$age, labels = ocme_2009$age) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2009 CT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 7)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2009

ggsave(plot=age_2009, filename = "images/age_2009.png", width = 10, height = 8)


sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(ocme_2009$age)$p.value) 

 
```

#### 2010

```{r age2010, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
# looks like there are some NAs in the 2010 data so I have to get rid of them
bandwidth <- bw.nrd0(na.omit(ocme_2010$age)) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(ocme_2010$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(ocme_2010$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
ocme_2010_young <- subset(ocme_2010, age <= antimodes[1])
ocme_2010_old <- subset(ocme_2010, age > antimodes[1])

# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

ocme_2010$aftermode <- ocme_2010$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=ocme_2010, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2010$age, labels = ocme_2010$age) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2010 CT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 7)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2010 <- ggplot(data=ocme_2010, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2010$age, labels = ocme_2010$age) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2010 CT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 7)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2010

ggsave(plot=age_2010, filename = "images/age_2010.png", width = 10, height = 8)


sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(ocme_2010$age)$p.value) 

 
```


#### 2011

```{r age2011, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
# looks like there are some NAs in the 2011 data so I have to get rid of them
bandwidth <- bw.nrd0(na.omit(ocme_2011$age)) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(ocme_2011$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(ocme_2011$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
ocme_2011_young <- subset(ocme_2011, age <= antimodes[1])
ocme_2011_old <- subset(ocme_2011, age > antimodes[1])

# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

ocme_2011$aftermode <- ocme_2011$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=ocme_2011, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2011$age, labels = ocme_2011$age) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2011 CT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 7)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2011 <- ggplot(data=ocme_2011, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2011$age, labels = ocme_2011$age) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2011 CT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 7)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2011

ggsave(plot=age_2011, filename = "images/age_2011.png", width = 10, height = 8)


sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(ocme_2011$age)$p.value) 

 
```


#### 2012

```{r age2012, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
# looks like there are some NAs in the 2012 data so I have to get rid of them
bandwidth <- bw.nrd0(na.omit(ocme_2012$age)) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(ocme_2012$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(ocme_2012$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
ocme_2012_young <- subset(ocme_2012, age <= antimodes[1])
ocme_2012_old <- subset(ocme_2012, age > antimodes[1])

# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

ocme_2012$aftermode <- ocme_2012$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=ocme_2012, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2012$age, labels = ocme_2012$age) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2012 CT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 7)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2012 <- ggplot(data=ocme_2012, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2012$age, labels = ocme_2012$age) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2012 CT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 7)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2012

ggsave(plot=age_2012, filename = "images/age_2012.png", width = 10, height = 8)


sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(ocme_2012$age)$p.value) 

 
```


#### 2013

```{r age2013, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
# looks like there are some NAs in the 2013 data so I have to get rid of them
bandwidth <- bw.nrd0(na.omit(ocme_2013$age)) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(ocme_2013$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(ocme_2013$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
ocme_2013_young <- subset(ocme_2013, age <= antimodes[1])
ocme_2013_old <- subset(ocme_2013, age > antimodes[1])

# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

ocme_2013$aftermode <- ocme_2013$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=ocme_2013, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2013$age, labels = ocme_2013$age) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2013 CT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 7)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2013 <- ggplot(data=ocme_2013, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2013$age, labels = ocme_2013$age) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2013 CT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 7)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2013

ggsave(plot=age_2013, filename = "images/age_2013.png", width = 10, height = 8)


sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(ocme_2013$age)$p.value) 

 
```


#### 2014

```{r age2014, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
# looks like there are some NAs in the 2014 data so I have to get rid of them
bandwidth <- bw.nrd0(na.omit(ocme_2014$age)) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(ocme_2014$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(ocme_2014$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
ocme_2014_young <- subset(ocme_2014, age <= antimodes[2])
ocme_2014_old <- subset(ocme_2014, age > antimodes[2])

# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

ocme_2014$aftermode <- ocme_2014$age > antimodes[2]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=ocme_2014, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2014$age, labels = ocme_2014$age) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2014 CT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 7)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2014 <- ggplot(data=ocme_2014, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2014$age, labels = ocme_2014$age) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2014 CT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 7)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2014

ggsave(plot=age_2014, filename = "images/age_2014.png", width = 10, height = 8)

sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(ocme_2014$age)$p.value) 

```
#### 2015


```{r age2015, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
# looks like there are some NAs in the 2015 data so I have to get rid of them
bandwidth <- bw.nrd0(na.omit(ocme_2015$age)) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(ocme_2015$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(ocme_2015$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
ocme_2015_young <- subset(ocme_2015, age <= antimodes[1])
ocme_2015_old <- subset(ocme_2015, age > antimodes[1])

# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

ocme_2015$aftermode <- ocme_2015$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=ocme_2015, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2015$age, labels = ocme_2015$age) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2015 CT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 7)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2015 <- ggplot(data=ocme_2015, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2015$age, labels = ocme_2015$age) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2015 CT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 7)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2015

ggsave(plot=age_2015, filename = "images/age_2015.png", width = 10, height = 8)


sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(ocme_2015$age)$p.value) 

 
```


#### 2016

```{r age2016, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
# looks like there are some NAs in the 2016 data so I have to get rid of them
bandwidth <- bw.nrd0(na.omit(ocme_2016$age)) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(ocme_2016$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(ocme_2016$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
ocme_2016_young <- subset(ocme_2016, age <= antimodes[1])
ocme_2016_old <- subset(ocme_2016, age > antimodes[1])

# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

ocme_2016$aftermode <- ocme_2016$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=ocme_2016, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2016$age, labels = ocme_2016$age) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2016 CT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 7)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2016 <- ggplot(data=ocme_2016, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2016$age, labels = ocme_2016$age) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2016 CT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 7)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2016

ggsave(plot=age_2016, filename = "images/age_2016.png", width = 10, height = 8)


sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(ocme_2016$age)$p.value) 

 
```


#### 2017

```{r age2017, echo=FALSE, message=FALSE, warning=FALSE}
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
# looks like there are some NAs in the 2017 data so I have to get rid of them
bandwidth <- bw.nrd0(na.omit(ocme_2017$age)) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(ocme_2017$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(ocme_2017$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
ocme_2017_young <- subset(ocme_2017, age <= antimodes[1])
ocme_2017_old <- subset(ocme_2017, age > antimodes[1])

# create a condition to use different colors based on the location of the antimode
# from https://stackoverflow.com/questions/13372523/altering-the-color-of-one-value-in-a-ggplot-histogram
# it just gives me a 0 for anything in the younger population, 1 for older

ocme_2017$aftermode <- ocme_2017$age > antimodes[1]

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid


# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=ocme_2017, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2017$age, labels = ocme_2017$age) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle("Age distribution for 2017 CT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 7)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2017 <- ggplot(data=ocme_2017, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(aftermode)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme_2017$age, labels = ocme_2017$age) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle("Age distribution for 2017 CT opioid-detected overdose deaths")+theme(plot.title = element_text(hjust = 0.5), legend.position="right", axis.text.x = element_text(size = 7)) + scale_fill_manual(name = "Age group", values = c("#00BFC4", "#F8766D"), labels = c("Younger", "Older"))+xlab("Age")+ylab("Number of decedents")

age_2017

ggsave(plot=age_2017, filename = "images/age_2017.png", width = 10, height = 8)


sprintf("Using Hartigan's dip test for bimodality, the resulting p value is %f.", dip.test(ocme_2017$age)$p.value) 

```

### Median age

```{r median, echo=FALSE, message=FALSE, warning=FALSE}
sprintf("The median age of decedents in 2009 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(ocme_2009$age), median(ocme_2009_young$age), median(ocme_2009_old$age))

sprintf("The median age of decedents in 2010 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(ocme_2010$age), median(ocme_2010_young$age), median(ocme_2010_old$age))

sprintf("The median age of decedents in 2011 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(ocme_2011$age), median(ocme_2011_young$age), median(ocme_2011_old$age))

sprintf("The median age of decedents in 2012 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(ocme_2012$age), median(ocme_2012_young$age), median(ocme_2012_old$age))

sprintf("The median age of decedents in 2013 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(ocme_2013$age), median(ocme_2013_young$age), median(ocme_2013_old$age))

sprintf("The median age of decedents in 2014 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(ocme_2014$age), median(ocme_2014_young$age), median(ocme_2014_old$age))

sprintf("The median age of decedents in 2015 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(ocme_2015$age), median(ocme_2015_young$age), median(ocme_2015_old$age))

sprintf("The median age of decedents in 2016 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(ocme_2016$age), median(ocme_2016_young$age), median(ocme_2016_old$age))

sprintf("The median age of decedents in 2017 was %f. For younger decedents the median age was %f and for older decedents the median age was %f.", median(ocme_2017$age), median(ocme_2017_young$age), median(ocme_2017_old$age))
```


### Proportion of older people by year {.tabset}

```{r propOld, echo=FALSE, message=FALSE, warning=FALSE}

sprintf("The proportion of older decedents in 2009 was %f, with %d out of %d deaths.", nrow(ocme_2009_old)/nrow(ocme_2009), nrow(ocme_2009_old), nrow(ocme_2009))
sprintf("The proportion of older decedents in 2010 was %f, with %d out of %d deaths.", nrow(ocme_2010_old)/nrow(ocme_2010), nrow(ocme_2010_old), nrow(ocme_2010))
sprintf("The proportion of older decedents in 2011 was %f, with %d out of %d deaths.", nrow(ocme_2011_old)/nrow(ocme_2011), nrow(ocme_2011_old), nrow(ocme_2011))
sprintf("The proportion of older decedents in 2012 was %f, with %d out of %d deaths.", nrow(ocme_2012_old)/nrow(ocme_2012), nrow(ocme_2012_old), nrow(ocme_2012))
sprintf("The proportion of older decedents in 2013 was %f, with %d out of %d deaths.", nrow(ocme_2013_old)/nrow(ocme_2013), nrow(ocme_2013_old), nrow(ocme_2013))
sprintf("The proportion of older decedents in 2014 was %f, with %d out of %d deaths.", nrow(ocme_2014_old)/nrow(ocme_2014), nrow(ocme_2014_old), nrow(ocme_2014))
sprintf("The proportion of older decedents in 2015 was %f, with %d out of %d deaths.", nrow(ocme_2015_old)/nrow(ocme_2015), nrow(ocme_2015_old), nrow(ocme_2015))
sprintf("The proportion of older decedents in 2016 was %f, with %d out of %d deaths.", nrow(ocme_2016_old)/nrow(ocme_2016), nrow(ocme_2016_old), nrow(ocme_2016))
sprintf("The proportion of older decedents in 2017 was %f, with %d out of %d deaths.", nrow(ocme_2017_old)/nrow(ocme_2017), nrow(ocme_2017_old), nrow(ocme_2017))

```

### OCME lists

```{r ocmeLists, echo=FALSE, message=FALSE, warning=FALSE}

# this will come in handy for the tox graphs
yearList <- c(2009, 2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017)

# make lists of the different OCME datasets so I can use them later
# create a list of dfs so I can iterate over them -- from https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
ocmelist <- list(ocme_2009, ocme_2010, ocme_2011, ocme_2012, ocme_2013, ocme_2014, ocme_2015, ocme_2016, ocme_2017)

ocmelistYoung <- list(ocme_2009_young, ocme_2010_young, ocme_2011_young, ocme_2012_young, ocme_2013_young, ocme_2014_young, ocme_2015_young, ocme_2016_young, ocme_2017_young)

ocmelistOld <- list(ocme_2009_old, ocme_2010_old, ocme_2011_old, ocme_2012_old, ocme_2013_old, ocme_2014_old, ocme_2015_old, ocme_2016_old, ocme_2017_old)
```



### Race of decedents {.tabset}

#### Young

```{r raceNumericYoung, echo=FALSE, message=FALSE, warning=FALSE}
for(i in 1:length(ocmelistYoung)){
  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  ocme <- ocmelistYoung[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- ocme$Year[1]
  
  
# see if I can automate the title, from https://stackoverflow.com/questions/59095237/automate-ggplots-while-using-variable-labels-as-title-and-axis-titles
racetitle <- paste("Racial distribution for", year, "CT opioid-detected overdose deaths in younger people")

race <- ggplot(data=ocme, aes(x=race_numeric_words, ..count..)) + 
      geom_bar(aes(fill = race_numeric_words), position = "dodge")  + scale_fill_manual(name = "Race", values = c("#E69F00", "#56B4E9", "#009E73", "#CC79A7"), labels = c("Black", "Hispanic", "Other", "White")) + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle(racetitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7)) 

# seem to need to use print() when in a for loop, from https://stackoverflow.com/questions/29053223/plots-not-working-in-for-loop
print(race)

# try to automate the filenames? from https://www.edureka.co/community/41963/r-programming-naming-the-output-file-using-a-variable
ggsave(plot=race, filename = paste0("images/race_", year, "_young.png"), width = 10, height = 8)

raceProp <- ggplot(data=ocme, aes(x=race_numeric_words, ..count..)) + 
      geom_bar(aes(fill = race_numeric_words), position = "dodge")  + scale_fill_manual(name = "Race", values = c("#E69F00", "#56B4E9", "#009E73", "#CC79A7"), labels = c("Black", "Hispanic", "Other", "White")) + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle(racetitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7)) 

print(raceProp)

#save this one for the presentation
ggsave(plot=raceProp, filename = paste0("images/race_proportions_", year, "_young.png"), width = 10, height = 8)
  
}

```

#### Old

```{r raceNumericOld, echo=FALSE, message=FALSE, warning=FALSE}
for(i in 1:length(ocmelistOld)){
  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  ocme <- ocmelistOld[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- ocme$Year[1]
  
  
# see if I can automate the title, from https://stackoverflow.com/questions/59095237/automate-ggplots-while-using-variable-labels-as-title-and-axis-titles
racetitle <- paste("Racial distribution for", year, "CT opioid-detected overdose deaths in older people")

race <- ggplot(data=ocme, aes(x=race_numeric_words, ..count..)) + 
      geom_bar(aes(fill = race_numeric_words), position = "dodge")  + scale_fill_manual(name = "Race", values = c("#E69F00", "#56B4E9", "#009E73", "#CC79A7"), labels = c("Black", "Hispanic", "Other", "White")) + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle(racetitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7)) 

# seem to need to use print() when in a for loop, from https://stackoverflow.com/questions/29053223/plots-not-working-in-for-loop
print(race)

# try to automate the filenames? from https://www.edureka.co/community/41963/r-programming-naming-the-output-file-using-a-variable
ggsave(plot=race, filename = paste0("images/race_", year, "_old.png"), width = 10, height = 8)

raceProp <- ggplot(data=ocme, aes(x=race_numeric_words, ..count..)) + 
      geom_bar(aes(fill = race_numeric_words), position = "dodge")  + scale_fill_manual(name = "Race", values = c("#E69F00", "#56B4E9", "#009E73", "#CC79A7"), labels = c("Black", "Hispanic", "Other", "White")) + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle(racetitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7)) 

print(raceProp)

#save this one for the presentation
ggsave(plot=raceProp, filename = paste0("images/race_proportions_", year, "_old.png"), width = 10, height = 8)
  
}

```


### Chi squared testing for race 


```{r chiSquaredRace, echo=FALSE, message=FALSE, warning=FALSE}
# initialize empty matrix, from https://stackoverflow.com/questions/32712301/create-empty-data-frame-with-column-names-by-assigning-a-string-vector
# this will hold all the p values

chisq_race <- data.frame(matrix(ncol = 2, nrow = 0))
cols <- c("Year", "p value")

for(i in 1:length(ocmelistYoung)){
  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  young <- ocmelistYoung[[i]] 
  old <- ocmelistOld[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- young$Year[1]
  
  racefreqtable <- young %>% group_by(race_numeric_words) %>% tally()
  names(racefreqtable)[names(racefreqtable) == 'n'] <- 'Young'

  racetableold <- old %>% group_by(race_numeric_words) %>% tally()
  names(racetableold)[names(racetableold) == 'n'] <- 'Old'

  racetable <- merge(racefreqtable, racetableold, by="race_numeric_words")
  
  # convert the "race" column into row names so I get a table with only numeric values, from   https://stackoverflow.com/questions/5555408/convert-the-values-in-a-column-into-row-names-in-an-existing-data-frame

  racemat <- racetable[,-1]
  rownames(racemat) <- racetable[,1]

  print(as.matrix(racemat))
  
  # append year and p value to a vector, rounding to make it more legible
  results <- c(year, round(chisq.test(as.matrix(racemat))$p.value, 4))
  
  # add this to the chi squared df 
  chisq_race <- rbind(chisq_race, results)

}

# for some reason it doesn't work if I do this before adding rows?
colnames(chisq_race) <- cols

chisq_race

```


### Sex of decedents {.tabset}

#### Young


```{r sexTableYoung, echo=FALSE, message=FALSE, warning=FALSE}
for(i in 1:length(ocmelistYoung)){

  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  ocme <- ocmelistYoung[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- ocme$Year[1]
  
  sextitle <- paste("Sex distribution for", year, "CT opioid-detected overdose deaths in younger people")

  
sex <- ggplot(data=ocme, aes(x=Sex_revised, ..count..)) + 
      geom_bar(aes(fill = Sex_revised), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle(sextitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

print(sex)

ggsave(plot=sex, filename = paste0("images/sex", year, "_young.png"), width = 10, height = 8)


sexProp <- ggplot(data=ocme, aes(x=Sex_revised, ..count..)) + 
      geom_bar(aes(fill = Sex_revised), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle(sextitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

print(sexProp)

ggsave(plot=sexProp, filename = paste0("images/sexProp", year, "_young.png"), width = 10, height = 8)

}

```


#### Old

```{r sexTableOld, echo=FALSE, message=FALSE, warning=FALSE}
for(i in 1:length(ocmelistOld)){

  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  ocme <- ocmelistOld[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- ocme$Year[1]
  
  sextitle <- paste("Sex distribution for", year, "CT opioid-detected overdose deaths in older people")

  
sex <- ggplot(data=ocme, aes(x=Sex_revised, ..count..)) + 
      geom_bar(aes(fill = Sex_revised), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black")+ ggtitle(sextitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

print(sex)

ggsave(plot=sex, filename = paste0("images/sex", year, "_old.png"), width = 10, height = 8)


sexProp <- ggplot(data=ocme, aes(x=Sex_revised, ..count..)) + 
      geom_bar(aes(fill = Sex_revised), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme), 3)), stat = "count", vjust = 0, colour = "black")+ ggtitle(sextitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none") 

print(sexProp)

ggsave(plot=sexProp, filename = paste0("images/sexProp", year, "_old.png"), width = 10, height = 8)

}

```


### Chi squared testing for sex 


```{r chiSquaredSex, echo=FALSE, message=FALSE, warning=FALSE}
# initialize empty matrix, from https://stackoverflow.com/questions/32712301/create-empty-data-frame-with-column-names-by-assigning-a-string-vector
# this will hold all the p values

chisq_sex <- data.frame(matrix(ncol = 2, nrow = 0))
cols <- c("Year", "p value")

for(i in 1:length(ocmelistYoung)){
  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  young <- ocmelistYoung[[i]] 
  old <- ocmelistOld[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- young$Year[1]
  
  sexfreqtable <- young %>% group_by(Sex_revised) %>% tally()
  names(sexfreqtable)[names(sexfreqtable) == 'n'] <- 'Young'

  sextableold <- old %>% group_by(Sex_revised) %>% tally()
  names(sextableold)[names(sextableold) == 'n'] <- 'Old'

  sextable <- merge(sexfreqtable, sextableold, by="Sex_revised")
  
  # convert the "race" column into row names so I get a table with only numeric values, from   https://stackoverflow.com/questions/5555408/convert-the-values-in-a-column-into-row-names-in-an-existing-data-frame

  sexmat <- sextable[,-1]
  rownames(sexmat) <- sextable[,1]

  print(as.matrix(sexmat))
  
  # append year and p value to a vector, rounding to make it more legible
  results <- c(year, round(chisq.test(as.matrix(sexmat))$p.value, 4))
  
  # add this to the chi squared df 
  chisq_sex <- rbind(chisq_sex, results)

}

# for some reason it doesn't work if I do this before adding rows?
colnames(chisq_sex) <- cols

chisq_sex

```

### Race and sex of decedents {.tabset}

#### Young


```{r sexRaceTableYoung, echo=FALSE, message=FALSE, warning=FALSE}
for(i in 1:length(ocmelistYoung)){

  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  ocme <- ocmelistYoung[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- ocme$Year[1]

raceSextitle <- paste("Race/ethnicity and sex distribution for", year, "CT opioid-detected \n overdose deaths in younger people")

  
raceSex <- ggplot(data=ocme, aes(x=race_sex, ..count..)) + 
      geom_bar(aes(fill = race_sex), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = -0.25, colour = "black", size = 7/.pt)+ ggtitle(raceSextitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7)) 

print(raceSex)

ggsave(plot=raceSex, filename = paste0("images/raceSex", year, "_young.png"), width = 10, height = 8)

raceSexProp <- ggplot(data=ocme, aes(x=race_sex, ..count..)) + 
      geom_bar(aes(fill = race_sex), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme), 3)), stat = "count", vjust = -0.25, colour = "black", size = 7/.pt)+ ggtitle(raceSextitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7)) 

print(raceSexProp)

ggsave(plot=raceSexProp, filename = paste0("images/raceSexProp", year, "_young.png"), width = 10, height = 8)



# get the % of each race sex category
ocme <- ocme %>% 
  group_by(race_sex) %>% 
  count() %>% 
  ungroup() %>% 
  mutate(per=`n`/sum(`n`)) %>% 
  arrange(desc(race_sex))
# create a label out of this
ocme$racesexlabel <- scales::percent(ocme$per)

# automatically assign the color based on race sex value -- not doing this in preprocessing because I don't want this to be a global change
# from https://stackoverflow.com/questions/39142596/r-ggplot-set-colour-for-specific-value

ocme$color <- ifelse(ocme$race_sex == "Black Female", "#e6d0a1", ifelse(ocme$race_sex == "Black Male", "#E69F00", ifelse(ocme$race_sex == "Hispanic Female", "#b3d4e8", ifelse(ocme$race_sex == "Hispanic Male", "#56B4E9", ifelse(ocme$race_sex == "Other Female", "#cca1b9", ifelse(ocme$race_sex == "Other Male", "#CC79A7", ifelse(ocme$race_sex == "White Female", "#a1a1a1", ifelse(ocme$race_sex == "White Male", "#000000", "#D9D9D9"))))))))


# plot into a pie chart -- for some reason I have to reverse the colors to get them to work right
piechart <- ggplot(data=ocme)+
  geom_bar(aes(x="", y=per, fill=race_sex), stat="identity", width = 1)+
  coord_polar("y", start=0)+
  theme_void()+ geom_label_repel(aes(label = racesexlabel, x = "", y = cumsum(per) - per/2), nudge_x = 0.02,
                   size = 3, show.legend = F)+  scale_fill_manual(name = "Race/ethnicity and sex", values=rev(ocme$color), labels = rev(ocme$race_sex))+ ggtitle(raceSextitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="right")  
# geom_text(aes(x=1, y = cumsum(per) - per/2, label=label)) for regular labels [not repelling]


print(piechart)

ggsave(plot=piechart, filename = paste0("images/pie", year, "_young.png"), width = 10, height = 8)





}
```


#### Old


```{r sexRaceTableOld, echo=FALSE, message=FALSE, warning=FALSE}
for(i in 1:length(ocmelistOld)){

  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  ocme <- ocmelistOld[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- ocme$Year[1]

raceSextitle <- paste("Race/ethnicity and sex distribution for", year, "CT opioid-detected \n overdose deaths in older people")

  
raceSex <- ggplot(data=ocme, aes(x=race_sex, ..count..)) + 
      geom_bar(aes(fill = race_sex), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = -0.25, colour = "black", size = 7/.pt)+ ggtitle(raceSextitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7)) 

print(raceSex)

ggsave(plot=raceSex, filename = paste0("images/raceSex", year, "_old.png"), width = 10, height = 8)

raceSexProp <- ggplot(data=ocme, aes(x=race_sex, ..count..)) + 
      geom_bar(aes(fill = race_sex), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme), 3)), stat = "count", vjust = -0.25, colour = "black", size = 7/.pt)+ ggtitle(raceSextitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7)) 

print(raceSexProp)

ggsave(plot=raceSexProp, filename = paste0("images/raceSexProp", year, "_old.png"), width = 10, height = 8)




# get the % of each race sex category
ocme <- ocme %>% 
  group_by(race_sex) %>% 
  count() %>% 
  ungroup() %>% 
  mutate(per=`n`/sum(`n`)) %>% 
  arrange(desc(race_sex))
# create a label out of this
ocme$racesexlabel <- scales::percent(ocme$per)

# automatically assign the color based on race sex value -- not doing this in preprocessing because I don't want this to be a global change
# from https://stackoverflow.com/questions/39142596/r-ggplot-set-colour-for-specific-value

ocme$color <- ifelse(ocme$race_sex == "Black Female", "#e6d0a1", ifelse(ocme$race_sex == "Black Male", "#E69F00", ifelse(ocme$race_sex == "Hispanic Female", "#b3d4e8", ifelse(ocme$race_sex == "Hispanic Male", "#56B4E9", ifelse(ocme$race_sex == "Other Female", "#cca1b9", ifelse(ocme$race_sex == "Other Male", "#CC79A7", ifelse(ocme$race_sex == "White Female", "#a1a1a1", ifelse(ocme$race_sex == "White Male", "#000000", "#D9D9D9"))))))))


# plot into a pie chart -- for some reason I have to reverse the colors to get them to work right
piechart <- ggplot(data=ocme)+
  geom_bar(aes(x="", y=per, fill=race_sex), stat="identity", width = 1)+
  coord_polar("y", start=0)+
  theme_void()+ geom_label_repel(aes(label = racesexlabel, x = "", y = cumsum(per) - per/2), nudge_x = 0.02,
                   size = 3, show.legend = F)+  scale_fill_manual(name = "Race/ethnicity and sex", values=rev(ocme$color), labels = rev(ocme$race_sex))+ ggtitle(raceSextitle)+ theme(plot.title = element_text(hjust = 0.5), legend.position="right")  
# geom_text(aes(x=1, y = cumsum(per) - per/2, label=label)) for regular labels [not repelling]


print(piechart)

ggsave(plot=piechart, filename = paste0("images/pie", year, "_old.png"), width = 10, height = 8)

}
```


### Chi squared testing for race and sex

```{r chiSquaredRaceSex, echo=FALSE, message=FALSE, warning=FALSE}
# initialize empty matrix, from https://stackoverflow.com/questions/32712301/create-empty-data-frame-with-column-names-by-assigning-a-string-vector
# this will hold all the p values

chisq_racesex <- data.frame(matrix(ncol = 2, nrow = 0))
cols <- c("Year", "p value")

for(i in 1:length(ocmelistYoung)){
  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  young <- ocmelistYoung[[i]] 
  old <- ocmelistOld[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- young$Year[1]
  
  racesexfreqtable <- young %>% group_by(race_sex) %>% tally()
  names(racesexfreqtable)[names(racesexfreqtable) == 'n'] <- 'Young'

  racesextableold <- old %>% group_by(race_sex) %>% tally()
  names(racesextableold)[names(racesextableold) == 'n'] <- 'Old'

  racesextable <- merge(racesexfreqtable, racesextableold, by="race_sex")
  
  # convert the "race" column into row names so I get a table with only numeric values, from   https://stackoverflow.com/questions/5555408/convert-the-values-in-a-column-into-row-names-in-an-existing-data-frame

  racesexmat <- racesextable[,-1]
  rownames(racesexmat) <- racesextable[,1]

  print(as.matrix(racesexmat))
  
  # append year and p value to a vector, rounding to make it more legible
  results <- c(year, round(chisq.test(as.matrix(racesexmat))$p.value, 4))
  
  # add this to the chi squared df 
  chisq_racesex <- rbind(chisq_racesex, results)

}

# for some reason it doesn't work if I do this before adding rows?
colnames(chisq_racesex) <- cols

chisq_racesex

```

### Day of week {.tabset}

#### Young

```{r weekYoung, echo=FALSE, message=FALSE, warning=FALSE}
for(i in 1:length(ocmelistYoung)){

  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  ocme <- ocmelistYoung[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- ocme$Year[1]

weekdaytitle <- paste("ODs per weekday for", year, "CT opioid-detected overdose deaths in younger people")

  
# ordering of variables on x axis from https://www.statology.org/ggplot-order-x-axis/
# make sure to use scale_x_discrete!

weekday <- ggplot(data=ocme, aes(x=factor(Weekday, level=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")), ..count..)) + geom_bar(aes(fill = as.character(Weekday)), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 0), breaks = ocme$Weekday, labels = ocme$Weekday) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle(weekdaytitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))+xlab("Weekday")

print(weekday)

ggsave(plot=weekday, filename = paste0("images/weekday_", year, "_young.png"), width = 10, height = 8)


weekday_prop <- ggplot(data=ocme, aes(x=factor(Weekday, level=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")), round(..count../nrow(ocme), 3))) + geom_bar(aes(fill = as.character(Weekday)), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 0), breaks = ocme$Weekday, labels = ocme$Weekday) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme), 3)), stat = "count", vjust = 0, colour = "black") + ggtitle(weekdaytitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))+xlab("Weekday")+ylab("Proportion")

print(weekday_prop)

ggsave(plot=weekday_prop, filename = paste0("images/weekday_", year, "_young_prop.png"), width = 10, height = 8)
}
```


#### Old

```{r weekOld, echo=FALSE, message=FALSE, warning=FALSE}
for(i in 1:length(ocmelistOld)){

  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  ocme <- ocmelistOld[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- ocme$Year[1]

weekdaytitle <- paste("ODs per weekday for", year, "CT opioid-detected overdose deaths in older people")

  
# ordering of variables on x axis from https://www.statology.org/ggplot-order-x-axis/
# make sure to use scale_x_discrete!

weekday <- ggplot(data=ocme, aes(x=factor(Weekday, level=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")), ..count..)) + geom_bar(aes(fill = as.character(Weekday)), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 0), breaks = ocme$Weekday, labels = ocme$Weekday) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle(weekdaytitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))+xlab("Weekday")

print(weekday)

ggsave(plot=weekday, filename = paste0("images/weekday_", year, "_old.png"), width = 10, height = 8)


weekday_prop <- ggplot(data=ocme, aes(x=factor(Weekday, level=c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")), round(..count../nrow(ocme), 3))) + geom_bar(aes(fill = as.character(Weekday)), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 0), breaks = ocme$Weekday, labels = ocme$Weekday) + theme_light() + geom_text(aes(label = round(..count../nrow(ocme), 3)), stat = "count", vjust = 0, colour = "black") + ggtitle(weekdaytitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))+xlab("Weekday")+ylab("Proportion")

print(weekday_prop)

ggsave(plot=weekday_prop, filename = paste0("images/weekday_", year, "_old_prop.png"), width = 10, height = 8)
}
```


### Chi squared testing for day of week

```{r chiSquaredWeekday, echo=FALSE, message=FALSE, warning=FALSE}
# initialize empty matrix, from https://stackoverflow.com/questions/32712301/create-empty-data-frame-with-column-names-by-assigning-a-string-vector
# this will hold all the p values

chisq_weekday <- data.frame(matrix(ncol = 2, nrow = 0))
cols <- c("Year", "p value")

for(i in 1:length(ocmelistYoung)){
  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  young <- ocmelistYoung[[i]] 
  old <- ocmelistOld[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- young$Year[1]
  
  weekdayfreqtable <- young %>% group_by(Weekday) %>% tally()
  names(weekdayfreqtable)[names(weekdayfreqtable) == 'n'] <- 'Young'

  weekdaytableold <- old %>% group_by(Weekday) %>% tally()
  names(weekdaytableold)[names(weekdaytableold) == 'n'] <- 'Old'

  weekdaytable <- merge(weekdayfreqtable, weekdaytableold, by="Weekday")
  
  # convert the "race" column into row names so I get a table with only numeric values, from   https://stackoverflow.com/questions/5555408/convert-the-values-in-a-column-into-row-names-in-an-existing-data-frame

  weekdaymat <- weekdaytable[,-1]
  rownames(weekdaymat) <- weekdaytable[,1]

  print(as.matrix(weekdaymat))
  
  # append year and p value to a vector, rounding to make it more legible
  results <- c(year, round(chisq.test(as.matrix(weekdaymat))$p.value, 4))
  
  # add this to the chi squared df 
  chisq_weekday <- rbind(chisq_weekday, results)

}

# for some reason it doesn't work if I do this before adding rows?
colnames(chisq_weekday) <- cols

chisq_weekday

```

### Month {.tabset}

#### Young

```{r month2018Young, echo=FALSE, message=FALSE, warning=FALSE}
for(i in 1:length(ocmelistYoung)){

  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  ocme <- ocmelistYoung[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- ocme$Year[1]
  
  monthtitle <- paste("ODs per month for", year, "CT opioid-detected overdose deaths in younger people")

  
month <- ggplot(data=ocme, aes(x=month, ..count..)) + geom_bar(aes(fill = as.character(month)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0), breaks = ocme$month, labels = ocme$month) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle(monthtitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))

print(month)

ggsave(plot=month, filename = paste0("images/month_", year, "_young.png"), width = 10, height = 8)

month_prop <- ggplot(data=ocme, aes(x=month, round(..count../nrow(ocme), 3))) + geom_bar(aes(fill = as.character(month)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0), breaks = ocme$month, labels = ocme$month) + theme_light() + geom_text(aes(label =  round(..count../nrow(ocme), 3)), stat = "count", vjust = 0, colour = "black") + ggtitle(monthtitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))+ylab("Proportion")

print(month_prop)

ggsave(plot=month_prop, filename = paste0("images/month_", year, "_young_prop.png"), width = 10, height = 8)


}

```

#### Old

```{r month2018Old, echo=FALSE, message=FALSE, warning=FALSE}
for(i in 1:length(ocmelistOld)){

  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  ocme <- ocmelistOld[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- ocme$Year[1]
  
  monthtitle <- paste("ODs per month for", year, "CT opioid-detected overdose deaths in older people")

  
month <- ggplot(data=ocme, aes(x=month, ..count..)) + geom_bar(aes(fill = as.character(month)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0), breaks = ocme$month, labels = ocme$month) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black") + ggtitle(monthtitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))

print(month)

ggsave(plot=month, filename = paste0("images/month_", year, "_old.png"), width = 10, height = 8)

month_prop <- ggplot(data=ocme, aes(x=month, round(..count../nrow(ocme), 3))) + geom_bar(aes(fill = as.character(month)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 0), breaks = ocme$month, labels = ocme$month) + theme_light() + geom_text(aes(label =  round(..count../nrow(ocme), 3)), stat = "count", vjust = 0, colour = "black") + ggtitle(monthtitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))+ylab("Proportion")

print(month_prop)

ggsave(plot=month_prop, filename = paste0("images/month_", year, "_old_prop.png"), width = 10, height = 8)


}

```


### Chi squared testing for month

```{r chiSquaredMonth, echo=FALSE, message=FALSE, warning=FALSE}
# initialize empty matrix, from https://stackoverflow.com/questions/32712301/create-empty-data-frame-with-column-names-by-assigning-a-string-vector
# this will hold all the p values

chisq_month <- data.frame(matrix(ncol = 2, nrow = 0))
cols <- c("Year", "p value")

for(i in 1:length(ocmelistYoung)){
  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  young <- ocmelistYoung[[i]] 
  old <- ocmelistOld[[i]] 
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- young$Year[1]
  
  monthfreqtable <- young %>% group_by(month) %>% tally()
  names(monthfreqtable)[names(monthfreqtable) == 'n'] <- 'Young'

  monthtableold <- old %>% group_by(month) %>% tally()
  names(monthtableold)[names(monthtableold) == 'n'] <- 'Old'

  monthtable <- merge(monthfreqtable, monthtableold, by="month")
  
  # convert the "race" column into row names so I get a table with only numeric values, from   https://stackoverflow.com/questions/5555408/convert-the-values-in-a-column-into-row-names-in-an-existing-data-frame

  monthmat <- monthtable[,-1]
  rownames(monthmat) <- monthtable[,1]

  print(as.matrix(monthmat))
  
  # append year and p value to a vector, rounding to make it more legible
  results <- c(year, round(chisq.test(as.matrix(monthmat))$p.value, 4))
  
  # add this to the chi squared df 
  chisq_month <- rbind(chisq_month, results)

}

# for some reason it doesn't work if I do this before adding rows?
colnames(chisq_month) <- cols

chisq_month

```


### Proportion of deaths involving fentanyl [will combine with the other plot!]


```{r propFent, echo=FALSE, message=FALSE, warning=FALSE}

# unfortunately there are different variable names for each ocme file so a for loop won't work :/

# get proportion of fent for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
fentyoung <- c(sum(ocme_2009_young$fent_or_frankens == 1)/nrow(ocme_2009_young), sum(ocme_2010_young$fent_or_frankens == 1)/nrow(ocme_2010_young), sum(ocme_2011_young$fent_or_frankens == 1)/nrow(ocme_2011_young), sum(ocme_2012_young$fent_or_frankens == 1)/nrow(ocme_2012_young), sum(ocme_2013_young$fent_or_frankens == 1)/nrow(ocme_2013_young), sum(ocme_2014_young$fent_or_frankens == 1)/nrow(ocme_2014_young), sum(ocme_2015_young$fent_or_frankens == 1)/nrow(ocme_2015_young), sum(ocme_2016_young$fent_or_frankens == 1)/nrow(ocme_2016_young), sum(ocme_2017_young$fent_or_frankens == 1)/nrow(ocme_2017_young))

# get proportion for older people
fentold <- c(sum(ocme_2009_old$fent_or_frankens == 1)/nrow(ocme_2009_old), sum(ocme_2010_old$fent_or_frankens == 1)/nrow(ocme_2010_old), sum(ocme_2011_old$fent_or_frankens == 1)/nrow(ocme_2011_old), sum(ocme_2012_old$fent_or_frankens == 1)/nrow(ocme_2012_old), sum(ocme_2013_old$fent_or_frankens == 1)/nrow(ocme_2013_old), sum(ocme_2014_old$fent_or_frankens == 1)/nrow(ocme_2014_old), sum(ocme_2015_old$fent_or_frankens == 1)/nrow(ocme_2015_old), sum(ocme_2016_old$fent_or_frankens == 1)/nrow(ocme_2016_old), sum(ocme_2017_old$fent_or_frankens == 1)/nrow(ocme_2017_old))

# combine vectors into df
propfent <- data.frame(yearList, fentyoung, fentold)

# change the col names to be more informative
names(propfent) <- c("Year", "Younger", "Older")

# create grouped bar plot, from https://stackoverflow.com/questions/72449952/making-a-grouped-bar-chart-in-r
# getting the labels in the right position from https://stackoverflow.com/questions/37587288/ggplot-label-bars-in-grouped-bar-plot
# info on changing legend title from here https://stackoverflow.com/questions/41649929/how-to-modify-a-legend-title-in-a-barplot-with-ggplot
# the scale x continuous is there because for some reason adding 2023 data messed up the x axis, see https://stackoverflow.com/questions/72051789/how-to-display-all-values-on-x-and-y-axes-in-r-ggplot2

pfent <- propfent |>
  # Reshape data to long format
  pivot_longer(Younger:Older) |> 
  ggplot(aes(x = Year, y = value, fill = name)) +
  # Implement a grouped bar chart
  geom_bar(stat = "identity", position = position_dodge(width= 1)) + geom_text(aes(label = round(value, 2)), stat = "identity", vjust = -0.1, colour = "black", size = 9/.pt, position = position_dodge(width= 1)) + theme_light() + ggtitle("Proportion of fentanyl-detected deaths by age group in CT, 2009-2017") + ylab("Proportion") + theme(plot.title = element_text(hjust = 0.5)) + scale_fill_discrete(name = "Age group") + scale_x_continuous(breaks=propfent$Year)+ coord_cartesian(ylim = c(0, 1)) 

pfent

ggsave(plot=pfent, filename = "images/prop_fent_old.png", width = 10, height = 8)
```


### Chi squared testing for the number of fentanyl-detected deaths

```{r chiSquaredFent, echo=FALSE, message=FALSE, warning=FALSE}
# get frequency of each race for old and young, from https://stackoverflow.com/questions/48562288/frequency-of-each-unique-combination-in-data-frame

# get proportion of fent for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
fentyoungcount <- c(sum(ocme_2009_young$fent_or_frankens == 1), sum(ocme_2010_young$fent_or_frankens == 1), sum(ocme_2011_young$fent_or_frankens == 1), sum(ocme_2012_young$fent_or_frankens == 1), sum(ocme_2013_young$fent_or_frankens == 1), sum(ocme_2014_young$fent_or_frankens == 1), sum(ocme_2015_young$fent_or_frankens == 1), sum(ocme_2016_young$fent_or_frankens == 1), sum(ocme_2017_young$fent_or_frankens == 1))

# get proportion for older people
fentoldcount <- c(sum(ocme_2009_old$fent_or_frankens == 1), sum(ocme_2010_old$fent_or_frankens == 1), sum(ocme_2011_old$fent_or_frankens == 1), sum(ocme_2012_old$fent_or_frankens == 1), sum(ocme_2013_old$fent_or_frankens == 1), sum(ocme_2014_old$fent_or_frankens == 1), sum(ocme_2015_old$fent_or_frankens == 1), sum(ocme_2016_old$fent_or_frankens == 1), sum(ocme_2017_old$fent_or_frankens == 1))

# combine vectors into df
fentcount <- data.frame(yearList, fentyoungcount, fentoldcount)

# change the col names to be more informative
names(fentcount) <- c("Year", "Younger", "Older")

# convert the "race_sex" column into row names so I get a table with only numeric values, from https://stackoverflow.com/questions/5555408/convert-the-values-in-a-column-into-row-names-in-an-existing-data-frame

fentchi <- fentcount[,-1]
rownames(fentchi) <- fentcount[,1]

as.matrix(fentchi)

# looks like I can just use any matrix for chisq.test, from https://researchguides.library.vanderbilt.edu/c.php?g=156859&p=3058795

chisq.test(as.matrix(fentchi))
```

### Proportion of deaths involving xylazine

```{r propXyl, echo=FALSE, message=FALSE, warning=FALSE}

# unfortunately there are different variable names for each ocme file so a for loop won't work :/

# get proportion of fent for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
xylyoung <- c(sum(ocme_2009_young$xylazine == 1)/nrow(ocme_2009_young), sum(ocme_2010_young$xylazine == 1)/nrow(ocme_2010_young), sum(ocme_2011_young$xylazine == 1)/nrow(ocme_2011_young), sum(ocme_2012_young$xylazine == 1)/nrow(ocme_2012_young), sum(ocme_2013_young$xylazine == 1)/nrow(ocme_2013_young), sum(ocme_2014_young$xylazine == 1)/nrow(ocme_2014_young), sum(ocme_2015_young$xylazine == 1)/nrow(ocme_2015_young), sum(ocme_2016_young$xylazine == 1)/nrow(ocme_2016_young), sum(ocme_2017_young$xylazine == 1)/nrow(ocme_2017_young))

# get proportion for older people
xylold <- c(sum(ocme_2009_old$xylazine == 1)/nrow(ocme_2009_old), sum(ocme_2010_old$xylazine == 1)/nrow(ocme_2010_old), sum(ocme_2011_old$xylazine == 1)/nrow(ocme_2011_old), sum(ocme_2012_old$xylazine == 1)/nrow(ocme_2012_old), sum(ocme_2013_old$xylazine == 1)/nrow(ocme_2013_old), sum(ocme_2014_old$xylazine == 1)/nrow(ocme_2014_old), sum(ocme_2015_old$xylazine == 1)/nrow(ocme_2015_old), sum(ocme_2016_old$xylazine == 1)/nrow(ocme_2016_old), sum(ocme_2017_old$xylazine == 1)/nrow(ocme_2017_old))


# combine vectors into df
propxyl <- data.frame(yearList, xylyoung, xylold)

# change the col names to be more informative
names(propxyl) <- c("Year", "Younger", "Older")

# create grouped bar plot, from https://stackoverflow.com/questions/72449952/making-a-grouped-bar-chart-in-r
# getting the labels in the right position from https://stackoverflow.com/questions/37587288/ggplot-label-bars-in-grouped-bar-plot
# info on changing legend title from here https://stackoverflow.com/questions/41649929/how-to-modify-a-legend-title-in-a-barplot-with-ggplot

pxyl <- propxyl |>
  # Reshape data to long format
  pivot_longer(Younger:Older) |> 
  ggplot(aes(x = Year, y = value, fill = name)) +
  # Implement a grouped bar chart
  geom_bar(stat = "identity", position = position_dodge(width= 1)) + geom_text(aes(label = round(value, 2)), stat = "identity", vjust = -0.1, colour = "black", size = 9/.pt, position = position_dodge(width= 1)) + theme_light() + ggtitle("Proportion of xylazine-detected deaths by age group in CT, 2009-2017") + ylab("Proportion") + theme(plot.title = element_text(hjust = 0.5)) + scale_fill_discrete(name = "Age group") + scale_x_continuous(breaks=propxyl$Year)+ coord_cartesian(ylim = c(0, 1))

pxyl

ggsave(plot=pxyl, filename = "images/prop_xyl_old.png", width = 10, height = 8)
```


### Chi squared testing for the number of xylazine-detected deaths

No point in doing it! No one tested positive in these years. 


### Proportion of deaths involving gabapentin

```{r propGab, echo=FALSE, message=FALSE, warning=FALSE}

# unfortunately there are different variable names for each ocme file so a for loop won't work :/

# get proportion of gabapentin for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
gabyoung <- c(sum(ocme_2009_young$gabapentin == 1)/nrow(ocme_2009_young), sum(ocme_2010_young$gabapentin == 1)/nrow(ocme_2010_young), sum(ocme_2011_young$gabapentin == 1)/nrow(ocme_2011_young), sum(ocme_2012_young$gabapentin == 1)/nrow(ocme_2012_young), sum(ocme_2013_young$gabapentin == 1)/nrow(ocme_2013_young), sum(ocme_2014_young$gabapentin == 1)/nrow(ocme_2014_young), sum(ocme_2015_young$gabapentin == 1)/nrow(ocme_2015_young), sum(ocme_2016_young$gabapentin == 1)/nrow(ocme_2016_young), sum(ocme_2017_young$gabapentin == 1)/nrow(ocme_2017_young))

# get proportion for older people
gabold <- c(sum(ocme_2009_old$gabapentin == 1)/nrow(ocme_2009_old), sum(ocme_2010_old$gabapentin == 1)/nrow(ocme_2010_old), sum(ocme_2011_old$gabapentin == 1)/nrow(ocme_2011_old), sum(ocme_2012_old$gabapentin == 1)/nrow(ocme_2012_old), sum(ocme_2013_old$gabapentin == 1)/nrow(ocme_2013_old), sum(ocme_2014_old$gabapentin == 1)/nrow(ocme_2014_old), sum(ocme_2015_old$gabapentin == 1)/nrow(ocme_2015_old), sum(ocme_2016_old$gabapentin == 1)/nrow(ocme_2016_old), sum(ocme_2017_old$gabapentin == 1)/nrow(ocme_2017_old))

# combine vectors into df
propgab <- data.frame(yearList, gabyoung, gabold)

# change the col names to be more informative
names(propgab) <- c("Year", "Younger", "Older")

# create grouped bar plot, from https://stackoverflow.com/questions/72449952/making-a-grouped-bar-chart-in-r
# getting the labels in the right position from https://stackoverflow.com/questions/37587288/ggplot-label-bars-in-grouped-bar-plot
# info on changing legend title from here https://stackoverflow.com/questions/41649929/how-to-modify-a-legend-title-in-a-barplot-with-ggplot

pgab <- propgab |>
  # Reshape data to long format
  pivot_longer(Younger:Older) |> 
  ggplot(aes(x = Year, y = value, fill = name)) +
  # Implement a grouped bar chart
  geom_bar(stat = "identity", position = position_dodge(width= 1)) + geom_text(aes(label = round(value, 2)), stat = "identity", vjust = -0.1, colour = "black", size = 9/.pt, position = position_dodge(width= 1)) + theme_light() + ggtitle("Proportion of gabapentin-detected deaths by age group in CT, 2009-2017") + ylab("Proportion") + theme(plot.title = element_text(hjust = 0.5)) + scale_fill_discrete(name = "Age group") + scale_x_continuous(breaks=propgab$Year)+ coord_cartesian(ylim = c(0, 1))

pgab

ggsave(plot=pgab, filename = "images/prop_gab_old.png", width = 10, height = 8)
```

### Chi squared testing for the number of gabepentin-detected deaths

```{r chiSquaredGab, echo=FALSE, message=FALSE, warning=FALSE}
# get frequency of each race for old and young, from https://stackoverflow.com/questions/48562288/frequency-of-each-unique-combination-in-data-frame

# get proportion of gab for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
gabyoungcount <- c(sum(ocme_2009_young$gabapentin == 1), sum(ocme_2010_young$gabapentin == 1), sum(ocme_2011_young$gabapentin == 1), sum(ocme_2012_young$gabapentin == 1), sum(ocme_2013_young$gabapentin == 1), sum(ocme_2014_young$gabapentin == 1), sum(ocme_2015_young$gabapentin == 1), sum(ocme_2016_young$gabapentin == 1), sum(ocme_2017_young$gabapentin == 1))

# get proportion for older people
gaboldcount <- c(sum(ocme_2009_old$gabapentin == 1), sum(ocme_2010_old$gabapentin == 1), sum(ocme_2011_old$gabapentin == 1), sum(ocme_2012_old$gabapentin == 1), sum(ocme_2013_old$gabapentin == 1), sum(ocme_2014_old$gabapentin == 1), sum(ocme_2015_old$gabapentin == 1), sum(ocme_2016_old$gabapentin == 1), sum(ocme_2017_old$gabapentin == 1))

# combine vectors into df
gabcount <- data.frame(yearList, gabyoungcount, gaboldcount)

# change the col names to be more informative
names(gabcount) <- c("Year", "Younger", "Older")

# convert the "race_sex" column into row names so I get a table with only numeric values, from https://stackoverflow.com/questions/5555408/convert-the-values-in-a-column-into-row-names-in-an-existing-data-frame

gabchi <- gabcount[,-1]
rownames(gabchi) <- gabcount[,1]

as.matrix(gabchi)

# looks like I can just use any matrix for chisq.test, from https://researchguides.library.vanderbilt.edu/c.php?g=156859&p=3058795
# getting rid of the nonzero entries for testing

chisq.test(tail(as.matrix(gabchi), -6))
```

### Proportion of deaths involving cocaine

```{r propCoke, echo=FALSE, message=FALSE, warning=FALSE}

# unfortunately there are different variable names for each ocme file so a for loop won't work :/

# get proportion of cocaine for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
cokeyoung <- c(sum(ocme_2009_young$cocaine == 1)/nrow(ocme_2009_young), sum(ocme_2010_young$cocaine == 1)/nrow(ocme_2010_young), sum(ocme_2011_young$cocaine == 1)/nrow(ocme_2011_young), sum(ocme_2012_young$cocaine == 1)/nrow(ocme_2012_young), sum(ocme_2013_young$cocaine == 1)/nrow(ocme_2013_young), sum(ocme_2014_young$cocaine == 1)/nrow(ocme_2014_young), sum(ocme_2015_young$cocaine == 1)/nrow(ocme_2015_young), sum(ocme_2016_young$cocaine == 1)/nrow(ocme_2016_young), sum(ocme_2017_young$cocaine == 1)/nrow(ocme_2017_young))

# get proportion for older people
cokeold <- c(sum(ocme_2009_old$cocaine == 1)/nrow(ocme_2009_old), sum(ocme_2010_old$cocaine == 1)/nrow(ocme_2010_old), sum(ocme_2011_old$cocaine == 1)/nrow(ocme_2011_old), sum(ocme_2012_old$cocaine == 1)/nrow(ocme_2012_old), sum(ocme_2013_old$cocaine == 1)/nrow(ocme_2013_old), sum(ocme_2014_old$cocaine == 1)/nrow(ocme_2014_old), sum(ocme_2015_old$cocaine == 1)/nrow(ocme_2015_old), sum(ocme_2016_old$cocaine == 1)/nrow(ocme_2016_old), sum(ocme_2017_old$cocaine == 1)/nrow(ocme_2017_old))

# combine vectors into df
propcoke <- data.frame(yearList, cokeyoung, cokeold)

# change the col names to be more informative
names(propcoke) <- c("Year", "Younger", "Older")

# create grouped bar plot, from https://stackoverflow.com/questions/72449952/making-a-grouped-bar-chart-in-r
# getting the labels in the right position from https://stackoverflow.com/questions/37587288/ggplot-label-bars-in-grouped-bar-plot
# info on changing legend title from here https://stackoverflow.com/questions/41649929/how-to-modify-a-legend-title-in-a-barplot-with-ggplot

pcoke <- propcoke |>
  # Reshape data to long format
  pivot_longer(Younger:Older) |> 
  ggplot(aes(x = Year, y = value, fill = name)) +
  # Implement a grouped bar chart
  geom_bar(stat = "identity", position = position_dodge(width= 1)) + geom_text(aes(label = round(value, 2)), stat = "identity", vjust = -0.1, colour = "black", size = 9/.pt, position = position_dodge(width= 1)) + theme_light() + ggtitle("Proportion of cocaine-detected deaths by age group in CT, 2009-2017") + ylab("Proportion") + theme(plot.title = element_text(hjust = 0.5)) + scale_fill_discrete(name = "Age group") + scale_x_continuous(breaks=propcoke$Year)+ coord_cartesian(ylim = c(0, 1))

pcoke

ggsave(plot=pcoke, filename = "images/prop_coke_old.png", width = 10, height = 8)
```


### Chi squared testing for the number of cocaine-detected deaths

```{r chiSquaredCoke, echo=FALSE, message=FALSE, warning=FALSE}
# get frequency of each race for old and young, from https://stackoverflow.com/questions/48562288/frequency-of-each-unique-combination-in-data-frame

# get proportion of coke for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
cokeyoungcount <- c(sum(ocme_2009_young$cocaine == 1), sum(ocme_2010_young$cocaine == 1), sum(ocme_2011_young$cocaine == 1), sum(ocme_2012_young$cocaine == 1), sum(ocme_2013_young$cocaine == 1), sum(ocme_2014_young$cocaine == 1), sum(ocme_2015_young$cocaine == 1), sum(ocme_2016_young$cocaine == 1), sum(ocme_2017_young$cocaine == 1))

# get proportion for older people
cokeoldcount <- c(sum(ocme_2009_old$cocaine == 1), sum(ocme_2010_old$cocaine == 1), sum(ocme_2011_old$cocaine == 1), sum(ocme_2012_old$cocaine == 1), sum(ocme_2013_old$cocaine == 1), sum(ocme_2014_old$cocaine == 1), sum(ocme_2015_old$cocaine == 1), sum(ocme_2016_old$cocaine == 1), sum(ocme_2017_old$cocaine == 1))

# combine vectors into df
cokecount <- data.frame(yearList, cokeyoungcount, cokeoldcount)

# change the col names to be more informative
names(cokecount) <- c("Year", "Younger", "Older")

# convert the "race_sex" column into row names so I get a table with only numeric values, from https://stackoverflow.com/questions/5555408/convert-the-values-in-a-column-into-row-names-in-an-existing-data-frame

cokechi <- cokecount[,-1]
rownames(cokechi) <- cokecount[,1]

as.matrix(cokechi)

# looks like I can just use any matrix for chisq.test, from https://researchguides.library.vanderbilt.edu/c.php?g=156859&p=3058795

chisq.test(as.matrix(cokechi))
```

### Proportion of deaths involving alcohol

```{r propEtOH, echo=FALSE, message=FALSE, warning=FALSE}

# unfortunately there are different variable names for each ocme file so a for loop won't work :/

# get proportion of alcohol for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
etohyoung <- c(sum(ocme_2009_young$et_oh == 1)/nrow(ocme_2009_young), sum(ocme_2010_young$et_oh == 1)/nrow(ocme_2010_young), sum(ocme_2011_young$et_oh == 1)/nrow(ocme_2011_young), sum(ocme_2012_young$et_oh == 1)/nrow(ocme_2012_young), sum(ocme_2013_young$et_oh == 1)/nrow(ocme_2013_young), sum(ocme_2014_young$et_oh == 1)/nrow(ocme_2014_young), sum(ocme_2015_young$et_oh == 1)/nrow(ocme_2015_young), sum(ocme_2016_young$et_oh == 1)/nrow(ocme_2016_young), sum(ocme_2017_young$et_oh == 1)/nrow(ocme_2017_young))

# get proportion for older people
etohold <- c(sum(ocme_2009_old$et_oh == 1)/nrow(ocme_2009_old), sum(ocme_2010_old$et_oh == 1)/nrow(ocme_2010_old), sum(ocme_2011_old$et_oh == 1)/nrow(ocme_2011_old), sum(ocme_2012_old$et_oh == 1)/nrow(ocme_2012_old), sum(ocme_2013_old$et_oh == 1)/nrow(ocme_2013_old), sum(ocme_2014_old$et_oh == 1)/nrow(ocme_2014_old), sum(ocme_2015_old$et_oh == 1)/nrow(ocme_2015_old), sum(ocme_2016_old$et_oh == 1)/nrow(ocme_2016_old), sum(ocme_2017_old$et_oh == 1)/nrow(ocme_2017_old))

# combine vectors into df
propetoh <- data.frame(yearList, etohyoung, etohold)

# change the col names to be more informative
names(propetoh) <- c("Year", "Younger", "Older")

# create grouped bar plot, from https://stackoverflow.com/questions/72449952/making-a-grouped-bar-chart-in-r
# getting the labels in the right position from https://stackoverflow.com/questions/37587288/ggplot-label-bars-in-grouped-bar-plot
# info on changing legend title from here https://stackoverflow.com/questions/41649929/how-to-modify-a-legend-title-in-a-barplot-with-ggplot

petoh <- propetoh |>
  # Reshape data to long format
  pivot_longer(Younger:Older) |> 
  ggplot(aes(x = Year, y = value, fill = name)) +
  # Implement a grouped bar chart
  geom_bar(stat = "identity", position = position_dodge(width= 1)) + geom_text(aes(label = round(value, 2)), stat = "identity", vjust = -0.1, colour = "black", size = 9/.pt, position = position_dodge(width= 1)) + theme_light() + ggtitle("Proportion of alcohol-detected deaths by age group in CT, 2009-2017") + ylab("Proportion") + theme(plot.title = element_text(hjust = 0.5)) + scale_fill_discrete(name = "Age group") + scale_x_continuous(breaks=propetoh$Year)+ coord_cartesian(ylim = c(0, 1))

petoh

ggsave(plot=petoh, filename = "images/prop_etoh_old.png", width = 10, height = 8)
```


### Chi squared testing for the number of alcohol-detected deaths

```{r chiSquaredEtoh, echo=FALSE, message=FALSE, warning=FALSE}
# get frequency of each race for old and young, from https://stackoverflow.com/questions/48562288/frequency-of-each-unique-combination-in-data-frame

# get proportion of etoh for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
etohyoungcount <- c(sum(ocme_2009_young$et_oh == 1), sum(ocme_2010_young$et_oh == 1), sum(ocme_2011_young$et_oh == 1), sum(ocme_2012_young$et_oh == 1), sum(ocme_2013_young$et_oh == 1), sum(ocme_2014_young$et_oh == 1), sum(ocme_2015_young$et_oh == 1), sum(ocme_2016_young$et_oh == 1), sum(ocme_2017_young$et_oh == 1))

# get proportion for older people
etoholdcount <- c(sum(ocme_2009_old$et_oh == 1), sum(ocme_2010_old$et_oh == 1), sum(ocme_2011_old$et_oh == 1), sum(ocme_2012_old$et_oh == 1), sum(ocme_2013_old$et_oh == 1), sum(ocme_2014_old$et_oh == 1), sum(ocme_2015_old$et_oh == 1), sum(ocme_2016_old$et_oh == 1), sum(ocme_2017_old$et_oh == 1))

# combine vectors into df
etohcount <- data.frame(yearList, etohyoungcount, etoholdcount)

# change the col names to be more informative
names(etohcount) <- c("Year", "Younger", "Older")

# convert the "race_sex" column into row names so I get a table with only numeric values, from https://stackoverflow.com/questions/5555408/convert-the-values-in-a-column-into-row-names-in-an-existing-data-frame

etohchi <- etohcount[,-1]
rownames(etohchi) <- etohcount[,1]

as.matrix(etohchi)

# looks like I can just use any matrix for chisq.test, from https://researchguides.library.vanderbilt.edu/c.php?g=156859&p=3058795

chisq.test(as.matrix(etohchi))
```

### Proportion of deaths involving buprenorphine/methadone

```{r propBupeMethadone, echo=FALSE, message=FALSE, warning=FALSE}

# unfortunately there are different variable names for each ocme file so a for loop won't work :/

# get proportion of cocaine for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
methbupeyoung <- c(sum(sum(ocme_2009_young$buprenorphine == 1), sum(ocme_2009_young$methadone == 1))/nrow(ocme_2009_young), sum(sum(ocme_2010_young$buprenorphine == 1), sum(ocme_2010_young$methadone == 1))/nrow(ocme_2010_young), sum(sum(ocme_2011_young$buprenorphine == 1), sum(ocme_2011_young$methadone == 1))/nrow(ocme_2011_young), sum(sum(ocme_2012_young$buprenorphine == 1), sum(ocme_2012_young$methadone == 1))/nrow(ocme_2012_young), sum(sum(ocme_2013_young$buprenorphine == 1), sum(ocme_2013_young$methadone == 1))/nrow(ocme_2013_young), sum(sum(ocme_2014_young$buprenorphine == 1), sum(ocme_2014_young$methadone == 1))/nrow(ocme_2014_young), sum(sum(ocme_2015_young$buprenorphine == 1), sum(ocme_2015_young$methadone == 1))/nrow(ocme_2015_young), sum(sum(ocme_2016_young$buprenorphine == 1), sum(ocme_2016_young$methadone == 1))/nrow(ocme_2016_young), sum(sum(ocme_2017_young$buprenorphine == 1), sum(ocme_2017_young$methadone == 1))/nrow(ocme_2017_young))

# get proportion for older people
methbupeold <- c(sum(sum(ocme_2009_old$buprenorphine == 1), sum(ocme_2009_old$methadone == 1))/nrow(ocme_2009_old), sum(sum(ocme_2010_old$buprenorphine == 1), sum(ocme_2010_old$methadone == 1))/nrow(ocme_2010_old), sum(sum(ocme_2011_old$buprenorphine == 1), sum(ocme_2011_old$methadone == 1))/nrow(ocme_2011_old), sum(sum(ocme_2012_old$buprenorphine == 1), sum(ocme_2012_old$methadone == 1))/nrow(ocme_2012_old), sum(sum(ocme_2013_old$buprenorphine == 1), sum(ocme_2013_old$methadone == 1))/nrow(ocme_2013_old), sum(sum(ocme_2014_old$buprenorphine == 1), sum(ocme_2014_old$methadone == 1))/nrow(ocme_2014_old), sum(sum(ocme_2015_old$buprenorphine == 1), sum(ocme_2015_old$methadone == 1))/nrow(ocme_2015_old), sum(sum(ocme_2016_old$buprenorphine == 1), sum(ocme_2016_old$methadone == 1))/nrow(ocme_2016_old), sum(sum(ocme_2017_old$buprenorphine == 1), sum(ocme_2017_old$methadone == 1))/nrow(ocme_2017_old))

# combine vectors into df
propmethbupe <- data.frame(yearList, methbupeyoung, methbupeold)

# change the col names to be more informative
names(propmethbupe) <- c("Year", "Younger", "Older")

# create grouped bar plot, from https://stackoverflow.com/questions/72449952/making-a-grouped-bar-chart-in-r
# getting the labels in the right position from https://stackoverflow.com/questions/37587288/ggplot-label-bars-in-grouped-bar-plot
# info on changing legend title from here https://stackoverflow.com/questions/41649929/how-to-modify-a-legend-title-in-a-barplot-with-ggplot

pmethbupe <- propmethbupe |>
  # Reshape data to long format
  pivot_longer(Younger:Older) |> 
  ggplot(aes(x = Year, y = value, fill = name)) +
  # Implement a grouped bar chart
  geom_bar(stat = "identity", position = position_dodge(width= 1)) + geom_text(aes(label = round(value, 2)), stat = "identity", vjust = -0.1, colour = "black", size = 9/.pt, position = position_dodge(width= 1)) + theme_light() + ggtitle("Proportion of methadone/buprenorphine-detected \n deaths by age group in CT, 2009-2017") + ylab("Proportion") + theme(plot.title = element_text(hjust = 0.5)) + scale_fill_discrete(name = "Age group") + scale_x_continuous(breaks=propmethbupe$Year)+ coord_cartesian(ylim = c(0, 1))

pmethbupe

ggsave(plot=pmethbupe, filename = "images/prop_meth_bupe_old.png", width = 10, height = 8)

```


### Chi squared testing for the number of methadone/buprenorphine-detected deaths

```{r chiSquaredMethBupe, echo=FALSE, message=FALSE, warning=FALSE}
# get frequency of each race for old and young, from https://stackoverflow.com/questions/48562288/frequency-of-each-unique-combination-in-data-frame

# get proportion of cocaine for each year for younger people -- from https://stackoverflow.com/questions/28195996/count-number-of-rows-matching-a-criteria
methbupeyoungcount <- c(sum(sum(ocme_2009_young$buprenorphine == 1), sum(ocme_2009_young$methadone == 1)), sum(sum(ocme_2010_young$buprenorphine == 1), sum(ocme_2010_young$methadone == 1)), sum(sum(ocme_2011_young$buprenorphine == 1), sum(ocme_2011_young$methadone == 1)), sum(sum(ocme_2012_young$buprenorphine == 1), sum(ocme_2012_young$methadone == 1)), sum(sum(ocme_2013_young$buprenorphine == 1), sum(ocme_2013_young$methadone == 1)), sum(sum(ocme_2014_young$buprenorphine == 1), sum(ocme_2014_young$methadone == 1)), sum(sum(ocme_2015_young$buprenorphine == 1), sum(ocme_2015_young$methadone == 1)), sum(sum(ocme_2016_young$buprenorphine == 1), sum(ocme_2016_young$methadone == 1)), sum(sum(ocme_2017_young$buprenorphine == 1), sum(ocme_2017_young$methadone == 1)))

# get proportion for older people
methbupeoldcount <- c(sum(sum(ocme_2009_old$buprenorphine == 1), sum(ocme_2009_old$methadone == 1)), sum(sum(ocme_2010_old$buprenorphine == 1), sum(ocme_2010_old$methadone == 1)), sum(sum(ocme_2011_old$buprenorphine == 1), sum(ocme_2011_old$methadone == 1)), sum(sum(ocme_2012_old$buprenorphine == 1), sum(ocme_2012_old$methadone == 1)), sum(sum(ocme_2013_old$buprenorphine == 1), sum(ocme_2013_old$methadone == 1)), sum(sum(ocme_2014_old$buprenorphine == 1), sum(ocme_2014_old$methadone == 1)), sum(sum(ocme_2015_old$buprenorphine == 1), sum(ocme_2015_old$methadone == 1)), sum(sum(ocme_2016_old$buprenorphine == 1), sum(ocme_2016_old$methadone == 1)), sum(sum(ocme_2017_old$buprenorphine == 1), sum(ocme_2017_old$methadone == 1)))

# combine vectors into df
methbupecount <- data.frame(yearList, methbupeyoungcount, methbupeoldcount)

# change the col names to be more informative
names(methbupecount) <- c("Year", "Younger", "Older")

# convert the "race_sex" column into row names so I get a table with only numeric values, from https://stackoverflow.com/questions/5555408/convert-the-values-in-a-column-into-row-names-in-an-existing-data-frame

methbupechi <- methbupecount[,-1]
rownames(methbupechi) <- methbupecount[,1]

as.matrix(methbupechi)

# looks like I can just use any matrix for chisq.test, from https://researchguides.library.vanderbilt.edu/c.php?g=156859&p=3058795

chisq.test(as.matrix(methbupechi))
```

### Top 5 substance combinations {.tabset}

### Injury location {.tabset}

### Town of injury {.tabset}

#### Young

Note: the "percent young" graph only shows towns where an OD occurred in a given year, which is not necessarily all CT towns. 

```{r townYoung, echo=FALSE, message=FALSE, warning=FALSE}
for(i in 1:length(ocmelistYoung)){

   # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  ocme <- ocmelistYoung[[i]] 
  
  # get a df for all OCME data, old and young
  
  ocmeall <- ocmelist[[i]]
  
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- ocme$Year[1]
  
  towntitle <- paste("City/town of OD for", year, "CT opioid-detected overdose deaths in younger people")
  
#fixed axis text size in theme() using https://stackoverflow.com/questions/65956833/ggplot2-how-to-dynamically-wrap-resize-rescale-x-axis-labels-so-they-wont-over

town_24h <- ggplot(data=ocme, aes(x=injury_city_revised, ..count..)) + 
      geom_bar(aes(fill = ..count..), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = -0.3, colour = "red", size = 3/.pt) + ggtitle(towntitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 3))

print(town_24h)

ggsave(plot=month, filename = paste0("images/overdoses_by_town_young_", year, ".png"), width = 10, height = 8)

#find the number of ODs in each town by aggregating the indicator var for each town, from https://stackoverflow.com/questions/60961960/how-do-i-sum-a-column-with-categorical-vairables-in-r

# create a new dataframe that we can manipulate without messing up within24h

youngprop <- ocme %>% group_by(injury_city_revised) %>% summarise_at(vars(fatal), sum)

# find towns that don't have anyone discovered within 24h and add them into the df
# i have to use unique() so I don't get repeats

missingtowns <- setdiff(unique(ocmeall$injury_city_revised), youngprop$injury_city_revised)

for (i in missingtowns) {
  
  # create row for that town and a 0 next to number of fatalities discovered in 24h
  newtown <- c(i, 0)
  
  # add to within24h2018prop
  youngprop = rbind(youngprop, newtown)
  
}

# convert the within24h2018prop fatals into numerics because for some reason they're strings
youngprop$fatal <- as.numeric(youngprop$fatal)

# aggregate fatalities for overall deaths [not just discovered in 24h]
ocme_agg <- ocmeall %>% group_by(injury_city_revised) %>% summarise_at(vars(fatal), sum)

# rename the danbury_agg24h "fatal" column because it'll get confusing otherwise when merged
names(ocme_agg)[names(ocme_agg) == 'fatal'] <- 'total'

# merge with within24h2018prop to have a column with the actual number of deaths
ocme_young_merged <- merge(youngprop, ocme_agg, by="injury_city_revised")

# get the proportion of fatalities discovered in 24h, multiply by 100 so we have a percentage
ocme_young_merged$Prop <- round((ocme_young_merged$fatal/ocme_young_merged$total)*100, digits=2)

# convert any NaN to 0 -- from https://stackoverflow.com/questions/18142117/how-to-replace-nan-value-with-zero-in-a-huge-data-frame

ocme_young_merged$Prop[is.nan(ocme_young_merged$Prop)] <- 0

# Plot the proportion of people discovered within 24h
# had an issue with the fill, edited with http://www.sthda.com/english/wiki/ggplot2-barplots-quick-start-guide-r-software-and-data-visualization
# needed to change the font size for the text overlay, used https://stackoverflow.com/questions/25061822/ggplot-geom-text-font-size-control

proptitle <- paste("Percent of fatalities in younger people by town in", year, "in CT")

town_young_capita <- ggplot(data=ocme_young_merged, aes(x=injury_city_revised, y=Prop, fill=Prop, position = "dodge")) + 
      geom_bar(stat="Identity") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = Prop), vjust = -0.3, colour = "red", size = 3/.pt) + ggtitle(proptitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 3))

print(town_young_capita)

ggsave(plot=town_young_capita, filename = paste0("images/overdoses_by_town_prop_young_", year, ".png"), width = 10, height = 8)

}

```


#### Old

```{r townOld, echo=FALSE, message=FALSE, warning=FALSE}
for(i in 1:length(ocmelistOld)){

   # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  ocme <- ocmelistOld[[i]] 
  
  # get a df for all OCME data, old and young
  
  ocmeall <- ocmelist[[i]]
  
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- ocme$Year[1]
  
  towntitle <- paste("City/town of OD for", year, "CT opioid-detected overdose deaths in older people")
  
#fixed axis text size in theme() using https://stackoverflow.com/questions/65956833/ggplot2-how-to-dynamically-wrap-resize-rescale-x-axis-labels-so-they-wont-over

town_24h <- ggplot(data=ocme, aes(x=injury_city_revised, ..count..)) + 
      geom_bar(aes(fill = ..count..), position = "dodge") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = ..count..), stat = "count", vjust = -0.3, colour = "red", size = 3/.pt) + ggtitle(towntitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 3))

print(town_24h)

ggsave(plot=month, filename = paste0("images/overdoses_by_town_old_", year, ".png"), width = 10, height = 8)

#find the number of ODs in each town by aggregating the indicator var for each town, from https://stackoverflow.com/questions/60961960/how-do-i-sum-a-column-with-categorical-vairables-in-r

# create a new dataframe that we can manipulate without messing up within24h

oldprop <- ocme %>% group_by(injury_city_revised) %>% summarise_at(vars(fatal), sum)

# find towns that don't have anyone discovered within 24h and add them into the df
# i have to use unique() so I don't get repeats

missingtowns <- setdiff(unique(ocmeall$injury_city_revised), oldprop$injury_city_revised)

for (i in missingtowns) {
  
  # create row for that town and a 0 next to number of fatalities discovered in 24h
  newtown <- c(i, 0)
  
  # add to within24h2018prop
  oldprop = rbind(oldprop, newtown)
  
}

# convert the within24h2018prop fatals into numerics because for some reason they're strings
oldprop$fatal <- as.numeric(oldprop$fatal)

# aggregate fatalities for overall deaths [not just discovered in 24h]
ocme_agg <- ocmeall %>% group_by(injury_city_revised) %>% summarise_at(vars(fatal), sum)

# rename the danbury_agg24h "fatal" column because it'll get confusing otherwise when merged
names(ocme_agg)[names(ocme_agg) == 'fatal'] <- 'total'

# merge with within24h2018prop to have a column with the actual number of deaths
ocme_old_merged <- merge(oldprop, ocme_agg, by="injury_city_revised")

# get the proportion of fatalities discovered in 24h, multiply by 100 so we have a percentage
ocme_old_merged$Prop <- round((ocme_old_merged$fatal/ocme_old_merged$total)*100, digits=2)

# convert any NaN to 0 -- from https://stackoverflow.com/questions/18142117/how-to-replace-nan-value-with-zero-in-a-huge-data-frame

ocme_old_merged$Prop[is.nan(ocme_old_merged$Prop)] <- 0

# Plot the proportion of people discovered within 24h
# had an issue with the fill, edited with http://www.sthda.com/english/wiki/ggplot2-barplots-quick-start-guide-r-software-and-data-visualization
# needed to change the font size for the text overlay, used https://stackoverflow.com/questions/25061822/ggplot-geom-text-font-size-control

proptitle <- paste("Percent of fatalities in older people by town in", year, "in CT")

town_old_capita <- ggplot(data=ocme_old_merged, aes(x=injury_city_revised, y=Prop, fill=Prop, position = "dodge")) + 
      geom_bar(stat="Identity") + scale_x_discrete(guide = guide_axis(angle = 45)) + theme_light() + geom_text(aes(label = Prop), vjust = -0.3, colour = "red", size = 3/.pt) + ggtitle(proptitle)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 3))

print(town_old_capita)

ggsave(plot=town_old_capita, filename = paste0("images/overdoses_by_town_prop_old_", year, ".png"), width = 10, height = 8)

}

```

### Maps of OD towns {.tabset} 

#### Young

```{r townMapYoung, echo=FALSE, message=FALSE, warning=FALSE}

for(i in 1:length(ocmelistYoung)){

  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  ocme <- ocmelistYoung[[i]]
    
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- ocme$Year[1]
  
#find the number of ODs in each town by aggregating the indicator var for each town, from https://stackoverflow.com/questions/60961960/how-do-i-sum-a-column-with-categorical-vairables-in-r

ocme_counts<- ocme %>% group_by(injury_city_revised) %>% summarise_at(vars(fatal), sum)

#rename the injury city variable so we can merge it with the ct_towns df

colnames(ocme_counts) <- c('NAME', 'Fatal')

# not all towns experienced an OD in 2020. Check which: https://stackoverflow.com/questions/72274729/find-elements-in-a-column-that-is-not-in-another-column-of-another-dataframe-in

ocme_odcounts <- data.frame(setdiff(ct_towns$NAME, ocme_counts$NAME), 0)

# give this the same column names as ocme_2020_counts

colnames(ocme_odcounts) <- c('NAME', 'Fatal')

# merge the two datasets

ocme_counts <- rbind(ocme_counts, ocme_odcounts)

#merge with ct_towns so that way we can use the fatal OD counts for the fill

ct_town_odcounts <- merge(ct_towns, ocme_counts,by="NAME")

# Plot! And use the fatal data to color the plot in. Info from https://geanders.github.io/navy_public_health/3-2-basic-mapping.html

maptitle <- paste("Fatal opioid-detected overdoses in younger people by town in", year, "in CT")


map <- ggplot(ocme) + theme_void() + geom_sf(data=ct_town_odcounts, size=0.1, aes(fill=Fatal))+ ggtitle(maptitle)+scale_fill_viridis_c(option="plasma")+theme(plot.title = element_text(hjust = 0.5), legend.position="right")

print(map)

#save this one for the presentation
ggsave(plot=map, filename =  paste0("images/map_", year, "_young.png"), width = 10, height = 8)

}

```


```{r townMapOld, echo=FALSE, message=FALSE, warning=FALSE}

for(i in 1:length(ocmelistOld)){

  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  ocme <- ocmelistOld[[i]]
    
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- ocme$Year[1]
  
#find the number of ODs in each town by aggregating the indicator var for each town, from https://stackoverflow.com/questions/60961960/how-do-i-sum-a-column-with-categorical-vairables-in-r

ocme_counts<- ocme %>% group_by(injury_city_revised) %>% summarise_at(vars(fatal), sum)

#rename the injury city variable so we can merge it with the ct_towns df

colnames(ocme_counts) <- c('NAME', 'Fatal')

# not all towns experienced an OD in 2020. Check which: https://stackoverflow.com/questions/72274729/find-elements-in-a-column-that-is-not-in-another-column-of-another-dataframe-in

ocme_odcounts <- data.frame(setdiff(ct_towns$NAME, ocme_counts$NAME), 0)

# give this the same column names as ocme_2020_counts

colnames(ocme_odcounts) <- c('NAME', 'Fatal')

# merge the two datasets

ocme_counts <- rbind(ocme_counts, ocme_odcounts)

#merge with ct_towns so that way we can use the fatal OD counts for the fill

ct_town_odcounts <- merge(ct_towns, ocme_counts,by="NAME")

# Plot! And use the fatal data to color the plot in. Info from https://geanders.github.io/navy_public_health/3-2-basic-mapping.html

maptitle <- paste("Fatal opioid-detected overdoses in older people by town in", year, "in CT")


map <- ggplot(ocme) + theme_void() + geom_sf(data=ct_town_odcounts, size=0.1, aes(fill=Fatal))+ ggtitle(maptitle)+scale_fill_viridis_c(option="plasma")+theme(plot.title = element_text(hjust = 0.5), legend.position="right")

print(map)

#save this one for the presentation
ggsave(plot=map, filename =  paste0("images/map_", year, "_old.png"), width = 10, height = 8)

}

```


### Injury town v. residence town {.tabset}

#### Young

```{r injuryResidenceYoung, echo=FALSE, message=FALSE, warning=FALSE}
# initialize empty matrix, from https://stackoverflow.com/questions/32712301/create-empty-data-frame-with-column-names-by-assigning-a-string-vector

out_of_town <- data.frame(matrix(ncol = 5, nrow = 0))
cols <- c("Town name", "Year", "Out-of-town ODs","Total ODs","Percent of out-of-town ODs")

for(i in 1:length(ocmelistYoung)){

  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  ocme <- ocmelistYoung[[i]]
    
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- ocme$Year[1]

# get rid of blank/unknown injury and residence locations, from https://www.r-bloggers.com/2022/05/subsetting-with-multiple-conditions-in-r/

ocme_noblank <- ocme %>% filter(residence_city_updated != "" & injury_city_revised!="Unknown")

# convert all Willimantic residences to Windham, because otherwise the out of town calculations get messed up
ocme_noblank$residence_city_updated <- ifelse(ocme_noblank$residence_city_updated == "Willimantic", "Windham", ocme_noblank$residence_city_updated)

# print a statement with some function output -- from https://stackoverflow.com/questions/13036649/how-do-i-concatenate-string-and-an-output-evaluated-from-a-function-in-r [the %f stands for an integer]

#sprintf("There were %f instances where a person's residence town didn't match up with their injury town, out of %f ODs in younger people in 2018 where the injury and residence town were both known. This is %f of all such ODs in this year.", nrow(subset(ocme_noblank, residence_city_updated != injury_city_revised)), nrow(ocme_noblank), nrow(subset(ocme_noblank, residence_city_updated != injury_city_revised))/nrow(ocme_noblank))

#print("Here are the towns where someone OD'd but didn't live there:")

#table(subset(ocme_noblank, residence_city_updated != injury_city_revised)$injury_city_revised)

#print("Here's a more in-depth look at out-of-town ODs for younger people in specific towns:")

# iteratively add info from these towns to the df, from https://www.projectpro.io/recipes/append-output-from-for-loop-dataframe-r

for (i in c("Danbury", "Waterbury", "Torrington", "New Haven", "West Haven", "Hartford", "Norwich", "New London", "Bridgeport", "New Britain", "Litchfield", "Meriden", "Norwalk", "Stamford", "Windham")) {
  
  # subset for the number of ODs in that town 
  
  ocme_injury <- subset(ocme_noblank, injury_city_revised == i)
  
  # create a results vector to append to out_of_town 2020 df
  # contains town name, number of ODs from out of town residents, total ODs and fraction of out of town ODs in that   order
  
  results <- c(i, year, nrow(subset(ocme_injury, residence_city_updated != injury_city_revised)), nrow(ocme_injury), round(nrow(subset(ocme_injury, residence_city_updated != injury_city_revised))/nrow(ocme_injury), 3)*100)

  # convert NaN to 0
  results[is.nan(results)] <- 0
  
  # add this to the out of town df 
  out_of_town = rbind(out_of_town, results)
}

# for some reason it doesn't work if I do this before adding rows?
colnames(out_of_town) <- cols

}

out_of_town
```

#### Old

```{r injuryResidenceOld, echo=FALSE, message=FALSE, warning=FALSE}
# initialize empty matrix, from https://stackoverflow.com/questions/32712301/create-empty-data-frame-with-column-names-by-assigning-a-string-vector

out_of_town <- data.frame(matrix(ncol = 5, nrow = 0))
cols <- c("Town name", "Year", "Out-of-town ODs","Total ODs","Percent of out-of-town ODs")

for(i in 1:length(ocmelistOld)){

  # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  ocme <- ocmelistOld[[i]]
    
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- ocme$Year[1]

# get rid of blank/unknown injury and residence locations, from https://www.r-bloggers.com/2022/05/subsetting-with-multiple-conditions-in-r/

ocme_noblank <- ocme %>% filter(residence_city_updated != "" & injury_city_revised!="Unknown")

# convert all Willimantic residences to Windham, because otherwise the out of town calculations get messed up
ocme_noblank$residence_city_updated <- ifelse(ocme_noblank$residence_city_updated == "Willimantic", "Windham", ocme_noblank$residence_city_updated)

# print a statement with some function output -- from https://stackoverflow.com/questions/13036649/how-do-i-concatenate-string-and-an-output-evaluated-from-a-function-in-r [the %f stands for an integer]

#sprintf("There were %f instances where a person's residence town didn't match up with their injury town, out of %f ODs in younger people in 2018 where the injury and residence town were both known. This is %f of all such ODs in this year.", nrow(subset(ocme_noblank, residence_city_updated != injury_city_revised)), nrow(ocme_noblank), nrow(subset(ocme_noblank, residence_city_updated != injury_city_revised))/nrow(ocme_noblank))

#print("Here are the towns where someone OD'd but didn't live there:")

#table(subset(ocme_noblank, residence_city_updated != injury_city_revised)$injury_city_revised)

#print("Here's a more in-depth look at out-of-town ODs for younger people in specific towns:")

# iteratively add info from these towns to the df, from https://www.projectpro.io/recipes/append-output-from-for-loop-dataframe-r

for (i in c("Danbury", "Waterbury", "Torrington", "New Haven", "West Haven", "Hartford", "Norwich", "New London", "Bridgeport", "New Britain", "Litchfield", "Meriden", "Norwalk", "Stamford", "Windham")) {
  
  # subset for the number of ODs in that town 
  
  ocme_injury <- subset(ocme_noblank, injury_city_revised == i)
  
  # create a results vector to append to out_of_town 2020 df
  # contains town name, number of ODs from out of town residents, total ODs and fraction of out of town ODs in that   order
  
  results <- c(i, year, nrow(subset(ocme_injury, residence_city_updated != injury_city_revised)), nrow(ocme_injury), round(nrow(subset(ocme_injury, residence_city_updated != injury_city_revised))/nrow(ocme_injury), 3)*100)

  # convert NaN to 0
  results[is.nan(results)] <- 0
  
  # add this to the out of town df 
  out_of_town = rbind(out_of_town, results)
}

# for some reason it doesn't work if I do this before adding rows?
colnames(out_of_town) <- cols

}

out_of_town
```

### How many people died in their own home? {.tabset}

#### Young

```{r ownHomeYoung, echo=FALSE, message=FALSE, warning=FALSE}

ownhome <- data.frame(matrix(ncol = 3, nrow = 0))
cols <- c("Year", "Total deaths", "Number of deaths in own residence", "Proportion of deaths in own residence","Percent of deaths in own residence out of all deaths in residence")

for(i in 1:length(ocmelistYoung)){

    # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  ocme <- ocmelistYoung[[i]]
    
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- ocme$Year[1]

  numdeaths <- sum(ocme$injury_address_updated == ocme$residence_address_updated)

  propdeaths <- sum(ocme$injury_address_updated == ocme$residence_address_updated)/nrow(ocme)
  
  propdeathsresidence <- sum(ocme$injury_address_updated == ocme$residence_address_updated)/sum(ocme$injury_location=="residence")*100

    
  results <- c(year, nrow(ocme), numdeaths, propdeaths, propdeathsresidence)
  
  # convert NaN to 0
  results[is.nan(results)] <- 0
  
  # add this to the out of town df 
  ownhome = rbind(ownhome, results)
  
  # for some reason I need to do this
  colnames(ownhome) <- cols
}

ownhome
```

#### Old

```{r ownHomeOld, echo=FALSE, message=FALSE, warning=FALSE}

ownhome <- data.frame(matrix(ncol = 3, nrow = 0))
cols <- c("Year", "Total number of deaths", "Number of deaths in own residence", "Proportion of deaths in own residence","Percent of deaths in own residence out of all deaths in residence")

for(i in 1:length(ocmelistOld)){

    # get the df you need -- you need the double brackets, see https://stackoverflow.com/questions/32007076/loop-through-list-of-data-frames
  ocme <- ocmelistOld[[i]]
    
  # grab the year -- you only need the first element in the year column since all of them have the same value
  year <- ocme$Year[1]

  numdeaths <- sum(ocme$injury_address_updated == ocme$residence_address_updated)

  propdeaths <- sum(ocme$injury_address_updated == ocme$residence_address_updated)/nrow(ocme)
  
  propdeathsresidence <- sum(ocme$injury_address_updated == ocme$residence_address_updated)/sum(ocme$injury_location=="residence")*100

    
  results <- c(year, nrow(ocme), numdeaths, propdeaths, propdeathsresidence)
  
  # convert NaN to 0
  results[is.nan(results)] <- 0
  
  # add this to the out of town df 
  ownhome = rbind(ownhome, results)
  
  # for some reason I need to do this
  colnames(ownhome) <- cols
}

ownhome
```

#### Tests


```{r age, echo=FALSE, message=FALSE, warning=FALSE}
# from https://stackoverflow.com/questions/71745093/using-a-loop-to-subset-data-by-year-and-store-the-subset-for-each-year-in-a-new
# seeing if I can use a for loop for this? Maybe not
#for(i in unique(ocme_old$Year)){

# initialize a table with dip test p values, from https://stackoverflow.com/questions/32712301/create-empty-data-frame-with-column-names-by-assigning-a-string-vector

diptest <- setNames(data.frame(matrix(ncol = 2, nrow = 0)), c("Year", "p value"))

# initialize a dataframe for old and young people, from https://www.geeksforgeeks.org/how-to-create-an-empty-dataframe-in-r/

old <- data.frame(matrix(ncol = 0, nrow = 0))
young <- data.frame(matrix(ncol = 0, nrow = 0))

# iterate over years to get the age dist/separate old and young

#for(i in unique(ocme_old$Year)){
for(i in c("2017")){

# only get data for year i -- need to do as.character(i) otherwise it throws a type error  
ocme <- subset(ocme_old, year == as.character(i))  
  
#there was an issue with the x axis not showing up when I coped the other histogram code -- resolved using https://stackoverflow.com/questions/42425195/ggplot-data-year-over-year-axis-labels-not-showing

#fixed issue with colors not showing -- turns out it was because the data was numeric: https://stackoverflow.com/questions/55400037/ggplot-geom-bar-fill-not-coloring-bars-on-plot

#calculate modes, see https://stackoverflow.com/questions/27418461/calculate-the-modes-in-a-multimodal-distribution-in-r

# Select an appropriate bandwidth. Here we use a "rule of thumb" method
# looks like there are some NAs in the 2009 data so I have to get rid of them
bandwidth <- bw.nrd0(na.omit(ocme$age)) 

# Other methods, e.g. bw.ucv(), will identify different bandwidths as optimal
# Consequently the number (and position) of modes may differ too

# Methods implemented in multimode are documented at ?bw.nrd

# Calculate number of modes
n <- nmodes(ocme$age, bw = bandwidth)

# Calculate location of these modes
loc <- locmodes(ocme$age, mod0 = n, display=TRUE)
# Remove `, display = TRUE` if you don't want to see the KDE
#loc

# Locations of modes are odd values in loc
modes <- loc$locations[seq(from = 1, by = 2, length.out = n)]
antimodes <- loc$locations[seq(from = 0, by = 2, length.out = n)]

# subset out "young" and "old" populations -- "young" is <= first antimode, "old" is > first antimode
ocme_young <- subset(ocme, age <= antimodes[1])
ocme_older <- subset(ocme, age > antimodes[1])

young <- rbind(young, ocme_young)
old <- rbind(old, ocme_older)

#added multiple mode lines to plot using https://stackoverflow.com/questions/71580196/add-horizontal-lines-from-a-vector-to-a-ggplot-uisng-facet-grid

# see if I can automate the title, from https://stackoverflow.com/questions/59095237/automate-ggplots-while-using-variable-labels-as-title-and-axis-titles
title <- paste("Age distribution for ", i, "CT opioid-detected overdose deaths")

# a plot without mode lines or anything, with the number of people dying for each age 
ggplot(data=ocme, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(age)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme$age, labels = ocme$age) + geom_text(aes(label = ..count..), stat = "count", vjust = 0, colour = "black", size = 7/.pt) + theme_light() + ggtitle(title)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))

age <- ggplot(data=ocme, aes(x=age, ..count..)) + 
      geom_bar(aes(fill = as.character(age)), position = "dodge") + scale_x_continuous(guide = guide_axis(angle = 90), breaks = ocme$age, labels = ocme$age) + theme_light() + geom_vline(xintercept = modes, size=1) + ggtitle(title)+theme(plot.title = element_text(hjust = 0.5), legend.position="none", axis.text.x = element_text(size = 7))

# seem to need to use print() when in a for loop, from https://stackoverflow.com/questions/29053223/plots-not-working-in-for-loop
print(age)

# try to automate the filenames? from https://www.edureka.co/community/41963/r-programming-naming-the-output-file-using-a-variable
ggsave(plot=age, filename = paste0("images/age_", i, ".png"), width = 10, height = 8)

# append year and dip test value to table
diptest[nrow(diptest) + 1,] = c(i, dip.test(ocme$age)$p.value)

}

diptest 
```


